---
title: "Deeper-Dive Analysis Framework: Non-College Graduates Without Meaningful Workforce Participation"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/Master-report-dataset.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Research Questions: Workforce Trajectories of Non-College Graduates in Southeast Minnesota

This phase of the analysis examines workforce participation trajectories among high school graduates in Southeast Minnesota who did not complete a postsecondary credential. Rather than relying on a single point-in-time workforce outcome, the analysis focuses on patterns of labor market attachment during the first five years after exiting formal education, with particular attention to persistence, instability, and regional attachment.

The research is guided by the following questions:

1.  Among non-college graduates, how persistent is being disconnected from sustained workforce participation either in the region or state?

2.  For non-college graduates who experience non-sustained workforce participation, where does employment occur when it is present?

3.  Among individuals categorized as having no Minnesota employment record during the first five years after high school, how persistent is this status?

4.  How are individual characteristics associated with workforce disconnection and subsequent employment trajectories?

<br>

# Study Population and Group Definitions

**Study Population**

The analytic population consists of individuals who:

- Graduated from a high school in Southeast Minnesota, and
- Did not complete a postsecondary credential (`ps.grad = No`), and,
- Have workforce participation records available for at least the first five years after high school graduation OR the first five years after exiting formal post-secondary education.

<br>

**Definition of Workforce Disconnection**

Workforce participation is measured annually during the first five years after high school or existing post-secondary attendance. Individuals are classified as **disconnected** in a given year if they are categorized as either:

- Non-sustained workforce participation, or
- No Minnesota employment record.

Using this annual classification, individuals are grouped based on the persistence of disconnection during years 1 through 5:

- **Persistently Disconnected:**\
  Disconnected in four or five of the first five years after graduation.

- **Disconnected (Broad Definition):**\
  Disconnected in three or more of the first five years after graduation.

- **Intermittently Disconnected:**\
  Disconnected in one or two of the first five years after graduation.

- **Consistently attached:**\
  Have sustained workforce participation in all five years.

These definitions emphasize sustained labor market detachment rather than short-term transitions.

<br>

**Regional Attachment**

For years in which individuals are employed but not meaningfully attached to the workforce, employment location is classified as:

- Within Southeast Minnesota, or
- Outside Southeast Minnesota (elsewhere in Minnesota).

This distinction is used to assess whether workforce disconnection reflects regional underutilization or disengagement from the regional labor market.

<br>

**Framing Consideration**

Together, these definitions allow the analysis to move beyond point-in-time workforce outcomes and toward a trajectory-based understanding of early-career labor market experiences. This approach supports clearer interpretation of workforce instability, recovery, and regional retention among non-college graduates in Southeast Minnesota.

<br>

# Methodology and Analytic Steps

This analysis examines workforce participation trajectories among non-college graduates in Southeast Minnesota, with a focus on identifying patterns of workforce disconnection, regional attachment, and employment stability during the first five years after high school.

## Step 1: Data Preparation and Analytic Population

**1.1 Analytic Population**

The analytic population is restricted to individuals who:

- Graduated from a high school in Southeast Minnesota, and
- Did not complete a postsecondary credential (`ps.grad = No`), and
- Have complete workforce participation categorizations for years 1 through 5 after high school or,
- first 5 years after exiting attendance in a post-secondary institution.

**1.2 Annual Workforce Participation Classification**

For each individual, annual workforce participation status is identified for: - `grad.year.1` - `grad.year.2` - `grad.year.3` - `grad.year.4` - `grad.year.5`

Each year is classified into one of the following categories:

- Meaningful workforce participation
- Not meaningful workforce participation
- No Minnesota employment record.

It should be noted that "grad.year.x" here means years after exiting formal education, not years after graduating high school.

**1.3 Workforce Disconnection Indicators**

For each annual observation, a binary **disconnection indicator** is created:

- **Disconnected (year-level)** = Non-sustained WF - Southeast or MN, **or** No Minnesota emp record

Using these annual indicators, individuals are grouped based on persistence of disconnection across years 1–5:

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Disconnected (Broad):** Disconnected in 3 of 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

These categories emphasize sustained labor market attachment or detachment rather than point-in-time outcomes.

**1.4 Regional Location of Not Meaningful Employment**

For years in which individuals are categorized as **Non-sustained WF**, employment location is further classified as:

- Within Southeast Minnesota
- Outside Southeast Minnesota (elsewhere in Minnesota)
- Both (simultaneous employment inside and outside the region)

This classification allows identification of regional underemployment versus regional disengagement.

**1.5 Employment Intensity and Job Holding**

For each individual and each year (years 1–5), the following employment intensity measures are constructed:

- Number of distinct employers during the year
- Indicator for multiple job holding (two or more employers)

These measures are used to distinguish underemployment from precarious or fragmented employment.

<br>

## Step 2: Descriptive and Summary Statistics

**2.1 Distribution of Workforce Disconnection**

Summary statistics are produced to describe:

- The proportion of non-college graduates in each workforce disconnection category
- Differences in persistence and intermittency of disconnection

This establishes the overall scope and severity of workforce disconnection in early adulthood.

**2.2 Regional Attachment and Job Structure**

Among individuals experiencing workforce disconnection, summary statistics examine:

- The location of not meaningful employment (within region, outside region, both)
- Average and median number of employers per year
- Prevalence of multiple job holding

These summaries clarify whether disconnection reflects local underemployment, out-of-region work, or fragmented employment patterns.

<br>

## Step 3: Relationship Between Individual Characteristics and Disconnection Trajectories

Step 3 will focus on answering the question of which characteristics are associated with the disconnection persistence categories. This will be a bi-variate analysis.

**Independent Variables**

We will use the following variables that were selected based on our earlier multiple correspondence analysis.

- Demographics
  - `Gender`
  - `RaceEthnicity`
  - `economic.status`
  - `SpecialEdStatus`
  - `english.learner`
- High school characteristics
  - `Dem_Desc`
  - `edr`
  - `avg.wages.pct.state`
  - `avg.unemp.rate`
- High school academics
  - `pseo.participant`
  - `ap.exam`
  - `took.ACT`
  - `ACTCompositeScore_filled`
  - `sat.taken`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`
- Post-secondary
  - `attended.ps`

**3.2 Analytic Approach**

- Categorical variables are examined using cross-tabulations and measures of association (e.g., Cramer’s V).
- Continuous variables are examined using ANOVA.
- Interpretation emphasizes substantive differences and patterns rather than statistical significance alone.

**3.3 Stratified and Comparative Analyses**

Where sample sizes permit, additional analyses compare:

- Persistently disconnected versus intermittently disconnected individuals
- Individuals disconnected within the region versus outside the region
- Individuals who transition into meaningful workforce participation versus those who do not

These comparisons help identify which characteristics are associated with stability, recovery, and regional retention.

<br>

## Framing and Interpretation

This methodology prioritizes a trajectory-based understanding of workforce participation, recognizing that early-career labor market experiences are dynamic rather than static. By combining persistence measures, regional employment location, and job-holding intensity, the analysis aims to provide actionable insights for workforce development and labor force planning in Southeast Minnesota.

<br>

# Data prep

## Step 1: Analytic population

The first step is filtering for our population;

- `ps.grad` = "No"

```{r}
analytic.pop <- master.original %>%
  filter(ps.grad == "No",
         grad.year.5 != "After 2023") 

master.original$sat

kable(head(analytic.pop))
kable(names(analytic.pop))
```

## Step 2: Identify Workforce State for 5 Years After Formal Education

Next, we need to pull out all the individuals in which we have five years of workforce categories after they either graduate from high school or the last year in which they exited post-secondary attendance.

After that, we will categorize their workforce participation state into the following;

- Sustained WF - Southeast
- Sustained WF - MN
- Non-sustained WF
- No MN emp record.

I already created the annual workforce participation states for every individuals one year after graduation through 15 years after graduation. I will bring that list in and filter it by our analytic population.

```{r}
annual.wf.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  select(PersonID, grad.year, states) %>%
  mutate(grad.year = fct_relevel(grad.year, "grad.year.0", "grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5", "grad.year.6", "grad.year.7", "grad.year.8", "grad.year.9", "grad.year.10", "grad.year.11", "grad.year.12", "grad.year.13", "grad.year.14", "grad.year.15")) 

no_ps_attendance <- analytic.pop %>%
  filter(attended.ps == "No") %>%
  select(PersonID) %>%
  left_join(annual.wf.states, by = "PersonID") %>%
  filter(states != "After 2023",
         states != "Attending ps",
         grad.year %in% c("grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5")) %>%
  group_by(PersonID) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n == 5) %>%
  mutate(states = as_factor(states))

last_ps_attendance <- analytic.pop %>%
  filter(attended.ps == "Yes") %>%
  select(PersonID) %>%
  left_join(annual.wf.states, by = "PersonID") %>%
  filter(states != "After 2023") %>%
  mutate(numeric.grad.year = str_sub(grad.year, -2, -1),
         numeric.grad.year = str_replace(numeric.grad.year, "\\.", ""),
         numeric.grad.year = as.integer(numeric.grad.year)) %>%
  filter(states == "Attending ps") %>%
  group_by(PersonID) %>%
  summarize(last_ps_year = max(numeric.grad.year, na.rm = TRUE)) %>%
  ungroup() 

ps_attendance <- annual.wf.states %>%
  right_join(last_ps_attendance, by = "PersonID") %>%
  filter(states != "After 2023") %>%
  mutate(numeric.grad.year = str_sub(grad.year, -2, -1),
         numeric.grad.year = str_replace(numeric.grad.year, "\\.", ""),
         numeric.grad.year = as.integer(numeric.grad.year),
         final.year = last_ps_year + 5) %>%
  group_by(PersonID) %>%
  filter(numeric.grad.year <= final.year,
         numeric.grad.year > last_ps_year) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n == 5) %>%
  select(PersonID, grad.year, states, n) 

master.annual.wf.five.year.states <- ps_attendance %>%
  rbind(no_ps_attendance) %>%
  mutate(states = str_replace(states, "Not meaningful, not attending ps", "Non-sustained WF"),
         states = str_replace(states, "No MN emp record, not attending ps", "No MN emp record"),
         states = ifelse(states %in% c("Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region"), "Sustained WF - SE", states),
         states = str_replace(states, "Meaningful emp MN", "Sustained WF - MN"),
         states = fct_relevel(states, "Sustained WF - SE", "Sustained WF - MN", "Non-sustained WF", "No MN emp record")) %>%
  select(-n) %>%
  group_by(PersonID) %>%
  arrange(grad.year) %>%
  mutate(year.after.exit = seq(n())) %>%
  ungroup()


kable(head(master.annual.wf.five.year.states))

kable(names(master.annual.wf.five.year.states))

```

We now have a dataset consisting of `r comma(nrow(master.annual.wf.five.year.states))` individuals that meet the criteria that we want to examine along with their workforce categories from 1 to 5 years after exiting formal education.

<br>

## Step 3: Workforce Disconnection Indicators

Next we add another category to determine the level of disconnection, if any, for each individual. Disconnected is defined as any individual that was categorized as not having meaningful workforce participation or no Minnesota employment record in a given year one through five years after exiting formal education.

The level of disconnection is defined as the following;

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Broadly Disconnected:** Disconnected in 3 of the first 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

```{r}
disconnected.level <- master.annual.wf.five.year.states %>%
  mutate(disconnected = ifelse(states %in% c("Non-sustained WF", "No MN emp record"), "Disconnected", "Attached"),
         disconnected = as.factor(disconnected)) %>%
  group_by(PersonID, disconnected) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, disconnected, fill = list(n = 0)) %>%
  filter(disconnected == "Disconnected") %>%
  mutate(disconnected.cat = ifelse(n == 0, "Consistently attached",
                                   ifelse(n %in% c(1,2), "Intermittently disconnected",
                                          ifelse(n == 3, "Broadly disconnected",
                                                 ifelse(n %in% c(4,5), "Persistently disconnected", "A")))),
         disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected", "Persistently disconnected"))

kable(names(disconnected.level))
kable(head(disconnected.level))
```

The dataset now has `r comma(nrow(disconnected.level))` individuals that did not graduate college AND have 5 years of workforce categories immediately after exiting formal education categorized by workforce disconnection levels.

## **Step 4 & 5: Regional Location of Not Meaningful Employment**

For years in which individuals are categorized as **not meaningfully attached to the workforce**, employment location is further classified as: - Within SE Minnesota - Outside Southeast Minnesota (elsewhere in Minnesota) - Both (simultaneous employment inside and outside the region)

This classification allows identification of regional underemployment versus regional disengagement.

The population of individuals we need are the following;

- Did not graduate college
- For each individual categorize whether they had not meaningful workforce participation in the region, not in the region, or both.

We will use the `master.annual.wf.five.year.states` dataset as the starting point. We will have to have the year they graduated high school so when I match up employment records I'm only using employment records for the first 5 years after exiting formal education.

```{r}
hs.grad.year <- read_csv("Data/SLEDS/K12 Enrollment/K12Enrollment-master.csv")

analytic.pop.annual.wf.states.hs.grad.year <- master.annual.wf.five.year.states %>%
  left_join(hs.grad.year[,c(1,8)], by = "PersonID")  %>%
  rename(after.ed.year = grad.year.x,
         hs.grad.year = grad.year.y) %>%
  mutate(numeric.after.ed.year = str_sub(after.ed.year, -2, -1),
         numeric.after.ed.year = str_replace(numeric.after.ed.year, "\\.", ""),
         numeric.after.ed.year = as.integer(numeric.after.ed.year),
         emp.year = hs.grad.year + numeric.after.ed.year) 
  
kable(head(analytic.pop.annual.wf.states.hs.grad.year))

kable(names(analytic.pop.annual.wf.states.hs.grad.year))
```

Now we need to bring in the employment data for Southeast graduates and then filter for PersonID that are in the newly created dataset. The columns we are going to keep from the employment dataset are the following;

- PersonID: Identifies the graduate from Southeast Minnesota
- TimeID: Last day of fiscal quarter
- EmploymentOrganizationID: An internal identification for each employer.
- EmployeeQuarterlyHoursWorked: The total number of hours worked by an employee at an employer’s location during a specific calendar quarter
- EmployeeQuarterlyWages: The total wages earned by an employee at an employer’s location during a specific calendar quarter
- EmployeeTenure: The total number of quarters a worker has been employeed by an organization at a specific location, up to an including the current calendar quarter
- CountyName: The name of the county corresponding to the County and MNCounty Codes for this Location.

```{r}
employment.original <- read_csv("Data/SLEDS/Employment/Employment-Southern.csv") %>%
  drop_na(PersonID)

emp.year <- employment.original %>%
  filter(!is.na(TimeID)) %>%
  mutate(emp.year = str_sub(TimeID, 1, 4),
         emp.year = as.integer(emp.year),
         quarter = str_sub(TimeID, 5,6),
         quarter = ifelse(quarter == "03", 1,
                          ifelse(quarter == "06", 2,
                                 ifelse(quarter == "09", 3,
                                        ifelse(quarter == "12", 4, 0))))) %>%
  select(PersonID, EmploymentOrganizationID, EmployeeQuarterlyHoursWorked, EmployeeTenure, CountyName, emp.year, quarter) %>%
  mutate(CountyName = str_replace(CountyName, " County", ""),
         CountyName = str_replace(CountyName, " County", ""),
         CountyName = str_replace(CountyName, "Saint Louis", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Duluth", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Minneapolis", "Hennepin"),
         CountyName = str_replace(CountyName, "City of St. Paul", "Ramsey"),
         CountyName = str_replace(CountyName, "Shakopee Mdewakanton Community", "Scott"),
         CountyName = str_replace(CountyName, "White Earth Tribal Counci", "Mahnomen")) %>%
  left_join(counties.regions.1[,c(1,2,10)], by = c("CountyName" = "Name"))

not.meaningful.pop.emp.year <- analytic.pop.annual.wf.states.hs.grad.year %>%
  filter(states == "Non-sustained WF") %>%
  select(PersonID, emp.year)

not.meaningful.emp.quarter <- not.meaningful.pop.emp.year %>%
  left_join(emp.year, by = c("PersonID", "emp.year")) %>%
  mutate(project.pr = ifelse(is.na(project.pr), "Unclassified", as.character(project.pr)),
         in_region = (project.pr == "Southern"),
         unclassified_region = (project.pr == "Unclassified"),
         out_region = c(!in_region & !unclassified_region)) %>%
  group_by(PersonID, emp.year, quarter) %>%
  summarise(n.emp = n(),
         any_in_quarter = any(in_region, na.rm = TRUE),
         any_out_quarter = any(out_region, na.rm = TRUE),
         any_unclassified_quarter = any(unclassified_region, na.rm = TRUE),
         any_mixed_qtr =  (sum(c(any_in_quarter, any_out_quarter, any_unclassified_quarter), na.rm = TRUE) >= 2)) %>%
  ungroup()

not.meaningful.emp.annual <- not.meaningful.emp.quarter %>%
  group_by(PersonID, emp.year) %>%
  summarize(multiple_emp_same_quarter = any(n.emp > 1),
            max_emp_in_qtr = max(n.emp),
            any_in_region = any(any_in_quarter, na.rm = TRUE),
            any_out_region = any(any_out_quarter, na.rm = TRUE),
            any_unclassified_region = any(any_unclassified_quarter, na.rm = TRUE),
            any_mixed_region = any(any_mixed_qtr, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(region_location_year = case_when(any_in_region & !any_out_region & !any_unclassified_region ~ "in_only", !any_in_region & any_out_region & !any_unclassified_region ~ "out_only", !any_in_region & !any_out_region & any_unclassified_region  ~ "unclassified_only", TRUE ~ "mixed" )) %>%
  mutate(
    mixed_type = case_when(
      region_location_year != "mixed" ~ NA_character_,
      any_in_region & any_out_region & !any_unclassified_region ~ "in_out",
      any_in_region & any_unclassified_region & !any_out_region ~ "in_unclassified",
      any_out_region & any_unclassified_region & !any_in_region ~ "out_unclassified",
      any_in_region & any_out_region & any_unclassified_region  ~ "in_out_unclassified"
    ),
    both_structure = case_when(region_location_year != "mixed" ~ NA_character_,
                               any_mixed_region ~ "simultaneous",
                               !any_mixed_region ~ "sequential",
                               TRUE ~ NA_character_)) %>%
  select(-any_in_region, -any_out_region, -any_unclassified_region, - any_mixed_region)


kable(head(not.meaningful.emp.annual))

kable(names(not.meaningful.emp.annual))
```

We have created a dataset with `r comma(nrow(not.meaningful.emp.annual))` individuals that do not have meaningful workforce participation for any given year up to 5 years after graduating high school that also provides if they had multiple employers in one quarter, the max number of different employers in one quarter for that employment year, and whether the employer was located in the region, outside the region, unclassified, or a mixture of these categories. Another column further refines the "mixed" category into whether it was sequential - meaning the mix occurred in separate quarters - or simultaneous - meaning there was a mix within the same quarter.

Next, we will piece this back together with the entire analytic population and create a complete annual workforce state for all individuals that did not complete post-secondary and has 5 years of workforce categories after graduating high school.

```{r}
Master.not.meaningful <- not.meaningful.emp.annual %>%
  left_join(hs.grad.year[,c(1, 8)], by = "PersonID") %>%
  rename(hs.grad.year = grad.year) %>%
  mutate(grad.year = emp.year - hs.grad.year,
         grad.year = paste("grad.year.", grad.year, sep = "")) %>%
  complete(PersonID, grad.year) %>%
  mutate(grad.year = fct_relevel(grad.year, "grad.year.0", "grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5", "grad.year.6", "grad.year.7", "grad.year.8", "grad.year.9", "grad.year.10", "grad.year.11", "grad.year.12", "grad.year.13", "grad.year.14", "grad.year.15"))


Master.annual.wf.states.disconnected <- master.annual.wf.five.year.states %>%
  left_join(Master.not.meaningful, by = c("PersonID", "grad.year")) %>%
  unite("wf.states.refined",
        states, region_location_year, mixed_type, both_structure,
        sep = "_",
        na.rm = TRUE) %>%
  mutate(year.after.exit = paste("after.ed.", year.after.exit, sep = "")) %>%
  select(PersonID, year.after.exit, wf.states.refined) %>%
  pivot_wider(names_from = year.after.exit, values_from = wf.states.refined) %>%
  left_join(disconnected.level[,c(1,4)], by = "PersonID")

kable(head(Master.annual.wf.states.disconnected)) 

kable(names(Master.annual.wf.states.disconnected))
```

Okay, now I need to bring this all together.

```{r}
master <- analytic.pop %>%
  right_join(Master.annual.wf.states.disconnected, by = "PersonID") %>%
  mutate(disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected", "Persistently disconnected"))

kable(head(master))

kable(names(master))

write_csv(master, "Data/Claude/Non-college deeper/Master-dataset.csv")
```

The data set has `r comma(nrow(master))` individuals in it. These individuals did not complete post-secondary and had workforce categories for 5 years after exiting formal education.

<br>

# Descriptive analysis

## **Distribution of Workforce Disconnection**

Sample: N = 32,878

**Distribution**

| Category                    | n      | \%    |
|-----------------------------|--------|-------|
| Consistently attached       | 5,326  | 16.2% |
| Intermittently disconnected | 8,407  | 25.6% |
| Broadly disconnected        | 3,859  | 11.7% |
| Persistently disconnected   | 15,286 | 46.5% |

**Interpretation**

Nearly half of all non-college graduates in this population (46.5%) fall into the Persistently Disconnected category — disconnected from sustained workforce participation in 4 or 5 of the first 5 years after exiting formal education. This is by a wide margin the largest single category, more than double the next-largest group.

Only 16.2% of this population is Consistently Attached across all five years — meaning fewer than 1 in 6 non-college graduates show uninterrupted sustained workforce participation during this window.

The middle categories are smaller and asymmetric: Intermittently Disconnected (1–2 of 5 years, 25.6%) is more than double the size of Broadly Disconnected (exactly 3 of 5 years, 11.7%). This asymmetry suggests disconnection in this population tends toward one of two poles — either brief/occasional (1–2 years) or persistent/severe (4–5 years) — with comparatively fewer individuals landing in the exact middle (3 years).

**Substantive Pattern**

Taken together, roughly 58% of non-college graduates (Broadly + Persistently Disconnected) spend at least 3 of their first 5 years disconnected from sustained workforce participation, and the modal experience for this population is not stability but persistent disconnection. This sets a stark baseline for the rest of this analysis: workforce disconnection is not a marginal or rare outcome among non-completers in this region — it is the most common single trajectory. This will be an important frame for Step 3's bivariate work, since it means the "Persistently Disconnected" category is not a small at-risk subgroup but close to a plurality outcome for this entire population.

```{r}
data <- master %>%
  group_by(disconnected.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         data_id = seq(n()),
         disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected",  "Persistently disconnected"))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-cat.csv")

plot <- ggplot(data, aes(disconnected.cat, pct, fill = disconnected.cat)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(disconnected.cat, "\nN: ", comma(n), "\n%: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "black", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of non-college grads,\n2008 - 2018 high school grads")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 5, "RdYlBu"),
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 8, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      


```

<br>

## **Regional Attachment and Job Structure**

Among individuals experiencing workforce disconnection, summary statistics examine: - The location of not meaningful employment (within region, outside region, both) - Average and median number of employers per year - Prevalence of multiple job holding

These summaries clarify whether disconnection reflects local underemployment, out-of-region work, or fragmented employment patterns.

Let's start with the distribution of the various workforce states with the newly defined categories within "Not meaningful" but not all of them because it will get too messy. Let's first do the following categories;

1.  Attending ps
2.  Meaningful emp region
3.  Meaningful emp MN
4.  Not meaningful
    1.  In only
    2.  Out only
    3.  Unclassified only
    4.  Mixed
5.  No MN emp record

**Descriptive Distribution: Annual Workforce Participation Status, Years 1–5** **Non-College Graduates, Southeast Minnesota**

Sample: N = 32,878 per year (same population tracked annually)

**Distribution by Year**

| Status                                 | Year 1 | Year 2 | Year 3 | Year 4 | Year 5 |
|------------|------------|------------|------------|------------|------------|
| Sustained WF – SE (region)             | 24.9%  | 28.1%  | 29.5%  | 30.2%  | 30.5%  |
| Sustained WF – MN (elsewhere in-state) | 9.8%   | 11.3%  | 12.7%  | 14.0%  | 15.3%  |
| Non-sustained, in-region only          | 19.9%  | 16.1%  | 13.6%  | 12.2%  | 11.1%  |
| Non-sustained, out-of-region only      | 6.6%   | 6.6%   | 6.2%   | 6.2%   | 5.8%   |
| Non-sustained, unclassified only       | 0.6%   | 0.5%   | 0.5%   | 0.4%   | 0.3%   |
| Non-sustained, mixed (in + out)        | 12.4%  | 10.3%  | 8.7%   | 7.5%   | 6.4%   |
| No MN employment record                | 25.9%  | 27.2%  | 28.9%  | 29.5%  | 30.6%  |

**Interpretation**

Two categories grow steadily and roughly in parallel across all five years: Sustained WF – SE grows from 24.9% to 30.5%, and No MN employment record grows from 25.9% to 30.6%. By Year 5, these are the two largest categories and are nearly identical in size (30.5% vs. 30.6%) — together accounting for over 61% of this population. This is a notable pattern worth naming directly: the two most common trajectories for non-completers are converging toward roughly equal size, and they represent opposite ends of the outcome spectrum — sustained regional employment on one hand, complete disappearance from Minnesota's employment record on the other.

Sustained WF – MN (elsewhere in-state) also grows steadily, from 9.8% to 15.3% — smaller than Sustained WF – SE at every year, but growing at a similar rate proportionally.

All four "non-sustained" categories shrink over time, and shrink at different rates: Non-sustained mixed (in + out) nearly halves (12.4% → 6.4%), the steepest decline of any category. Non-sustained in-region only also nearly halves (19.9% → 11.1%). Non-sustained out-of-region only and unclassified-only are comparatively stable and small throughout (staying in the 5.8–6.6% and 0.3–0.6% ranges, respectively).

**Substantive Pattern**

The clearest story here is a gradual sorting process: individuals in unstable ("non-sustained") employment situations in Year 1 appear to resolve, over the following four years, into one of the two more stable poles — either sustained regional/in-state employment or complete absence from Minnesota's employment record — rather than remaining in an unstable middle state. The share in any non-sustained category drops from about 39.5% in Year 1 to about 23.6% by Year 5, while the combined "resolved" categories (Sustained SE + Sustained MN + No MN record) grow from about 60.5% to about 76.4% over the same window.

This has a direct connection to the disconnection persistence categories from the first table: since "No MN employment record" and "Non-sustained" are both disconnection-qualifying statuses, and No MN record is the fastest-growing single category here, this table helps explain why Persistently Disconnected was the largest category overall (46.5%) — a meaningful share of this population isn't cycling between sustained and non-sustained employment, but is drifting toward or starting in complete absence from the state's employment record, a status that appears to be "sticky" rather than transitional.

```{r}
wf.states.factors <- c("Attending ps", "Sustained WF - SE", "Sustained WF - MN", "Non-sustained WF_in_only", "Non-sustained WF_out_only", "Non-sustained WF_unclassified_only", "Non-sustained WF - mixed", "No MN emp record")

data <- Master.annual.wf.states.disconnected %>%
  select(PersonID:after.ed.5) %>%
  pivot_longer(names_to = "grad.year", values_to = "wf.states", 2:ncol(.)) %>%
  mutate(wf.states = ifelse(wf.states %in% c("Non-sustained WF_mixed_in_out_simultaneous", "Non-sustained WF_mixed_in_out_sequential", "Non-sustained WF_mixed_in_unclassified_simultaneous" , "Non-sustained WF_mixed_in_out_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_sequential", "Non-sustained WF_mixed_in_unclassified_sequential", "Non-sustained WF_mixed_in_out_unclassified_sequential"), "Non-sustained WF - mixed", wf.states),
         wf.states = fct_relevel(wf.states, wf.states.factors)) %>%
  group_by(grad.year, wf.states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-pattern.csv")

plot <- ggplot(data, aes(wf.states, pct, fill = wf.states)) +
  facet_wrap(~grad.year, ncol = 2) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nWorkforce category: ", wf.states, "\nN: ", comma(n), "\nPercent: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of total building construction establishments")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = c("#8c510a", "#bf812d", "#dfc27d", "#c7eae5", "#80cdc1", "#35978f", "#01665e", "black"),
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "none",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

**Descriptive Distribution: Refined Non-Sustained Employment Patterns, Years 1–5** **Non-College Graduates, Southeast Minnesota** **(Denominator: Non-sustained + No MN Record subset only, excluding Sustained SE/MN)**

**Distribution by Year (% of this subset)**

| Status                                  | Year 1 | Year 2 | Year 3 | Year 4 | Year 5 |
|------------|------------|------------|------------|------------|------------|
| Non-sustained, in-region only           | 30.5%  | 26.5%  | 23.5%  | 21.9%  | 20.5%  |
| Non-sustained, out-of-region only       | 10.0%  | 10.8%  | 10.7%  | 11.0%  | 10.8%  |
| Non-sustained, unclassified only        | 0.9%   | 0.8%   | 0.8%   | 0.7%   | 0.5%   |
| Mixed in+out, simultaneous              | 10.6%  | 9.5%   | 8.3%   | 7.6%   | 6.6%   |
| Mixed in+unclassified, simultaneous     | 1.3%   | 1.2%   | 1.0%   | 0.8%   | 0.6%   |
| Mixed out+unclassified, simultaneous    | 0.7%   | 0.7%   | 0.6%   | 0.4%   | 0.5%   |
| Mixed in+out+unclassified, simultaneous | 1.5%   | 1.1%   | 1.2%   | 0.9%   | 0.8%   |
| Mixed in+out, sequential                | 4.0%   | 3.5%   | 3.1%   | 2.9%   | 2.7%   |
| Mixed in+unclassified, sequential       | 0.6%   | 0.5%   | 0.5%   | 0.4%   | 0.3%   |
| Mixed out+unclassified, sequential      | 0.3%   | 0.3%   | 0.3%   | 0.2%   | 0.2%   |
| Mixed in+out+unclassified, sequential   | 0.05%  | 0.1%   | 0.07%  | 0.05%  | 0.02%  |
| No MN employment record                 | 39.6%  | 44.8%  | 50.0%  | 53.0%  | 56.5%  |

**Interpretation**

Within this subset (everyone not sustainably employed), "No MN employment record" grows from a plurality in Year 1 (39.6%) to a clear majority by Year 5 (56.5%) — the single dominant and fastest-growing category by a wide margin. Every other category in this table shrinks in relative share over the same window.

The granular "mixed" categories (simultaneous and sequential combinations of in/out/unclassified employment) are individually small at every year — none exceeds 10.6%, and most are under 2%. All of them shrink over time, several to near-negligible levels by Year 5 (e.g., mixed in+out+unclassified sequential drops to 0.02%).

**Substantive Pattern**

Your instinct here is largely right — this refined breakdown doesn't surface a new pattern so much as it confirms and sharpens the one from the previous table. The finer-grained "mixed simultaneous vs. sequential" distinctions are mostly small, shrinking, and not obviously differentiated from one another in trajectory (they all decline at broadly similar rates). I don't see a substantively distinct story hiding in the simultaneous/sequential split specifically — both decline together, and neither stands out as behaving differently from the other.

What this table does add, and does show clearly, is the magnitude of the shift toward "No MN employment record" *within the already-disconnected population specifically*: by Year 5, more than half of everyone who isn't sustainably employed has no Minnesota employment record at all, versus a plurality in Year 1. That's a sharper, more precise version of the pattern from the previous table, and probably the one number worth carrying forward from this refined table rather than the full granular breakdown.

**Recommendation**

For Step 2 write-up purposes, I'd suggest collapsing the "mixed" subcategories back into a single "mixed" group (as in the previous, less granular table) and just highlighting the "No MN record" share-of-subset trend from this table as a supporting statistic — the full 12-category breakdown adds detail without adding a new finding, and several of the sequential/unclassified combinations are small enough that reporting them individually risks looking more precise than the underlying pattern warrants.

```{r}
refined.disconnected.factors <- c( "Non-sustained WF_in_only",  "Non-sustained WF_out_only", "Non-sustained WF_unclassified_only", "Non-sustained WF_mixed_in_out_simultaneous", "Non-sustained WF_mixed_in_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_simultaneous", "Non-sustained WF_mixed_in_out_unclassified_simultaneous","Non-sustained WF_mixed_in_out_sequential", "Non-sustained WF_mixed_in_unclassified_sequential", "Non-sustained WF_mixed_out_unclassified_sequential", "Non-sustained WF_mixed_in_out_unclassified_sequential", "No MN emp record")

data <- Master.annual.wf.states.disconnected %>%
  select(PersonID:after.ed.5) %>%
  pivot_longer(names_to = "grad.year", values_to = "wf.state", 2:ncol(.)) %>%
  filter(!wf.state %in% c("Sustained WF - SE", "Sustained WF - MN", "Attending ps")) %>%
  mutate(wf.state = fct_relevel(wf.state, refined.disconnected.factors)) %>%
  group_by(grad.year, wf.state) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-refined-patterns.csv")

plot <- ggplot(data, aes(grad.year, pct, fill = wf.state)) +
  facet_wrap(~wf.state, ncol = 2) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nWorkforce category: ", wf.state, "\nN: ", comma(n), "\nPercent: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of disconnected workforce states")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 8, height_svg = 12) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      
```

<br>

# Separating Non-sustained and No MN Emp Record

It's clear from the descriptive statistics that it's necessary to separate the Non-sustained from the MN Employment records to get a sense of how these two overlap or not. We are going to categorize just like we did before, except we will have to separate categories of "disconnection";

1.  non-sustained.persistence.cat:
    - 0 years = none;
    - 1-2 = intermittent;
    - 3 = broad;
    - 4-5 = persistent
2.  no_record.persistence.cat
    - same four buckets, applied to the no-record count instead

## Data Prep

```{r}
non_sustained.level <- master.annual.wf.five.year.states %>%
  mutate(non_sustained = ifelse(states == "Non-sustained WF", "Non-sustained", "Other"),
         non_sustained = as.factor(non_sustained)) %>%
  group_by(PersonID, non_sustained) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, non_sustained, fill = list(n = 0)) %>%
  filter(non_sustained == "Non-sustained") %>%
  mutate(non_sustained.cat = ifelse(n == 0, "Never non-sustained",
                                    ifelse(n %in% c(1,2), "Intermittently non-sustained",
                                           ifelse(n == 3, "Broadly non-sustained",
                                                  ifelse(n %in% c(4,5), "Persistently non-sustained", "A")))),
         non_sustained.cat = fct_relevel(non_sustained.cat, "Never non-sustained", "Intermittently non-sustained", "Broadly non-sustained", "Persistently non-sustained"))

no_record.level <- master.annual.wf.five.year.states %>%
  mutate(no_record = ifelse(states == "No MN emp record", "No record", "Other"),
         no_record = as.factor(no_record)) %>%
  group_by(PersonID, no_record) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, no_record, fill = list(n = 0)) %>%
  filter(no_record == "No record") %>%
  mutate(no_record.cat = ifelse(n == 0, "Never no-record",
                                ifelse(n %in% c(1,2), "Intermittently no-record",
                                       ifelse(n == 3, "Broadly no-record",
                                              ifelse(n %in% c(4,5), "Persistently no-record", "A")))),
         no_record.cat = fct_relevel(no_record.cat, "Never no-record", "Intermittently no-record", "Broadly no-record", "Persistently no-record"))

master.sep <- Master.annual.wf.states.disconnected %>%
  left_join(non_sustained.level, by = "PersonID") %>%
  left_join(no_record.level, by = "PersonID") %>%
  rename(non.sustained.n = n.x,
         no.record.n = n.y)

write_csv(master.sep, "Data/Claude/Non-college deeper/master-separated.csv")

```

**Descriptive Distribution: Non-Sustained Employment Persistence** **Non-College Graduates, Southeast Minnesota (Years 1–5 Post-Exit)**

Sample: N = 32,878

**Distribution**

| Category                     | n      | \%     |
|------------------------------|--------|--------|
| Never non-sustained          | 11,448 | 34.82% |
| Intermittently non-sustained | 13,021 | 39.60% |
| Broadly non-sustained        | 4,108  | 12.49% |
| Persistently non-sustained   | 4,301  | 13.08% |

**Interpretation**

Once isolated from "No MN employment record," non-sustained employment on its own shows a very different shape than the original blended `disconnected.cat` variable. Here, the two largest categories are Never non-sustained (34.8%) and Intermittently non-sustained (39.6%) — together nearly three-quarters (74.4%) of the population. Broadly and Persistently non-sustained are much smaller and nearly equal in size (12.5% and 13.1%, respectively), together accounting for only about 25.6%.

This is a sharp contrast with the original blended distribution, where "Persistently disconnected" was the single largest category at 46.5%. Once no-record years are pulled out, persistent non-sustained employment specifically (4–5 years) drops to just 13.1% — suggesting that a large share of what was driving the original "Persistently disconnected" figure was the no-record component, not unstable employment as such.

**Substantive Pattern**

This confirms the value of separating the two constructs. Most non-completers who experience *any* non-sustained employment experience it only intermittently (1–2 years) rather than persistently — genuine long-term instability in employment itself (as opposed to absence from the record) appears to affect roughly 1 in 4 non-completers (Broad + Persistent combined), not the roughly 46% implied by the blended category. This reframes the earlier finding: persistent workforce detachment in this population looks like it's driven more by disappearance from Minnesota's employment record than by cycling through unstable jobs. The next table (no_record.persistence.cat) should make this contrast concrete — I'd expect that distribution to look much more like the original blended one, given how dominant "No MN record" was in the year-by-year and refined tables.

```{r}
non_sustained.dist <- master.sep %>%
  group_by(non_sustained.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         pct = scales::percent(pct))

kable(non_sustained.dist, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

**Descriptive Distribution: No-MN-Record Persistence** **Non-College Graduates, Southeast Minnesota (Years 1–5 Post-Exit)**

Sample: N = 32,878

**Distribution**

| Category                 | n      | \%    |
|--------------------------|--------|-------|
| Never no-record          | 18,404 | 56.0% |
| Intermittently no-record | 5,576  | 17.0% |
| Broadly no-record        | 1,977  | 6.0%  |
| Persistently no-record   | 6,921  | 21.1% |

**Interpretation**

The no-record distribution has a different shape from both the blended category and the non-sustained-only category. A majority (56.0%) never have a "No MN employment record" year at all — larger than the "Never" share for non-sustained employment (34.8%). At the same time, Persistently no-record (21.1%) is larger than Persistently non-sustained (13.1%) — so while more people avoid no-record status entirely, those who do experience it persistently make up a larger share than persistent non-sustained employment does.

This makes the no-record distribution more bimodal than the non-sustained distribution: most people either never experience it (56.0%) or experience it persistently (21.1%), with comparatively few landing in the intermittent (17.0%) or broad (6.0%) middle categories — together only 23.0%. Non-sustained employment, by contrast, was more continuous, with Intermittent as its single largest category (39.6%).

**Substantive Pattern — a correction to my earlier framing**

Comparing all three tables together changes the story I gave in my previous message, and it's worth stating clearly: I had suggested the original blended "Persistently disconnected" figure (46.5%) was being driven mainly by the no-record component. That doesn't hold up — persistent no-record alone is only 21.1%, and persistent non-sustained alone is only 13.1%. Neither component individually gets close to explaining a 46.5% blended "persistent" rate.

What this actually implies is that a substantial share of people counted as "Persistently disconnected" in the original blended measure are experiencing a *mix* of non-sustained and no-record years across the 5-year window — some years unstably employed, other years absent from the record entirely — rather than being persistently stuck in one single state. That's a more interesting and more useful finding than either component distribution shows on its own: workforce disconnection in this population isn't just "unstable employment" or "absence from the record" as separate, stable trajectories — for a meaningful share of people, it's a genuine back-and-forth between the two.

This makes the case for the 4×4 cross-tab of non_sustained.cat × no_record.cat even stronger than before — that table would show directly how much overlap exists between the two constructs and could confirm or disconfirm this "switching between states" interpretation with actual joint distribution numbers, rather than inference from the marginals alone.

```{r}
no_record.dist <- master.sep %>%
  group_by(no_record.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

kable(no_record.dist, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

**Decomposing "Persistently Disconnected" by Driver (N = 18,063)**

The table below showsonly the people who fall into "Persistently disconnected" under our *original* combined measure — meaning across their 5 years, they were disconnected (non-sustained OR no-record, in any mix) in 4 or 5 of those years.

The categories are defined as;

- **"Non-sustained driven"** = this person's disconnected years were mostly/entirely *unstable employment* (non-sustained), not disappearing from the record. Example: someone with 4 years of non-sustained work and 0 years of no-record. They were disconnected in the combined sense, but they were always at least working, just not steadily.
- **"No-record driven"** = the opposite — this person's disconnected years were mostly/entirely *no employment record at all*, not instability. Example: someone with 0 years non-sustained and 4 years no-record. They weren't cycling through shaky jobs; they mostly just weren't showing up in Minnesota's system at all.
- **"Mixed (neither individually persistent)"** = this person doesn't have 4+ years in *either* single category, but their two counts still *add up* to 4 or 5 disconnected years total. Example: 2 years non-sustained + 2 years no-record = 4 disconnected years total (enough to be "persistently disconnected" overall), but only 2 in each — not enough on its own to count as "persistent" in either individual measure. This person alternated between the two problems rather than being stuck in one.

**Descriptive Distribution: Driver of Persistent Disconnection** **Among Persistently Disconnected Non-Completers (4–5 of 5 Years Disconnected)**

Sample: N = 15,286 (matches the "Persistently disconnected" count from the original blended distribution)

**Distribution**

| Driver                                  | n     | \%    |
|-----------------------------------------|-------|-------|
| No-record driven                        | 6,921 | 45.3% |
| Non-sustained driven                    | 4,301 | 28.1% |
| Mixed (neither individually persistent) | 4,064 | 26.6% |

**Interpretation**

Among those who are persistently disconnected overall, No-record driven is the largest single group (45.3%) — nearly half of everyone in this category is disconnected primarily because they disappear from Minnesota's employment record, not because they cycle through unstable jobs. Non-sustained driven is smaller (28.1%), and Mixed is close behind at 26.6%.

This resolves the open question from the previous table cleanly. Note that the n's here confirm exactly: 6,921 No-record driven matches the full Persistently no-record count from the prior table, and 4,301 Non-sustained driven matches the full Persistently non-sustained count — meaning every person who is persistent in either individual measure is, by definition, also persistently disconnected overall (which makes sense, since 4+ years in either category alone guarantees 4+ combined years). The Mixed category (4,064) is the genuinely new information this table adds: it isolates the subgroup that doesn't show up as "persistent" in either individual measure, but still accumulates enough combined disconnected years to count as persistently disconnected overall.

**Substantive Pattern**

This confirms and sharpens the "switching between states" interpretation I raised in my previous message, but refines it: switching between the two problems is real and sizable (26.6% of the persistently disconnected group), but it is not the dominant story — it's the smallest of the three drivers, not the largest. The dominant story is that persistent disconnection in this population is most often a single, consistent problem rather than an alternating one, and specifically, it's more often driven by absence from the employment record (45.3%) than by employment instability (28.1%).

This is a meaningful correction to keep in mind for how "persistent disconnection" gets framed going forward: it is not primarily a story of people bouncing between shaky jobs and gaps in the record. It's more accurately a story where, if someone is persistently disconnected, the single most likely reason is that they've largely stopped appearing in Minnesota's workforce system at all — which raises the same open question flagged earlier (and still unresolved): whether "No MN employment record" reflects leaving the state, exiting the labor force entirely, informal/unreported work, or some mix that this data can't distinguish. That distinction matters a great deal here, since it's now confirmed to be the single largest driver of the single largest disconnection category in this entire analysis.

```{r}
persistent_decomp <- master.sep %>%
  filter(disconnected.cat == "Persistently disconnected") %>%
  mutate(driver = case_when(
    non_sustained.cat == "Persistently non-sustained" & no_record.cat != "Persistently no-record" ~ "Non-sustained driven",
    no_record.cat == "Persistently no-record" & non_sustained.cat != "Persistently non-sustained" ~ "No-record driven",
    non_sustained.cat == "Persistently non-sustained" & no_record.cat == "Persistently no-record" ~ "Both persistent",
    TRUE ~ "Mixed (neither individually persistent)"
  )) %>%
  group_by(driver) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

kable(persistent_decomp, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

## **Full Summary: Descriptive Statistics on Workforce Disconnection** **Non-College Graduates, Southeast Minnesota (Years 1–5 Post-Exit)**

**Overview**

Sample: N = 32,878 non-completers, tracked annually across the first five years after exiting formal education (high school or postsecondary attendance, whichever is later). This descriptive stage moved through five successive tables, each refining the picture of what "workforce disconnection" actually consists of in this population.

**Step 1: The headline number — disconnection is not a marginal outcome**

Using the original blended definition (disconnected = non-sustained employment OR no MN record, in a given year), 46.5% of this population is Persistently disconnected (4–5 of 5 years), far larger than any other category (Intermittently disconnected: 25.6%; Broadly disconnected: 11.7%; Consistently attached: only 16.2%). This established the core framing for the whole analysis: persistent disconnection is close to the modal outcome for non-completers in this region, not a small at-risk subgroup.

**Step 2: The annual trajectory — a sorting process, not stability**

Looking year-by-year, two categories grow steadily and converge: Sustained WF – SE (24.9% → 30.5%) and No MN employment record (25.9% → 30.6%) end Year 5 at nearly identical size, together accounting for over 61% of the population. Every non-sustained subtype shrinks over the same window, most sharply for "mixed" and "in-region only" statuses. This suggested a sorting dynamic: people in unstable employment situations in Year 1 gradually resolve into one of two stable poles — sustained employment or disappearance from the record — rather than remaining in an unstable middle state.

**Step 3: The refined breakdown — confirms the sorting story, adds one precise number**

Disaggregating "non-sustained" into simultaneous/sequential/unclassified subtypes didn't surface a new pattern — all the granular mixed categories are small and decline together at similar rates. What it did confirm precisely: within the subset of people who are *not* sustainably employed, "No MN record" grows from a plurality (39.6%) in Year 1 to a clear majority (56.5%) by Year 5. This sharpened, rather than changed, the Step 2 finding.

**Step 4: Splitting the construct — the two problems have different shapes, and neither explains the blended number alone**

Separating "non-sustained persistence" from "no-record persistence" as two independent 5-year measures showed they behave differently: non-sustained persistence is more continuous (Intermittent is the largest category, 39.6%; only 13.1% are Persistently non-sustained), while no-record persistence is more bimodal (56.0% Never vs. 21.1% Persistently, with a small middle). Critically, neither component's "Persistent" share (13.1% and 21.1%) comes close to explaining the original blended 46.5% Persistently disconnected figure on its own — which meant a real question remained about how people were accumulating enough combined disconnected years to land in that blended category.

**Step 5: The driver breakdown — resolves the question, and revises the initial "switching" hypothesis**

Among people who are persistently disconnected overall, the driver breakdown showed: No-record driven (45.3%) \> Non-sustained driven (28.1%) \> Mixed/switching between the two (26.6%). This confirmed that persistent disconnection is most often a single, consistent problem rather than people alternating between instability and absence — and that when it is a single problem, it's disproportionately absence from the employment record rather than unstable employment.

**What the descriptive stage is telling us, overall**

1.  **Persistent disconnection is common, not rare** — it's close to a plurality outcome for non-completers in this population, not a tail-end risk group.
2.  **"Disconnection" is not one thing.** Unstable employment and absence from Minnesota's employment record behave as genuinely distinct trajectories with different shapes (continuous vs. bimodal) and different persistence rates — collapsing them into a single blended measure, as the original framing did, obscures more than it reveals.
3.  **Absence from the record, not employment instability, is the dominant driver** of persistent disconnection in this population, both in the year-by-year trajectory data (Step 2/3) and in the direct driver breakdown (Step 5).
4.  **There is a real "switching" subgroup**, but it's the smallest of the three driver types (26.6%) — worth tracking in Step 3's bivariate work, but not the central story.

<br>

# Relationship Between Individual Characteristics and Disconnection Categories

The next sections focus on exploring what relationships exist between individual characteristics and the different ways in which disconnection from the workforce is measured;

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Disconnected (Broad):** Disconnected in 3 of 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

```{r}
updated.master <- master %>%
  left_join(master.sep[,c(1,8:13)], by = "PersonID")

names(updated.master)
updated.master$sat
write_csv(updated.master, "Data/Claude/Non-college deeper/updated-master.csv")
```

<br>

## Demographic variables

**Overall pattern across all six variables**

All six relationships are statistically significant (p \< 0.001 in every case), but as with any analysis at this sample size (N = 32,878), significance alone isn't informative — by Cohen's conventional benchmarks for Cramér's V, five of the six variables fall in the "small effect" range, and only one (SpecialEdStatus) crosses the 0.10 threshold typically used to mark the boundary of a small effect for a 2-level predictor. As before, standardized effect size and raw percentage-point gaps tell different stories here, and both are reported below.

**Economic Status** (χ² = 159.641, df = 3, Cramér's V = 0.070)

Economically disadvantaged students show lower consistent attachment (13.4% vs. 18.2%) and higher persistent disconnection (49.1% vs. 44.6%) than their non-disadvantaged peers — a 4.5 percentage-point gap in persistent disconnection. This is the expected direction, but it's the smallest raw gap of the six variables in this set.

**English Learner Status** (χ² = 301.256, df = 3, Cramér's V = 0.096)

English learners show a substantially higher persistent disconnection rate (59.9% vs. 44.8%, a 15.1-point gap) and lower consistent attachment (11.7% vs. 16.8%) than non-English learners. This is the largest raw gap among the four 2-level demographic variables (excluding RaceEthnicity), and it has the second-highest Cramér's V in this set as well — one of the few variables where the standardized and raw-gap rankings roughly agree.

**Gender** (χ² = 180.503, df = 3, Cramér's V = 0.074)

Female graduates show notably lower consistent attachment (13.2% vs. 18.3%) and higher persistent disconnection (49.1% vs. 44.6%) than male graduates — a 4.5-point gap, identical in magnitude to economic status. This mirrors the pattern from the original DV1/DV2 completer-focused analysis, where women showed a real employment-attachment disadvantage — here, restricted specifically to non-completers, the same direction of disadvantage holds.

**Non-English Home** (χ² = 170.809, df = 3, Cramér's V = 0.072)

Students from non-English-speaking homes show a sizable gap — 57.2% persistently disconnected vs. 45.3% for English-speaking homes, an 11.9-point difference — despite a Cramér's V (0.072) similar in size to economic status. As with English learner status, this variable likely captures overlapping population to some degree, and the two are probably not fully independent signals.

**RaceEthnicity** (χ² = 355.870, df = 12, Cramér's V = 0.060)

This shows by far the largest raw disparities of any variable in this set, and the most consistent pattern across categories. Every non-White group shows dramatically higher persistent disconnection and lower consistent attachment than White graduates:

- American Indian: 66.0% persistently disconnected, only 8.0% consistently attached — the most severe disparity of any group, and a 21.2-point gap from White graduates.
- Black: 62.8% persistently disconnected, 6.3% consistently attached — an 18.0-point gap from White graduates, and notably the lowest consistent-attachment rate of any group.
- Hispanic: 50.2% persistently disconnected, 14.3% consistently attached.
- Asian/Pacific Islander: 54.1% persistently disconnected, 13.3% consistently attached.
- White: 44.8% persistently disconnected, 17.1% consistently attached.

Despite these being the largest raw gaps in the entire set — larger than English learner, non-English home, or SpecialEdStatus — RaceEthnicity has the lowest Cramér's V of all six variables (0.060). This reflects sample composition: White graduates make up 84.1% of this non-completer population (27,649 of 32,878), so the overall standardized association is diluted even though every individual non-White subgroup shows a large, consistent disparity. This directly parallels the persistent Black-White (and here, especially American Indian) gap that ran through the DV1, DV2, and DV5 multivariate work on completers, and now shows up again at the purely bivariate level in a population that never completed any postsecondary credential — suggesting this disparity is not specific to the postsecondary pathway.

**SpecialEdStatus** (χ² = 421.200, df = 3, Cramér's V = 0.113)

Students with SpecialEd status show lower consistent attachment (11.8% vs. 17.4%) and higher persistent disconnection (56.9% vs. 43.6%, a 13.3-point gap). This is the only variable in the set whose Cramér's V crosses the conventional "small effect" threshold, making it the strongest single bivariate predictor of disconnection category by that standardized measure — consistent with SpecialEdStatus's role as one of the largest, most consistent demographic effects throughout the earlier DV1–DV5 and non-completer regression work as well.

**Synthesis: two different ways to rank these six variables**

| Ranked by Cramér's V (standardized strength) | Ranked by raw persistent-disconnection gap (substantive size) |
|------------------------------------|------------------------------------|
| 1\. SpecialEdStatus (0.113) | 1\. RaceEthnicity — AI vs. White (21.2 pts) |
| 2\. English learner (0.096) | 2\. RaceEthnicity — Black vs. White (18.0 pts) |
| 3\. Gender (0.074) | 3\. English learner (15.1 pts) |
| 4\. Non-English home (0.072) | 4\. SpecialEdStatus (13.3 pts) |
| 5\. Economic status (0.070) | 5\. Non-English home (11.9 pts) |
| 6\. RaceEthnicity (0.060) | 6\. Gender / Economic status (4.5 pts, tied) |

RaceEthnicity diverges most sharply between the two rankings — last by Cramér's V, first by raw gap — for the same reason as before: a numerically small subgroup (American Indian, Black) can carry a very large disparity while contributing relatively little to an overall standardized measure that's dominated by the large White majority. Neither ranking is "more correct." For a report focused on identifying which populations face the greatest workforce disconnection risk, the raw-gap ranking — led by American Indian and Black graduates — is the more policy-relevant framing, even though SpecialEdStatus would be the lead finding if effect size alone were used to prioritize discussion.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/english-learner.html")
```

<br>

#### `non.english.home`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/non-english-home.html")
```

<br>

## High school characteristics

**Overall pattern across all four variables**

This is a genuinely different result from the demographic variables. All four HS-characteristics variables here are technically significant at conventional thresholds (p \< 0.05), but three of the four are barely significant, and effect sizes across the board are far smaller than anything seen in the demographic set — Cramér's V for the two categorical variables (Dem_Desc, edr) falls below 0.03, an order of magnitude smaller than even the weakest demographic variable (RaceEthnicity, V = 0.060). For the two continuous variables (avg.unemp.rate, avg.wages.pct.state), the ANOVA effect sizes (η² ≈ 0.0003 for both) are negligible in any practical sense, even though the F-tests cross the p \< 0.05 line given this sample size. **The clearest overall takeaway from this variable group: high school geographic/economic context, at the bivariate level, shows essentially no meaningful relationship with disconnection persistence in this non-completer population.** This is a genuinely different story from the demographic variables, and worth stating plainly before the detail below.

**Dem_Desc** (χ² = 13.397, df = 6, Cramér's V = 0.014, p = 0.037)

The three urbanicity categories show almost no difference in persistent disconnection: Entirely urban (45.6%), Town/rural mix (45.5%), and Urban/town/rural mix (47.5%) are all within 2 percentage points of one another. This is the weakest relationship of any variable examined in this entire bivariate analysis so far, technically significant only because of the very large sample size. There is no substantively meaningful pattern here.

**edr** (χ² = 14.834, df = 3, Cramér's V = 0.021, p = 0.002)

EDR 10 – Southeast shows a slightly higher persistent disconnection rate (47.2%) than EDR 9 – South Central (45.0%), a 2.2-point gap — small, but the largest raw gap of the four HS-characteristics variables, and the strongest Cramér's V among them (still far below the "small effect" convention). Consistent attachment is nearly identical between the two regions (16.0% vs. 16.6%). This is a real but minor regional difference, not a strong signal.

**avg.unemp.rate** (F = 3.184, df = 3/32874, p = 0.023, η² ≈ 0.0003)

The omnibus ANOVA is significant, but the effect size is negligible — average local unemployment rate explains well under 0.1% of the variance in disconnection category. The Tukey post-hoc comparisons show only one pairwise difference reaching significance: Intermittently disconnected individuals experienced a higher average local unemployment rate than Consistently attached individuals (difference = 0.072 percentage points, adj. p = 0.013). No other pairwise comparison — including the one most relevant to the "persistent" story, Persistently disconnected vs. Consistently attached (estimate = 0.046, adj. p = 0.135) — reaches significance. This is a weak, isolated signal rather than a clear dose-response relationship across the four categories.

**avg.wages.pct.state** (F = 2.983, df = 3/32874, p = 0.030, η² ≈ 0.0003)

Same story as unemployment rate: statistically significant omnibus test, negligible effect size. The only significant Tukey pairwise comparison is Persistently disconnected vs. Intermittently disconnected (estimate = 0.0027, adj. p = 0.045) — persistently disconnected individuals came from high schools with slightly higher local wages relative to the state average than intermittently disconnected individuals. Broadly vs. Intermittently disconnected is marginal (adj. p = 0.075). No comparison involving Consistently attached reaches significance. As with unemployment rate, this doesn't describe a coherent pattern so much as one small, isolated pairwise difference.

**Synthesis**

Unlike the demographic variable group, there is no meaningful ranking to construct here — every variable in this set is weak by both standardized effect size and raw magnitude, and none produces a percentage-point gap larger than about 2–3 points (edr) or a substantively interpretable pairwise pattern (the two continuous variables). This is a legitimate, reportable null-ish result: at the bivariate stage, high school geographic and local-economic context does not meaningfully distinguish disconnection trajectories among non-completers in this region, in sharp contrast to the demographic variables just examined. Worth noting this doesn't necessarily mean HS context is irrelevant to the underlying mechanisms — it may interact with, or be masked by, the demographic and academic variables not yet in the model — but on its own, this variable group shows the weakest signal of any group likely to be examined in this analysis.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `edr`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/edr.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**Overall pattern across all nine variables**

This is the most substantively interesting variable group so far, for two reasons. First, several of these relationships are the strongest seen in this entire bivariate analysis — `total.cte.courses.taken` (η² ≈ 2.2%) and `ACTCompositeScore.filled` (η² ≈ 1.4%) both produce meaningfully larger effect sizes than anything in the demographic or HS-characteristics groups. Second, and more importantly, this group contains the first clear **directional reversal** in the whole analysis: `ACTCompositeScore.filled` shows persistently disconnected individuals having *higher* average ACT scores than consistently attached individuals, while `MCA.M` shows the opposite — persistently disconnected individuals have *lower* average math scores. These two academic-achievement measures point in opposite directions, and that divergence is worth holding onto as we move through the rest of this group.

**Categorical HS Academics**

**ap.exam** (χ² = 45.523, df = 3, Cramér's V = 0.037)

Counterintuitively, AP exam participants show *higher* persistent disconnection (51.9% vs. 45.9%, a 6.0-point gap) and *lower* consistent attachment (13.4% vs. 16.5%) than non-participants. This is a genuinely surprising direction — in every prior completer-focused analysis in this project, AP exam participation was one of the clearest markers of the "academic/mobility" pathway associated with leaving the region, not with worse workforce outcomes generally. Here, in a population that never completed a postsecondary credential, taking an AP exam is associated with *worse* disconnection outcomes. This may reflect a specific subgroup: students academically prepared enough to attempt AP coursework but who didn't go on to complete a credential could be a distinct, higher-risk group rather than a generically "more prepared" one. Worth flagging as a genuine departure from the completer-population pattern rather than smoothing it into the usual academic-orientation story.

**cte.achievement** (χ² = 635.315, df = 6, Cramér's V = 0.098)

A clean, monotonic dose-response: persistent disconnection is lowest among CTE Concentrators/Completers (40.9%), higher among CTE Participants (45.5%), and highest among those with No CTE (56.1%) — a 15.2-point spread from top to bottom. Consistent attachment follows the same clean gradient in reverse (20.0% → 15.3% → 11.4%). This is the second-strongest categorical relationship in the entire bivariate analysis to date (behind only SpecialEdStatus at 0.113), and it's the cleanest dose-response pattern seen anywhere in this project — CTE engagement, even among people who never completed any postsecondary credential, shows a strong, orderly association with staying attached to the workforce.

**pseo.participant** (χ² = 7.171, df = 3, Cramér's V = 0.018, p = 0.067)

Not significant at conventional thresholds. The gap between participants and non-participants is negligible (44.8% vs. 46.3% persistent disconnection, consistent attachment nearly identical at 16.2% vs. 16.7%). This is a real departure from PSEO's role throughout the regression work, where it was consistently one of the stronger positive predictors of both postsecondary completion and regional employment attachment. Here, restricted to people who never completed a credential, PSEO participation shows essentially no bivariate relationship with disconnection persistence — worth remembering that this subset (n = 21,890, smaller than the full 32,878) already excludes anyone missing PSEO data, which may itself be a non-random restriction worth keeping in mind.

**sat.taken** (χ² = 60.869, df = 3, Cramér's V = 0.043)

SAT takers show a striking gap — 69.2% persistently disconnected vs. 46.3% for non-takers, a 22.9-point difference, the largest raw gap of any categorical variable in this entire academics group. However, this needs an explicit cell-size caveat: only 279 of 32,878 non-completers (0.8%) took the SAT, and the "Consistently attached" cell for SAT-takers is just 19 people. A gap of this size built on such a small subgroup should be treated as unstable rather than a confirmed finding, similar to the SAT cell-size caution flagged earlier in the DV1 regression work.

**took.ACT** (χ² = 201.773, df = 3, Cramér's V = 0.078)

ACT test-takers show *lower* persistent disconnection (42.4% vs. 49.8%, a 7.4-point gap) and higher consistent attachment (18.4% vs. 14.5%) than non-takers. This is the third-strongest categorical relationship in this academics group and is directionally consistent with `took.ACT`'s role throughout the earlier regression work as a strong, independent positive predictor of workforce attachment — the one clean piece of continuity between the "participation" side of the ACT story and this non-completer bivariate analysis.

**Continuous HS Academics (ANOVA)**

**ACTCompositeScore.filled** (F = 70.615, df = 3/14,683, p \< 0.001, η² ≈ 1.4%)

This produces the cleanest, most fully monotonic pattern of any continuous variable so far, and it runs in the *counterintuitive* direction relative to `took.ACT` above: Tukey comparisons show a strict ordering where average composite score rises steadily from Consistently attached → Intermittently disconnected → Broadly disconnected → Persistently disconnected, with every pairwise comparison against Consistently attached significant (Intermittent: +0.43, p = 0.001; Broad: +0.66, p \< 0.001; Persistent: +1.41, p \< 0.001). In other words, among those who took the ACT, *higher scorers* are more likely to be persistently disconnected from the workforce, not less. This mirrors the "two forces" pattern from the completer-focused regression work (participation positive, score itself negative for regional attachment) — but here it shows up as a *workforce disconnection* outcome specifically, in a population that never completed college, which is a new and noteworthy extension of that pattern.

**MCA.M** (F = 19.871, df = 3/30,058, p \< 0.001, η² ≈ 0.2%)

Runs in the *opposite* direction from ACT composite score: persistently disconnected individuals have *lower* average MCA Math scores than consistently attached individuals (−0.130, p \< 0.001), with a roughly monotonic pattern across categories. This is a much smaller effect size than ACT composite score, but the directional reversal between these two "achievement" measures is itself the notable finding here — worth being cautious in reporting, per the methodology's own caution that MCA missingness is confounded with English learner status; this bivariate relationship may be partly reflecting that confound rather than a clean achievement effect.

**MCA.R** (F = 6.700, df = 3/29,643, p \< 0.001, η² ≈ 0.07%)

A much weaker version of the MCA.M pattern — only two pairwise comparisons reach significance (Persistently vs. Intermittently disconnected: −0.055, p = 0.001; Persistently vs. Broadly disconnected: −0.063, p = 0.005), and the comparison most people would care about (Persistently vs. Consistently attached) is not significant (p = 0.230). This is a weak, partial signal rather than a clean pattern.

**MCA.S** (F = 2.156, df = 3/22,539, p = 0.091)

Not significant. No pairwise comparison reaches significance. This is the one HS-academics variable that shows essentially no bivariate relationship with disconnection category at all.

**total.cte.courses.taken** (F = 246.028, df = 3/32,874, p \< 0.001, η² ≈ 2.2%)

This is the strongest bivariate relationship of any variable examined in this entire analysis to date — larger than ACT composite score, larger than cte.achievement, larger than every demographic variable's Cramér's V-equivalent magnitude. The pattern is fully monotonic and in the expected direction: each successive disconnection category takes fewer CTE courses on average, and every pairwise Tukey comparison is significant (Consistently attached has the most courses; Persistently disconnected has the fewest, a full 1.31-course average difference between the two extremes). This reinforces and sharpens the `cte.achievement` finding above — CTE engagement, measured either categorically or by course count, is the single strongest predictor identified anywhere in this bivariate analysis so far.

**Synthesis**

| Ranked by effect size (η² for continuous / Cramér's V for categorical) | Direction |
|------------------------------------|------------------------------------|
| 1\. total.cte.courses.taken (η² ≈ 2.2%) | More courses → less disconnection (expected direction) |
| 2\. ACTCompositeScore.filled (η² ≈ 1.4%) | Higher score → *more* disconnection (counterintuitive, mirrors "two forces" pattern) |
| 3\. cte.achievement (V = 0.098) | More CTE engagement → less disconnection (expected direction, clean dose-response) |
| 4\. took.ACT (V = 0.078) | Took ACT → less disconnection (expected direction) |
| 5\. sat.taken (V = 0.043) | Took SAT → *more* disconnection (counterintuitive, but tiny subgroup — unstable) |
| 6\. ap.exam (V = 0.037) | Took AP exam → *more* disconnection (counterintuitive, departs from completer-population pattern) |
| 7\. MCA.M (η² ≈ 0.2%) | Higher score → less disconnection (expected direction, but small and EL-confound caution applies) |
| 8\. MCA.R (η² ≈ 0.07%) | Weak, partial version of MCA.M pattern |
| 9\. pseo.participant (V = 0.018, ns) | No meaningful relationship |
| 10\. MCA.S (ns) | No meaningful relationship |

The clearest overall story: **CTE engagement (both achievement level and course count) is the strongest and most consistently positive predictor in this entire bivariate analysis**, exceeding every demographic variable examined earlier. The second clearest story is a genuine **split within "academic achievement/orientation" measures**: ACT composite score, SAT-taking, and AP exam participation all point toward *more* disconnection among non-completers, while MCA scores point toward *less* — this is not a contradiction to explain away but a real signal that different measures of "academic strength" are capturing different things in this specific population, and it deserves its own discussion in any final write-up rather than being collapsed into a single "academic ability" narrative.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `sat.taken`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/sat-taken.html")
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

**attended.ps** (χ² = 1094.651, df = 3, Cramér's V = 0.182)

Sample: N = 32,878 (19,792 never attended postsecondary; 13,086 attended without completing)

Individuals who attended postsecondary education without completing show substantially better workforce outcomes than those who never attended at all: 36.5% persistently disconnected vs. 53.1% for non-attenders — a 16.6-point gap, and consistent attachment is nearly double (22.0% vs. 12.4%).

**This is the strongest bivariate relationship found anywhere in this entire descriptive/bivariate analysis.** Cramér's V = 0.182 clears the conventional "small effect" threshold (0.10) by a wide margin — nearly double SpecialEdStatus (0.113), the previous strongest variable, and more than eight times RaceEthnicity's standardized value (0.060). It's also the single largest raw percentage-point gap in persistent disconnection found among any 2-level variable in this project, exceeding English learner status (15.1 points) and SpecialEdStatus (13.3 points) from the demographic set.

**Substantive Pattern**

This directly replicates and reinforces the single largest effect found throughout the entire non-completer regression work (Part 1 and Part 2 of the prior logistic analysis), where `attended.ps` consistently produced large, significant positive coefficients at every time point, even as its magnitude faded from Year 1 to Year 10. Here, at the bivariate level and using a 5-year trajectory-based outcome rather than a single time-point measure, the same relationship shows up as the strongest single predictor identified in this entire descriptive stage — stronger than any demographic variable, any HS characteristic, and any HS academic variable examined so far, including `total.cte.courses.taken` and `cte.achievement`, the previous leaders.

This is a genuinely important finding for how the whole non-completer disconnection story gets framed: even without completing a credential, *any* exposure to postsecondary education is associated with a meaningfully better long-term workforce trajectory than no exposure at all. Combined with the earlier finding that `attended.ps`'s regression coefficient fades over time (Year 1 → Year 10), this bivariate result — built on a full 5-year persistence measure rather than a single year — suggests the postsecondary-exposure advantage is durable enough to show up clearly even when averaged across the entire post-exit window, not just immediately after exit.

This is very likely the variable most worth leading with when this bivariate stage gets synthesized, alongside the two CTE variables already flagged as unusually strong.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of Overall Disconnection Persistence (`disconnected.cat`)

**Overview**

This stage examined the bivariate relationship between disconnection persistence category (Consistently attached / Intermittently / Broadly / Persistently disconnected) and sixteen individual characteristics across four groups: demographics, high school characteristics, high school academics, and postsecondary attendance. Cross-tabulations with Cramér's V were used for categorical predictors; one-way ANOVA with Tukey post-hoc comparisons for continuous predictors. Consistent with the emphasis on substantive interpretation over significance alone, effect sizes varied enormously across groups — from negligible (high school characteristics) to the strongest relationship found anywhere in this analysis (postsecondary attendance).

**Ranking across all sixteen variables (by standardized effect size)**

| Rank | Variable | Group | Effect size | Direction |
|---------------|---------------|---------------|---------------|---------------|
| 1 | attended.ps | Postsecondary | V = 0.182 | Attended → less disconnection |
| 2 | total.cte.courses.taken | HS academics | η² ≈ 2.2% | More courses → less disconnection |
| 3 | ACTCompositeScore.filled | HS academics | η² ≈ 1.4% | Higher score → *more* disconnection |
| 4 | SpecialEdStatus | Demographics | V = 0.113 | SpecialEd → more disconnection |
| 5 | cte.achievement | HS academics | V = 0.098 | More CTE engagement → less disconnection |
| 6 | English learner | Demographics | V = 0.096 | EL → more disconnection |
| 7 | took.ACT | HS academics | V = 0.078 | Took ACT → less disconnection |
| 8 | Gender | Demographics | V = 0.074 | Female → more disconnection |
| 9 | Non-English home | Demographics | V = 0.072 | Non-English home → more disconnection |
| 10 | Economic status | Demographics | V = 0.070 | Disadvantaged → more disconnection |
| 11 | RaceEthnicity | Demographics | V = 0.060 | Every non-White group worse than White |
| 12 | sat.taken | HS academics | V = 0.043 | Took SAT → more disconnection (unstable, n=279) |
| 13 | ap.exam | HS academics | V = 0.037 | Took AP → more disconnection (counterintuitive) |
| 14 | edr | HS characteristics | V = 0.021 | Minimal regional difference |
| 15 | MCA.M | HS academics | η² ≈ 0.2% | Higher score → less disconnection |
| 16 | Dem_Desc | HS characteristics | V = 0.014 | No meaningful pattern |

(pseo.participant, MCA.R, MCA.S, avg.unemp.rate, avg.wages.pct.state either not significant or negligible — see below)

**The four headline findings**

**1. Postsecondary exposure — even without completion — is the single strongest predictor found in this entire bivariate stage.** `attended.ps` (V = 0.182) exceeds every other variable by a wide margin, nearly double the next-strongest relationship. This directly replicates the largest effect found throughout the earlier non-completer regression work, now confirmed using a full 5-year persistence outcome rather than a single time point.

**2. CTE engagement is the second-clearest story, and unusually strong at the bivariate level.** Both `total.cte.courses.taken` (η² ≈ 2.2%, the single strongest continuous relationship) and `cte.achievement` (V = 0.098) show clean, monotonic, dose-response patterns — more CTE engagement, by either measure, consistently predicts less disconnection. This is consistent with CTE's role throughout the entire project (DV1–DV5, and both Parts of the non-completer regression work) as one of the most durable positive predictors of workforce attachment.

**3. Race/ethnicity shows the largest raw disparities in the entire analysis, despite the lowest standardized effect size among demographic variables.** American Indian graduates show a 21.2-point gap in persistent disconnection relative to White graduates (66.0% vs. 44.8%), and Black graduates show an 18.0-point gap (62.8%) — both larger than any other raw gap in the demographic set. This directly parallels the unexplained Black-White (and here, especially American Indian) gap that ran through the DV1, DV2, and DV5 multivariate work on completers, now confirmed at the purely bivariate level in a population that never completed any postsecondary credential — this disparity is not specific to the postsecondary pathway.

**4. Academic achievement measures split into two opposing camps, and this split is genuinely new to this project.** ACT composite score, SAT-taking, and AP exam participation all point toward *more* disconnection among higher scorers/participants, while MCA scores point toward *less*. This is the first time in the whole body of work (completer regressions, non-completer regressions, and now this bivariate stage) that "academic strength" measures have split so cleanly in opposite directions rather than moving together. It likely reflects something specific about non-completers who show strong test performance but never finish a credential — this subgroup may face a distinct set of risks not captured elsewhere in this analysis.

**Where the signal is weak or absent**

High school characteristics (Dem_Desc, edr, avg.unemp.rate, avg.wages.pct.state) show uniformly negligible relationships with disconnection category — the weakest group in this entire analysis by a wide margin, in sharp contrast to the demographic and academic groups. PSEO participation and two of the three MCA subjects (MCA.R weakly, MCA.S not at all) also show minimal-to-no bivariate relationship, a notable departure from PSEO's stronger role in the completer-population regression work.

**Connecting back to the rest of the project**

This bivariate stage largely reinforces two of the most important findings from the entire body of work to date: the outsized importance of any postsecondary exposure (echoing `attended.ps` and `ps.grad` throughout the regression models) and the persistent, unexplained racial disparity — especially for American Indian and Black graduates — that shows up regardless of population (completers or non-completers), regardless of outcome (employment attachment, credential completion, or disconnection persistence), and regardless of analytic method (multivariate logistic regression or simple bivariate cross-tabulation. The one genuinely new finding is the academic-achievement split (ACT/SAT/AP vs. MCA), which hasn't appeared in this form anywhere else in the project and is worth flagging as a distinct area for further exploration rather than folding into the existing "two forces" or "unexplained disparity" narratives.

**Open items carried forward**

The ap.exam and sat.taken counterintuitive directions, and the MCA-vs-ACT/SAT split, remain unexplained at the bivariate level — these associations may offer exactly the kind of clue you mentioned earlier that could shed light on what "No MN employment record" represents, once compared against the persistent non-sustained employment breakdown coming next.

<br>

# Relationship Between Individual Characteristics and Non-sustaining Categories

The next sections focus on exploring what relationships exist between individual characteristics and the frequency of having non-sustained workforce participation in the first 5 years after formal education.

- **Never non-sustained** — 0 of the first 5 years (post-high-school or post-postsecondary-exit) were classified as "Non-sustained WF"
- **Intermittently non-sustained** — 1-2 of the first 5 years were "Non-sustained WF"
- **Broadly non-sustained** — exactly 3 of the first 5 years were "Non-sustained WF"
- **Persistently non-sustained** — 4-5 of the first 5 years were "Non-sustained WF"

<br>

## Demographic variables

**Overall pattern across all six variables — and a striking shift from the blended `disconnected.cat` results**

All six relationships remain statistically significant, but the *pattern of which variables are strong* has shifted substantially now that non-sustained employment is isolated from no-record status. Some variables (Gender, Economic status) hold roughly steady or even strengthen; others (English learner, Non-English home, RaceEthnicity) weaken dramatically, in several cases losing most of the raw gap that made them so prominent in the blended analysis. This divergence is the most important finding in this table — it starts to answer the open question about what's driving disconnection for different groups.

**Gender** (χ² = 272.661, df = 3, Cramér's V = 0.091)

Female graduates show notably higher persistent non-sustained employment (15.5% vs. 11.4%, a 4.1-point gap) and much lower "Never non-sustained" rates (31.0% vs. 37.5%, a 6.5-point gap) than male graduates. This is the **strongest relationship in this table**, and notably, Gender's Cramér's V *increased* relative to the blended disconnected.cat result (0.074 → 0.091) — the opposite of what happened to several other variables here. This suggests the gender disadvantage in workforce disconnection is concentrated specifically in employment instability, not in disappearing from the record.

**Economic Status** (χ² = 216.948, df = 3, Cramér's V = 0.081)

Economically disadvantaged students show higher persistent non-sustained employment (15.2% vs. 11.6%, a 3.6-point gap). Cramér's V here (0.081) is essentially unchanged from the blended result (0.070) — if anything, slightly stronger — suggesting economic status's disconnection effect operates fairly evenly across both non-sustained employment and no-record status, unlike several other variables in this set.

**SpecialEdStatus** (χ² = 192.125, df = 3, Cramér's V = 0.076)

Students with SpecialEd status show higher persistent non-sustained employment (17.8% vs. 11.7%, a 6.1-point gap) — the largest raw gap of any variable in this table. However, Cramér's V dropped meaningfully from the blended result (0.113 → 0.076), and the "Never non-sustained" rates are nearly identical between groups (33.5% vs. 35.2%) — suggesting SpecialEdStatus's overall disconnection effect is partly, but not entirely, about employment instability specifically.

**English Learner Status** (χ² = 131.161, df = 3, Cramér's V = 0.063)

This shows a **dramatic weakening** relative to the blended result. The persistent non-sustained gap is now just 2.4 points (15.2% vs. 12.8%), compared to a 13.9-point gap in the blended disconnected.cat table. Even more tellingly, English learners are now *more* likely to be "Never non-sustained" than non-learners (41.4% vs. 34.0%) — the opposite direction from what the blended category showed. This strongly suggests English learners' overall disconnection disadvantage is concentrated almost entirely in disappearing from Minnesota's employment record, not in cycling through unstable jobs.

**Non-English Home** (χ² = 124.025, df = 3, Cramér's V = 0.061)

An even more striking version of the same pattern. The persistent non-sustained gap essentially vanishes and reverses (12.3% vs. 13.2% — non-English home students are marginally *less* likely to be persistently non-sustained), while "Never non-sustained" is far higher for this group (43.4% vs. 33.9%). Recall this variable showed the second-largest raw gap in the entire blended analysis (14.5 points). That gap has now almost entirely disappeared once non-sustained employment is isolated — this is close to definitive evidence that non-English-home students' disconnection is driven almost exclusively by no-record status.

**RaceEthnicity** (χ² = 196.443, df = 12, Cramér's V = 0.045)

The most dramatic shift of all six variables. In the blended table, American Indian graduates showed a 21.2-point gap in persistent disconnection relative to White graduates — here, the gap in persistent *non-sustained* employment is only 2.3 points (15.1% vs. 12.8%). Black graduates' gap similarly shrinks from 18.0 points (blended) to 7.8 points (non-sustained only) — still real, but less than half its previous size. This is a significant finding: the large racial disparities that stood out so clearly in the blended disconnection measure appear to be driven substantially — for American Indian graduates, almost entirely — by no-record status rather than employment instability.

**Synthesis: comparing Cramér's V, blended vs. non-sustained-only**

| Variable | Blended disconnected.cat V | Non-sustained-only V | Change |
|------------------|------------------|------------------|------------------|
| Gender | 0.074 | 0.091 | Strengthens |
| Economic status | 0.070 | 0.081 | Roughly stable/slight increase |
| SpecialEdStatus | 0.113 | 0.076 | Weakens substantially |
| English learner | 0.096 | 0.063 | Weakens substantially |
| Non-English home | 0.072 | 0.061 | Weakens, raw gap nearly vanishes |
| RaceEthnicity | 0.060 | 0.045 | Weakens, raw gaps shrink by 60-90% |

**Substantive Pattern**

This table provides the clearest evidence yet on the open question flagged earlier: several of the demographic disparities that looked large in the blended disconnection measure — especially RaceEthnicity (particularly American Indian), English learner status, and Non-English home — are substantially or almost entirely driven by no-record status, not by cycling through unstable employment. By contrast, Gender and Economic status show the opposite pattern: their disconnection effects hold steady or even strengthen when non-sustained employment is isolated, suggesting these two variables operate more through genuine employment instability.

This is a meaningful, substantive finding on its own, and it sets up a clear prediction for the next table: I'd expect the no-record persistence cross-tabs to show *larger* gaps for RaceEthnicity, English learner, and Non-English home than what appeared here, and *smaller or reversed* gaps for Gender and Economic status. Worth testing that prediction directly once you share the no-record table.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/english-learner.html")
```

<br>

#### `non.english.home`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/non-english-home.html")
```

<br>

## High school characteristics

**Overall pattern across all four variables**

This is now the second variable group in this analysis (after HS characteristics against the blended `disconnected.cat`) to show uniformly weak-to-null relationships — and this time, two of the four don't reach significance at all. Where HS characteristics showed *some* weak-but-significant signal against the blended disconnection measure, against non-sustained employment specifically they show essentially nothing.

**Dem_Desc** (χ² = 24.098, df = 6, Cramér's V = 0.019, p = 0.001)

Technically significant, but the raw gaps are trivial: persistent non-sustained employment ranges only from 12.6% to 13.8% across the three urbanicity categories, and "Never non-sustained" ranges from 33.1% to 35.8%. This is marginally larger than the blended-category Dem_Desc result (V = 0.014) but still far below any threshold of substantive interest.

**edr** (χ² = 1.458, df = 3, Cramér's V = 0.007, p = 0.692)

**Not significant.** EDR 9 and EDR 10 are nearly identical on every category (persistent non-sustained: 13.4% vs. 12.9%; never non-sustained: 34.9% vs. 34.8%). This is a genuine null result, and notably weaker than edr's already-weak relationship with the blended disconnection measure (V = 0.021, p = 0.002) — the small regional difference that existed in the blended table has essentially vanished once no-record status is removed from the outcome.

**avg.unemp.rate** (F = 1.716, df = 3/32,874, p = 0.161)

**Not significant.** Unlike the blended-category result (which was significant, p = 0.023, albeit with a negligible effect size), average local unemployment rate shows no relationship with non-sustained employment persistence at all once isolated from no-record status. No Tukey comparison approaches significance (smallest adj. p = 0.125).

**avg.wages.pct.state** (F = 0.611, df = 3/32,874, p = 0.608)

**Not significant.** Same pattern as unemployment rate — the weak signal present in the blended disconnection measure (p = 0.030) disappears entirely here. No Tukey comparison approaches significance.

**Synthesis**

This table effectively confirms the same story that emerged from the demographic table: the already-weak relationships between HS characteristics and the blended disconnection measure appear to have been driven substantially by the no-record component, not by employment instability. Both continuous local-economy variables (unemployment rate, wages) and one of the two categorical geography variables (edr) lose significance entirely once isolated to non-sustained employment. Only Dem_Desc retains a technically significant but substantively negligible relationship.

**This strengthens the prediction from the previous table.** If HS-context and local-economy variables show no relationship with employment instability but *did* show a weak signal in the blended measure, the no-record persistence table should show these same variables (especially avg.unemp.rate and avg.wages.pct.state) as more clearly significant — local economic conditions plausibly relate more to whether someone shows up in Minnesota's employment system at all than to whether their employment, once present, is stable. Worth testing this directly against the next table.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `edr`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/edr.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**Overall pattern — a mix of confirmed predictions and genuine surprises**

This table offers the clearest test yet of the prediction from the demographics/HS-characteristics tables: several variables that showed strong relationships with the *blended* disconnection measure weaken dramatically here, confirming those effects were driven substantially by no-record status. But two variables (MCA.R, MCA.S) actually *strengthen* when isolated to non-sustained employment — the opposite of the pattern elsewhere — and pseo.participant flips from a null result in the blended table to a significant one here. This group is more mixed than the previous two, and worth walking through carefully rather than assuming a single direction of change.

**Variables that weakened substantially (confirming the no-record-driven prediction)**

**total.cte.courses.taken** — the most dramatic weakening in this entire analysis. Blended: F = 246.03, η² ≈ 2.2% (the single strongest relationship found anywhere in Step 3). Here: F = 18.17, η² ≈ 0.17% — more than a 12-fold reduction in effect size. The Tukey pattern is also less clean: only Broadly and Persistently non-sustained differ significantly from Never (and from Intermittently), while Intermittently vs. Never is not significant (p = 0.092). This strongly suggests CTE course-taking's powerful blended relationship with disconnection was operating mainly through no-record status, not employment stability.

**cte.achievement** — similarly weakens substantially: V = 0.098 (blended) → V = 0.041 here. The clean 15.2-point dose-response gradient from the blended table (40.9% → 45.5% → 56.1%) collapses to a 2.7-point spread (11.6% → 14.0% → 14.3%). CTE achievement level, like course count, appears to matter mainly for whether someone shows up in the employment record at all, not for whether their employment is stable once present.

**ACTCompositeScore.filled** — this is the single largest reversal in significance found in this entire analysis. Blended: F = 70.62, p \< 0.001, η² ≈ 1.4%. Here: F = 1.82, **p = 0.141 — not significant at all.** No Tukey comparison approaches significance. The entire "higher scorers show more disconnection" pattern flagged as counterintuitive in the blended table appears to be exclusively a no-record phenomenon — academically stronger test-takers who don't complete a credential are apparently *not* more likely to cycle through unstable jobs; they're more likely to disappear from the employment record entirely. This is an important clarification of that earlier counterintuitive finding.

**sat.taken** — also drops to non-significance (V = 0.043, p \< 0.001 blended → V = 0.010, p = 0.317 here). The dramatic 22.9-point gap from the blended table essentially vanishes (13.1% vs. 10.4% persistent non-sustained, direction even reverses slightly). Combined with the earlier cell-size caution (only 279 SAT-takers), this result is now doubly uncertain — both underpowered and apparently driven by no-record status rather than employment instability.

**Variables that strengthened or newly emerged (breaking the pattern)**

**MCA.S** — genuinely surprising. Not significant in the blended table (p = 0.091, no Tukey comparisons significant). Here: F = 27.63, **p \< 0.001**, with a full, clean monotonic pattern — every pairwise Tukey comparison is significant except Intermittent vs. Never. Persistently non-sustained individuals score 0.155 points lower than Never non-sustained. This is the only HS-academics variable in this whole group whose relationship *emerged* rather than weakened once isolated to employment instability — MCA Science scores appear to specifically predict non-sustained employment, a signal that was fully masked in the blended measure.

**MCA.R** — also strengthens: η² ≈ 0.07% (blended, weak/partial) → η² ≈ 0.14% here, now with a cleaner pattern (Persistently non-sustained significantly lower than all three other categories).

**MCA.M** — roughly stable (η² ≈ 0.2% blended vs. 0.25% here), remaining the most consistent of the three MCA subjects across both outcome specifications.

**pseo.participant** — flips from non-significant in the blended table (V = 0.018, p = 0.067) to significant here (V = 0.042, p \< 0.001). But the direction is counterintuitive: PSEO participants are *less* likely to be "Never non-sustained" (30.6% vs. 35.7%) and *more* likely to be "Intermittently non-sustained" (43.1% vs. 38.7%), while the "Persistently non-sustained" rate is essentially identical (13.3% vs. 13.3%). This isn't a simple "PSEO helps or hurts" story — it looks like PSEO participation shifts people out of the "Never" category and into "Intermittent" specifically, without affecting the persistent end at all. Worth flagging as a genuinely different shape than any other variable examined so far.

**took.ACT** — weakens modestly (V = 0.078 blended → V = 0.047 here) and shows a similarly complex shape: ACT-takers are actually *less* likely to be "Never non-sustained" (32.6% vs. 36.6%) but *more* likely to be "Intermittently non-sustained" (41.8% vs. 37.8%), with persistent rates nearly identical (12.9% vs. 13.2%). Like PSEO, this suggests ACT participation's blended-table advantage was concentrated in the no-record dimension, and its relationship with non-sustained employment specifically is more about shifting people from "never unstable" to "occasionally unstable" than about preventing persistent instability.

**ap.exam** — weakens somewhat (V = 0.037 → 0.030) but retains the same counterintuitive direction seen in the blended table: AP exam participants show higher persistent non-sustained employment (14.5% vs. 12.9%).

**Synthesis**

| Variable | Blended V/η² | Non-sustained-only V/η² | Pattern |
|------------------|------------------|------------------|------------------|
| total.cte.courses.taken | η² ≈ 2.2% | η² ≈ 0.17% | Weakens dramatically |
| ACTCompositeScore.filled | η² ≈ 1.4% | ns (p=0.141) | Loses significance entirely |
| cte.achievement | V = 0.098 | V = 0.041 | Weakens substantially |
| sat.taken | V = 0.043 | ns (p=0.317) | Loses significance entirely |
| took.ACT | V = 0.078 | V = 0.047 | Weakens, shape changes |
| ap.exam | V = 0.037 | V = 0.030 | Weakens modestly, same direction |
| MCA.M | η² ≈ 0.2% | η² ≈ 0.25% | Roughly stable |
| MCA.R | η² ≈ 0.07% | η² ≈ 0.14% | Strengthens |
| MCA.S | ns (p=0.091) | η² ≈ 0.37%, p\<0.001 | Emerges from null |
| pseo.participant | ns (p=0.067) | V = 0.042, p\<0.001 | Emerges, but complex shape |

**Substantive Pattern**

This table adds real nuance to the "no-record vs. non-sustained" question. Most HS-academics variables confirm the prediction — ACT score, SAT-taking, CTE achievement, and CTE course count all show their blended-table relationships substantially or entirely driven by no-record status. But the MCA subjects, especially MCA.S, tell the opposite story: these relationships were being *suppressed* in the blended measure and only become visible once non-sustained employment is isolated. This is a genuinely important distinction to carry forward — it suggests MCA performance (particularly Science) may be a more specific marker of employment *instability* risk, while ACT/SAT/CTE measures are more specific markers of employment-record *presence* risk. The next table (no-record persistence) should make this contrast even sharper — I'd expect ACT composite score, SAT-taking, and CTE variables to show their strongest relationships there, while MCA.S and MCA.R may show weaker or null relationships in that table by comparison.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `sat.taken`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/sat-taken.html")
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

**attended.ps** (χ² = 257.260, df = 3, Cramér's V = 0.088)

Sample: N = 32,878 (19,792 never attended postsecondary; 13,086 attended without completing)

Individuals who attended postsecondary education without completing show lower persistent non-sustained employment (9.9% vs. 15.2%, a 5.3-point gap) and modestly higher "Never non-sustained" rates (36.0% vs. 34.0%) than non-attenders. Interestingly, attenders show *higher* rates of Intermittent non-sustained employment (42.8% vs. 37.5%) — so the advantage of postsecondary attendance shows up specifically at the persistent end of the spectrum, not by making people uniformly less likely to ever experience instability.

**Substantial weakening relative to the blended measure**

This is the clearest confirmation yet of the no-record-driven pattern established across the last three tables. In the blended `disconnected.cat` analysis, `attended.ps` was the single strongest relationship found anywhere in the entire Step 3 analysis (V = 0.182, more than double any other variable). Here, isolated to non-sustained employment specifically, it drops to V = 0.088 — less than half its blended strength, and now roughly on par with Gender (0.091) and Economic status (0.081) rather than standing alone at the top.

**Substantive Pattern**

This strongly suggests that postsecondary attendance's massive blended-table advantage was driven substantially by no-record status — attenders are much less likely to disappear from Minnesota's employment record entirely — while its relationship with employment *stability* specifically, though still real and significant, is far more modest. This is an important qualification to the earlier finding: "any postsecondary exposure helps" remains true, but the mechanism looks like it operates mainly through keeping people visible in the employment system, not through making their employment more stable once they're in it.

This fits the pattern now established across every variable group in this non-sustained-only analysis: attended.ps, total.cte.courses.taken, cte.achievement, ACTCompositeScore, sat.taken, and most HS-characteristics variables all weakened substantially or lost significance entirely once no-record status was removed from the outcome — while Gender, Economic status, SpecialEdStatus (partially), MCA.R, MCA.S, and pseo.participant held steady or strengthened. This is shaping up to be the central organizing finding of the whole non-sustained vs. no-record comparison, and the no-record table should make it definitive.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of Non-Sustained Employment Persistence (`non_sustained.cat`)

**Overview**

This stage repeated the full Step 3 bivariate analysis using `non_sustained.cat` — employment instability persistence, isolated from no-record status — as the outcome, in place of the blended `disconnected.cat` measure used previously. The central purpose was to test whether the strong relationships found in the blended analysis were driven by employment instability, by no-record status, or both. The results are decisive: **most variables that showed strong blended-table relationships weakened substantially or lost significance entirely once isolated to non-sustained employment specifically** — but a distinct minority of variables held steady or even strengthened, and that split is itself the most important finding of this stage.

**Variables whose blended effect was substantially or entirely driven by no-record status**

| Variable | Blended effect | Non-sustained-only effect | Change |
|------------------|------------------|------------------|------------------|
| attended.ps | V = 0.182 (strongest in project) | V = 0.088 | Weakens by more than half |
| total.cte.courses.taken | η² ≈ 2.2% | η² ≈ 0.17% | Weakens \~12-fold |
| cte.achievement | V = 0.098 | V = 0.041 | Weakens substantially |
| ACTCompositeScore.filled | η² ≈ 1.4%, p\<0.001 | p = 0.141 | **Loses significance entirely** |
| sat.taken | V = 0.043, p\<0.001 | p = 0.317 | **Loses significance entirely** |
| SpecialEdStatus | V = 0.113 | V = 0.076 | Weakens substantially |
| English learner | V = 0.096 | V = 0.063 | Weakens, raw gap shrinks from 13.9 to 2.4 pts |
| Non-English home | V = 0.072 | V = 0.061 | Weakens, raw gap nearly vanishes |
| RaceEthnicity | V = 0.060 | V = 0.045 | Weakens, AI/Black raw gaps shrink 60-90% |
| edr | V = 0.021, p=0.002 | p = 0.692 | **Loses significance entirely** |
| avg.unemp.rate | η² negligible, p=0.023 | p = 0.161 | **Loses significance entirely** |
| avg.wages.pct.state | η² negligible, p=0.030 | p = 0.608 | **Loses significance entirely** |

This is the majority pattern across the whole analysis — twelve of sixteen variables examined show this shape. Race/ethnicity is the most consequential example: American Indian graduates' 21.2-point blended gap shrinks to 2.3 points once isolated to non-sustained employment, and English learner's 13.9-point gap shrinks to 2.4 points. These findings indicate that for these groups, the blended "disconnection" story was overwhelmingly a story about disappearing from Minnesota's employment record, not about cycling through unstable jobs.

**Variables that held steady, strengthened, or newly emerged**

| Variable | Blended effect | Non-sustained-only effect | Change |
|------------------|------------------|------------------|------------------|
| Gender | V = 0.074 | V = 0.091 | **Strengthens** |
| Economic status | V = 0.070 | V = 0.081 | Roughly stable/slight increase |
| MCA.M | η² ≈ 0.2% | η² ≈ 0.25% | Roughly stable |
| MCA.R | η² ≈ 0.07%, weak | η² ≈ 0.14% | Strengthens |
| MCA.S | Not significant (p=0.091) | η² ≈ 0.37%, p\<0.001 | **Emerges from null** |
| pseo.participant | Not significant (p=0.067) | V = 0.042, p\<0.001 | **Emerges from null** |

Gender is the clearest example of a variable whose disconnection effect operates specifically through employment instability rather than record absence — its standardized effect actually grows once isolated. MCA.S is the most striking reversal in the entire analysis: completely null in the blended measure, but a clean, fully monotonic, highly significant predictor of non-sustained employment specifically. This means the blended measure was fully masking a real relationship for this variable — the opposite problem from what happened to ACT score, CTE, and SAT.

**The clearest overall finding**

Two genuinely distinct mechanisms are operating in this population, and this table set is the first place in the whole analysis where they can be cleanly separated:

1.  **A "record-absence" cluster**: postsecondary exposure (attended.ps), CTE engagement (both measures), ACT/SAT performance, race/ethnicity, English learner and non-English home status, and local geography/economy variables all relate mainly to whether someone shows up in Minnesota's employment system at all, not to whether their employment, once present, is stable.

2.  **A "genuine instability" cluster**: Gender, Economic status, and — surprisingly — MCA Science and Reading scores (but not composite ACT) relate specifically to employment instability itself, largely independent of record-presence.

**pseo.participant and took.ACT deserve a specific note**: both show a complex, non-monotonic shape here — participants/test-takers are *less* likely to be "Never non-sustained" but *not* less likely to be "Persistently non-sustained," suggesting these variables shift people from stability into occasional instability without protecting against the most severe, persistent form of it. This is a different shape from every other variable examined and doesn't fit neatly into either cluster above.

**Connecting back to the open question**

This entire table set is the strongest evidence yet on what "No MN employment record" represents. If record-absence is disproportionately tied to CTE/ACT/postsecondary exposure, race/ethnicity, and English learner/non-English-home status — while genuine employment instability is tied more to Gender, Economic status, and MCA performance — that's a meaningful clue rather than a resolved answer: it suggests "no record" may be capturing something like out-of-state migration or a shift to non-covered/informal work among people who otherwise show markers of academic engagement or are part of specific demographic groups, while genuine cycling through unstable jobs looks more tied to structural economic disadvantage and gender. This is exactly the kind of pattern-based clue you anticipated earlier — worth carrying forward as a working hypothesis, not a conclusion, once the no-record persistence tables are run and can be compared directly against this one.

**Bottom line**

The single most important finding from this stage: **disconnection in this population is not one phenomenon with many correlates — it is (at least) two distinct phenomena with substantially different correlates**, and the original decision to split `disconnected.cat` into `non_sustained.cat` and `no_record.cat` was analytically necessary to see this. The next step — running the same sixteen variables against `no_record.cat` — should now be treated as close to confirmatory rather than exploratory, given how clearly this table set has already outlined what that comparison is likely to show.

<br>

# Relationship Between Individual Characteristics and having No Minnesota Employment Categories

The next sections focus on exploring what relationships exist between individual characteristics and the frequency of having no Minnesota employment record in the first 5 years after formal education.

- **Never no-record** — 0 of the first 5 years (post-high-school or post-postsecondary-exit) had no Minnesota employment record
- **Intermittently no-record** — 1-2 of the first 5 years had no Minnesota employment record
- **Broadly no-record** — exactly 3 of the first 5 years had no Minnesota employment record
- **Persistently no-record** — 4-5 of the first 5 years had no Minnesota employment record

<br>

## Demographic variables

**Overall pattern — this table confirms the predictions from the non-sustained analysis directly**

This table lets us close the loop opened by the last two tables. The prediction was that variables whose blended effect weakened when isolated to non-sustained employment should now show *stronger* relationships here, and vice versa. That prediction holds cleanly for four of the six variables, with two genuinely interesting exceptions.

**English Learner Status** (χ² = 378.567, df = 3, Cramér's V = 0.107)

This is now the **strongest relationship in this table**, and it confirms the prediction almost perfectly. Persistent no-record: 33.4% for English learners vs. 19.5% for non-learners — a **13.9-point gap**, nearly identical to the original blended gap (13.9 points) and far larger than the 2.4-point gap found for non-sustained employment specifically. This is close to definitive: English learner status's entire disconnection disadvantage runs through no-record status, essentially none of it through employment instability.

**Non-English Home** (χ² = 342.585, df = 3, Cramér's V = 0.102)

Nearly as clean a confirmation. Persistent no-record: 33.5% vs. 19.7% — a 13.8-point gap, almost exactly matching the original blended gap (14.5 points), and vastly larger than the non-sustained-only gap (which had nearly vanished, 12.3% vs. 13.2%). Combined with English learner status above, this table confirms that both of these related variables' disconnection stories are overwhelmingly about record-absence, not job instability.

**RaceEthnicity** (χ² = 252.101, df = 12, Cramér's V = 0.051)

A more nuanced picture than the two variables above. American Indian graduates show a persistent no-record rate of 34.4% vs. 19.8% for White graduates — a 14.6-point gap, which accounts for the large majority (about two-thirds) of AI's original 21.2-point blended gap, with the remaining share falling in the non-sustained table (2.3 points). Black graduates show a smaller no-record gap (24.8% vs. 19.8%, 5.0 points) that's roughly comparable in size to their non-sustained gap (7.8 points, from the earlier table) — suggesting Black graduates' disconnection, unlike American Indian graduates', is more evenly split between the two mechanisms rather than dominated by either one.

**SpecialEdStatus** (χ² = 145.661, df = 3, Cramér's V = 0.067)

Persistent no-record: 25.7% vs. 19.7% — a 6.0-point gap, almost exactly matching the 6.1-point gap found for non-sustained employment. This is the clearest example of a variable whose blended effect (V = 0.113, the strongest demographic relationship in the whole project) splits roughly evenly between the two mechanisms, rather than being dominated by one — genuinely different from every other demographic variable examined so far.

**Gender** (χ² = 8.890, df = 3, Cramér's V = 0.016, p = 0.031)

Barely significant, and the weakest relationship in this entire table. Persistent no-record: 21.2% (male) vs. 20.8% (female) — a negligible 0.4-point gap. This confirms the prediction dramatically: Gender's disconnection effect, which *strengthened* to become the strongest relationship in the non-sustained table (V = 0.091), essentially disappears here. Gender's disadvantage in this population is specifically and almost exclusively about employment instability, not record-absence.

**Economic Status** (χ² = 34.876, df = 3, Cramér's V = 0.033)

Also weak, and the direction is worth flagging: persistent no-record is *slightly lower* for economically disadvantaged individuals (20.4% vs. 21.5%). Combined with economic status's relatively stronger relationship with non-sustained employment (V = 0.081 in the earlier table), this confirms Economic status is, like Gender, primarily an employment-instability variable rather than a record-absence one.

**Synthesis: the full three-way comparison, now complete**

| Variable | Blended V | Non-sustained V | No-record V | Primary mechanism |
|---------------|---------------|---------------|---------------|---------------|
| English learner | 0.096 | 0.063 | **0.107** | No-record (almost entirely) |
| Non-English home | 0.072 | 0.061 | **0.102** | No-record (almost entirely) |
| SpecialEdStatus | 0.113 | 0.076 | 0.067 | Both, roughly evenly split |
| RaceEthnicity (AI specifically) | 0.060 | 0.045 | 0.051 | No-record (\~2/3 of AI's gap) |
| RaceEthnicity (Black specifically) | — | — | — | Both, roughly evenly split |
| Economic status | 0.070 | **0.081** | 0.033 | Non-sustained (primarily) |
| Gender | 0.074 | **0.091** | 0.016 | Non-sustained (almost entirely) |

**Substantive Pattern — this is now a fully resolved, three-way finding**

Every demographic variable examined across all three outcome specifications now has a clear, specific mechanism attached to it, rather than a single blended "disconnection" story:

- **Gender and Economic status** operate almost entirely through genuine employment instability.
- **English learner status and Non-English home** operate almost entirely through disappearance from Minnesota's employment record.
- **SpecialEdStatus** is genuinely mixed, splitting close to evenly between the two mechanisms.
- **RaceEthnicity** is group-specific: American Indian graduates' disparity is disproportionately about record-absence, while Black graduates' disparity is more evenly split — meaning "racial disparity in workforce disconnection" is not a single phenomenon even within this one demographic variable.

**Connecting back to the open question**

This is meaningful, if still indirect, evidence on what "No MN employment record" represents. The fact that it's overwhelmingly concentrated among English learners and non-English-home students — populations where out-of-state relocation (including to a different state or country) or gaps tied to language/documentation barriers to formal employment are both plausible — leans toward "no record" capturing something like migration or exit from Minnesota's tracked labor market, rather than simple non-participation in the labor force generally. That's still a hypothesis, not a conclusion, but it's now supported by a specific, consistent pattern across two closely related variables rather than a single ambiguous finding.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/english-learner.html")
```

<br>

#### `non.english.home`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/non-english-home.html")
```

<br>

## High school characteristics

**Overall pattern — partial confirmation, one genuine surprise**

The categorical variables here behave as predicted from the non-sustained-only table: both edr and Dem_Desc strengthen somewhat when isolated to no-record status. But the two continuous local-economy variables produce a genuine surprise — neither reaches significance here either, meaning their weak blended-table significance doesn't cleanly belong to *either* mechanism. This is worth sitting with rather than smoothing over.

**Dem_Desc** (χ² = 44.294, df = 6, Cramér's V = 0.026, p \< 0.001)

Strengthens modestly relative to both prior tables (blended: 0.014; non-sustained-only: 0.019). Persistent no-record shows a small gradient: Entirely urban (19.5%) \< Town/rural mix (20.4%) \< Urban/town/rural mix (22.3%) — a 2.8-point spread, still modest in absolute terms but the clearest pattern this variable has shown across any of the three outcome specifications.

**edr** (χ² = 37.195, df = 3, Cramér's V = 0.034, p \< 0.001)

This is the clearest confirmation of the prediction in this table. edr was significant but weak in the blended measure (V = 0.021), then dropped to complete non-significance in the non-sustained-only table (p = 0.692) — and now returns as the strongest relationship in this HS-characteristics group (V = 0.034). EDR 10 shows higher persistent no-record status than EDR 9 (21.3% vs. 20.4%), and — more tellingly — EDR 9 shows notably higher "Never no-record" (58.4% vs. 54.9%). This confirms that edr's original blended relationship was entirely a no-record phenomenon, with zero contribution from employment instability, a clean example of the pattern this whole three-way comparison was designed to detect.

**avg.unemp.rate** (F = 1.677, df = 3/32,874, p = 0.170)

**Not significant** — and this breaks the pattern rather than confirming it. This variable was weakly significant in the blended table (p = 0.023) and non-significant in the non-sustained-only table (p = 0.161). If the prediction held, it should have reached significance here. It didn't. Local unemployment rate shows no clear relationship with either specific disconnection mechanism — its weak blended-table significance appears to have been a product of very large sample size rather than a real, attributable effect belonging to either pathway.

**avg.wages.pct.state** (F = 0.926, df = 3/32,874, p = 0.427)

**Also not significant**, and also breaks the pattern the same way. Weakly significant in the blended table (p = 0.030), non-significant in non-sustained-only (p = 0.608), and now non-significant here as well. Same conclusion as unemployment rate: this variable's blended significance does not appear to be genuinely attributable to either the non-sustained or no-record mechanism specifically.

**Synthesis: revising the three-way framework**

| Variable | Blended V/p | Non-sustained V/p | No-record V/p | Pattern |
|---------------|---------------|---------------|---------------|---------------|
| edr | 0.021, p=.002 | ns, p=.692 | 0.034, p\<.001 | Clean confirmation — no-record driven |
| Dem_Desc | 0.014, p=.037 | 0.019, p=.001 | 0.026, p\<.001 | Modest strengthening for no-record |
| avg.unemp.rate | negligible, p=.023 | ns, p=.161 | ns, p=.170 | **Neither mechanism — likely spurious** |
| avg.wages.pct.state | negligible, p=.030 | ns, p=.608 | ns, p=.427 | **Neither mechanism — likely spurious** |

**Substantive Pattern**

This table requires a genuine revision to the framework established over the last several tables. It's not simply true that every blended relationship decomposes cleanly into "belongs to non-sustained" or "belongs to no-record" — for `avg.unemp.rate` and `avg.wages.pct.state`, the honest conclusion is that their weak blended-table significance likely reflects the very large sample size (N = 32,878) crossing the p \< 0.05 threshold on noise, rather than a real relationship with disconnection through either specific pathway. This is a useful, if less exciting, finding in its own right: local labor-market context at the high-school level shows essentially no meaningful bivariate relationship with any disconnection outcome examined in this entire analysis, however the outcome is sliced.

By contrast, edr and Dem_Desc both confirm the broader pattern nicely — modest geography effects that are specifically about record-presence, not employment stability.

**Connecting back to the open question**

edr's clean no-record-driven pattern adds a small but real data point to the "what does no-record mean" question: if leaving the region (or the specific EDR) for elsewhere is part of what drives no-record status, a real EDR-level difference in no-record rates is consistent with a migration-related interpretation, though it's a modest effect and shouldn't be overweighted on its own.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `edr`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school characteristics/edr.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**Overall pattern — this table completes the three-way academic picture, and reveals a genuine split within the MCA subjects**

This is the most conclusive table in the entire three-way comparison. Every variable that weakened dramatically in the non-sustained-only table (ACT composite score, CTE achievement, CTE course count, SAT-taking) comes roaring back here at or above its original blended strength — clean confirmation that their disconnection story runs through record-absence. But the three MCA subjects split apart from each other in a way they hadn't before: MCA.M and MCA.R, which were significant for non-sustained employment, are **not significant here at all** — the mirror image of the ACT/CTE/SAT pattern. MCA.S is the only variable in this entire academics group significant in *both* mechanism-specific tables.

**Variables confirmed as no-record-driven**

**ACTCompositeScore.filled** (F = 95.286, df = 3/14,683, p \< 0.001, η² ≈ 1.91%)

Stronger here than in the blended table (η² ≈ 1.4%) and vastly stronger than the non-sustained-only result (not significant, p = 0.141). Every Tukey comparison against "Never no-record" is significant and positive — higher ACT scores associate with *more* no-record persistence, confirming the counterintuitive blended-table direction was entirely a record-absence phenomenon. Academically stronger test-takers who don't complete a credential are specifically disappearing from Minnesota's employment record, not cycling through unstable jobs.

**total.cte.courses.taken** (F = 170.986, df = 3/32,874, p \< 0.001, η² ≈ 1.54%)

The large majority of this variable's blended effect (η² ≈ 2.2%) lives here (η² ≈ 1.54%), with only a small residual in the non-sustained table (η² ≈ 0.17%). Fully confirms the prediction from the earlier tables.

**cte.achievement** (χ² = 462.675, df = 6, Cramér's V = 0.084)

Similarly, the large majority of the blended effect (V = 0.098) is recovered here (V = 0.084), versus only V = 0.041 in the non-sustained table. The persistent no-record gradient (17.9% → 19.1% → 27.9% across Concentrator/Completor → Participant → No CTE) is far steeper than the equivalent non-sustained gradient (11.6% → 14.0% → 14.3%), confirming CTE's protective effect is primarily about staying visible in the employment record.

**sat.taken** (χ² = 83.372, df = 3, Cramér's V = 0.050)

This fully resolves the earlier ambiguity. SAT-taking was non-significant for non-sustained employment (p = 0.317) but is significant and large here — persistent no-record is 38.4% for SAT-takers vs. 20.9% for non-takers, and "Never no-record" is 31.9% vs. 56.2%, a 24.3-point gap. The dramatic blended-table finding was entirely a no-record phenomenon. The same cell-size caution still applies (only 279 SAT-takers, n = 59 in the persistent no-record cell), but the direction is now unambiguous.

**took.ACT** (χ² = 300.282, df = 3, Cramér's V = 0.096)

The strongest categorical relationship in this entire academics group across all three tables, and stronger here (V = 0.096) than in the blended table (V = 0.078) or the non-sustained table (V = 0.047). ACT test-takers show a 7.7-point lower persistent no-record rate (16.8% vs. 24.5%) — this is where the bulk of `took.ACT`'s protective effect lives.

**ap.exam** (χ² = 55.387, df = 3, Cramér's V = 0.041)

Strengthens modestly relative to both prior tables (blended 0.037, non-sustained 0.030), though the pattern here is non-monotonic: persistent no-record rates are nearly identical between AP-takers and non-takers (21.0% vs. 21.1%), with the real difference showing up in the intermittent and broadly categories instead. This doesn't fit as cleanly into the simple "no-record-driven" story as the other variables.

**Variables that split apart from each other — the MCA subjects**

**MCA.M** (F = 2.083, df = 3/30,058, p = 0.100) — **Not significant.** This is the mirror image of the ACT/CTE/SAT pattern: MCA.M was significant in both the blended table (η² ≈ 0.2%) and the non-sustained table (η² ≈ 0.25%), but has no relationship with no-record status at all. MCA Math's disconnection story runs entirely through employment instability, not record-absence.

**MCA.R** (F = 0.689, df = 3/29,643, p = 0.558) — **Not significant.** Same pattern as MCA.M: significant for non-sustained employment (η² ≈ 0.14%) but null here. Both MCA.M and MCA.R appear to be genuine employment-instability markers, not record-absence markers — the only HS-academics variables in this entire analysis to show that specific pattern.

**MCA.S** (F = 10.662, df = 3/22,539, p \< 0.001, η² ≈ 0.14%) — Significant in both tables (non-sustained: η² ≈ 0.37%; no-record: η² ≈ 0.14%), though the effect is somewhat larger for non-sustained employment. MCA Science stands alone as the one academic variable with a genuine relationship to *both* mechanisms.

**pseo.participant** (χ² = 40.355, df = 3, Cramér's V = 0.043)

Similar magnitude to the non-sustained table (V = 0.042), but a cleaner, more classically protective shape here: PSEO participants show lower persistent no-record status (17.5% vs. 21.3%), unlike the non-sustained table where PSEO mainly shifted people from "Never" into "Intermittent" without touching the persistent extreme. PSEO appears to operate somewhat differently across the two mechanisms rather than showing one consistent shape.

**Synthesis: the complete three-way academic picture**

| Variable | Blended | Non-sustained | No-record | Mechanism |
|---------------|---------------|---------------|---------------|---------------|
| took.ACT | V=0.078 | V=0.047 | **V=0.096** | No-record (majority) |
| ACTCompositeScore.filled | η²≈1.4% | ns | **η²≈1.91%** | No-record (essentially entirely) |
| total.cte.courses.taken | η²≈2.2% | η²≈0.17% | **η²≈1.54%** | No-record (majority) |
| cte.achievement | V=0.098 | V=0.041 | **V=0.084** | No-record (majority) |
| sat.taken | V=0.043 | ns | **V=0.050** | No-record (essentially entirely) |
| ap.exam | V=0.037 | V=0.030 | V=0.041 | Weakly no-record-leaning, non-monotonic |
| pseo.participant | ns | V=0.042 | V=0.043 | Both, different shapes |
| MCA.M | η²≈0.2% | η²≈0.25% | ns | **Non-sustained (essentially entirely)** |
| MCA.R | η²≈0.07% | η²≈0.14% | ns | **Non-sustained (essentially entirely)** |
| MCA.S | ns | η²≈0.37% | η²≈0.14% | Both mechanisms |

**Substantive Pattern**

This table set completes a genuinely important structural finding for this whole HS-academics group: **standardized test performance splits cleanly by which test.** ACT composite score and SAT-taking are both almost purely record-absence markers. MCA Math and Reading are almost purely employment-instability markers. MCA Science sits in between, relating to both. This is a much more specific and useful finding than the earlier "academic achievement measures split into two camps" observation — it's not really about achievement level at all, but about which specific assessment is involved, and that distinction maps cleanly onto the two disconnection mechanisms this whole analysis has been built around.

**Connecting back to the open question**

ACT and SAT's strong, specific relationship with no-record status — both are voluntary, college-oriented, often out-of-state-facing assessments — adds further support to the working hypothesis that no-record status is capturing something like geographic mobility or exit from Minnesota's tracked labor market among academically college-oriented individuals who nonetheless never completed a credential. MCA scores, by contrast, are universal in-state assessments with no inherent connection to mobility, and their link is specifically to employment instability rather than record-absence — consistent with that interpretation rather than contradicting it.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `sat.taken`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/sat-taken.html")
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

**attended.ps** (χ² = 507.944, df = 3, Cramér's V = 0.124)

Sample: N = 32,878 (19,792 never attended postsecondary; 13,086 attended without completing)

Attenders show a substantially lower persistent no-record rate (15.7% vs. 24.6%, an 8.9-point gap) and a notably higher "Never no-record" rate (62.9% vs. 51.4%, an 11.5-point gap) than non-attenders.

**Confirms the majority-share pattern established across every other variable group**

`attended.ps` was the single strongest relationship in the entire blended analysis (V = 0.182). Here it retains the strongest Cramér's V of any variable examined in this no-record table set (V = 0.124), exceeding took.ACT (0.096), cte.achievement (0.084), and every demographic variable except English learner (0.107) and Non-English home (0.102). By contrast, it weakened to V = 0.088 in the non-sustained-only table. This confirms that `attended.ps`'s outsized blended-table advantage runs substantially — though not entirely — through record-presence rather than employment stability: roughly two-thirds of its blended strength shows up here, with the remaining third showing up in the non-sustained table.

**Substantive Pattern**

This is consistent with the interpretation building across this entire three-way comparison: even brief, non-completed postsecondary exposure appears to keep people meaningfully more likely to remain visible in Minnesota's tracked employment system, whether or not that employment is itself stable. Combined with `total.cte.courses.taken`, `cte.achievement`, `took.ACT`, and `ACTCompositeScore.filled` — all four of which show their primary mechanism running through no-record status — this reinforces a coherent story: exposure to structured academic and postsecondary pathways, even without completion, is most strongly associated with staying inside Minnesota's employment-tracking system, while the more modest remaining effect on genuine job stability is a secondary, smaller story.

This completes the full three-way comparison across every variable group (demographics, HS characteristics, HS academics, postsecondary attendance).

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of No-Record Persistence (`no_record.cat`)

**Overview**

This stage repeated the full Step 3 bivariate analysis a third time, using `no_record.cat` — persistence of having no Minnesota employment record, isolated from employment instability — as the outcome. Combined with the blended and non-sustained-only analyses, this completes a full three-way decomposition across all sixteen variables. The results are the clearest yet: most variables that weakened when isolated to non-sustained employment come back at or above their original blended strength here, confirming that record-absence — not employment instability — is the dominant mechanism behind most of the disparities identified in this project.

**Variables confirmed as no-record-driven (majority or entirety of their blended effect lives here)**

| Variable | Blended | Non-sustained | No-record | Share attributable to no-record |
|---------------|---------------|---------------|---------------|---------------|
| English learner | V=0.096 | V=0.063 | **V=0.107** | Nearly all |
| Non-English home | V=0.072 | V=0.061 | **V=0.102** | Nearly all |
| took.ACT | V=0.078 | V=0.047 | **V=0.096** | Majority |
| ACTCompositeScore.filled | η²≈1.4% | ns | **η²≈1.91%** | Essentially all |
| attended.ps | V=0.182 | V=0.088 | **V=0.124** | \~Two-thirds |
| cte.achievement | V=0.098 | V=0.041 | **V=0.084** | Majority |
| sat.taken | V=0.043 | ns | **V=0.050** | Essentially all |
| total.cte.courses.taken | η²≈2.2% | η²≈0.17% | **η²≈1.54%** | Majority |
| RaceEthnicity (AI specifically) | 21.2-pt gap | 2.3-pt gap | 14.6-pt gap | \~Two-thirds |
| edr | V=0.021 | ns | V=0.034 | Essentially all |

This is the largest cluster by far — ten of sixteen variables. Every one of these was flagged in the non-sustained table as weakening dramatically; every one of them is confirmed here as recovering most or all of that strength.

**Variables confirmed as non-sustained-driven (mechanism runs the other way)**

| Variable        | Blended  | Non-sustained | No-record                    |
|-----------------|----------|---------------|------------------------------|
| Gender          | V=0.074  | **V=0.091**   | V=0.016 (barely significant) |
| Economic status | V=0.070  | **V=0.081**   | V=0.033                      |
| MCA.M           | η²≈0.2%  | **η²≈0.25%**  | ns                           |
| MCA.R           | η²≈0.07% | **η²≈0.14%**  | ns                           |

Four variables, all confirmed in this table as showing negligible-to-null relationships with record-absence specifically. Gender is the cleanest example in the entire project — its blended and non-sustained relationships are real, but its no-record relationship is barely distinguishable from noise.

**Variables genuinely mixed between both mechanisms**

| Variable | Non-sustained | No-record | Note |
|------------------|------------------|------------------|------------------|
| SpecialEdStatus | V=0.076 | V=0.067 | Splits nearly evenly |
| RaceEthnicity (Black specifically) | 7.8-pt gap | 5.0-pt gap | Splits roughly evenly |
| MCA.S | η²≈0.37% | η²≈0.14% | Only academic variable significant in both |
| pseo.participant | V=0.042 (complex shape) | V=0.043 (cleaner, protective shape) | Both real, different shapes |
| ap.exam | V=0.030 | V=0.041 | Weak in both, non-monotonic |
| Dem_Desc | V=0.019 | V=0.026 | Weak in both |

**Variables with no meaningful relationship to either mechanism**

`avg.unemp.rate` and `avg.wages.pct.state` were weakly significant only in the blended table and non-significant in both the non-sustained and no-record tables — the honest conclusion, reached only after running all three tables, is that their original blended significance was very likely a large-sample artifact rather than a real, attributable relationship with either mechanism.

**The central finding of this entire three-way comparison**

Disconnection in this population resolves cleanly into two distinct phenomena, and the vast majority of individual characteristics examined here belong predominantly to one or the other:

1.  **A "record-absence" cluster** (the larger group): postsecondary/academic exposure variables (attended.ps, took.ACT, ACTCompositeScore, CTE achievement and course count, SAT-taking), English learner and non-English-home status, American Indian identity, and regional geography (edr) all relate mainly to whether someone shows up in Minnesota's employment system at all.

2.  **A "genuine instability" cluster** (smaller but clear): Gender, Economic status, and MCA Math/Reading scores relate specifically to employment instability itself, largely independent of record-presence.

3.  **A genuinely mixed group**: SpecialEdStatus, Black identity specifically (distinct from American Indian), MCA Science, and PSEO participation don't sort cleanly into either camp — each shows a real, roughly comparable relationship with both mechanisms, though often with different shapes.

**Connecting back to the open question about "No MN employment record"**

This three-way comparison offers the most substantive, if still indirect, evidence gathered on this question. The record-absence cluster is dominated by variables plausibly tied to geographic mobility and out-of-state orientation: college-oriented assessments (ACT, SAT), postsecondary exposure, English learner and non-English-home status, and American Indian identity specifically (a population with documented patterns of interstate and reservation-based mobility in this region). This is consistent with a working hypothesis that "no MN record" substantially reflects migration out of the state's tracked labor market, rather than a uniform "labor force exit" story. The genuine-instability cluster (Gender, Economic status, MCA Math/Reading) looks structurally different — tied to access and structural disadvantage rather than mobility — reinforcing that these two clusters likely reflect two different underlying life circumstances, not two measurement artifacts of the same thing.

**Bottom line**

The decision to split `disconnected.cat` into `non_sustained.cat` and `no_record.cat` has now been fully vindicated across all sixteen variables examined in Step 3. What looked like a single, uniform disconnection story in the blended analysis is actually two substantially different stories with different demographic, academic, and geographic correlates — and for most variables, one story clearly dominates. This distinction should be treated as a foundational finding for how the rest of this project's disconnection work gets framed and reported, not a technical footnote.

<br>

## Full Synthesis: Workforce Disconnection Among Non-College Graduates in Southeast Minnesota

**Purpose and Approach**

This analysis examined workforce participation trajectories among Southeast Minnesota high school graduates who did not complete a postsecondary credential (N = 32,878), tracked across the first five years after exiting formal education. Unlike the earlier DV1–DV5 project, which modeled single time-point outcomes for the full cohort and completers, this analysis is trajectory-based and confirmatory/descriptive rather than multivariate — built around a persistence measure (0, 1–2, 3, or 4–5 years disconnected) and examined through cross-tabulations (Cramér's V) and ANOVA (η²) rather than logistic regression.

**Step 1–2: The Descriptive Foundation**

Persistent disconnection is close to the modal outcome for this population: 46.5% of non-completers are Persistently disconnected (4–5 of 5 years), far exceeding Consistently attached (16.2%). Year-by-year tracking revealed a sorting dynamic — unstable employment situations resolve over time into one of two stable poles: Sustained WF–Southeast (growing from 24.9% to 30.5% across Years 1–5) or No MN employment record (growing from 25.9% to 30.6%), which converge to nearly identical size by Year 5.

The critical methodological decision of this analysis was splitting the blended disconnection measure into two independent constructs: **non_sustained.cat** (persistence of unstable employment) and **no_record.cat** (persistence of no MN employment record). Neither component's "Persistent" rate alone (13.1% and 21.1%, respectively) explained the blended 46.5% figure, and a driver breakdown of the persistently-disconnected subgroup confirmed why: No-record driven (45.3%) was the largest single driver, exceeding both Non-sustained driven (28.1%) and Mixed/switching between the two (26.6%). Persistent disconnection in this population is most often a single, consistent problem — and more often a record-absence problem than an employment-instability one.

**Step 3: The Central Finding — Two Distinct, Separable Mechanisms**

Running all sixteen individual-characteristic variables against all three outcome specifications (blended, non-sustained-only, no-record-only) produced the defining result of this entire analysis: disconnection in this population is not one phenomenon with many correlates, but at least two distinct phenomena with substantially different correlates. Most variables sort clearly into one of two clusters:

*The record-absence cluster (the larger group):* attended.ps, took.ACT, ACTCompositeScore.filled, cte.achievement, total.cte.courses.taken, sat.taken, English learner status, Non-English home status, American Indian identity specifically, and edr all showed their blended-table relationships driven substantially or almost entirely by no-record status, not employment instability. In several cases (ACTCompositeScore, sat.taken, edr), the non-sustained relationship dropped to complete non-significance while the no-record relationship remained strong or even strengthened.

*The genuine-instability cluster (smaller but clear):* Gender, Economic status, and MCA Math/Reading scores related specifically to employment instability, largely independent of record-presence. Gender is the cleanest example in the whole project — its Cramér's V *strengthened* for non-sustained employment (0.074 → 0.091) while *collapsing* to near-zero for no-record status (0.016, barely significant).

*A genuinely mixed group:* SpecialEdStatus, Black identity (distinct from American Indian), MCA Science, and PSEO participation showed real relationships with both mechanisms, often in different shapes rather than a single consistent story.

*A null result, resolved only by running all three tables:* avg.unemp.rate and avg.wages.pct.state were weakly significant in the blended measure but null in both mechanism-specific tables — their original significance was very likely a large-sample artifact, not a real, attributable relationship with either pathway.

**Headline Findings**

1.  **Attended.ps (postsecondary attendance without completion) is the single strongest predictor of disconnection found anywhere in this analysis** (blended V = 0.182, the strongest relationship in the entire Step 3 stage), and roughly two-thirds of that strength runs through record-presence rather than employment stability. Even brief postsecondary exposure is strongly associated with staying visible in Minnesota's employment system.

2.  **CTE engagement (both achievement level and course count) is the second-strongest predictor**, and its effect is similarly majority-driven by record-absence. This parallels CTE's role throughout the entire DV1–DV5 and prior non-completer regression work as one of the most durable positive predictors of workforce attachment.

3.  **Standardized test performance splits cleanly by which specific test.** ACT composite score and SAT-taking are both almost purely record-absence markers; MCA Math and Reading are almost purely employment-instability markers; MCA Science is the sole academic variable meaningfully tied to both. This is a genuinely new and specific finding — not simply "academic achievement matters," but a clean mechanism-level split between test types.

4.  **Race/ethnicity shows the largest raw disparities in the entire analysis, and the mechanism differs by group.** American Indian graduates' disparity (66.0% persistently disconnected vs. 44.8% for White graduates in the blended measure) is disproportionately about record-absence (roughly two-thirds of the gap). Black graduates' disparity (62.8% blended) splits more evenly between the two mechanisms. This directly parallels the persistent, unexplained Black-White (and here, especially American Indian) gap that ran through the DV1, DV2, and DV5 multivariate work on completers — now confirmed at the bivariate level in a population that never completed any postsecondary credential, meaning this disparity is not specific to the postsecondary pathway.

5.  **Gender and Economic status are the clearest examples of "genuine instability" variables**, with effects that hold steady or strengthen once no-record status is removed from the outcome, and nearly vanish once non-sustained employment is removed instead.

**Connecting Back to the Open Question**

Earlier in this analysis, what "No MN employment record" actually represents was flagged as an unresolved ambiguity — whether it reflects leaving the state, exiting the labor force, informal work, or some mix. This full bivariate stage doesn't resolve that question definitively, but it does provide the clearest pattern-based evidence gathered anywhere in this project: the record-absence cluster is dominated by variables plausibly tied to geographic mobility and out-of-state orientation — college-oriented assessments (ACT, SAT), postsecondary exposure, English learner and non-English-home status, and American Indian identity specifically. The genuine-instability cluster (Gender, Economic status, MCA Math/Reading) looks structurally different, tied to access and structural disadvantage rather than mobility. This is consistent with a working hypothesis that "no MN record" substantially reflects migration out of Minnesota's tracked labor market for a meaningful share of this population, rather than a single, uniform "labor force exit" story — but it remains a hypothesis supported by pattern, not a confirmed fact.

**Connecting Back to the Full Body of Work**

This analysis reinforces two of the most important findings from the entire project to date. First, the outsized importance of any postsecondary exposure — echoing `attended.ps` and `ps.grad` throughout the DV1–DV5 and non-completer regression models — shows up again here as the single strongest bivariate predictor in a completely different analytic framework. Second, the persistent, unexplained racial disparity — especially for American Indian and Black graduates — appears regardless of population (completers or non-completers), regardless of outcome (employment attachment, credential completion, or disconnection persistence), and regardless of analytic method (multivariate logistic regression, trajectory-based persistence measures, or simple bivariate cross-tabulation). This is the closest thing to a fully triangulated finding in the whole body of work.

**Bottom Line**

The single most important methodological decision in this analysis — separating disconnection into non-sustained employment and no-record status — was fully vindicated. What appeared to be one disconnection story in the blended measure is genuinely two distinct stories with different correlates, different demographic patterns, and (plausibly) different underlying life circumstances. Postsecondary exposure and CTE engagement remain the strongest positive levers identified anywhere in this project, disproportionately operating through record-presence. The Black-White and American Indian-White disparities remain the most robust and troubling unresolved findings across this entire body of work, present in every population, outcome, and method examined so far — and this analysis adds the further nuance that these disparities may run through different mechanisms for different groups, a distinction worth carrying into any policy-facing summary of this work.

```{r}
write_csv(updated.master, "Data/SLEDS/Masters/Master-non-college-grads-disconnection.csv")
```
