---
title: "Post-secondary pathways"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv") %>%
  mutate(ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "1", "Public, 4-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "4", "Public, 2-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "11", "Other", as.character(ps.grad.InstitutionSector)),
         ps.grad.location = ifelse(ps.grad.location == "PS grad EDR", "PS grad region", ps.grad.location))

kable(names(master.original))

kable(head(master.original))
```

# Introduction

The previous analysis focused on how demographics, high school characteristics, accomplishments, and postsecondary experiences predict meaningful employment in the region versus disconnection or statewide mobility. The findings highlighted postsecondary experiences as the most powerful predictors of long-term outcomes, with clear dual pipelines: **CTE and sub-baccalaureate pathways anchoring students regionally**, and **academic rigor with higher credentials leading to statewide mobility**.

Given the centrality of postsecondary variables in shaping workforce attachment and mobility, the next phase of analysis shifts focus. Specifically, we examine how **demographic background, high school characteristics, and high school accomplishments** influence key postsecondary outcomes such as:

- Whether students attend a postsecondary institution.
- Whether they graduate from postsecondary.
- The sector and location of their postsecondary institution.
- The highest credential level attained.

By analyzing these linkages, we can identify which student attributes and high school experiences most strongly predict postsecondary pathways, and in turn, how these pathways shape regional workforce retention. This step deepens the understanding of **how early life circumstances and schooling experiences channel students into postsecondary trajectories** that later drive employment outcomes.

<br>

# Methodology

To evaluate the relationships between student attributes and postsecondary experiences, the analysis uses a combination of **cross-tabulations, Cramer’s V effect size, and ANOVA tests**, following the same approach applied in the prior employment-focused analysis.

1.  **Cross-tabulations** were constructed between demographic variables (e.g., gender, race/ethnicity, economic status, special education), high school characteristics (e.g., rural/metro classification, county wage/unemployment context), and accomplishments (e.g., CTE engagement, ACT/AP participation, MCA performance) against postsecondary outcomes. These tables provide a descriptive view of group-level differences.

2.  **Cramer’s V** was calculated for categorical relationships to assess the strength of association between predictor variables and postsecondary outcomes. Effect sizes were interpreted on a conventional scale (weak, moderate, strong) to prioritize meaningful predictors.

3.  **ANOVA tests** were used where continuous predictors (e.g., test scores, number of CTE courses taken) were compared across postsecondary outcome categories. This allowed for assessing mean differences and the statistical significance of those differences.

Results will again be summarized across three tiers of predictive value: **strong predictors** (clear, consistent associations), **moderate predictors** (some association, less consistent), and **weak or inconsistent predictors** (low effect sizes, candidates for exclusion). This structure provides both clarity and comparability with the employment outcome analysis, while highlighting the mechanisms by which students are sorted into postsecondary pathways that ultimately drive workforce outcomes.

Each dependent variable has the following categories;

```{r dependent variable categories}
data <- master.original %>%
  select(ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location) %>%
  mutate(ps.pathway = case_when(
    
    # Never attended or attended but did not graduate
    highest.cred.level == "Did not graduate ps" & 
      ps.grad.location == "Did not grad" ~ "Did not graduate ps",
    highest.cred.level == "No college"         ~ "Did not graduate ps",
    
    # Sub-baccalaureate by location
    highest.cred.level == "Less than Bachelors" & 
      ps.grad.location == "PS grad region"      ~ "Sub-baccalaureate - region",
    highest.cred.level == "Less than Bachelors" & 
      ps.grad.location == "PS grad MN"          ~ "Sub-baccalaureate - MN",
    highest.cred.level == "Less than Bachelors" & 
      ps.grad.location == "PS grad outside MN"  ~ "Sub-baccalaureate - outside MN",
    
    # Bachelor or higher by location
    highest.cred.level == "Bachelors or higher" & 
      ps.grad.location == "PS grad region"      ~ "Bachelor or higher - region",
    highest.cred.level == "Bachelors or higher" & 
      ps.grad.location == "PS grad MN"          ~ "Bachelor or higher - MN",
    highest.cred.level == "Bachelors or higher" & 
      ps.grad.location == "PS grad outside MN"  ~ "Bachelor or higher - outside MN",
    
    # Currently attending - treat as its own category
    ps.grad.location == "Attending ps"          ~ "Attending ps",
    highest.cred.level == "Attending ps"        ~ "Attending ps",
    
    TRUE ~ NA_character_),
    across(c("ps.grad":"ps.grad.location"), factor),
    ps.pathway = fct_relevel(ps.pathway, "Did not graduate ps", "Sub-baccalaureate - region", "Bachelor or higher - region", "Sub-baccalaureate - MN", "Bachelor or higher - MN", "Sub-baccalaureate - outside MN", "Bachelor or higher - outside MN"))

```

**ps.grad**

```{r echo = FALSE}
levels(data$ps.grad)
```

**ps.grad.InstitutionSector**

```{r echo = FALSE}
levels(data$ps.grad.InstitutionSector)
```

**highest.cred.level**

```{r echo = FALSE}
levels(data$highest.cred.level)
```

**ps.grad.location**

```{r echo = FALSE}
levels(data$ps.grad.location)
```

**ps.pathway**

```{r echo = FALSE}
levels(data$ps.pathway)
```

For the `ps.grad` outcomes, we will explore relationships with indpendent variables among the full population. But for the remaining dependent variables, we will subset the population to explore college completers only.

<br>

# Predictors

For all of these, we will use the predictors that were in Tier 1 and Tier 2 of the previous cross-tab analysis.

- Demographics
  - `Gender`
  - `economic.status`
  - `RaceEthnicity`
  - `SpecialEdStatus`
- High school characteristics
  - `edr`
  - `avg.unemp.rate`
- High school accomplishments
  - `took.ACT`
  - `ACTCompositeScore`
  - `ap.exam`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`

<br>

# `ps.grad` - full population

## Demographics

Here are the variables we will be exploring;

- `Gender`
- `economic.status`
- `RaceEthnicity`
- `SpecialEdStatus`

**DEMOGRAPHIC VARIABLES AND POSTSECONDARY GRADUATION**

Overall, 46% of graduates in this analysis completed a postsecondary credential, with attending ps graduates excluded. Demographic variables show moderate to strong associations with credential completion, and the patterns are consistent with what was observed in the workforce outcome bivariate stage.

Gender shows a meaningful association (φ=0.134). Females complete postsecondary credentials at a substantially higher rate (52.9%) than males (39.5%), a gap of 13.4 percentage points. This is consistent with the Stage 3 finding that females attend postsecondary at higher rates and remain enrolled longer, and confirms that the female postsecondary advantage extends through credential completion.

Economic status is a strong predictor of credential completion (φ=0.269). Non-low-income graduates complete at 55.1% compared to just 26.2% for low-income graduates — a gap of nearly 29 percentage points. This is the largest gap of any binary demographic variable and reinforces the Stage 3 finding that economic disadvantage is the most consequential demographic barrier to postsecondary pathway success.

Special education status shows a similarly strong association (φ=0.255). Only 14.4% of special education graduates complete a postsecondary credential compared to 51.1% of non-special education graduates — a gap of 36.7 percentage points, the largest absolute gap of any demographic variable examined. This confirms the Stage 3 finding that special education status is associated with dramatically lower postsecondary participation and extends it to show that completion rates are even more severely affected than attendance rates alone would suggest.

Race and ethnicity shows the strongest overall association (V=0.299) and substantial variation across groups. White graduates show the highest completion rate (53.1%), followed closely by Asian/PI graduates (51.8%). Black (28.9%), Hispanic (26.0%), and American Indian (14.6%) graduates show substantially lower completion rates. The Unknown category shows a completion rate of only 8.1%, which is anomalously low and likely reflects a data coding artifact or a specific subpopulation rather than a meaningful racial or ethnic group — this category should be interpreted with caution and flagged in any formal reporting. American Indian graduates show the lowest completion rate of any identified racial or ethnic group, though cell sizes remain small (n=553) and estimates should be interpreted cautiously.

Across the demographic group, the clearest finding is that economic disadvantage and special education status are the two strongest barriers to postsecondary credential completion in Southeast Minnesota, with gaps exceeding 29 and 36 percentage points respectively. The gender gap favoring females is consistent and meaningful. Racial and ethnic disparities are substantial, with Black, Hispanic, and American Indian graduates completing at roughly half the rate of White and Asian/PI graduates. These patterns will be important context for interpreting the postsecondary pathway distributions in the multivariate stages.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/Demographics/gender.html")
```

### `economic.status`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/Demographics/economic-status.html")

```

### `RaceEthnicity`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

- `edr`
- `avg.unemp.rate`

**HIGH SCHOOL CHARACTERISTICS AND POSTSECONDARY GRADUATION**

EDR shows no meaningful association with postsecondary credential completion (φ=0.004, p=0.199). Completion rates are essentially identical across EDR 10 (46.2%) and EDR 9 (45.7%). Whether a graduate attended high school in the Southeast or South Central economic development region has no bearing on whether they complete a postsecondary credential. This is consistent with the Stage 3 finding that EDR differences in workforce outcomes emerge after credential completion through geographic sorting, not before it through differential completion rates.

Average local unemployment rate shows a statistically significant and substantively meaningful association with credential completion (F=895.6, p=0.000). Postsecondary graduates come from communities with meaningfully higher average unemployment (difference=0.252) than non-completers. This is a notable finding: higher local unemployment is associated with greater likelihood of credential completion, suggesting that weak local labor markets increase the relative attractiveness of postsecondary investment — graduates from communities with fewer immediate employment opportunities are more likely to pursue and complete credentials rather than entering the labor market directly. This inverts the Stage 3 pattern, where higher unemployment was associated with out-of-region employment and labor force marginality among those who entered the workforce directly.

::: panel-tabset
### Dem_Desc

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school characteristics/edr.html")
```

### avg.unemp.rate

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad, data = data)

summary <- aov(avg.unemp.rate ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

- High school accomplishments
  - `took.ACT`
  - `ACTCompositeScore`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`

**HIGH SCHOOL ACCOMPLISHMENT VARIABLES AND POSTSECONDARY GRADUATION**

The high school accomplishment variables show the strongest and most consistent associations with postsecondary credential completion of any variable group, and the direction of associations reveals a clear academic orientation versus CTE engagement divide that mirrors the workforce outcome findings.

took.ACT is the strongest categorical predictor of credential completion (φ=0.383). ACT takers complete postsecondary credentials at 59.7% compared to just 19.4% for non-takers — a gap of 40.3 percentage points. This is the largest single gap of any accomplishment variable and confirms that ACT participation is functioning as a near-perfect proxy for postsecondary intent and follow-through in Southeast Minnesota.

ap.exam shows a very strong association with credential completion (φ=0.299). AP exam takers complete postsecondary credentials at 74.9% compared to just 38.4% for non-takers — a gap of 36.5 percentage points. This is the second strongest categorical predictor of credential completion after took.ACT (φ=0.383), and the two variables are measuring closely related dimensions of academic orientation and postsecondary intent. Only 25.1% of AP exam takers fail to complete a credential, compared to 61.6% of non-takers. AP participation in high school is among the strongest available signals that a graduate will successfully navigate the postsecondary pathway to credential completion.

ACT Composite Score shows the strongest F-statistic of any continuous variable (F=9081.7). Postsecondary graduates score 3.49 composite points above non-completers — a substantively large difference on a 36-point scale. Academic achievement as measured by the ACT is a powerful predictor of credential completion, operating independently of and likely additively with ACT participation itself.

MCA Math shows a strong association (F=8689.3) with graduates scoring 0.588 standard deviations above non-completers. MCA Reading is similarly strong (F=10459.2) with graduates scoring 0.570 above non-completers. The two measures are essentially equivalent in their association with credential completion and both confirm that academic achievement in high school is a robust predictor of postsecondary success.

cte.achievement shows a moderate but noteworthy association (V=0.124), and the direction is the inverse of all other accomplishment variables. No CTE graduates show the highest completion rate (54.3%), CTE participants are intermediate (47.4%), and CTE concentrators and completors show the lowest completion rate (39.5%). This monotonic negative gradient confirms that CTE engagement and postsecondary credential completion are in tension — the workforce-first pathway and the college completer pathway are selecting different students, and those most committed to CTE are least likely to complete postsecondary credentials.

Total CTE courses taken reinforces this finding (F=2264.7). Postsecondary graduates take on average 1.005 fewer CTE courses than non-graduates — approximately one full course less. The direction is unambiguous and consistent with cte.achievement: greater CTE investment in high school is associated with lower likelihood of postsecondary credential completion.

Taken together, the accomplishment group findings clarify the mechanism underlying the three-pathway structure. Academic achievement variables — took.ACT, ap.exam, ACT score, MCA Reading, MCA Math — are positively and strongly associated with credential completion, while CTE engagement variables are negatively associated with it. These are not simply weak students versus strong students — they represent genuinely different orientations toward post-high school life that are already visible in high school course-taking and testing behavior.

::: panel-tabset
### `took.ACT`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/took-ACT.html")
```

### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `ap.exam`

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/ap-exam.html")
```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### MCA.R

```{r}
summary <- aov(MCA.R ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### MCA.S

```{r}
summary <- aov(MCA.S ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

### cte.achievement

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/cte-achievement.html")
```

### total.cte.courses.taken

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```
:::

<br>

# `ps.grad.InstitutionSector` - college completers only

## Demographics

Here are the variables we will be exploring;

- `Gender`
- `economic.status`
- `RaceEthnicity`
- `SpecialEdStatus`

**DEMOGRAPHIC VARIABLES AND POSTSECONDARY INSTITUTION SECTOR (Among credential completers only, n=47,959)**

Among the 47,959 graduates who completed a postsecondary credential, the distribution across institution sectors is: public 4-year (44.5%), public 2-year (25.7%), private not-for-profit 4-year (18.3%), multiple sectors (9.3%), and other (2.2%).

Gender shows a moderate association with institution sector (V=0.134). The most striking difference is in public 2-year completion — males complete at public 2-year institutions at 30.8% compared to 21.6% for females, a gap of 9.2 percentage points. Females show higher rates of multiple sector completion (11.0% vs. 7.2%) and other institution completion (3.1% vs. 1.1%). Public 4-year completion is essentially identical across genders (44.5% for both), and private not-for-profit 4-year completion slightly favors females (19.9% vs. 16.3%). The gender divide between public 2-year and private not-for-profit 4-year is the clearest signal — males disproportionately complete sub-baccalaureate credentials at community and technical colleges, while females show greater representation at private 4-year institutions.

Economic status shows a strong association (V=0.182). Low-income graduates complete at public 2-year institutions at 41.1% compared to just 22.3% for non-low-income graduates — a gap of nearly 19 percentage points. Non-low-income graduates dominate public 4-year completion (47.2% vs. 32.2%) and private not-for-profit 4-year completion (19.4% vs. 13.1%). The economic status gradient in institution sector is monotonic and consistent: as economic disadvantage increases, credential completion shifts from 4-year toward 2-year institutions. This directly anticipates the workforce outcome implications — low-income completers are more likely to hold sub-baccalaureate credentials associated with regional employment, while non-low-income completers are more likely to hold 4-year credentials associated with out-of-region careers.

Special education status shows a similarly strong association (V=0.184). Special education graduates who complete credentials do so overwhelmingly at public 2-year institutions (62.9% vs. 24.0% for non-special education graduates), and at substantially lower rates at public 4-year (17.6% vs. 45.7%) and private not-for-profit 4-year (9.1% vs. 18.7%) institutions. The pattern mirrors economic status but is even more pronounced — among those who complete credentials, special education graduates are concentrated almost exclusively in public 2-year institutions.

Race and ethnicity shows a moderate association (V=0.111) with substantively important within-group variation. The Unknown category again shows an anomalous pattern — 90.0% public 2-year completion — which reinforces the interpretation that this category captures a specific non-traditional student population rather than simply missing data. Among identified groups, Hispanic graduates show the highest public 2-year completion rate (44.6%), followed by Black graduates (33.7%) and American Indian graduates (33.3%), all substantially above White graduates (23.9%). Asian/PI graduates show the highest private not-for-profit 4-year completion rate (23.0%) and the highest public 4-year rate (47.6%). White graduates show the highest public 4-year rate among larger groups (45.9%). These patterns are consistent with economic access constraints directing minority graduates toward more accessible 2-year institutions, though the Asian/PI pattern suggests a distinctly different credential pathway.

Across the demographic group, the dominant finding is that economic disadvantage, special education status, male gender, and minority racial and ethnic status are all associated with public 2-year completion at the expense of 4-year completion. Given the strong workforce outcome associations shown in Stage 3 — where public 2-year completion predicted regional employment and 4-year completion predicted out-of-region careers — these demographic patterns have direct implications for understanding which graduates end up in which workforce pathway.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(!ps.grad.InstitutionSector %in% c("Attending ps", "Did not grad", "Never attended ps"))

tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/Demographics/Gender.html")
```

### `economic.status`

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/Demographics/economic-status.html")
```

### `RaceEthnicity`

```{r}

tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

- `Dem_Desc`
- `avg.unemp.rate`

**HIGH SCHOOL CHARACTERISTICS AND POSTSECONDARY INSTITUTION SECTOR (Among credential completers only, n=47,959)**

EDR shows a weak but statistically significant association with institution sector (V=0.077). The most notable differences are in multiple sector completion (10.0% for EDR 10 vs. 7.5% for EDR 9) and the other institution category (1.6% for EDR 10 vs. 3.7% for EDR 9). Public 2-year completion is identical across EDRs (25.7% both), and public 4-year completion is very similar (44.0% vs. 45.7%). The overall picture is that where a graduate attended high school within Southeast Minnesota has essentially no bearing on what type of institution they complete a credential from — the sector distribution is near-equivalent across EDR 9 and EDR 10.

Average local unemployment rate shows a statistically significant but modest association with institution sector (F=45.1). The Tukey HSD comparisons reveal a clear ordering: graduates from the other institution category come from the highest unemployment communities, followed by multiple sector (10), then private not-for-profit 4-year (2), then public 2-year and public 4-year — which are not significantly different from each other (difference=-0.010, p=0.963). The other and multiple sector categories represent less conventional credential paths and their association with higher local unemployment is consistent with graduates from weaker labor markets pursuing less traditional postsecondary routes. The absence of a meaningful difference between public 2-year and public 4-year completers in local unemployment rates is notable — it suggests that the choice between 2-year and 4-year public institutions is not primarily driven by local labor market conditions but rather by other factors, likely academic preparation and economic resources as seen in the demographic findings above.

::: panel-tabset
### `edr`

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school characteristics/edr.csv")
```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

- `took.ACT`

- `ACTCompositeScore`

- `MCA.M`

- `MCA.R`

- `MCA.S`

- `cte.achievement`

- `total.cte.courses.taken`

**HIGH SCHOOL ACCOMPLISHMENT VARIABLES AND POSTSECONDARY INSTITUTION SECTOR (Among credential completers only, n=47,959)**

The high school accomplishment variables reveal a sharp and consistent academic achievement gradient across institution sectors, with CTE engagement showing the opposite pattern.

took.ACT is the strongest categorical predictor of institution sector (V=0.357). Non-ACT takers complete at public 2-year institutions at a dramatically higher rate (62.1%) than ACT takers (19.7%), while ACT takers dominate public 4-year (49.0% vs. 17.1%) and private not-for-profit 4-year (20.1% vs. 7.6%) completion. The other institution category is notable — 4.6% of non-ACT takers versus 1.8% of ACT takers — consistent with non-traditional credential paths being more common among those who never pursued academic preparation for four-year institutions.

ap.exam shows the strongest association with institution sector of any categorical accomplishment variable (V=0.330). AP exam takers complete at private not-for-profit 4-year institutions at 28.3% compared to just 13.1% for non-takers — more than double — and at public 4-year institutions at 54.8% versus 39.1%. By contrast, non-takers complete at public 2-year institutions at 34.9% compared to just 8.0% for AP exam takers — a gap of nearly 27 percentage points. AP participation is the strongest single predictor of 4-year over 2-year credential completion among all categorical variables examined, and the private not-for-profit 4-year advantage is particularly pronounced. The other institution category shows a higher non-AP-taker rate (2.9% vs. 1.0%), consistent with non-traditional credential paths being concentrated among those who did not pursue advanced academic coursework in high school.

ACT Composite Score shows the largest mean differences of any continuous variable (F=2391.4). The ordering is clear: private not-for-profit 4-year graduates score highest, followed by public 4-year, then multiple sectors, then public 2-year, then other. Public 4-year graduates score 4.38 composite points above public 2-year graduates — a substantively large difference — and private not-for-profit 4-year graduates score an additional 1.49 points above public 4-year graduates. The other category scores lowest by a considerable margin, 2.40 points below multiple sectors and 5.32 points below private not-for-profit 4-year.

MCA Math (F=1325.3), MCA Reading (F=1550.4), and MCA Science (F=1298.4) all show the same sector ordering and the pattern is highly consistent across measures. In all three assessments: private not-for-profit 4-year graduates score highest, public 4-year second, multiple sectors third, public 2-year fourth, and other lowest. The public 4-year versus public 2-year gaps are approximately 0.62 (Math), 0.53 (Reading), and 0.57 (Science) standard deviations — meaningful differences that confirm the academic achievement gradient operates similarly across all three assessments.

cte.achievement shows a strong association (V=0.177) in the opposite direction from academic achievement. CTE concentrators and completors complete at public 2-year institutions at 37.3% — more than double the rate for No CTE graduates (14.4%) — while No CTE graduates dominate private not-for-profit 4-year completion (26.7% vs. 11.8% for concentrators). Public 4-year completion is relatively similar across CTE categories (39.9%, 46.3%, 47.7%), with No CTE graduates slightly more represented. The CTE concentrator advantage in public 2-year completion is the clearest expression of the workforce-first pathway structure within the credential completion population — those most committed to CTE earn sub-baccalaureate credentials at community and technical colleges.

Total CTE courses taken (F=1290.2) confirms this with a clear dose-response relationship. Public 2-year graduates take the most CTE courses on average, followed by other, multiple sectors, public 4-year, and private not-for-profit 4-year. Public 2-year graduates take approximately 1.92 more CTE courses than public 4-year graduates and 2.80 more than private not-for-profit 4-year graduates — the largest continuous variable gaps in this analysis.

The accomplishment variable findings for institution sector are among the most analytically important in this postsecondary stage. Academic achievement sorts graduates decisively toward 4-year institutions, while CTE engagement sorts them toward 2-year institutions — and these two sorting mechanisms operate in opposite directions within the credential completer population. Combined with the workforce outcome findings from Stage 3, this establishes a clear chain: high CTE engagement → public 2-year completion → regional employment, and high academic achievement → private or public 4-year completion → out-of-region employment.

::: panel-tabset
### `took.ACT`

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/took-ACT.html")

```

### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `ap.exam`

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/ap-exam.html")
```

### cte.achievement

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/cte-achievement.html")

```

### total.cte.courses.taken

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

### MCA.M

```{r}
summary <- aov(MCA.M ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### MCA.R

```{r}
summary <- aov(MCA.R ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### MCA.S

```{r}
summary <- aov(MCA.S ~ ps.grad.InstitutionSector, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-InstitutionSector/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# `ps.grad.location` - college completers only

## Demographics

Here are the variables we will be exploring;

- `Gender`
- `economic.status`
- `RaceEthnicity`
- `SpecialEdStatus`

**DEMOGRAPHIC VARIABLES AND POSTSECONDARY GRADUATION LOCATION (Among credential completers only, n=47,959)**

Among credential completers, 40.5% graduated from a regional institution, 30.3% from outside Minnesota, and 29.1% from elsewhere in Minnesota.

Gender shows no meaningful association with graduation location (V=0.011, p=0.065). The distribution across regional, MN, and outside MN graduation is essentially identical for females (41.0%, 28.9%, 30.1%) and males (39.9%, 29.5%, 30.6%). Where a graduate completes their credential is not associated with gender — this contrasts with the institution sector findings where gender showed a meaningful association with 2-year versus 4-year completion. The implication is that gender differences in institution type do not translate into meaningful geographic differences in where credentials are earned.

Economic status shows a moderate association with graduation location (V=0.126). Low-income graduates complete regional credentials at a substantially higher rate (52.6% vs. 37.9% for non-low-income graduates) and are considerably less likely to complete outside Minnesota (19.8% vs. 32.6%). MN graduation rates are modestly lower for low-income graduates (27.5% vs. 29.5%). The economic gradient in graduation location is clear and consistent with what the institution sector findings would predict — low-income graduates who complete credentials are more likely to do so at nearby regional institutions, while non-low-income graduates have greater access to out-of-state institutions. This has direct workforce outcome implications given the strong association between regional graduation and regional employment shown in Stage 3.

Special education status shows a weak but significant association (V=0.059). Special education completers graduate from regional institutions at a higher rate (54.0% vs. 39.9%) and from outside Minnesota at a substantially lower rate (21.5% vs. 30.7%). MN graduation is modestly lower for special education completers (24.5% vs. 29.4%). The pattern mirrors economic status — among those who complete credentials, special education graduates are more geographically concentrated in regional institutions, likely reflecting both access constraints and the higher representation of special education graduates in public 2-year institutions shown in the institution sector analysis.

Race and ethnicity shows a moderate association (V=0.066) with substantively important variation. Hispanic (57.8%) and Unknown (58.5%) graduates show the highest regional completion rates, followed by Black (52.2%) and American Indian (32.1%) graduates, all above White graduates (39.5%). Outside MN graduation is highest among White (31.2%) and Asian/PI (29.9%) graduates, with substantially lower rates for Black (20.1%), Hispanic (19.5%), and Unknown (16.0%) graduates. MN graduation is highest among American Indian (43.2%) and Asian/PI (33.7%) graduates. The racial and ethnic variation in graduation location is consistent with the economic access constraints identified above — minority graduates who complete credentials are more likely to do so at nearby regional or in-state institutions, while White and Asian/PI graduates show greater geographic mobility in credential attainment.

The demographic findings for graduation location tell a coherent story. Economic disadvantage and special education status concentrate credential completion at regional institutions — not because of deliberate regional preference but because geographic mobility in credential pursuit is constrained by resources. These graduates will subsequently be more likely to achieve regional employment, not primarily through active regional orientation but through geographic proximity to the regional labor market at the point of credential completion. Gender, by contrast, is irrelevant to graduation location once credential completion has occurred — the gender gap in institution sector does not translate into geographic differences in where credentials are earned.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(!ps.grad.location %in% c("Attending ps", "Did not graduate ps"))

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/Demographics/gender.html")
```

### `economic.status`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/Demographics/economic-status.html")
```

### `RaceEthnicity`

```{r}

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

- `edr`
- `avg.unemp.rate`

**HIGH SCHOOL CHARACTERISTICS AND POSTSECONDARY GRADUATION LOCATION (Among credential completers only, n=47,959)**

EDR shows a meaningful association with graduation location (V=0.084). EDR 10 graduates complete credentials outside Minnesota at a substantially higher rate (32.7% vs. 24.4% for EDR 9) and at a lower rate within MN outside the region (27.6% vs. 33.0%). Regional completion rates are similar across EDRs (39.7% vs. 42.6%). This is a notable finding — EDR 10 graduates who complete credentials are more geographically mobile, with a greater tendency to pursue and complete out-of-state credentials. EDR 9 graduates who complete credentials are more likely to do so at in-state but out-of-region institutions, consistent with the Stage 3 finding that EDR 9 showed stronger meaningful MN employment at Years 5 and 10. The EDR difference in graduation location is the likely mechanism underlying that workforce outcome gap — EDR 9 completers earn credentials in Minnesota and stay in Minnesota, while EDR 10 completers are more likely to leave the state entirely.

Average local unemployment rate shows a statistically significant but very modest association with graduation location (F=19.4). MN graduates come from the highest unemployment communities, and both outside MN and regional graduates come from lower unemployment areas — with regional and outside MN graduates essentially indistinguishable from each other (difference=-0.006, p=0.907). The mean differences are small in absolute terms, and the practical significance of this finding is limited. Higher local unemployment is modestly associated with completing credentials within Minnesota rather than outside it, which may reflect that graduates from weaker labor markets are more likely to attend in-state institutions out of cost and access constraints — but the effect is too small to carry much interpretive weight.

::: panel-tabset
### `edr`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school characteristics/edr.html")
```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)

```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

- `took.ACT`

- `ACTCompositeScore`

- `ap.exam`

- `cte.achievement`

- `total.cte.courses.taken`

- `MCA.M`

- `MCA.R`

- `MCA.S`

**HIGH SCHOOL ACCOMPLISHMENT VARIABLES AND POSTSECONDARY GRADUATION LOCATION (Among credential completers only, n=47,959)**

The accomplishment variables reveal a consistent and analytically important pattern: academic achievement and orientation predict out-of-region and out-of-state credential completion, while CTE engagement predicts regional completion — the same sorting mechanism seen throughout the analysis, now operating within the credential completer population.

took.ACT shows a moderate association with graduation location (V=0.102). Non-ACT takers complete regional credentials at 51.7% compared to 38.7% for ACT takers, while ACT takers complete outside Minnesota at 31.9% versus just 20.6% for non-takers. MN completion is similar across groups (29.4% vs. 27.6%). ACT participation is associated with geographic mobility in credential completion — those who prepared academically for four-year institutions are more likely to complete credentials outside the region and outside the state.

ap.exam shows the strongest categorical association with graduation location (V=0.183). AP exam takers complete outside Minnesota at 38.5% compared to 26.1% for non-takers — a gap of 12.4 percentage points — and complete regionally at just 28.3% compared to 46.9% for non-takers. MN completion is modestly higher for AP takers (33.2% vs. 27.1%). AP participation is the strongest academic achievement indicator of out-of-state credential completion among all categorical variables examined, consistent with AP takers being among the most academically prepared and geographically mobile graduates.

ACT Composite Score shows the largest mean differences across graduation locations (F=1537.1). Outside MN graduates score 0.47 composite points above MN graduates, while regional graduates score 2.24 points below MN graduates and 2.72 points below outside MN graduates. The academic achievement gradient across graduation location is steep and consistent — every composite point increase is associated with a shift away from regional toward MN and out-of-state completion.

MCA Math (F=618.9), MCA Reading (F=553.8), and MCA Science (F=771.0) all show the same ordering: outside MN graduates score highest, MN graduates second, and regional graduates lowest. The regional versus outside MN gaps are approximately 0.33 (Math), 0.25 (Reading), and 0.35 (Science) standard deviations — meaningful differences that are consistent across all three assessments. Outside MN graduates score modestly but significantly above MN graduates on all three measures, suggesting that even within the out-of-region completer population, academic achievement continues to sort graduates geographically.

cte.achievement shows a moderate association in the expected direction (V=0.082). CTE concentrators and completors graduate regionally at 45.9% compared to 32.8% for No CTE graduates — a gap of 13.1 percentage points. The gradient is monotonic: concentrators show the highest regional completion, participants intermediate (42.8%), and No CTE graduates the lowest. Outside MN completion shows the reverse pattern — No CTE graduates complete outside the state at 35.0% compared to 26.9% for concentrators. This confirms that CTE engagement is associated with geographic proximity in credential completion just as it is in workforce attachment, and that the absence of CTE engagement predicts geographic departure at the credential stage as well as the employment stage.

The accomplishment findings for graduation location are among the most analytically important across all postsecondary dependent variables. They establish that the geographic sorting visible in workforce outcomes at Years 5 and 10 begins at the credential completion stage — high academic achievement and low CTE engagement predict out-of-state completion, while CTE engagement predicts regional completion. The chain from high school to workforce is now fully traceable: CTE engagement → regional credential completion → regional employment; high academic achievement → out-of-state credential completion → out-of-state employment or no MN record.

::: panel-tabset
### `took.ACT`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school accomplishments/took-ACT.html")
```

### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)

```

### `ap.exam`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
        file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school accomplishments/ap-exam.html")
```

### `cte.achievement`

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school accomplishments/cte-achievement.html")
```

### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school characteristics/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### `MCA.R`

```{r}
summary <- aov(MCA.R ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### `MCA.S`

```{r}
summary <- aov(MCA.S ~ ps.grad.location, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-grad-location/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# `highest.cred.level` - college completers only

## Demographics

Here are the variables we will be exploring;

- `Gender`
- `economic.status`
- `RaceEthnicity`
- `SpecialEdStatus`

**DEMOGRAPHIC VARIABLES AND HIGHEST CREDENTIAL LEVEL (Among credential completers only, n=48,162)**

Among credential completers, 68.2% hold a bachelor's degree or higher and 31.8% hold a sub-baccalaureate credential. Note the sample size here is 48,162 rather than 47,959 — the attending ps category is included here since highest.cred.level captures credential status at time of observation rather than final credential location.

Gender shows a modest but meaningful association with credential level (φ=0.078). Females complete bachelor's or higher credentials at a higher rate (71.4% vs. 64.1% for males), while males are more likely to hold sub-baccalaureate credentials (35.9% vs. 28.6%). This is directly consistent with the institution sector finding — males' greater concentration in public 2-year institutions translates into higher sub-baccalaureate credential attainment. Given the Stage 3 finding that sub-baccalaureate completion predicts regional employment, this gender difference in credential level is part of the mechanism behind males' modest advantage in meaningful regional employment throughout the observation window.

Economic status shows a strong association with credential level (φ=0.174). Non-low-income graduates complete bachelor's or higher credentials at 72.0% compared to 50.8% for low-income graduates — a gap of 21.2 percentage points. Low-income completers hold sub-baccalaureate credentials at nearly twice the rate of non-low-income completers (49.2% vs. 28.0%). This finding reinforces the economic access narrative — among those who complete credentials, economic disadvantage is strongly associated with stopping at the sub-baccalaureate level rather than continuing to a bachelor's degree or higher.

Special education status shows an equally strong association with credential level (φ=0.174). Special education completers hold sub-baccalaureate credentials at 70.0% — more than double the rate for non-special education completers (30.1%). Only 30.0% of special education completers reach bachelor's or higher, compared to 69.9% of non-special education completers. This is one of the starkest demographic contrasts in the entire postsecondary analysis and confirms that when special education graduates do complete credentials, they do so overwhelmingly at the sub-baccalaureate level.

Race and ethnicity shows the strongest demographic association with credential level (V=0.203). The Unknown category again shows an anomalous pattern — 95.4% sub-baccalaureate — which further confirms this group captures a specific non-traditional student population. Among identified groups, Asian/PI graduates show the highest bachelor's or higher completion rate (77.0%), followed by White (70.0%), American Indian (58.0%), Black (59.2%), and Hispanic (49.4%) graduates. Hispanic graduates are the only identified racial or ethnic group where a majority of completers hold sub-baccalaureate credentials (50.6%). The racial and ethnic gradient in credential level is consistent with the economic access constraints and institution sector patterns identified above — minority graduates who complete credentials do so at sub-baccalaureate levels at substantially higher rates than White and Asian/PI graduates.

The demographic findings for credential level tell a consistent and reinforcing story. Economic disadvantage, special education status, male gender, and most minority racial and ethnic identities are all associated with sub-baccalaureate rather than baccalaureate credential attainment among completers. Given the non-linear relationship between credential level and regional employment identified in Stage 3 — where sub-baccalaureate completion predicts the highest regional employment rates — these patterns indicate that demographic disadvantage, while limiting access to four-year credentials, does not prevent credential completers from achieving regional workforce attachment. In fact, it may channel them toward the credential type most strongly associated with it.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Attending ps", "Did not graduate ps"))

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/Demographics/gender.html")

```

### `economic.status`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/Demographics/economic-status.html")
```

### `RaceEthnicity`

```{r}

tab_xtab(var.row = data$highest.cred.level,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

- `Dem_Desc`
- `avg.unemp.rate`

**HIGH SCHOOL CHARACTERISTICS AND HIGHEST CREDENTIAL LEVEL (Among credential completers only, n=48,162)**

EDR shows a weak but statistically significant association with credential level (φ=0.042). EDR 10 graduates complete bachelor's or higher credentials at a modestly higher rate (69.5% vs. 65.1% for EDR 9), while EDR 9 graduates are somewhat more likely to hold sub-baccalaureate credentials (34.9% vs. 30.5%). The difference is small in practical terms but consistent with the institution sector and graduation location patterns seen earlier — EDR 9 graduates who complete credentials are somewhat more concentrated in sub-baccalaureate pathways, which in turn are associated with MN rather than out-of-state employment, again reinforcing the mechanism behind the EDR 9 meaningful MN employment advantage seen in Stage 3.

Average local unemployment rate shows a statistically significant but negligible association with credential level (F=5.9, p=0.015). Sub-baccalaureate completers come from communities with very slightly lower average unemployment than bachelor's or higher completers (difference=-0.032), but the practical magnitude is essentially zero. Local unemployment rate has no meaningful bearing on what level of credential a graduate ultimately earns. This is consistent with the broader pattern across the high school characteristics group — place-based factors shape whether graduates attempt postsecondary and where they complete credentials, but have little influence on credential level conditional on completion.

::: panel-tabset
### `edr`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school characteristics/edr.html")

```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)

```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

- `took.ACT`

- `ACTCompositeScore`

- `ap.exam`

- `cte.achievement`

- `total.cte.courses.taken`

- `MCA.M`

- `MCA.R`

- `MCA.S`

**HIGH SCHOOL ACCOMPLISHMENT VARIABLES AND HIGHEST CREDENTIAL LEVEL (Among credential completers only, n=48,162)**

The accomplishment variables show the strongest and most consistent associations with credential level of any variable group, and the academic achievement versus CTE engagement divide is at its sharpest here.

took.ACT shows a very strong association with credential level (φ=0.332). ACT takers complete bachelor's or higher credentials at 74.5% compared to just 30.4% for non-takers — a gap of 44.1 percentage points. Non-takers hold sub-baccalaureate credentials at 69.6% versus 25.5% for ACT takers. Among all categorical accomplishment variables, ACT participation is the single strongest predictor of whether a completer reaches the baccalaureate level.

ap.exam shows an even stronger association (φ=0.292). AP exam takers complete bachelor's or higher at 87.1% — the highest rate of any categorical predictor examined — compared to 58.4% for non-takers. Only 12.9% of AP exam takers stop at the sub-baccalaureate level. AP participation is the clearest high school signal that a graduate will pursue and complete a four-year credential.

ACT Composite Score shows the largest continuous variable gap (F=6632.2). Bachelor's or higher completers score 3.98 composite points above sub-baccalaureate completers — a very large difference on a 36-point scale. This is the strongest single continuous variable association in the entire postsecondary analysis.

MCA Math (F=4166.8), MCA Reading (F=5113.5), and MCA Science (F=3871.3) all show large and consistent gaps. Sub-baccalaureate completers score approximately 0.56 standard deviations below bachelor's or higher completers on Math, 0.48 below on Reading, and 0.52 below on Science. The consistency across all three assessments reinforces the finding that academic achievement is a powerful and pervasive predictor of credential level among completers.

cte.achievement shows a strong association in the opposite direction (V=0.213). CTE concentrators and completors hold sub-baccalaureate credentials at 43.7% — more than double the rate for No CTE graduates (20.0%) and well above CTE participants (30.7%). The gradient is monotonic and steep. No CTE graduates complete bachelor's or higher at 80.0% compared to just 56.3% for CTE concentrators. The association between CTE concentration and sub-baccalaureate credential completion is the strongest inverse relationship in the accomplishment group and directly mirrors the positive association with regional employment seen throughout the analysis.

Total CTE courses taken shows a large and highly significant association (F=3780.6). Sub-baccalaureate completers take on average 1.84 more CTE courses than bachelor's or higher completers — nearly two full courses more. This is the largest continuous variable mean difference for any accomplishment variable predicting credential level and confirms the dose-response nature of the CTE engagement and sub-baccalaureate credential relationship.

The accomplishment findings for credential level complete the chain of evidence established across the postsecondary analysis. High academic achievement in high school — ACT participation, AP exam participation, high MCA scores — strongly and consistently predicts bachelor's or higher credential attainment. CTE engagement — particularly at the concentrator level and measured as total courses taken — strongly and consistently predicts sub-baccalaureate credential attainment. These two sorting mechanisms operate in opposite directions and are essentially the same mechanisms that sort graduates into out-of-region versus regional employment outcomes a decade later.

::: panel-tabset
### `took.ACT`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/took-ACT.html")
```

### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `ap.exam`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/ap-exam.html")
```

### `cte.achievement`

```{r}
tab_xtab(var.row = data$highest.cred.level,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/cte-achievement.html")

```

### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### `MCA.R`

```{r}
summary <- aov(MCA.R ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### `MCA.S`

```{r}
summary <- aov(MCA.S ~ highest.cred.level, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/highest-cred-level/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# `ps.pathways` - college completers only

## Demographics

Here are the variables we will be exploring;

- `Gender`
- `economic.status`
- `RaceEthnicity`
- `SpecialEdStatus`

**DEMOGRAPHIC VARIABLES AND POSTSECONDARY PATHWAY (Among credential completers only, n=47,959)**

The ps.pathway variable combines credential level and graduation location into six categories: sub-baccalaureate region (19.2%), bachelor or higher region (21.3%), sub-baccalaureate MN (7.6%), bachelor or higher MN (21.5%), sub-baccalaureate outside MN (4.8%), and bachelor or higher outside MN (25.5%). This is the most analytically complete postsecondary outcome variable and the one that most directly maps onto the three workforce pathway framework.

Gender shows a moderate association with ps.pathway (V=0.086). The most notable differences are in sub-baccalaureate regional completion — males complete at 20.9% compared to 17.9% for females — and bachelor or higher regional completion — females at 23.1% compared to 19.1% for males. Females show modestly higher bachelor or higher MN completion (22.4% vs. 20.3%). The out-of-state pathways are essentially equivalent across genders (bachelor or higher outside MN: 26.0% female vs. 24.9% male). The gender difference is primarily in the regional pathways — males are more concentrated in sub-baccalaureate regional completion while females are more concentrated in bachelor or higher regional completion — rather than in geographic mobility.

Economic status shows a strong association with ps.pathway (V=0.191). Low-income graduates are dramatically more concentrated in sub-baccalaureate regional completion (32.0% vs. 16.4%) and sub-baccalaureate MN completion (10.9% vs. 6.9%), while non-low-income graduates dominate bachelor or higher outside MN completion (28.1% vs. 13.8%) and bachelor or higher MN completion (22.6% vs. 16.6%). Bachelor or higher regional completion is essentially equal across economic status groups (21.5% vs. 20.6%). The economic gradient is concentrated in credential level and out-of-state geographic mobility — low-income completers are channeled toward sub-baccalaureate regional and MN pathways, while non-low-income completers access four-year and out-of-state pathways at substantially higher rates.

Special education status shows a strong association with ps.pathway (V=0.175). Special education completers are concentrated in sub-baccalaureate regional completion (42.7% vs. 18.1%) and sub-baccalaureate MN completion (17.6% vs. 7.2%), while showing very low rates of bachelor or higher MN (7.0% vs. 22.2%) and bachelor or higher outside MN (11.9% vs. 26.1%) completion. The pattern is the most extreme of any demographic variable — among special education graduates who complete credentials, over 60% do so through sub-baccalaureate pathways, and the majority of those complete regionally.

Race and ethnicity shows a moderate association (V=0.097) with substantial within-group variation. The Unknown category again shows an anomalous pattern — 57.2% sub-baccalaureate regional and 23.5% sub-baccalaureate MN — consistent with a non-traditional student population concentrated in community and technical colleges. Among identified groups, Hispanic graduates show the highest sub-baccalaureate regional rate (38.5%) followed by Black (29.0%) and American Indian (22.2%) graduates, all substantially above White (17.7%) and Asian/PI (17.1%) graduates. Bachelor or higher outside MN is highest for Asian/PI (27.5%) and White (26.5%) graduates, with substantially lower rates for Black (15.7%), Hispanic (15.9%), and American Indian (16.0%) graduates. American Indian graduates show the highest bachelor or higher MN rate (32.1%), followed by Asian/PI (30.2%), consistent with in-state four-year credential completion being a relatively common pathway for these groups among those who reach the baccalaureate level.

The ps.pathway findings synthesize the patterns seen across the individual postsecondary dependent variables. Economic disadvantage and special education status concentrate graduates in sub-baccalaureate regional pathways — the credential completion route most directly associated with regional workforce attachment in Stage 3. Non-low-income and non-special-education graduates access four-year and out-of-state pathways at substantially higher rates, which the Stage 3 findings show are associated with out-of-region employment and geographic departure from Minnesota. Gender differences are modest and concentrated in the regional pathways rather than geographic mobility. Racial and ethnic minority graduates — particularly Hispanic and Black — show elevated sub-baccalaureate regional completion rates that, combined with the Stage 3 workforce findings, suggest these graduates are more likely to achieve regional workforce attachment through the college completer pathway than their White and Asian/PI peers.

::: panel-tabset
### `Gender`

```{r}
data <- master.original %>%
  filter(!highest.cred.level %in% c("Did not graduate ps", "Attending ps"),
         ps.grad.location != "Did not graduate ps") %>%
  mutate(highest.cred.level = as_factor(highest.cred.level),
         ps.grad.location = as_factor(ps.grad.location),
         ps.pathway = case_when(
    
    # Never attended or attended but did not graduate
    highest.cred.level == "Did not graduate ps" & 
      ps.grad.location == "Did not grad" ~ "Did not graduate ps",
    highest.cred.level == "No college"         ~ "Did not graduate ps",
    
    # Sub-baccalaureate by location
    highest.cred.level == "Less than Bachelors" & 
      ps.grad.location == "PS grad region"      ~ "Sub-baccalaureate - region",
    highest.cred.level == "Less than Bachelors" & 
      ps.grad.location == "PS grad MN"          ~ "Sub-baccalaureate - MN",
    highest.cred.level == "Less than Bachelors" & 
      ps.grad.location == "PS grad outside MN"  ~ "Sub-baccalaureate - outside MN",
    
    # Bachelor or higher by location
    highest.cred.level == "Bachelors or higher" & 
      ps.grad.location == "PS grad region"      ~ "Bachelor or higher - region",
    highest.cred.level == "Bachelors or higher" & 
      ps.grad.location == "PS grad MN"          ~ "Bachelor or higher - MN",
    highest.cred.level == "Bachelors or higher" & 
      ps.grad.location == "PS grad outside MN"  ~ "Bachelor or higher - outside MN",
    
    # Currently attending - treat as its own category
    ps.grad.location == "Attending ps"          ~ "Attending ps",
    highest.cred.level == "Attending ps"        ~ "Attending ps",
    
    TRUE ~ NA_character_),
    across(c("ps.grad":"ps.grad.location"), factor),
    ps.pathway = fct_relevel(ps.pathway, "Sub-baccalaureate - region", "Bachelor or higher - region", "Sub-baccalaureate - MN", "Bachelor or higher - MN", "Sub-baccalaureate - outside MN", "Bachelor or higher - outside MN"))

tab_xtab(var.row = data$ps.pathway,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/Demographics/gender.html")

```

### `economic.status`

```{r}
tab_xtab(var.row = data$ps.pathway,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/Demographics/economic.status.html")
```

### `RaceEthnicity`

```{r}

tab_xtab(var.row = data$ps.pathway,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/Demographics/RaceEthnicity.html")
```

### `SpecialEdStatus`

```{r}
tab_xtab(var.row = data$ps.pathway,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/Demographics/SpecialEdStatus.html")
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

- `Dem_Desc`
- `avg.unemp.rate`

**HIGH SCHOOL CHARACTERISTICS AND POSTSECONDARY PATHWAY (Among credential completers only, n=47,959)**

EDR shows a meaningful association with ps.pathway (V=0.125). The most striking difference is in sub-baccalaureate MN completion — EDR 9 graduates complete at 11.6% compared to just 6.0% for EDR 10 graduates, nearly double the rate. EDR 10 graduates show substantially higher bachelor or higher outside MN completion (27.9% vs. 19.8%), consistent with the geographic mobility pattern identified in earlier analyses. Bachelor or higher regional completion is modestly higher for EDR 9 (24.3% vs. 20.1%), while sub-baccalaureate regional completion is similar across EDRs (19.5% vs. 18.4%). Bachelor or higher MN completion is essentially identical (21.6% vs. 21.3%). The EDR pattern clarifies the mechanism underlying the Stage 3 workforce outcome differences — EDR 9 graduates who complete credentials are more likely to do so at sub-baccalaureate and baccalaureate levels within Minnesota, producing stronger in-state out-of-region employment, while EDR 10 graduates are more likely to complete credentials outside Minnesota and leave the state entirely.

Average local unemployment rate shows a statistically significant association with ps.pathway (F=49.1) with a clear ordering. Sub-baccalaureate regional graduates come from communities with the lowest average unemployment of any pathway group. All other pathways show significantly higher unemployment rates, with sub-baccalaureate MN (difference=0.315), sub-baccalaureate outside MN (difference=0.291), and bachelor or higher region (difference=0.227) all substantially above the regional sub-baccalaureate group. Sub-baccalaureate MN and sub-baccalaureate outside MN are not significantly different from each other (difference=-0.024, p=0.987). Bachelor or higher outside MN (difference=0.095) is significantly above sub-baccalaureate regional but below all other sub-baccalaureate pathways.

This is a counterintuitive finding worth careful interpretation. Graduates who complete sub-baccalaureate credentials regionally come from the strongest local labor markets — not the weakest. This suggests that regional sub-baccalaureate completion is not primarily driven by economic necessity or lack of alternatives, but rather by a genuine regional orientation that coexists with relatively strong local employment conditions. Graduates from higher-unemployment communities are more likely to pursue credentials outside the region — either at in-state institutions in stronger labor markets or out of state entirely — consistent with weaker local conditions pushing credential-seeking graduates toward more geographically distant opportunities.

::: panel-tabset
### `edr`

```{r}
tab_xtab(var.row = data$ps.pathway,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school characteristics/edr.html")

```

### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ ps.pathway, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)

```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

- `took.ACT`

- `ACTCompositeScore`

- `ap.exam`

- `cte.achievement`

- `total.cte.courses.taken`

- `MCA.M`

- `MCA.R`

- `MCA.S`

**HIGH SCHOOL ACCOMPLISHMENT VARIABLES AND POSTSECONDARY PATHWAY (Among credential completers only, n=47,959)**

The accomplishment variables show their strongest and most differentiated associations with ps.pathway, which as the most complete postsecondary outcome variable reveals the full sorting mechanism operating from high school through credential completion.

took.ACT is the strongest categorical predictor of ps.pathway (V=0.333). Non-ACT takers are concentrated in sub-baccalaureate regional (40.3%) and sub-baccalaureate MN (19.3%) completion, with very low bachelor or higher MN (8.4%) and bachelor or higher outside MN (10.8%) rates. ACT takers show the opposite profile — dominated by bachelor or higher outside MN (27.9%), bachelor or higher MN (23.7%), and bachelor or higher regional (23.0%), with low sub-baccalaureate regional (15.7%) and very low sub-baccalaureate MN (5.7%) rates. ACT participation is functioning as a near-perfect sorter between sub-baccalaureate and baccalaureate credential pathways across all geographic locations.

ap.exam shows the second strongest association (V=0.322). AP exam takers are heavily concentrated in bachelor or higher outside MN (36.6%) and bachelor or higher MN (30.6%), with very low sub-baccalaureate regional (8.5%) and sub-baccalaureate MN (2.5%) rates. Non-AP takers show the reverse — sub-baccalaureate regional (24.8%) is their largest single pathway category, and bachelor or higher outside MN (19.8%) is their second largest. AP participation is the strongest predictor of the bachelor or higher MN pathway specifically (30.6% vs. 16.8%) — a finding with direct workforce implications given that bachelor or higher MN graduates show the highest meaningful MN employment rates at Years 5 and 10 in Stage 3.

ACT Composite Score shows very large mean differences across pathways (F=1777.0). The ordering is clear and consistent: bachelor or higher outside MN and bachelor or higher MN score highest and are not significantly different from each other (difference=0.038, p=0.986). Bachelor or higher regional scores approximately 2.32 points below both. Sub-baccalaureate regional scores the lowest among single-state pathways, and sub-baccalaureate MN and sub-baccalaureate outside MN score approximately 0.45 and 0.26 points below sub-baccalaureate regional respectively — though sub-baccalaureate outside MN is not significantly different from sub-baccalaureate regional (p=0.240). The convergence of bachelor or higher MN and bachelor or higher outside MN at nearly identical ACT scores is analytically important — geographic destination after completing a four-year credential is not primarily driven by academic achievement but by other factors.

MCA Math (F=968.4), MCA Reading (F=1087.8), and MCA Science (F=937.3) all show the same two-cluster structure at the pathway level. The bachelor or higher MN and bachelor or higher outside MN pathways are not significantly different from each other on any of the three assessments (Math p=0.439, Reading p=0.992, Science p=1.000). Sub-baccalaureate regional and sub-baccalaureate outside MN are not significantly different on Math (p=0.572) or Science (p=0.800), and only marginally different on Reading (p=0.791). This two-cluster structure — sub-baccalaureate pathways at lower scores, bachelor or higher pathways at higher scores — is the clearest expression of credential level as the primary sorting mechanism, with geographic location playing a secondary role within each credential level cluster.

cte.achievement shows a strong association with ps.pathway (V=0.162). CTE concentrators are most concentrated in sub-baccalaureate regional (25.4%) completion — more than double the rate for No CTE graduates (12.3%) — and show the lowest bachelor or higher outside MN rate (20.1%) of any CTE category. No CTE graduates dominate bachelor or higher outside MN (31.9%) and bachelor or higher MN (27.5%). Bachelor or higher regional completion is essentially flat across CTE categories (20.4%, 23.2%, 20.6%), confirming that CTE engagement does not meaningfully influence whether a bachelor's graduate stays in the region — it operates primarily through the credential level and sub-baccalaureate regional sorting mechanism. The sub-baccalaureate outside MN and sub-baccalaureate MN gaps between concentrators and No CTE graduates (6.8% vs. 3.1% and 11.3% vs. 4.6%) suggest that even among sub-baccalaureate completers, CTE engagement is associated with more geographically proximate credential completion.

Total CTE courses taken (F=834.9) confirms the dose-response pattern throughout. Sub-baccalaureate regional graduates take the most CTE courses, significantly above all other pathways. Sub-baccalaureate MN and sub-baccalaureate outside MN graduates also take substantially more courses than bachelor or higher graduates of any geographic type. Bachelor or higher MN and bachelor or higher outside MN graduates are not significantly different from each other (difference=0.161, p=0.001 — marginally significant but very small in practical terms), again confirming that geographic destination within the four-year pathway is not meaningfully driven by CTE engagement.

The ps.pathway accomplishment findings are the most analytically complete picture of how high school experiences sort graduates into postsecondary pathways. The primary sorting mechanism is credential level — academic achievement predicts four-year completion while CTE engagement predicts sub-baccalaureate completion — and this sorting operates across all geographic destinations. The secondary sorting mechanism within credential level is geographic — for sub-baccalaureate completers, CTE engagement predicts regional completion; for four-year completers, academic achievement above the already-high threshold does not further differentiate geographic destination. These two mechanisms together produce the workforce pathway structure identified throughout the analysis.

::: panel-tabset
### `took.ACT`

```{r}
tab_xtab(var.row = data$ps.pathway,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/took-ACT.html")
```

### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ ps.pathway, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

### `ap.exam`

```{r}
tab_xtab(var.row = data$ps.pathway,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/ap-exam.html")
```

### `cte.achievement`

```{r}
tab_xtab(var.row = data$ps.pathway,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/cte-achievement.html")

```

### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ ps.pathway, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

### `MCA.M`

```{r}
summary <- aov(MCA.M ~ ps.pathway, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

### `MCA.R`

```{r}
summary <- aov(MCA.R ~ ps.pathway, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)

```

### `MCA.S`

```{r}
summary <- aov(MCA.S ~ ps.pathway, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_post_secondary_pathways_ct/ps-pathway/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```
:::

<br>

# POSTSECONDARY PATHWAY ANALYSIS — SUMMARY AND NARRATIVE Southeast Minnesota (EDR 9 and EDR 10)

**OVERVIEW**

The postsecondary analysis examined relationships between high school experiences and four postsecondary outcomes — credential completion (ps.grad), institution sector (ps.grad.InstitutionSector), graduation location (ps.grad.location), credential level (highest.cred.level), and the combined pathway variable (ps.pathway) — among 104,610 Southeast Minnesota high school graduates. The analysis was conducted in two stages: relationships with ps.grad were examined for the full population, and relationships with all other outcomes were examined among the 47,959 credential completers only. Across all dependent variables and all independent variable groups, two sorting mechanisms emerge with remarkable consistency. Academic achievement and orientation predict four-year credential completion and geographic departure from the region. CTE engagement predicts sub-baccalaureate credential completion and geographic proximity in both credential attainment and subsequent workforce attachment. These two mechanisms operate independently, in opposite directions, and are visible at every stage of the postsecondary pathway.

**WHO COMPLETES A POSTSECONDARY CREDENTIAL**

Forty-six percent of Southeast Minnesota graduates complete a postsecondary credential by 2023. The barriers to completion are concentrated among specific demographic groups and are strongly associated with high school academic orientation.

Economic disadvantage is the most powerful demographic barrier — low-income graduates complete credentials at 26.2% compared to 55.1% for non-low-income graduates, a gap of nearly 29 percentage points. Special education status is equally consequential — only 14.4% of special education graduates complete credentials. These two variables represent the strongest demographic associations with credential completion in the analysis. Gender shows a meaningful gap favoring females (52.9% vs. 39.5% for males), and substantial racial and ethnic disparities exist — White (53.1%) and Asian/PI (51.8%) graduates complete at rates roughly double those of Black (28.9%), Hispanic (26.0%), and American Indian (14.6%) graduates. Higher local unemployment rates are modestly associated with higher completion rates, suggesting that weak local labor markets increase the relative attractiveness of postsecondary investment as an alternative to limited immediate employment opportunities.

The high school accomplishment variables tell the most important part of the credential completion story. ACT participation is the single strongest predictor — non-ACT takers complete at just 19.4% compared to 59.7% for ACT takers. AP exam participation is nearly as powerful (74.9% vs. 38.4%). All three MCA assessments show large score gaps favoring completers. By contrast, CTE engagement is negatively associated with completion — CTE concentrators complete at 39.5% compared to 54.3% for No CTE graduates, and each additional CTE course taken is associated with lower completion likelihood. The academic orientation versus CTE engagement divide that defines the workforce pathway structure is already fully visible at the credential completion stage.

**WHAT KIND OF CREDENTIAL AND FROM WHERE**

Among the 47,959 graduates who complete credentials, the analysis reveals a second stage of sorting that is equally consequential for long-run workforce outcomes.

Credential level splits 68.2% bachelor or higher versus 31.8% sub-baccalaureate. The same demographic and accomplishment variables that predict credential completion predict credential level within completers. Low-income completers hold sub-baccalaureate credentials at 49.2% — nearly double the rate for non-low-income completers (28.0%). Special education completers hold sub-baccalaureate credentials at 70.0%. Males complete sub-baccalaureate credentials at 35.9% compared to 28.6% for females. Hispanic (50.6%) and Black (40.8%) completers are more likely to hold sub-baccalaureate credentials than White (30.0%) and Asian/PI (23.0%) completers. ACT participation, AP exam participation, and high MCA scores all strongly predict bachelor or higher completion, while CTE concentration and higher total CTE course-taking strongly predict sub-baccalaureate completion. The dose-response relationship between total CTE courses taken and sub-baccalaureate completion is among the largest continuous variable associations in the entire analysis — sub-baccalaureate completers take on average 1.84 more CTE courses than bachelor or higher completers.

Graduation location sorts completers into three geographic destinations — region (40.5%), outside Minnesota (30.3%), and elsewhere in Minnesota (29.1%). Gender is the striking non-finding here — it has essentially no association with graduation location (V=0.011, p=0.065). All other demographic variables show meaningful associations: low-income graduates complete regionally at 52.6% compared to 37.9% for non-low-income graduates, and special education completers complete regionally at 54.0% compared to 39.9%. These are access-constraint effects — economic and institutional barriers concentrate disadvantaged graduates at nearby regional institutions regardless of their credential level preferences. Among accomplishment variables, CTE engagement is the strongest predictor of regional graduation, while AP exam participation and high ACT scores predict out-of-state graduation. The geographic sorting that produces the workforce outcome differences seen in Stage 3 begins here, at the point of credential completion.

EDR is the most interesting high school characteristics finding in this stage. EDR 10 graduates complete outside Minnesota at 32.7% compared to just 24.4% for EDR 9 graduates, while EDR 9 graduates complete sub-baccalaureate credentials within Minnesota at nearly double the EDR 10 rate (11.6% vs. 6.0%). This EDR difference in credential completion geography is the mechanism behind the workforce outcome divergence identified in Stage 3 — EDR 9 completers earn credentials in Minnesota and stay in Minnesota for employment, while EDR 10 completers are more likely to earn credentials out of state and leave entirely.

**THE COMBINED PATHWAY PICTURE**

The ps.pathway variable — crossing credential level with graduation location — produces the clearest and most analytically complete picture of how high school experiences sort graduates into long-run workforce trajectories. The six pathway categories sort graduates as follows: bachelor or higher outside MN (25.5%), bachelor or higher regional (21.3%), bachelor or higher MN (21.5%), sub-baccalaureate regional (19.2%), sub-baccalaureate MN (7.6%), and sub-baccalaureate outside MN (4.8%).

The academic achievement variables show a two-cluster structure across ps.pathway that is consistent across all measures. Bachelor or higher MN and bachelor or higher outside MN graduates score at virtually identical levels on ACT composite, MCA Math, MCA Reading, and MCA Science — and are not significantly different from each other on any assessment. Sub-baccalaureate regional and sub-baccalaureate outside MN graduates similarly converge at lower score levels. This two-cluster structure means that academic achievement predicts credential level — the primary sorting decision — but does not meaningfully predict geographic destination within each credential level cluster. Whether a four-year graduate stays in Minnesota or leaves is not primarily a function of how academically accomplished they were in high school.

CTE engagement shows the complementary pattern. Sub-baccalaureate regional graduates take the most CTE courses of any pathway group, and the sub-baccalaureate pathways as a whole take substantially more courses than the bachelor or higher pathways. Within the sub-baccalaureate cluster, CTE engagement predicts regional over out-of-state completion. Within the bachelor or higher cluster, CTE engagement shows only marginal differentiation between geographic destinations. CTE engagement is primarily a sub-baccalaureate and regional sorting mechanism — it channels graduates toward the credential type and geographic proximity most associated with regional workforce attachment.

**THE STORY THE DATA TELLS**

Three distinct postsecondary stories emerge from Southeast Minnesota's high school graduating cohorts.

The first story is the college completer pathway in its two expressions. For graduates who complete four-year credentials, the critical variable is not academic achievement — which is high across all four-year completers — but geography. Those who complete regional four-year credentials are substantially more likely to achieve regional employment. Those who complete outside Minnesota — a group characterized by high AP exam participation and who come from relatively stronger local labor markets — are the most likely to establish careers entirely outside Minnesota's formal employment records. The bachelor or higher MN pathway represents an intermediate group who complete credentials in Minnesota but outside the region and subsequently show the strongest meaningful MN employment rates in Stage 3. For sub-baccalaureate completers, the story is cleaner — regional completion strongly predicts regional employment, and the strongest predictor of regional sub-baccalaureate completion is CTE engagement in high school.

The second story is the workforce-first pathway. Graduates who do not complete credentials — more than half the population — enter the labor market directly from high school. Among these graduates, CTE engagement is the dominant predictor of regional workforce success, as established in Stage 3. The postsecondary analysis adds an important dimension to this story: the decision not to complete a credential is not random. It is concentrated among graduates with high CTE engagement, lower academic achievement, lower economic status, special education status, and certain racial and ethnic minority backgrounds. These graduates were systematically selected away from postsecondary completion by both their own orientations — CTE engagement reflects a deliberate choice toward workforce preparation — and by structural barriers including economic constraints and institutional access limitations.

The third story is the disconnected pathway. Graduates who neither complete credentials nor achieve sustained employment are concentrated among those with low CTE engagement, low academic achievement, and high levels of economic disadvantage and special education status — the group that falls between both pathways rather than choosing one. The postsecondary analysis clarifies that this group is not simply defined by postsecondary non-completion. Many non-completers achieve sustained regional employment through the workforce-first pathway. The disconnected group is defined by the absence of both postsecondary engagement and CTE investment — graduates who neither developed vocational skills for regional employment nor academic skills for postsecondary success.

**IMPLICATIONS FOR THE MULTIVARIATE STAGES**

The postsecondary analysis provides strong and specific expectations for the multivariate confirmation stages. The ps.pathway combined variable should show the strongest and most differentiated associations with sustained regional employment at Year 5 — with sub-baccalaureate regional completion showing the highest odds among completers, followed by bachelor or higher regional, and with outside MN pathways showing the lowest odds. The CTE engagement variables should retain strong independent associations with regional employment among non-college completers even after controlling for demographic characteristics and high school characteristics. Economic disadvantage and special education status should show attenuated but persistent associations reflecting both direct barriers and their channeling effects through sub-baccalaureate credential pathways. The EDR effect should be modest but present, reflecting the geographic sorting of EDR 9 graduates toward in-state pathways and EDR 10 graduates toward out-of-state departure.
