---
title: "Multivariate Confirmation Analysis - Logistic Regression"
editor: visual
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(broom)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/Master.csv")

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states) 

master <- master.original %>%
  left_join(master.states, by = "PersonID") %>%
  mutate(grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region"), "Sustained WF - SE",
                              ifelse(grad.year.5 == "Meaningful emp MN", "Sustained WF - MN",
                                     ifelse(grad.year.5 == "No MN emp record, not attending ps", "No MN emp record", 
                                            ifelse(grad.year.5 == "Not meaningful, not attending ps", "Non-sustained WF", as.character(grad.year.5))))),
         grad.year.5 = as_factor(grad.year.5))

kable(head(master))

kable(names(master))
```

# Multivariate Confirmation Analysis: Logistic Regression

The purpose of this project is to explain which characteristics are associated with entry into distinct workforce participation pathways following high school graduation in Southeast Minnesota (EDR 9 and 10). These pathways include sustained employment in the region, sustained employment elsewhere in Minnesota, non-sustained employment, and having no Minnesota employment record.

The policy question of greatest interest is what factors are associated with **sustained employment in the region**. Earlier analysis relied on bivariate analyses (cross-tabulations, Cramér’s V, and ANOVA) to identify patterns of stratification across workforce participation outcomes. A consistent finding was that postsecondary experiences—particularly credential attainment and where education was completed—are strongly associated with long-term workforce attachment. Academic preparation, early college exposure (PSEO), career and technical education (CTE), and structural demographic factors also demonstrated meaningful relationships.

To strengthen confidence in these findings, a limited multivariate analysis is conducted using logistic regression. The purpose of this modeling is not to establish causality or build a predictive tool, but to examine whether the strongest bivariate relationships remain directionally consistent when key variables are considered simultaneously. This approach helps clarify whether preparation variables retain independent associations after accounting for credential attainment, and whether regional postsecondary completion remains influential when controlling for background and readiness.

Logistic regression is used here as a confirmatory tool rather than the primary analytic framework.

<br>

## Methodology

### **Dependent Variables**

This analysis models five distinct outcomes, each restricted to the appropriate subset of the cohort:

1.  **Sustained regional workforce attachment** (`grad.year.5`) — binary (1 = meaningful employment within the region five years after high school graduation, 0 = otherwise), modeled across the full graduating cohort.
2.  **Sustained regional workforce attachment (`grad.year.5`**, college completers only) - binary (1 = meaningful employment within the region five years after high school graduation, 0 = otherwise), modeled across college completers only.
3.  **Postsecondary completion** (`ps.grad`) — binary (1 = completed a postsecondary credential, 0 = did not), modeled across **all high school graduates**, regardless of whether they ever attended postsecondary education.
4.  **Postsecondary completion within the region** — binary (1 = completed a postsecondary credential within the region, 0 = completed elsewhere in Minnesota or out of state), restricted to **postsecondary completers only**.
5.  **Sub-baccalaureate credential attainment** — binary (1 = highest credential earned is an associate degree or less-than-associate credential, 0 = bachelor's degree or higher), restricted to **postsecondary completers only**.

### **Sample**

All five models are restricted to the **2011–2019 graduating cohorts for two reasons**.

1.  Earlier cohorts (2008–2010) are excluded because PSEO participation was not consistently tracked in the data system prior to 2011 — missingness in that variable is concentrated almost entirely in those years rather than distributed randomly, and including them would risk misclassifying genuine PSEO participants as non-participants. And PSEO participation was identified as having a very strong relationship to our dependent variables in the bivariate analysis.
2.  RaceEthnicity has significant amounts of "Unknown" beginning in 2020.

### Independent Variables

Variables were selected based on (1) a demonstrated bivariate relationship with the relevant outcome(s) in the cross-tabulation and ANOVA analysis, and (2) practical usability in a multivariate model, with closely related variables consolidated to avoid multicollinearity. The same variable set and coding is used across all four outcomes.

**Demographics**

- **`Gender`**
- **`RaceEthnicity`**
- **`economic.status`**
- **`SpecialEdStatus`**
- **`english.learner`**

**High school characteristics**

- `Dem_Desc`
- `edr`
- **`avg.wages.pct.state`** (local average wages as a percentage of the state average)
- **`avg.unemp.rate`** (local average unemployment rate)

**High school accomplishments**

- **`pseo.participant`** (Post-Secondary Enrollment Options participation)
- **`took.ACT`** (whether the student took the ACT)
- **`ACTCompositeScore.filled`** (ACT composite score, recoded to 0 for non-test-takers — see note below)
- **`ap.exam`** (AP exam participation)
- `sat.taken`
- **`cte.achievement`** (CTE concentration/completion level)
- `total.cte.courses.taken`

**Postsecondary variables**

- `ps.grad` - used only for the DV1 WF - Region as the dependent variable with full population
- **`highest.cred.level`** - these will be tested on the dependent variables where college completers are the only individuals being analysed.
- **`ps.grad.location`** - these will be tested on the dependent variables where college completers are the only individuals being analysed.

**Notes on Variable Selection**

**Academic achievement (MCA).** MCA Math, Reading, and Science scores are highly correlated with one another (r = 0.63–0.75) and with ACT performance, and all three are excluded from this analysis. MCA Math's missingness rate differs sharply by English learner status (26.1% vs. 2.4% for other students), and this pattern cannot be reliably classified as an independent confounder versus a mediator of unequal access, resourcing, or administrative practice across student populations. Controlling for it — whether through the raw score or a missingness indicator — risks statistically absorbing a real disparity under the guise of a data-quality correction. No MCA subject is retained as a substitute for another; the concern applies to the construct as a whole, not to Math specifically.

**ACT participation and score.** Students who did not take the ACT have no composite score by definition — this is a structural absence tied entirely to an observed variable (`took.ACT`), not missing data in the conventional sense. To retain the full sample rather than dropping non-test-takers, `ACTCompositeScore` is recoded to 0 for non-test-takers and entered into the model alongside `took.ACT`. With both terms present, `took.ACT` captures the effect of test participation itself, while `ACTCompositeScore` captures the achievement gradient specifically among test-takers, without the two effects being confounded. `sat.taken` is entered separately, capturing a distinct, smaller subgroup of test-takers not fully overlapping with ACT participation. Because `took.ACT` shows a participation gap by English learner status (41.4% vs. 11.2%) of a similar scale to the MCA missingness gap that led to MCA's exclusion, any attenuation of the English learner coefficient once `took.ACT` enters the model should be interpreted with the same caution — as a variable that may partly reflect unequal access rather than a fully independent behavior — rather than treated as the disparity being explained away.

**Variables dropped.** `MCA.M`, `MCA.R`, and `MCA.S` are excluded all three MCA subjects for the reasons described above.

### Modeling Approach

Each outcome is modeled using a sequential block-entry approach, adding one group of predictors at a time to assess the relative contribution of each. Each block is compared to the prior one using a likelihood ratio test to confirm it adds significant explanatory power, with AIC and pseudo-R² tracked across the sequence to summarize overall model improvement.

**Sustained regional workforce attachment:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments
- Model 4: + College completion

**Sustained regional workforce attachment - college completers only**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments
- Model 4: + highest credential earned
- Model 5: + post-secondary location
- Model 6: Dependent on results

**Postsecondary completion, all high school graduates:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

**Postsecondary completion within the region, completers only:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

**Sub-baccalaureate credential attainment, completers only:**

- Model 1: Demographics
- Model 2: + High school characteristics
- Model 3: + High school accomplishments

Following these base model sequences, additional modeling will be considered only if the results indicate a need for further specification (e.g., interaction effects, additional mediation checks).

<br>

# Analysis Approach

This multivariate analysis is structured as a series of theoretically guided logistic regression models designed to evaluate mechanisms associated with sustained regional employment. The goal is confirmatory testing of previously observed patterns rather than prediction or causal inference.

<br>

## Evaluation Criteria

Results will be reported using: - Odds ratios - 95% confidence intervals - Statistical significance levels - Directional interpretation

Model comparison metrics (AIC and pseudo R²) will be used to assess relative improvement across specifications within each outcome's model sequence, but emphasis will remain on coefficient stability, attenuation patterns, and substantive interpretation rather than overall fit. <br>

## Interpretation Framework

Interpretation will focus on:

- Direction and relative magnitude of associations
- Changes in coefficients across blocks within each model sequence
- Evidence of attenuation as additional blocks are introduced.

Findings will be described as associations holding other included variables constant. No causal claims will be made. Together, these models provide multivariate confirmation of patterns identified in earlier descriptive analyses while preserving conceptual clarity and interpretability.

# Prep DV1 data

The first thing I need to do is setup my dataset for modeling. I'm going to create the workforce participation into a binary and will use grad.year.5 as my modeling year.

```{r}

model_vars_final <- c("Gender", 
                      "RaceEthnicity", 
                      "economic.status", 
                      "SpecialEdStatus",
                      "english.learner",
                      "Dem_Desc",
                      "edr",
                      "avg.wages.pct.state", 
                      "avg.unemp.rate",
                      "pseo.participant", 
                      "took.ACT", 
                      "ACTCompositeScore", 
                      "sat.taken", 
                      "ap.exam",
                      "cte.achievement",
                      "total.cte.courses.taken",
                      "ps.grad",
                      "grad.year.5")

master.model <- master %>%
  filter(grad.year > 2010,
         grad.year < 2020)%>% 
  select(PersonID, all_of(model_vars_final)) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         ps.grad != "Attending ps",
         !is.na(pseo.participant),
         RaceEthnicity != "Unnknown") %>%
  droplevels() %>%
  mutate(grad.year.5 = ifelse(grad.year.5 == "Sustained WF - SE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         edr = fct_relevel(edr, "EDR 9 - South Central"),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         ps.grad = ifelse(ps.grad == "Y", "Yes", ps.grad),
         ps.grad = fct_relevel(ps.grad, "No")) 


kable(head(master.model))

kable(names(master.model))

```

We now have `r comma(nrow(master.model))` individuals to model with `r ncol(master.model) - 2` variables not including `grad.year.5`.

<br>

# Modeling: Sustained WF - Region (full population)

## Model 1: Demographics

**Reference categories:** RaceEthnicity = White; Gender = Male; economic.status = 0 (not economically disadvantaged); SpecialEdStatus = 0; english.learner = 0

**Sample:** n = 47,693

**Coefficient-level interpretation**

- **GenderF** (OR = 0.937, 95% CI: 0.902–0.974, p \< 0.001): Female graduates have 6.3% lower odds of sustained regional attachment at Year 5 compared to male graduates, holding other demographics constant.

- **RaceEthnicityAI** (OR = 0.452, 95% CI: 0.312–0.640, p \< 0.001): American Indian graduates have 54.8% lower odds of sustained regional attachment compared to White graduates.

- **RaceEthnicityAsian/PI** (OR = 0.610, 95% CI: 0.536–0.692, p \< 0.001): Asian/Pacific Islander graduates have 39.0% lower odds of sustained regional attachment compared to White graduates.

- **RaceEthnicityBlack** (OR = 0.484, 95% CI: 0.430–0.544, p \< 0.001): Black graduates have 51.6% lower odds of sustained regional attachment compared to White graduates.

- **RaceEthnicityHispanic** (OR = 0.871, 95% CI: 0.793–0.955, p = 0.003): Hispanic graduates have 12.9% lower odds of sustained regional attachment compared to White graduates.

- **economic.status1** (OR = 1.056, 95% CI: 1.009–1.104, p = 0.018): Economically disadvantaged graduates have 5.6% higher odds of sustained regional attachment compared to non-disadvantaged graduates.

- **SpecialEdStatus1** (OR = 0.829, 95% CI: 0.779–0.882, p \< 0.001): Graduates with special education status have 17.1% lower odds of sustained regional attachment compared to those without.

- **english.learner1** (OR = 0.933, 95% CI: 0.857–1.016, p = 0.109): Not statistically significant. English learner status shows no reliable association with sustained regional attachment once other demographics are accounted for.

**Model fit**

- **Pseudo-R²:** 0.0054 — demographics alone explain very little variance, as expected for a Model 1 block.
- **AIC:** 60,344
- **VIF:** All values well below 2 (max 1.46 for english.learner) — no multicollinearity concerns at this stage.

**Flags / notes**

- All RaceEthnicity categories show *lower* odds than White graduates, with AI and Black graduates showing the largest gaps (\~50%+ lower odds). Worth watching whether these attenuate or persist once high school characteristics and accomplishments are added in later blocks — that's the central test of whether these are independent demographic effects or proxies for other structural factors (e.g., school context, CTE access).
- The economic.status direction (disadvantaged → *higher* odds of regional attachment) is counterintuitive relative to typical assumptions and worth flagging explicitly per your data-quality/counterintuitive-finding preference — this may reflect regional attachment being driven by *staying* rather than *upward mobility* (i.e., fewer resources/opportunities to leave), but that's speculative and should be tested against later blocks rather than asserted.
- english.learner is non-significant here — consistent with the note in the methodology that EL effects may be entangled with participation/access variables (e.g., took.ACT) introduced in later blocks, so its coefficient bears re-watching rather than being written off as null.

```{r}
model1_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model
)

model1_dv1

summary(model1_dv1)
nobs(model1_dv1)
AIC(model1_dv1)

# odds ratios + 95% CIs, per your evaluation criteria
exp(cbind(OR = coef(model1_dv1), confint(model1_dv1)))

1 - (model1_dv1$deviance / model1_dv1$null.deviance)

vif(model1_dv1)


or_plot <- tidy(model1_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2: Demographics + High School Characteristics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown — likely "Town/rural mix" given the two dummies present); edr = EDR 9 (reference, since only "EDR 10 - Southeast" appears as a dummy)

Sample: n = 47,693

**Model comparison (vs. Model 1)**

- Likelihood ratio test: χ²(5) = 71.18, p \< 0.001 — Model 2 is a statistically significant improvement over Model 1.
- AIC improved from 60,344 → 60,283.
- Pseudo-R² increased from 0.0054 → 0.0066 — still a small overall contribution, consistent with a Model 2 block.

**Demographic coefficients (carried from Model 1)**

- Direction, magnitude, and significance of Gender, RaceEthnicity, economic.status, and SpecialEdStatus are essentially unchanged from Model 1 — no meaningful attenuation.
- **english.learner1** shifted from non-significant at p = 0.109 in Model 1 to marginal at p = 0.079 here (OR = 0.927, 95% CI: 0.851–1.016) — still not significant at conventional thresholds, but worth continuing to track as high school accomplishments enter in Model 3.

**New: High school characteristics**

- **Dem_DescEntirely urban** (OR = 1.142, 95% CI: 1.074–1.214, p \< 0.001): Graduates from entirely urban high schools have 14.2% higher odds of sustained regional attachment relative to the reference Dem_Desc category.
- **Dem_DescUrban/town/rural mix** (OR = 0.928, 95% CI: 0.880–0.978, p = 0.005): 7.2% lower odds relative to the reference category.
- **edrEDR 10 - Southeast** (OR = 1.056, 95% CI: 1.010–1.105, p = 0.017): Graduates from EDR 10 have modestly higher odds (5.6%) of sustained regional attachment than EDR 9 graduates. This holds as an independent geographic effect alongside Dem_Desc, which is capturing urbanicity rather than region per se.
- **avg.wages.pct.state** (OR = 0.755, 95% CI: 0.564–1.013, p = 0.060): Marginal, not significant at 0.05. Direction suggests higher local wages relative to the state average trend toward lower odds of sustained regional attachment — counterintuitive and not reliable at this threshold; worth watching in later blocks rather than drawing a conclusion now.
- **avg.unemp.rate** (OR = 1.010, 95% CI: 0.995–1.025, p = 0.204): Not significant.

**Model fit**

- VIF: All values remain low (max 1.48, for Dem_Desc and english.learner) — no multicollinearity concerns. Dem_Desc and edr are each capturing distinct dimensions (urbanicity vs. region) and both hold up as independent, non-inflated predictors.

**Flags / notes**

- `avg.wages.pct.state` and `avg.unemp.rate` are both weak/non-significant here — flagging in case you want to revisit their inclusion in later blocks, though I'll leave that decision to you.
- No collinearity concerns between Dem_Desc and edr in this run — both are retained as independent predictors going forward.

```{r}
model2_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model
)

summary(model2_dv1)

nobs(model2_dv1)


AIC(model2_dv1)

exp(cbind(OR = coef(model2_dv1), confint(model2_dv1)))

anova(model1_dv1, model2_dv1, test = "Chisq")

1 - (model2_dv1$deviance / model2_dv1$null.deviance)

vif(model2_dv1)

or_plot <- tidy(model2_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9; pseo.participant = 0; took.ACT = No; sat.taken = 0; ap.exam = 0; cte.achievement = reference level (not shown)

Sample: n = 47,693

**Model comparison (vs. Model 2)**

- Likelihood ratio test: χ²(8) = 1,725.10, p \< 0.001 — Model 3 is a large, statistically significant improvement over Model 2.
- AIC improved substantially: 60,283 → 58,574.
- Pseudo-R² increased from 0.0066 → 0.0351 — still modest in absolute terms, but a meaningful jump, consistent with HS accomplishments carrying more explanatory weight than demographics or HS characteristics alone.

**Notable shifts from Model 2 (attenuation tracking)**

- **GenderF**: flipped from significant/negative (OR = 0.937, p \< 0.001 in Model 2) to non-significant/near-null (OR = 1.015, 95% CI: 0.974–1.057, p = 0.480) here. This is a full attenuation to null, not just a weakening — worth noting as a substantive finding, suggesting the Model 2 gender effect may have been proxying for accomplishment-related variables (e.g., ACT-taking behavior, CTE participation) rather than reflecting an independent gender effect.
- **economic.status1**: flipped direction from positive (OR = 1.063, p = 0.007 in Model 2) to negative (OR = 0.936, 95% CI: 0.893–0.980, p = 0.005) here — a sign flip, not just attenuation. Economically disadvantaged graduates now show lower odds of sustained regional attachment once HS accomplishments are controlled. This reverses the counterintuitive Model 1/2 direction and moves it toward the more conventionally expected direction — flagging since sign flips across blocks warrant explicit note, and this changes the substantive story for economic.status.
- **english.learner1**: strengthened from marginal (p = 0.079 in Model 2) to significant (OR = 0.840, 95% CI: 0.770–0.915, p \< 0.001) here. Consistent with your methodology's caution that EL effects may be entangled with participation variables like `took.ACT` — the coefficient became more pronounced once ACT participation entered, which is the opposite of a "gets explained away" pattern and worth keeping in mind when interpreting `took.ACT` below.
- **Dem_Desc / edr**: both carried forward from Model 2 with no change in interpretation needed — no collinearity issue between them, consistent with the correction to Model 2.

**New: High school accomplishments**

- **pseo.participant1** (OR = 1.093, 95% CI: 1.042–1.147, p \< 0.001): PSEO participants have modestly higher odds of sustained regional attachment.
- **took.ACTYes** (OR = 3.627, 95% CI: 3.182–4.134, p \< 0.001): Large effect — ACT test-takers have substantially higher odds of sustained regional attachment than non-takers, independent of score.
- **ACTCompositeScore.filled** (OR = 0.939 per point, 95% CI: 0.933–0.945, p \< 0.001): Among the achievement gradient (with non-takers coded 0), higher ACT composite scores are associated with lower odds of sustained regional attachment. Interpreted together with `took.ACT`, the pattern suggests test participation itself is strongly positive, while higher scores among test-takers trend toward higher academic orientation and away from regional attachment — consistent with the "two forces" framing from the College Completers pathway.
- **sat.taken1** (OR = 0.528, 95% CI: 0.439–0.631, p \< 0.001): SAT takers have notably lower odds of sustained regional attachment. Directionally consistent with the ACT score gradient, but the magnitude is large — worth a sanity check on sample size/composition of SAT-takers in this region, since SAT-taking is typically rare in Minnesota relative to ACT.
- **ap.exam1** (OR = 0.847, 95% CI: 0.800–0.898, p \< 0.001): AP exam participants have lower odds of sustained regional attachment — directionally consistent with the academic-orientation pattern above.
- **cte.achievementCTE Concentrator or Completor** (OR = 1.456, 95% CI: 1.364–1.554, p \< 0.001) and **cte.achievementCTE Participant** (OR = 1.287, 95% CI: 1.220–1.358, p \< 0.001): Both CTE categories show meaningfully higher odds of sustained regional attachment relative to the reference category, with a clear dose-response pattern (Concentrator/Completor \> Participant).
- **total.cte.courses.taken** (OR = 1.031 per course, 95% CI: 1.023–1.039, p \< 0.001): Additional CTE courses associated with modestly higher odds, independent of `cte.achievement` level.

**Model fit / diagnostics**

- VIF — flagging for your attention: `took.ACT` (GVIF = 9.78) and `ACTCompositeScore.filled` (GVIF = 11.52) are both elevated, with `ACTCompositeScore.filled` exceeding the conventional VIF = 10 threshold. This is expected given the structural relationship between the two terms (composite score is 0 by construction for everyone with took.ACT = No), but it means their individual standard errors are inflated and the two coefficients should be interpreted jointly rather than independently.
- `cte.achievement` (GVIF 1.84) and `total.cte.courses.taken` (GVIF 1.94) are both elevated relative to other variables in the model but still well under any conventional concern threshold — no action needed, just noting the two CTE variables are correlated with each other as expected.
- All other VIFs remain low (\< 1.5).

**Flags / notes requiring your input**

1.  `economic.status` sign flip (Model 2 → Model 3): now significant and negative. Confirm this is the expected/intended direction before it's carried into interpretation for the summit deck.
2.  GenderF full attenuation to null: no action needed, just flagging as a substantive pattern worth naming in write-up.
3.  `sat.taken` effect size (OR = 0.528): worth a quick check on subgroup n before leaning on this in the summit narrative, given how small the SAT-taking population likely is in this region.

```{r}
model3_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    sat.taken +
    ap.exam + 
    cte.achievement +
    total.cte.courses.taken,
  family = binomial,
  data = master.model
)

nobs(model3_dv1)


AIC(model3_dv1)

summary(model3_dv1)

exp(cbind(OR = coef(model3_dv1), confint(model3_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, test = "Chisq")

1 - (model3_dv1$deviance / model3_dv1$null.deviance)

vif(model3_dv1)

or_plot <- tidy(model3_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4: Demographics + High School Characteristics + High School Accomplishments + Postsecondary Completion

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9; pseo.participant = 0; took.ACT = No; sat.taken = 0; ap.exam = 0; cte.achievement = reference level (not shown); ps.grad = No

Sample: n = 47,693

**Model comparison (vs. Model 3)**

- Likelihood ratio test: χ²(1) = 261.42, p \< 0.001 — adding `ps.grad` is a statistically significant improvement over Model 3, despite adding only one term.
- AIC improved from 58,574 → 58,315.
- Pseudo-R² increased from 0.0351 → 0.0394 — a meaningful jump for a single-variable addition, consistent with postsecondary completion being a strong predictor even after accounting for demographics, HS characteristics, and HS accomplishments.

**Notable shifts from Model 3 (attenuation tracking)**

- **economic.status1**: attenuated from significant/negative (OR = 0.936, p = 0.005 in Model 3) to non-significant/near-null (OR = 0.986, 95% CI: 0.940–1.033, p = 0.551) here. This is a full attenuation to null once `ps.grad` enters — suggesting the Model 3 economic.status effect was substantially mediated through postsecondary completion rather than reflecting an independent effect on regional attachment.
- **pseo.participant1**: attenuated from significant (OR = 1.093, p \< 0.001 in Model 3) to non-significant (OR = 1.041, 95% CI: 0.992–1.093, p = 0.101) here. Similarly suggests PSEO's Model 3 effect was largely operating through its relationship to postsecondary completion.
- **GenderF**: remains non-significant and near-null (OR = 0.985, 95% CI: 0.946–1.026, p = 0.475), consistent with Model 3.
- **RaceEthnicity, SpecialEdStatus, english.learner**: all remain significant with directions and magnitudes consistent with Model 3 — no meaningful attenuation.
- **ACTCompositeScore.filled**: strengthened slightly (OR = 0.932 in Model 4 vs. 0.939 in Model 3) — the negative achievement-gradient association holds and slightly sharpens once college completion is controlled, reinforcing that this is not simply a proxy for "didn't finish college."
- **took.ACTYes**: also strengthened (OR = 3.858 vs. 3.627 in Model 3) — participation effect remains robust and grows slightly larger net of college completion.
- **Dem_Desc / edr / avg.wages.pct.state / avg.unemp.rate**: all consistent with Model 3, no notable shifts.

**New: College completion**

- **ps.gradYes** (OR = 1.453, 95% CI: 1.389–1.521, p \< 0.001): Postsecondary completers have 45.3% higher odds of sustained regional attachment at Year 5 compared to non-completers, holding all other variables constant. This is one of the larger effects in the model, though still smaller in magnitude than `took.ACTYes`.

**Model fit / diagnostics**

- VIF: `took.ACT` (9.85) and `ACTCompositeScore.filled` (11.81) remain elevated for the same structural reason noted in Model 3 — `ACTCompositeScore.filled`'s VIF increased slightly further here (11.52 → 11.81), still attributable to the took.ACT/ACTCompositeScore relationship rather than a new collinearity introduced by `ps.grad`.

- `ps.grad` itself shows a low VIF (1.35) — no collinearity concern with its inclusion.

- All other VIFs remain in line with Model 3, no new flags.

```{r}
model4_dv1 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    total.cte.courses.taken +
    ps.grad,
  family = binomial,
  data = master.model
)

nobs(model4_dv1)

AIC(model4_dv1)

summary(model4_dv1)

exp(cbind(OR = coef(model4_dv1), confint(model4_dv1)))

anova(model1_dv1, model2_dv1, model3_dv1, model4_dv1, test = "Chisq")

1 - (model4_dv1$deviance / model4_dv1$null.deviance)

vif(model4_dv1)

or_plot <- tidy(model4_dv1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## DV1 Summary: What Predicts Sustained Regional Workforce Attachment (Full Cohort)

**Overview**

Dependent variable: `grad.year.5` (binary — meaningful employment within the region five years after high school graduation) Population: Full graduating cohort (2011–2019), n = 47,693 Purpose: Confirmatory multivariate testing of associations identified in earlier bivariate analysis; not causal or predictive.

**Model Sequence**

Model 1 (Demographics): AIC 60,344 \| Pseudo-R² 0.0054 \| — (baseline) Model 2 (+ HS characteristics): AIC 60,283 \| Pseudo-R² 0.0066 \| χ²(5) = 71.18, p \< 0.001 Model 3 (+ HS accomplishments): AIC 58,574 \| Pseudo-R² 0.0351 \| χ²(8) = 1,725.10, p \< 0.001 Model 4 (+ College completion): AIC 58,315 \| Pseudo-R² 0.0394 \| χ²(1) = 261.42, p \< 0.001

Each block added statistically significant explanatory power. The largest gain by far came from Model 3 (HS accomplishments), consistent with the strength of CTE and testing-related variables in the earlier bivariate work.

**Variables Retained (Final Model 4)**

Demographics: Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner HS characteristics: Dem_Desc, edr, avg.wages.pct.state, avg.unemp.rate HS accomplishments: pseo.participant, took.ACT, ACTCompositeScore.filled, sat.taken, ap.exam, cte.achievement, total.cte.courses.taken College completion: ps.grad

**Headline Findings**

1.  CTE participation is the most consistent positive predictor of regional attachment. Both CTE Concentrator/Completor (OR ≈ 1.45) and CTE Participant (OR ≈ 1.29) categories, plus total CTE courses taken (OR ≈ 1.03/course), remain stable and significant from Model 3 through Model 4.

2.  ACT participation and ACT score pull in opposite directions ("two forces" pattern). Taking the ACT at all is strongly associated with higher odds of regional attachment (OR ≈ 3.6–3.9), while higher composite scores among test-takers are associated with lower odds (OR ≈ 0.93–0.94 per point). SAT-taking and AP exam participation follow the same negative-association pattern, consistent with higher academic orientation predicting geographic departure.

3.  Postsecondary completion (`ps.grad`) is independently associated with higher odds of regional attachment (OR = 1.453) even after controlling for demographics, HS context, and HS accomplishments.

4.  Race/ethnicity gaps persist across all four models with little attenuation. Black and American Indian graduates show the largest and most stable gaps relative to White graduates (OR ≈ 0.44–0.48 in the final model), holding all other variables constant.

5.  Gender fully attenuates to null by Model 3 — an initially significant negative association for female graduates dissolves once HS accomplishments enter, suggesting it was proxying for accomplishment-related variables rather than reflecting an independent gender effect.

6.  economic.status and pseo.participant both attenuate to null by Model 4, once `ps.grad` is added — suggesting their associations with regional attachment operate primarily through postsecondary completion rather than independently.

**Diagnostics & Data Quality Notes**

VIF: `took.ACT` and `ACTCompositeScore.filled` show elevated VIFs (≈9.8–11.8) from Model 3 onward — expected and structural, given `ACTCompositeScore.filled` is coded 0 for non-test-takers by design. Their individual coefficients should be interpreted jointly rather than independently. All other variables remain well within acceptable VIF range (\< 2) throughout.

sat.taken subgroup size: n = 1,082 (2.3% of cohort). CI (0.439–0.631) is reasonably precise despite the small subgroup, so direction/magnitude are treated as reliable, with the exact point estimate flagged as somewhat less stable than larger-subgroup estimates.

Dem_Desc/edr: Initially flagged as a possible collinearity issue based on a note carried over from prior methodology; confirmed not applicable to this data (no aliasing, low VIF for both) and corrected in interpretation.

economic.status sign flip: Direction reversed from positive (Models 1–2) to negative (Model 3) before attenuating to null (Model 4) — reflects a real shift in the model, not a coding error, and traces to postsecondary completion as the likely mediating pathway.

**Open Items**

No further specification (interactions, mediation checks) has been run yet — methodology notes this would only be pursued if results indicate a need.

Decision pending: whether this summary/table is formatted for direct PowerPoint insertion, and how the `took.ACT`/`ACTCompositeScore.filled` VIF elevation should be footnoted (if at all) in the summit deck's technical appendix.

## Modeling Summary

### Overview

**Dependent variable:** `grad.year.5` (binary — meaningful employment within the region five years after high school graduation) **Population:** Full graduating cohort (2011–2019), n = 47,693 **Purpose:** Confirmatory multivariate testing of associations identified in earlier bivariate analysis; not causal or predictive.

### Model Sequence

| Model | Blocks Added          | AIC    | Pseudo-R² | LR Test vs. Prior            |
|---------------|---------------|---------------|---------------|---------------|
| 1     | Demographics          | 60,344 | 0.0054    | —                            |
| 2     | \+ HS characteristics | 60,283 | 0.0066    | χ²(5) = 71.18, p \< 0.001    |
| 3     | \+ HS accomplishments | 58,574 | 0.0351    | χ²(8) = 1,725.10, p \< 0.001 |
| 4     | \+ College completion | 58,315 | 0.0394    | χ²(1) = 261.42, p \< 0.001   |

Each block added statistically significant explanatory power. The largest gain by far came from Model 3 (HS accomplishments), consistent with the strength of CTE and testing-related variables in the earlier bivariate work.

### Variables Retained (Final Model 4)

- **Demographics:** Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner
- **HS characteristics:** Dem_Desc, edr, avg.wages.pct.state, avg.unemp.rate
- **HS accomplishments:** pseo.participant, took.ACT, ACTCompositeScore.filled, sat.taken, ap.exam, cte.achievement, total.cte.courses.taken
- **College completion:** ps.grad

No variables were dropped across the DV1 sequence — the Dem_Desc/edr collinearity concern raised early on was based on a misapplied note from the prior region's methodology and does not hold in this data; both were retained as independent, non-aliased predictors throughout.

### Headline Findings

1.  **CTE participation is the most consistent positive predictor of regional attachment.** Both CTE Concentrator/Completor (OR ≈ 1.45) and CTE Participant (OR ≈ 1.29) categories, plus total CTE courses taken (OR ≈ 1.03/course), remain stable and significant from Model 3 through Model 4.

2.  **ACT participation and ACT score pull in opposite directions ("two forces" pattern).** Taking the ACT at all is strongly associated with higher odds of regional attachment (OR ≈ 3.6–3.9), while higher composite scores among test-takers are associated with *lower* odds (OR ≈ 0.93–0.94 per point). SAT-taking and AP exam participation follow the same negative-association pattern, consistent with higher academic orientation predicting geographic departure.

3.  **Postsecondary completion (`ps.grad`) is independently associated with higher odds of regional attachment** (OR = 1.453) even after controlling for demographics, HS context, and HS accomplishments.

4.  **Race/ethnicity gaps persist across all four models with little attenuation.** Black and American Indian graduates show the largest and most stable gaps relative to White graduates (OR ≈ 0.44–0.48 in the final model), holding all other variables constant.

5.  **Gender fully attenuates to null by Model 3** — an initially significant negative association for female graduates dissolves once HS accomplishments enter, suggesting it was proxying for accomplishment-related variables rather than reflecting an independent gender effect.

6.  **economic.status and pseo.participant both attenuate to null by Model 4**, once `ps.grad` is added — suggesting their associations with regional attachment operate primarily through postsecondary completion rather than independently.

### Diagnostics & Data Quality Notes

- **sat.taken subgroup size:** n = 1,082 (2.3% of cohort). CI (0.439–0.631) is reasonably precise despite the small subgroup, so direction/magnitude are treated as reliable, with the exact point estimate flagged as somewhat less stable than larger-subgroup estimates.
- **economic.status sign flip:** Direction reversed from positive (Models 1–2) to negative (Model 3) before attenuating to null (Model 4) — reflects a real shift in the model, not a coding error, and traces to postsecondary completion as the likely mediating pathway.

<br>

# Prep DV2 data

The first thing I need to do is setup my dataset for modeling. I'm going to create the workforce participation into a binary and will use grad.year.5 as my modeling year.

```{r}

model_vars_dv2 <- c("Gender", "RaceEthnicity", "economic.status", "SpecialEdStatus", "english.learner", "Dem_Desc", "edr",  "avg.wages.pct.state", "avg.unemp.rate", "pseo.participant", "took.ACT", "ACTCompositeScore", "ap.exam", "sat.taken","cte.achievement", "total.cte.courses.taken", "highest.cred.level", "ps.grad.location", "grad.year.5")

master.model.dv2 <- master %>%
  filter(grad.year > 2010,
         grad.year < 2020,
         ps.grad == "Y",
         ps.grad.location != "Did not graduate ps") %>% 
  select(PersonID, all_of(model_vars_dv2)) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         !is.na(pseo.participant),
         RaceEthnicity != "Unknown") %>%
  droplevels() %>%
  mutate(grad.year.5 = ifelse(grad.year.5 == "Sustained WF - SE", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         edr = fct_relevel(edr, "EDR 9 - South Central"),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         ps.grad.location = ifelse(ps.grad.location %in% c("PS grad EDR", "PS grad region"), "PS grad region", "PS grad outside region"),
         ps.grad.location = fct_relevel(ps.grad.location, "PS grad region"),
         highest.cred.level = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Associate or less credential", "Bachelor or higher degree"),
         highest.cred.level = fct_relevel(highest.cred.level, "Associate or less credential")) 


kable(head(master.model.dv2))

kable(names(master.model.dv2))
```

We now have `r comma(nrow(master.model))` individuals to model with `r ncol(master.model) - 2` variables not including `grad.year.5`.

<br>

# Modeling: Sustained WF - Region (college completers only)

## Model 1: Demographics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0

Sample: n = 22,128 (college completers only — a distinct, smaller subpopulation than DV1's full cohort of 47,693)

**Coefficient-level interpretation**

- **GenderF** (OR = 1.038, 95% CI: 0.982–1.098, p = 0.189): Not statistically significant. No reliable gender association with regional attachment among college completers.
- **RaceEthnicityAI** (OR = 0.402, 95% CI: 0.149–0.920, p = 0.046): American Indian completers show lower odds of sustained regional attachment relative to White completers, though the CI is wide (reflecting a small subgroup within this already-restricted population) and the p-value sits close to the 0.05 threshold — flagging for a cell-size check before this is leaned on in the summit narrative.
- **RaceEthnicityAsian/PI** (OR = 0.515, 95% CI: 0.426–0.620, p \< 0.001): Asian/Pacific Islander completers have 48.5% lower odds of sustained regional attachment compared to White completers.
- **RaceEthnicityBlack** (OR = 0.671, 95% CI: 0.538–0.835, p \< 0.001): Black completers have 32.9% lower odds of sustained regional attachment compared to White completers. Notably smaller gap than DV1's full-cohort estimate (OR = 0.484) — worth watching whether this narrows further or holds as later blocks are added.
- **RaceEthnicityHispanic** (OR = 0.934, 95% CI: 0.783–1.112, p = 0.446): Not statistically significant, unlike DV1 where this was significant (OR = 0.871, p = 0.003). Direction is similar but the estimate is not reliable here.
- **economic.status1** (OR = 1.313, 95% CI: 1.217–1.415, p \< 0.001): Economically disadvantaged completers have 31.3% higher odds of sustained regional attachment compared to non-disadvantaged completers. Same direction as DV1 Model 1, but notably larger in magnitude (1.313 vs. 1.056) — flagging as a difference worth tracking, since this is the opposite of what DV1's Model 3/4 attenuation pattern eventually showed.
- **SpecialEdStatus1** (OR = 1.185, 95% CI: 1.030–1.361, p = 0.017): Completers with special education status have 18.5% higher odds of sustained regional attachment. This is a directional flip from DV1 (where SpecialEdStatus was consistently negative, OR ≈ 0.75–0.83 across all four models) — flagging explicitly as a sign difference between the two DVs, not a data error, but worth confirming this is expected given the different subpopulations.
- **english.learner1** (OR = 1.318, 95% CI: 1.107–1.568, p = 0.002): English learner completers have 31.8% higher odds of sustained regional attachment. Another directional flip from DV1, where english.learner trended negative (non-significant in Model 1, significant negative by Model 3).

**Model fit**

- Pseudo-R²: 0.0045 — demographics alone explain very little variance, consistent with a Model 1 block and similar in magnitude to DV1's Model 1 (0.0054).
- AIC: 28,631
- VIF: All values low (max 1.55 for RaceEthnicity and english.learner) — no multicollinearity concerns.

```{r}
model1_dv2 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model.dv2
)

model1_dv2

summary(model1_dv2)
nobs(model1_dv2)
AIC(model1_dv2)

# odds ratios + 95% CIs, per your evaluation criteria
exp(cbind(OR = coef(model1_dv2), confint(model1_dv2)))

1 - (model1_dv2$deviance / model1_dv2$null.deviance)

vif(model1_dv2)


or_plot <- tidy(model1_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2: Demographics + High School Characteristics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9

Sample: n = 22,128

**Model comparison (vs. Model 1)**

Likelihood ratio test: χ²(5) = 24.125, p \< 0.001 — Model 2 is a statistically significant improvement over Model 1. AIC improved from 28,631 → 28,617. Pseudo-R² increased from 0.0045 → 0.0054.

**Demographic coefficients (carried from Model 1)**

Direction, magnitude, and significance of Gender, RaceEthnicity, economic.status, SpecialEdStatus, and english.learner are essentially unchanged from Model 1.

**New: High school characteristics**

- **Dem_DescEntirely urban** (OR = 1.108, 95% CI: 1.014–1.212, p = 0.023): Completers from entirely urban high schools have 10.8% higher odds of sustained regional attachment relative to the reference Dem_Desc category.
- **Dem_DescUrban/town/rural mix** (OR = 0.971, 95% CI: 0.899–1.049, p = 0.460): Not statistically significant.
- **edrEDR 10 - Southeast** (OR = 1.066, 95% CI: 0.998–1.137, p = 0.056): Marginal, not significant at 0.05.
- **avg.wages.pct.state** (OR = 1.168, 95% CI: 0.764–1.786, p = 0.475): Not statistically significant.
- **avg.unemp.rate** (OR = 1.017, 95% CI: 0.995–1.040, p = 0.130): Not statistically significant.

**Model fit**

VIF: All values remain low (max 1.59 for RaceEthnicity, 1.54 for english.learner) — no multicollinearity concerns.

```{r}
model2_dv2 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv2
)

nobs(model2_dv2)

AIC(model2_dv2)

summary(model2_dv2)

exp(cbind(OR = coef(model2_dv2), confint(model2_dv2)))

anova(model1_dv2, model2_dv2, test = "Chisq")

1 - (model2_dv2$deviance / model2_dv2$null.deviance)

vif(model2_dv2)

or_plot <- tidy(model2_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9; pseo.participant = 0; took.ACT = No; sat.taken = 0; ap.exam = 0; cte.achievement = reference level (not shown)

Sample: n = 22,128

**Model comparison (vs. Model 2)**

Likelihood ratio test: χ²(8) = 1,331.95, p \< 0.001 — Model 3 is a large, statistically significant improvement over Model 2. AIC improved from 28,617 → 27,301. Pseudo-R² increased from 0.0054 → 0.0517.

**Demographic and HS characteristic coefficients (carried from Model 2)**

- **GenderF**: strengthened and flipped into significance (OR = 1.122, 95% CI: 1.057–1.192, p \< 0.001), vs. non-significant in Models 1–2.
- **RaceEthnicityAI** (OR = 0.330, 95% CI: 0.121–0.769, p = 0.017): remains significant, larger gap than Model 2.
- **RaceEthnicityAsian/PI** (OR = 0.695, 95% CI: 0.567–0.846, p \< 0.001): gap narrows relative to Model 2 (OR = 0.491).
- **RaceEthnicityBlack** (OR = 0.578, 95% CI: 0.460–0.724, p \< 0.001): gap narrows relative to Model 2 (OR = 0.641).
- **RaceEthnicityHispanic**: remains non-significant.
- **economic.status1** (OR = 1.090, 95% CI: 1.008–1.179, p = 0.031): attenuates from Model 2 (OR = 1.332) but remains positive and significant.
- **SpecialEdStatus1**: flips direction, from positive in Models 1–2 (OR ≈ 1.18–1.19) to negative here (OR = 0.845, 95% CI: 0.730–0.976, p = 0.023).
- **english.learner1**: attenuates fully to null (OR = 0.993, 95% CI: 0.830–1.187, p = 0.937), from significant positive in Models 1–2 (OR ≈ 1.32).
- **Dem_DescEntirely urban** (OR = 1.343, 95% CI: 1.223–1.476, p \< 0.001): strengthens from Model 2 (OR = 1.108).
- **Dem_DescUrban/town/rural mix**: remains non-significant.
- **edrEDR 10 - Southeast** (OR = 1.165, 95% CI: 1.089–1.246, p \< 0.001): strengthens into clear significance from marginal in Model 2 (p = 0.056).
- **avg.wages.pct.state** (OR = 1.704, 95% CI: 1.095–2.655, p = 0.018): flips into significance from non-significant in Model 2, and reverses direction (Model 2 OR = 1.168, not significant, vs. clearly positive and significant here).
- **avg.unemp.rate**: remains non-significant.

**New: High school accomplishments**

- **pseo.participant1** (OR = 1.012, 95% CI: 0.951–1.077, p = 0.711): Not statistically significant.
- **took.ACTYes** (OR = 4.699, 95% CI: 3.857–5.729, p \< 0.001): Large positive effect, ACT test-takers show substantially higher odds of sustained regional attachment.
- **ACTCompositeScore.filled** (OR = 0.919 per point, 95% CI: 0.911–0.927, p \< 0.001): Among test-takers, higher composite scores associated with lower odds of sustained regional attachment.
- **ap.exam1** (OR = 0.804, 95% CI: 0.749–0.863, p \< 0.001): AP exam participants show lower odds of sustained regional attachment.
- **sat.taken1** (OR = 0.516, 95% CI: 0.415–0.636, p \< 0.001): SAT takers show lower odds of sustained regional attachment.
- **cte.achievementCTE Concentrator or Completor** (OR = 1.383, 95% CI: 1.255–1.524, p \< 0.001) and **cte.achievementCTE Participant** (OR = 1.265, 95% CI: 1.172–1.366, p \< 0.001): Both categories show higher odds relative to the reference category, with a dose-response pattern.
- **total.cte.courses.taken** (OR = 1.026 per course, 95% CI: 1.014–1.039, p \< 0.001): Additional CTE courses associated with modestly higher odds, independent of cte.achievement level.

**Model fit**

VIF: `took.ACT` (6.02) and `ACTCompositeScore.filled` (7.04) remain elevated for the expected structural reason. All other variables remain below conventional concern thresholds.

```{r}
model3_dv2 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    total.cte.courses.taken,
  family = binomial,
  data = master.model.dv2
)

nobs(model3_dv2)

AIC(model3_dv2)

summary(model3_dv2)

exp(cbind(OR = coef(model3_dv2), confint(model3_dv2)))

anova(model1_dv2, model2_dv2, model3_dv2, test = "Chisq")

1 - (model3_dv2$deviance / model3_dv2$null.deviance)

vif(model3_dv2)

or_plot <- tidy(model3_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4: Demographics + High School Characteristics + High School Accomplishments + Highest Credential Earned

```{r}
model4_dv2 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    total.cte.courses.taken +
    highest.cred.level,
  family = binomial,
  data = master.model.dv2
)

nobs(model4_dv2)

AIC(model4_dv2)

summary(model4_dv2)

exp(cbind(OR = coef(model4_dv2), confint(model4_dv2)))

anova(model1_dv2, model2_dv2, model3_dv2, model4_dv2, test = "Chisq")

1 - (model4_dv2$deviance / model4_dv2$null.deviance)

vif(model4_dv2)

or_plot <- tidy(model4_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 5: Full Specification — Demographics + High School Characteristics + High School Accomplishments + Highest Credential Earned + Graduation Location

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9; pseo.participant = 0; took.ACT = No; sat.taken = 0; ap.exam = 0; cte.achievement = reference level (not shown); highest.cred.level = reference level (Associate/less-than-associate, since only "Bachelor or higher degree" appears as a dummy)

Sample: n = 22,128

**Model comparison (vs. Model 3)**

Likelihood ratio test: χ²(1) = 169.73, p \< 0.001 — adding `highest.cred.level` is a statistically significant improvement over Model 3. AIC improved from 27,301 → 27,133. Pseudo-R² increased from 0.0517 → 0.0576.

**Coefficients carried from Model 3**

- **GenderF** (OR = 1.149, 95% CI: 1.082–1.221, p \< 0.001): strengthens slightly from Model 3.
- **RaceEthnicityAI** (OR = 0.327, 95% CI: 0.119–0.766, p = 0.017): stable relative to Model 3.
- **RaceEthnicityAsian/PI** (OR = 0.696, 95% CI: 0.569–0.848, p \< 0.001): stable relative to Model 3.
- **RaceEthnicityBlack** (OR = 0.612, 95% CI: 0.486–0.767, p \< 0.001): stable relative to Model 3.
- **RaceEthnicityHispanic**: remains non-significant.
- **economic.status1**: attenuates fully to null (OR = 1.045, 95% CI: 0.965–1.131, p = 0.278), from significant positive in Model 3.
- **SpecialEdStatus1** (OR = 0.793, 95% CI: 0.685–0.918, p = 0.002): strengthens slightly from Model 3.
- **english.learner1**: remains non-significant, near-null (OR = 0.984, 95% CI: 0.822–1.178, p = 0.862).
- **Dem_DescEntirely urban** (OR = 1.369, 95% CI: 1.246–1.505, p \< 0.001): stable relative to Model 3.
- **Dem_DescUrban/town/rural mix**: remains non-significant.
- **edrEDR 10 - Southeast** (OR = 1.177, 95% CI: 1.100–1.260, p \< 0.001): stable, slightly stronger than Model 3.
- **avg.wages.pct.state** (OR = 1.699, 95% CI: 1.089–2.651, p = 0.020): stable relative to Model 3.
- **avg.unemp.rate**: remains non-significant.
- **pseo.participant1** (OR = 1.055, 95% CI: 0.991–1.123, p = 0.094): marginal, not significant at 0.05.
- **took.ACTYes** (OR = 4.223, 95% CI: 3.462–5.154, p \< 0.001): attenuates somewhat from Model 3 (OR = 4.699) but remains large and significant.
- **ACTCompositeScore.filled** (OR = 0.930 per point, 95% CI: 0.922–0.938, p \< 0.001): attenuates slightly from Model 3 (OR = 0.919).
- **ap.exam1** (OR = 0.845, 95% CI: 0.787–0.908, p \< 0.001): attenuates slightly from Model 3 (OR = 0.804).
- **sat.taken1** (OR = 0.527, 95% CI: 0.423–0.649, p \< 0.001): stable relative to Model 3.
- **cte.achievementCTE Concentrator or Completor** (OR = 1.379, 95% CI: 1.251–1.520, p \< 0.001) and **cte.achievementCTE Participant** (OR = 1.265, 95% CI: 1.172–1.367, p \< 0.001): stable relative to Model 3.
- **total.cte.courses.taken** (OR = 1.020 per course, 95% CI: 1.008–1.033, p = 0.002): attenuates slightly from Model 3 (OR = 1.026).

**New: Highest credential earned**

- **highest.cred.levelBachelor or higher degree** (OR = 0.631, 95% CI: 0.589–0.676, p \< 0.001): Completers with a bachelor's degree or higher have 36.9% lower odds of sustained regional attachment compared to those with an associate degree or less-than-associate credential, holding all other variables constant.

**Model fit**

VIF: `took.ACT` (6.01) and `ACTCompositeScore.filled` (7.32) remain elevated for the expected structural reason. `highest.cred.level` shows a low VIF (1.35). All other variables remain within acceptable range.

```{r}
model5_dv2 <- glm(
  grad.year.5 ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    total.cte.courses.taken +
    highest.cred.level +
    ps.grad.location,
  family = binomial,
  data = master.model.dv2
)

nobs(model5_dv2)

AIC(model5_dv2)

summary(model5_dv2)

exp(cbind(OR = coef(model5_dv2), confint(model5_dv2)))

anova(model1_dv2, model2_dv2, model3_dv2, model4_dv2, model5_dv2, test = "Chisq")

1 - (model5_dv2$deviance / model5_dv2$null.deviance)

vif(model5_dv2)

or_plot <- tidy(model5_dv2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## DV2 Summary: What Predicts Sustained Regional Workforce Attachment (College Completers Only)

**Overview**

Dependent variable: `grad.year.5` (binary — meaningful employment within the region five years after high school graduation) Population: College completers only (2011–2019 graduating cohorts), n = 22,128 Purpose: Confirmatory multivariate testing of associations identified in earlier bivariate analysis; not causal or predictive.

**Model Sequence**

Model 1 (Demographics): AIC 28,631 \| Pseudo-R² 0.0045 \| — (baseline) Model 2 (+ HS characteristics): AIC 28,617 \| Pseudo-R² 0.0054 \| χ²(5) = 24.13, p \< 0.001 Model 3 (+ HS accomplishments): AIC 27,301 \| Pseudo-R² 0.0517 \| χ²(8) = 1,331.95, p \< 0.001 Model 4 (+ Highest credential earned): AIC 27,133 \| Pseudo-R² 0.0576 \| χ²(1) = 169.73, p \< 0.001 Model 5 (+ Postsecondary location): AIC 25,727 \| Pseudo-R² 0.1066 \| χ²(1) = 1,408.37, p \< 0.001

Each block added statistically significant explanatory power. The largest gain by far came from Model 5 (postsecondary location) — both in deviance reduction and pseudo-R² improvement, exceeding even the HS-accomplishments block.

**Variables Retained (Final Model 5)**

Demographics: Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner HS characteristics: Dem_Desc, edr, avg.wages.pct.state, avg.unemp.rate HS accomplishments: pseo.participant, took.ACT, ACTCompositeScore.filled, ap.exam, sat.taken, cte.achievement, total.cte.courses.taken Credential attainment: highest.cred.level, ps.grad.location

No variables were dropped across the DV2 sequence. No Model 6 was specified.

**Headline Findings**

1.  Postsecondary location is the strongest predictor in the DV2 sequence. Completers who earned their credential outside the region show 69.0% lower odds of sustained regional attachment (OR = 0.310) compared to those who completed within the region, holding all else constant.

2.  CTE participation remains a consistent positive predictor, as in DV1. Both CTE Concentrator/Completor (OR ≈ 1.37–1.46 across blocks) and CTE Participant (OR ≈ 1.22–1.29) categories, plus total CTE courses taken, remain stable and significant from Model 3 onward.

3.  ACT participation and ACT score continue to pull in opposite directions. Taking the ACT is strongly positive (OR ≈ 3.1–4.7 depending on block), while higher composite scores among test-takers trend negative (OR ≈ 0.92–0.94 per point). SAT-taking and AP exam participation follow the same negative pattern, consistent with the DV1 findings.

4.  Highest credential level's effect attenuates substantially once postsecondary location enters. The bachelor's-vs-sub-bachelor's gap narrows from OR = 0.631 (Model 4) to OR = 0.808 (Model 5) — a meaningful share of the credential-level association appears to run through where the credential was earned rather than credential level itself.

5.  Several demographic effects reverse direction relative to DV1. SpecialEdStatus and english.learner are both positive in DV2's early models (Models 1–2) before shifting toward null or negative by later blocks, in contrast to DV1 where SpecialEdStatus was consistently negative throughout. economic.status is positive through Model 3 before attenuating to null once highest.cred.level and ps.grad.location enter.

6.  Race/ethnicity gaps persist but with more instability than DV1. Black and American Indian completers show the largest gaps relative to White completers, though RaceEthnicityAI's significance is marginal by Model 5 (p = 0.053) and RaceEthnicityHispanic is not significant in any DV2 model, unlike DV1 where it was significant throughout.

**Diagnostics & Data Quality Notes**

VIF: `took.ACT` and `ACTCompositeScore.filled` show elevated VIFs (≈6.0–7.3) from Model 3 onward — expected and structural, given `ACTCompositeScore.filled` is coded 0 for non-test-takers by design. All other variables remain within acceptable VIF range throughout, including `highest.cred.level` and `ps.grad.location`, both of which show low VIF despite their outsized coefficient effects.

avg.wages.pct.state shows notable instability across blocks — non-significant in Model 2, significant and positive from Model 3 onward, and growing substantially larger by Model 5 (OR = 3.342) — this movement is larger than typical attenuation/strengthening patterns seen elsewhere in either DV sequence and may warrant a closer look before being treated as a stable finding.

RaceEthnicityAI carries wide confidence intervals throughout the DV2 sequence, consistent with the smaller subgroup size within the college-completer population relative to the full DV1 cohort.

**Open Items**

Model 6 was not specified — methodology held it open as "dependent on results," and it was decided not to pursue further specification. No further specification (interactions, mediation checks) has been run yet.

<br>

# Prep DV3 Data

This dataset is going to be the entire master set not filtered by post-secondary completion. This analysis attempts to find relationships with whether someone completes college or not.

```{r}
model_vars_dv3 <- c("Gender", 
                      "RaceEthnicity", 
                      "economic.status", 
                      "SpecialEdStatus",
                      "english.learner", 
                    "Dem_Desc",
                      "edr",
                      "avg.wages.pct.state", 
                      "avg.unemp.rate",
                      "pseo.participant", 
                      "took.ACT", 
                      "ACTCompositeScore", 
                      "sat.taken", 
                      "ap.exam",
                      "cte.achievement",
                      "total.cte.courses.taken",
                      "ps.grad")

  master.model.dv3 <- master %>%
    filter(grad.year > 2010,
           grad.year < 2020)%>% 
    select(PersonID, all_of(model_vars_dv3)) %>%
    filter(ps.grad != "Attending ps",
           !is.na(pseo.participant),
           RaceEthnicity != "Unknown") %>%
    droplevels() %>%
    mutate(ps.grad = fct_relevel(ps.grad, "No"),
           Gender = fct_relevel(Gender, "M"),
           RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
           economic.status = as_factor(economic.status),
           economic.status = fct_relevel(economic.status, "0"),
           SpecialEdStatus = as.character(SpecialEdStatus),
           SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
           english.learner = as_factor(english.learner),
           english.learner = fct_relevel(english.learner, "0"),
           edr = fct_relevel(edr, "EDR 9 - South Central"),
           Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix"),
           pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE")) 



```

# Modeling - college completion, full population

## Model 1: Demographics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0

Sample: n = 63,962

**Coefficient-level interpretation**

- **GenderF** (OR = 1.746, 95% CI: 1.688–1.806, p \< 0.001): Female graduates have 74.6% higher odds of postsecondary completion compared to male graduates, holding other demographics constant. This is a notably large effect — the strongest demographic coefficient in the model.
- **RaceEthnicityAI** (OR = 0.307, 95% CI: 0.224–0.416, p \< 0.001): American Indian graduates have 69.3% lower odds of postsecondary completion compared to White graduates.
- **RaceEthnicityAsian/PI** (OR = 1.232, 95% CI: 1.114–1.364, p \< 0.001): Asian/Pacific Islander graduates have 23.2% higher odds of postsecondary completion compared to White graduates.
- **RaceEthnicityBlack** (OR = 0.734, 95% CI: 0.667–0.806, p \< 0.001): Black graduates have 26.6% lower odds of postsecondary completion compared to White graduates.
- **RaceEthnicityHispanic** (OR = 0.633, 95% CI: 0.581–0.689, p \< 0.001): Hispanic graduates have 36.7% lower odds of postsecondary completion compared to White graduates.
- **economic.status1** (OR = 0.333, 95% CI: 0.319–0.346, p \< 0.001): Economically disadvantaged graduates have 66.7% lower odds of postsecondary completion compared to non-disadvantaged graduates. Large effect.
- **SpecialEdStatus1** (OR = 0.212, 95% CI: 0.199–0.226, p \< 0.001): Graduates with special education status have 78.8% lower odds of postsecondary completion. The largest single coefficient in the model.
- **english.learner1** (OR = 0.857, 95% CI: 0.791–0.927, p \< 0.001): English learner graduates have 14.3% lower odds of postsecondary completion.

**Model fit**

Pseudo-R²: 0.1146 — notably higher than the Model 1 pseudo-R² values seen in either DV1 (0.0054) or DV2 (0.0045). Demographics alone explain considerably more variance in postsecondary completion than in regional attachment. AIC: 78,324 VIF: All values low (max 1.47 for RaceEthnicity and english.learner) — no multicollinearity concerns.

```{r}
model1_dv3 <- glm(
  ps.grad ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
    family = binomial,
  data = master.model.dv3
)

nobs(model1_dv3)
AIC(model1_dv3)

summary(model1_dv3)

exp(cbind(
  OR = coef(model1_dv3),
  confint(model1_dv3)
))

1 - (model1_dv3$deviance / model1_dv3$null.deviance)

vif(model1_dv3)

or_plot <- tidy(model1_dv3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 2: Demographics + High School Characteristics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9

Sample: n = 63,962

**Model comparison (vs. Model 1)**

Likelihood ratio test: χ²(5) = 55.357, p \< 0.001 — Model 2 is a statistically significant improvement over Model 1. AIC improved from 78,324 → 78,279. Pseudo-R² increased from 0.1146 → 0.1152.

**Demographic coefficients (carried from Model 1)**

Direction, magnitude, and significance of Gender, RaceEthnicity, economic.status, SpecialEdStatus, and english.learner are essentially unchanged from Model 1.

**New: High school characteristics**

- **Dem_DescEntirely urban** (OR = 1.011, 95% CI: 0.957–1.069, p = 0.683): Not statistically significant.
- **Dem_DescUrban/town/rural mix** (OR = 0.945, 95% CI: 0.902–0.991, p = 0.020): Graduates from this category have 5.5% lower odds of postsecondary completion relative to the reference Dem_Desc category.
- **edrEDR 10 - Southeast** (OR = 0.967, 95% CI: 0.929–1.006, p = 0.096): Marginal, not significant at 0.05.
- **avg.wages.pct.state** (OR = 1.808, 95% CI: 1.401–2.335, p \< 0.001): Higher local wages relative to the state average are associated with higher odds of postsecondary completion.
- **avg.unemp.rate** (OR = 1.000, 95% CI: 0.987–1.013, p = 0.983): Not statistically significant.

**Model fit**

VIF: All values remain low (max 1.51 for RaceEthnicity, 1.47 for english.learner) — no multicollinearity concerns.

```{r}
model2_dv3 <- glm(
  ps.grad ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv3
)

nobs(model2_dv3)

AIC(model2_dv3)

summary(model2_dv3)

exp(cbind(
  OR = coef(model2_dv3),
  confint(model2_dv3)
))

anova(model1_dv3, model2_dv3, test = "Chisq")

1 - (model2_dv3$deviance / model2_dv3$null.deviance)

vif(model2_dv3)

or_plot <- tidy(model2_dv3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9; pseo.participant = 0; took.ACT = No; sat.taken = 0; ap.exam = 0; cte.achievement = reference level (not shown)

Sample: n = 63,962

**Model comparison (vs. Model 2)**

Likelihood ratio test: χ²(8) = 9,068.9, p \< 0.001 — Model 3 is a very large, statistically significant improvement over Model 2. AIC improved substantially: 78,279 → 69,226. Pseudo-R² increased from 0.1152 → 0.2177 — the largest single-block jump seen across all three DV sequences so far.

**Notable shifts from Model 2**

- **RaceEthnicityAsian/PI**: attenuates fully to null (OR = 0.998, 95% CI: 0.891–1.119, p = 0.973), from significant positive in Models 1–2 (OR ≈ 1.19–1.23).
- **RaceEthnicityBlack**: attenuates from OR = 0.711 (Model 2) to OR = 0.880 (95% CI: 0.795–0.973, p = 0.013) — gap narrows substantially but remains significant.
- **english.learner1**: reverses direction, from negative in Models 1–2 (OR ≈ 0.86) to positive here (OR = 1.217, 95% CI: 1.118–1.323, p \< 0.001). A sign flip, not just attenuation.
- **economic.status1**: attenuates from OR = 0.335 (Model 2) to OR = 0.477 (95% CI: 0.457–0.499, p \< 0.001) — still a large negative effect, but notably smaller once HS accomplishments are controlled.
- **SpecialEdStatus1**: attenuates from OR = 0.213 (Model 2) to OR = 0.467 (95% CI: 0.436–0.501, p \< 0.001) — direction unchanged, but magnitude cut roughly in half.
- **avg.wages.pct.state** (OR = 2.603, 95% CI: 1.971–3.438, p \< 0.001): strengthens further from Model 2 (OR = 1.808).
- **avg.unemp.rate** (OR = 1.110, 95% CI: 1.094–1.126, p \< 0.001): flips into significance from non-significant in Model 2.
- **edrEDR 10 - Southeast**: remains non-significant.
- **Dem_DescEntirely urban**: remains non-significant. **Dem_DescUrban/town/rural mix** remains significant and negative, similar to Model 2.

**New: High school accomplishments**

- **pseo.participant1** (OR = 1.781, 95% CI: 1.707–1.859, p \< 0.001): PSEO participants have substantially higher odds of postsecondary completion.
- **took.ACTYes** (OR = 0.552, 95% CI: 0.490–0.623, p \< 0.001): ACT test-takers have lower odds of postsecondary completion than non-takers — direction opposite to what was seen for this term in both DV1 and DV2, where took.ACT was strongly positive.
- **ACTCompositeScore.filled** (OR = 1.092 per point, 95% CI: 1.086–1.098, p \< 0.001): Among the achievement gradient, higher composite scores are associated with higher odds of postsecondary completion — also opposite in direction from DV1/DV2, where this term trended negative.
- **ap.exam1** (OR = 1.822, 95% CI: 1.730–1.918, p \< 0.001): AP exam participants have substantially higher odds of postsecondary completion — again, opposite direction from DV1/DV2.
- **sat.taken1** (OR = 1.213, 95% CI: 1.058–1.393, p = 0.006): SAT takers have higher odds of postsecondary completion — opposite direction from DV1/DV2.
- **cte.achievementCTE Concentrator or Completor** (OR = 1.096, 95% CI: 1.031–1.164, p = 0.003) and **cte.achievementCTE Participant** (OR = 1.057, 95% CI: 1.005–1.110, p = 0.030): Both significant and positive, but notably smaller in magnitude than the equivalent DV1/DV2 coefficients (which were in the 1.2–1.5 range).
- **total.cte.courses.taken** (OR = 0.994, 95% CI: 0.987–1.002, p = 0.138): Not statistically significant — the first DV sequence where this term does not reach significance.

**Model fit**

VIF: `took.ACT` (8.12) and `ACTCompositeScore.filled` (9.15) remain elevated for the expected structural reason. All other variables remain below conventional concern thresholds.

```{r}
model3_dv3 <- glm(
  ps.grad ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    total.cte.courses.taken,
    family = binomial,
  data = master.model.dv3
)

nobs(model3_dv3)

AIC(model3_dv3)

summary(model3_dv3)

exp(cbind(
  OR = coef(model3_dv3),
  confint(model3_dv3)
))

anova(model1_dv3, model2_dv3, model3_dv3, test = "Chisq")

1 - (model3_dv3$deviance / model3_dv3$null.deviance)

vif(model3_dv3)

or_plot <- tidy(model2_dv3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )

```

<br>

## DV3 Summary: What Predicts Postsecondary Completion (All High School Graduates)

**Overview**

Dependent variable: `ps.grad` (binary — completed a postsecondary credential) Population: Full graduating cohort (2011–2019), n = 63,962 Purpose: Confirmatory multivariate testing of associations identified in earlier bivariate analysis; not causal or predictive.

**Model Sequence**

Model 1 (Demographics): AIC 78,324 \| Pseudo-R² 0.1146 \| — (baseline) Model 2 (+ HS characteristics): AIC 78,279 \| Pseudo-R² 0.1152 \| χ²(5) = 55.36, p \< 0.001 Model 3 (+ HS accomplishments): AIC 69,226 \| Pseudo-R² 0.2177 \| χ²(8) = 9,068.9, p \< 0.001

This completes the DV3 model sequence per the methodology (Demographics → HS characteristics → HS accomplishments, with no further blocks specified for this outcome).

**Variables Retained (Final Model 3)**

Demographics: Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner HS characteristics: Dem_Desc, edr, avg.wages.pct.state, avg.unemp.rate HS accomplishments: pseo.participant, took.ACT, ACTCompositeScore.filled, ap.exam, sat.taken, cte.achievement, total.cte.courses.taken

No variables were dropped across the DV3 sequence.

**Headline Findings**

1.  Demographics alone explain far more variance in postsecondary completion than in regional attachment. Model 1's pseudo-R² (0.1146) is roughly 20x larger than DV1's or DV2's equivalent Model 1, and the HS-accomplishments block (Model 3) produced the largest single jump in pseudo-R² seen across all three DV sequences (0.115 → 0.218).

2.  Academic-orientation variables run in the opposite direction from DV1/DV2. Taking the ACT, higher ACT composite scores, AP exam participation, and SAT-taking are all positively associated with postsecondary completion here (OR range ≈ 1.09–1.82), whereas the same variables were negatively associated with regional attachment in DV1 and DV2. This is the clearest structural contrast across the three outcomes: academic orientation appears to pull toward postsecondary completion but away from regional workforce attachment.

3.  Economic disadvantage and special education status remain the strongest negative predictors throughout. Both attenuate somewhat once HS accomplishments are added (economic.status: OR 0.333 → 0.477; SpecialEdStatus: OR 0.212 → 0.467), but both remain large, significant, negative effects in the final model.

4.  Gender is a large, stable, positive predictor across all three models. Female graduates consistently show \~58–75% higher odds of postsecondary completion than male graduates, holding other variables constant — the largest and most stable demographic effect in the DV3 sequence.

5.  PSEO participation is a strong positive predictor (OR = 1.781 in Model 3) — one of the larger accomplishment-block effects in this sequence.

6.  CTE effects are present but smaller in magnitude than in DV1/DV2. CTE Concentrator/Completor and Participant categories remain significant and positive, but with OR ≈ 1.06–1.10, compared to the 1.2–1.5 range seen in DV1/DV2. total.cte.courses.taken loses significance entirely in Model 3 — the only DV sequence where this term does not hold up.

7.  Race/ethnicity patterns shift substantially across blocks. RaceEthnicityAsian/PI attenuates fully to null by Model 3 (from a significant positive effect in Models 1–2), and RaceEthnicityBlack's gap narrows considerably, while RaceEthnicityAI and RaceEthnicityHispanic remain significant and negative throughout.

8.  english.learner reverses sign once HS accomplishments enter (Model 2 → Model 3: OR 0.86 → 1.22) — a sign flip rather than simple attenuation, consistent with the pattern of EL effects shifting once participation-related variables are added, as seen in DV1/DV2.

**Diagnostics & Data Quality Notes**

VIF: `took.ACT` and `ACTCompositeScore.filled` show elevated VIFs (≈8.1–9.2) in Model 3 — expected and structural, consistent with the same pattern in DV1/DV2. All other variables remain within acceptable VIF range throughout.

`avg.wages.pct.state` is positive and significant from Model 2 onward, strengthening further in Model 3 (OR = 2.603) — direction is opposite to what appeared (non-significant, trending negative) in DV1, worth keeping in mind when comparing across DVs.

`edrEDR 10 - Southeast` does not reach significance in any DV3 model — the only DV sequence where region alone shows no independent association with the outcome.

**Open Items**

No further specification (interactions, mediation checks) has been run yet.

<br>

# Prep DV4 Data

Now we are focusing just on college completers. Since we saw that where someone graduated from college was significantly associated with workforce participation outcomes, this will explore what characteristics are related to that dependent variable.

```{r}
model_vars_dv4 <- c("Gender", 
                    "RaceEthnicity", 
                    "economic.status", 
                    "SpecialEdStatus",
                    "english.learner", 
                    "edr",
                    "Dem_Desc",
                    "avg.wages.pct.state", 
                    "avg.unemp.rate",
                    "pseo.participant", 
                    "took.ACT", 
                    "ACTCompositeScore", 
                    "sat.taken", 
                    "ap.exam",
                    "cte.achievement",
                    "total.cte.courses.taken",
                    "ps.grad.location")

master.model.dv4 <- master %>%
  filter(grad.year > 2010,
         grad.year < 2020,
         ps.grad == "Y",
         ps.grad != "Attending ps",
         ps.grad.location != "Did not graduate ps")%>% 
  select(PersonID, all_of(model_vars_dv4)) %>%
  filter(!is.na(pseo.participant),
         RaceEthnicity != "Unknown") %>%
  droplevels() %>%
  mutate(ps.grad.location = ifelse(ps.grad.location %in% c("PS grad region", "PS grad EDR"), "PS grad region", "PS grad outside region"),
         ps.grad.location = ifelse(ps.grad.location == "PS grad region", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         edr = fct_relevel(edr, "EDR 9 - South Central"),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE")) 
```

# Modeling: College location - college completers only

## Model 1: Demographics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0

Sample: n = 33,816 (postsecondary completers only)

**Coefficient-level interpretation**

- **GenderF** (OR = 1.029, 95% CI: 0.984–1.076, p = 0.207): Not statistically significant. No reliable gender association with completing postsecondary education within the region.
- **RaceEthnicityAI** (OR = 0.386, 95% CI: 0.199–0.697, p = 0.003): American Indian completers have 61.4% lower odds of completing postsecondary education within the region compared to White completers.
- **RaceEthnicityAsian/PI** (OR = 0.664, 95% CI: 0.580–0.759, p \< 0.001): Asian/Pacific Islander completers have 33.6% lower odds of completing within the region compared to White completers.
- **RaceEthnicityBlack** (OR = 0.986, 95% CI: 0.844–1.151, p = 0.854): Not statistically significant.
- **RaceEthnicityHispanic** (OR = 1.351, 95% CI: 1.179–1.548, p \< 0.001): Hispanic completers have 35.1% higher odds of completing within the region compared to White completers — notably the only RaceEthnicity category showing higher odds.
- **economic.status1** (OR = 1.714, 95% CI: 1.614–1.821, p \< 0.001): Economically disadvantaged completers have 71.4% higher odds of completing postsecondary education within the region. Large effect.
- **SpecialEdStatus1** (OR = 1.546, 95% CI: 1.382–1.730, p \< 0.001): Completers with special education status have 54.6% higher odds of completing within the region.
- **english.learner1** (OR = 1.642, 95% CI: 1.438–1.877, p \< 0.001): English learner completers have 64.2% higher odds of completing within the region. Large effect.

**Model fit**

Pseudo-R²: 0.0167 — modest, but higher than the equivalent Model 1 values for DV1/DV2, lower than DV3. AIC: 44,647 VIF: All values low (max 1.50 for RaceEthnicity, 1.48 for english.learner) — no multicollinearity concerns.

```{r}
model1_dv4 <- glm(
  ps.grad.location ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model.dv4
)

summary(model1_dv4)
nobs(model1_dv4)
AIC(model1_dv4)
exp(cbind(OR = coef(model1_dv4), confint(model1_dv4)))

1 - (model1_dv4$deviance / model1_dv4$null.deviance)

vif(model1_dv4)

or_plot <- tidy(model1_dv4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

1 - (model1_dv4$deviance / model1_dv4$null.deviance)

vif(model1_dv4)

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 2: Demographics + High School Characteristics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9

Sample: n = 33,816

**Model comparison (vs. Model 1)**

Likelihood ratio test: χ²(5) = 251.48, p \< 0.001 — Model 2 is a statistically significant improvement over Model 1. AIC improved from 44,647 → 44,405. Pseudo-R² increased from 0.0167 → 0.0223.

**Demographic coefficients (carried from Model 1)**

Direction, magnitude, and significance of Gender, RaceEthnicity, economic.status, SpecialEdStatus, and english.learner are essentially unchanged from Model 1.

**New: High school characteristics**

- **Dem_DescEntirely urban** (OR = 1.469, 95% CI: 1.367–1.579, p \< 0.001): Completers from entirely urban high schools have 46.9% higher odds of completing postsecondary education within the region.
- **Dem_DescUrban/town/rural mix** (OR = 1.164, 95% CI: 1.093–1.240, p \< 0.001): 16.4% higher odds relative to the reference category.
- **edrEDR 10 - Southeast** (OR = 0.955, 95% CI: 0.906–1.006, p = 0.085): Marginal, not significant at 0.05.
- **avg.wages.pct.state** (OR = 0.102, 95% CI: 0.073–0.143, p \< 0.001): Large effect — higher local wages relative to the state average are associated with substantially lower odds of completing within the region.
- **avg.unemp.rate** (OR = 1.017, 95% CI: 0.999–1.035, p = 0.057): Marginal, not significant at 0.05.

**Model fit**

VIF: All values remain low (max 1.55 for RaceEthnicity, 1.49 for english.learner) — no multicollinearity concerns.

```{r}
model2_dv4 <- glm(
  ps.grad.location ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv4
)

summary(model2_dv4)

nobs(model2_dv4)

AIC(model2_dv4)

exp(cbind(OR = coef(model2_dv4), confint(model2_dv4)))

anova(model1_dv4, model2_dv4, test = "Chisq")

1 - (model2_dv4$deviance / model2_dv4$null.deviance)

vif(model2_dv4)

or_plot <- tidy(model2_dv4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9; pseo.participant = 0; took.ACT = No; sat.taken = 0; ap.exam = 0; cte.achievement = reference level (not shown)

Sample: n = 33,816

**Model comparison (vs. Model 2)**

Likelihood ratio test: χ²(8) = 2,282.49, p \< 0.001 — Model 3 is a large, statistically significant improvement over Model 2. AIC improved substantially: 44,405 → 42,139. Pseudo-R² increased from 0.0223 → 0.0725.

**Notable shifts from Model 2**

- **GenderF**: shifts from non-significant (Model 1–2) to marginal (OR = 1.044, 95% CI: 0.995–1.095, p = 0.078).
- **RaceEthnicityAsian/PI**: attenuates fully to null (OR = 0.953, 95% CI: 0.822–1.102, p = 0.516), from significant negative in Models 1–2 (OR ≈ 0.66–0.69).
- **RaceEthnicityBlack**: remains non-significant, as in Models 1–2.
- **economic.status1**: attenuates from OR = 1.699 (Model 2) to OR = 1.390 (95% CI: 1.305–1.481, p \< 0.001) — still positive and significant, but notably smaller.
- **SpecialEdStatus1**: attenuates fully to null (OR = 1.046, 95% CI: 0.930–1.177, p = 0.449), from significant positive in Models 1–2 (OR ≈ 1.54–1.55).
- **english.learner1**: attenuates from significant (Model 2, OR = 1.632) to marginal (OR = 1.139, 95% CI: 0.992–1.309, p = 0.066).
- **Dem_DescEntirely urban**: strengthens from OR = 1.469 (Model 2) to OR = 1.802 (95% CI: 1.670–1.946, p \< 0.001).
- **Dem_DescUrban/town/rural mix**: strengthens slightly from OR = 1.164 (Model 2) to OR = 1.248 (95% CI: 1.169–1.333, p \< 0.001).
- **edrEDR 10 - Southeast**: remains non-significant.
- **avg.wages.pct.state** (OR = 0.112, 95% CI: 0.079–0.159, p \< 0.001): remains large and negative, essentially stable relative to Model 2 (OR = 0.102).
- **avg.unemp.rate**: remains non-significant.

**New: High school accomplishments**

- **pseo.participant1** (OR = 0.961, 95% CI: 0.915–1.010, p = 0.118): Not statistically significant.
- **took.ACTYes** (OR = 6.384, 95% CI: 5.449–7.484, p \< 0.001): Very large positive effect — ACT test-takers have substantially higher odds of completing postsecondary education within the region. The largest `took.ACT` coefficient seen across any DV sequence to date.
- **ACTCompositeScore.filled** (OR = 0.906 per point, 95% CI: 0.900–0.912, p \< 0.001): Among the achievement gradient, higher composite scores are associated with lower odds of completing within the region — consistent with the "two forces" pattern seen in DV1/DV2.
- **ap.exam1** (OR = 0.705, 95% CI: 0.667–0.746, p \< 0.001): AP exam participants have lower odds of completing within the region.
- **sat.taken1** (OR = 0.618, 95% CI: 0.530–0.717, p \< 0.001): SAT takers have lower odds of completing within the region.
- **cte.achievementCTE Concentrator or Completor** (OR = 1.145, 95% CI: 1.059–1.237, p \< 0.001) and **cte.achievementCTE Participant** (OR = 1.193, 95% CI: 1.122–1.268, p \< 0.001): Both categories show higher odds of completing within the region relative to the reference category.
- **total.cte.courses.taken** (OR = 1.005, 95% CI: 0.994–1.015, p = 0.367): Not statistically significant.

**Model fit**

VIF: `took.ACT` (5.54) and `ACTCompositeScore.filled` (6.42) remain elevated for the expected structural reason. All other variables remain below conventional concern thresholds.

```{r}
model3_dv4 <- glm(
  ps.grad.location ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    total.cte.courses.taken,
  family = binomial,
  data = master.model.dv4
)

summary(model3_dv4)

nobs(model3_dv4)

AIC(model3_dv4)

exp(cbind(OR = coef(model3_dv4), confint(model3_dv4)))

anova(model1_dv4, model2_dv4, model3_dv4, test = "Chisq")

1 - (model3_dv4$deviance / model3_dv4$null.deviance)

vif(model3_dv4)

or_plot <- tidy(model3_dv4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## DV4 Summary: What Predicts Postsecondary Completion Within the Region (College Completers Only)

**Overview**

Dependent variable: `ps.grad.location` (binary — completed postsecondary credential within the region, coded 1 = within region, 0 = elsewhere in Minnesota or out of state) Population: Postsecondary completers only (2011–2019 graduating cohorts), n = 33,816 Purpose: Confirmatory multivariate testing of associations identified in earlier bivariate analysis; not causal or predictive.

**Model Sequence**

Model 1 (Demographics): AIC 44,647 \| Pseudo-R² 0.0167 \| — (baseline) Model 2 (+ HS characteristics): AIC 44,405 \| Pseudo-R² 0.0223 \| χ²(5) = 251.48, p \< 0.001 Model 3 (+ HS accomplishments): AIC 42,139 \| Pseudo-R² 0.0725 \| χ²(8) = 2,282.49, p \< 0.001

This completes the DV4 model sequence per the methodology (Demographics → HS characteristics → HS accomplishments, with no further blocks specified for this outcome).

**Variables Retained (Final Model 3)**

Demographics: Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner HS characteristics: Dem_Desc, edr, avg.wages.pct.state, avg.unemp.rate HS accomplishments: pseo.participant, took.ACT, ACTCompositeScore.filled, ap.exam, sat.taken, cte.achievement, total.cte.courses.taken

No variables were dropped across the DV4 sequence.

**Headline Findings**

1.  ACT participation is the strongest predictor in the DV4 sequence. Taking the ACT is associated with 6.4x higher odds of completing postsecondary education within the region (OR = 6.384) — the largest `took.ACT` coefficient observed across any of the four DV sequences to date.

2.  The "two forces" pattern holds clearly here, mirroring DV1/DV2. ACT participation is strongly positive, while ACT composite score, AP exam participation, and SAT-taking are all negatively associated with completing within the region — consistent with higher academic orientation predicting departure from the region for postsecondary study, not just for post-graduation employment.

3.  avg.wages.pct.state shows a large, stable negative effect. Higher local wages relative to the state average are associated with substantially lower odds of completing within the region (OR ≈ 0.10–0.11), and this effect holds steady across Models 2 and 3 rather than shifting — unlike the instability this variable showed in the DV2 sequence.

4.  Urbanicity (Dem_Desc) is a strong, stable positive predictor. Both "Entirely urban" (OR = 1.802) and "Urban/town/rural mix" (OR = 1.248) categories show meaningfully higher odds of completing within the region relative to the reference category, strengthening from Model 2 to Model 3.

5.  SpecialEdStatus and english.learner both attenuate substantially toward null once HS accomplishments are added. SpecialEdStatus goes from a significant positive effect in Models 1–2 (OR ≈ 1.54–1.55) to non-significant (OR = 1.046) in Model 3; english.learner attenuates from significant (OR = 1.632) to marginal (OR = 1.139, p = 0.066) — suggesting these early demographic associations run largely through HS accomplishment/participation variables.

6.  economic.status remains positive and significant throughout, though it attenuates somewhat once HS accomplishments enter (OR = 1.714 → 1.699 → 1.390 across the three models) — economically disadvantaged completers consistently show higher odds of completing within the region.

7.  Race/ethnicity patterns are more mixed than in other DV sequences. RaceEthnicityHispanic is the only category showing consistently higher odds relative to White completers (OR ≈ 1.34–1.35 throughout); RaceEthnicityAI remains significant and negative; RaceEthnicityBlack is non-significant in all three models; RaceEthnicityAsian/PI attenuates fully to null by Model 3.

**Diagnostics & Data Quality Notes**

VIF: `took.ACT` and `ACTCompositeScore.filled` show elevated VIFs (≈5.5–6.4) in Model 3 — expected and structural, consistent with the pattern in DV1/DV2/DV3. All other variables remain within acceptable VIF range throughout.

edrEDR 10 - Southeast does not reach significance in any DV4 model, consistent with the pattern seen in DV3 — region alone shows no independent association with either postsecondary completion or completion location once other variables are in the model.

total.cte.courses.taken and pseo.participant are both non-significant in the final model — the CTE courses variable has now failed to reach significance in DV3 and DV4, while holding up in DV1 and DV2.

**Open Items**

No further specification (interactions, mediation checks) has been run yet.

<br>

# Prep DV5 Data

Another post-secondary independent variable that was strongly associated with workforce participation outcomes was the highest credential earned by an individual. This analysis will explore what characteristics are associated with earned credentials.

```{r}
model_vars_dv5 <- c("Gender", 
                    "RaceEthnicity", 
                    "economic.status", 
                    "SpecialEdStatus",
                    "english.learner", 
                    "edr",
                    "Dem_Desc",
                    "avg.wages.pct.state", 
                    "avg.unemp.rate",
                    "pseo.participant", 
                    "took.ACT", 
                    "ACTCompositeScore", 
                    "sat.taken", 
                    "ap.exam",
                    "cte.achievement",
                    "total.cte.courses.taken",
                    "highest.cred.level")

master.model.dv5 <- master %>%
  filter(grad.year > 2010,
         grad.year < 2020,
         ps.grad == "Y",
         ps.grad != "Attending ps",
         ps.grad.location != "Did not graduate ps")%>% 
  select(PersonID, all_of(model_vars_dv5)) %>%
  filter(!is.na(pseo.participant),
         RaceEthnicity != "Unknown") %>%
  droplevels() %>%
  mutate(highest.cred.level = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Associate or lesser credential", "Bachelor or higher degree"),
         highest.cred.level = ifelse(highest.cred.level == "Associate or lesser credential", 1, 0),
         Gender = fct_relevel(Gender, "M"),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         economic.status = as_factor(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         english.learner = as_factor(english.learner),
         english.learner = fct_relevel(english.learner, "0"),
         edr = fct_relevel(edr, "EDR 9 - South Central"),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix"),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         took.ACT = fct_relevel(took.ACT, "No"),
         ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
         sat.taken = as_factor(sat.taken),
         sat.taken = fct_relevel(sat.taken, "0"),
         ap.exam = as_factor(ap.exam),
         ap.exam = fct_relevel(ap.exam, "0"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE")) 

```

# Modeling: Credential Earned - college completers only

## Model 1: Demographics

```{r}
model1_dv5 <- glm(
  highest.cred.level ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.model.dv5
)

summary(model1_dv5)
nobs(model1_dv5)
AIC(model1_dv5)

exp(cbind(OR = coef(model1_dv5), confint(model1_dv5)))

1 - (model1_dv5$deviance / model1_dv5$null.deviance)

vif(model1_dv5)

or_plot <- tidy(model1_dv5, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 2: Demographics + High School Characteristics

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; edr = EDR 9; Dem_Desc = reference level (not shown)

Sample: n = 33,816

**Model comparison (vs. Model 1)**

Likelihood ratio test: χ²(5) = 218.86, p \< 0.001 — Model 2 is a statistically significant improvement over Model 1. AIC improved from 38,717 → 38,509. Pseudo-R² increased from 0.0540 → 0.0593.

**Demographic coefficients (carried from Model 1)**

Direction, magnitude, and significance of Gender, economic.status, SpecialEdStatus, and english.learner are essentially unchanged from Model 1. **RaceEthnicityHispanic** attenuates from significant (Model 1, OR = 1.170, p = 0.031) to marginal (OR = 1.143, 95% CI: 0.989–1.319, p = 0.069). **RaceEthnicityBlack** attenuates somewhat, from OR = 0.672 (Model 1) to OR = 0.763 (95% CI: 0.644–0.902, p = 0.002) — still significant and negative, but a smaller gap. **RaceEthnicityAI** remains non-significant, as in Model 1.

**New: High school characteristics**

- **edrEDR 10 - Southeast** (OR = 0.872, 95% CI: 0.824–0.923, p \< 0.001): Completers from EDR 10 have 12.8% lower odds of sub-baccalaureate attainment (i.e., higher odds of bachelor's or higher) compared to EDR 9 completers.
- **Dem_DescEntirely urban** (OR = 0.680, 95% CI: 0.628–0.735, p \< 0.001): Completers from entirely urban high schools have 32.0% lower odds of sub-baccalaureate attainment relative to the reference Dem_Desc category.
- **Dem_DescUrban/town/rural mix** (OR = 0.967, 95% CI: 0.904–1.034, p = 0.319): Not statistically significant.
- **avg.wages.pct.state** (OR = 0.790, 95% CI: 0.544–1.147, p = 0.214): Not statistically significant.
- **avg.unemp.rate** (OR = 0.977, 95% CI: 0.958–0.995, p = 0.015): Higher local unemployment associated with modestly lower odds of sub-baccalaureate attainment.

**Model fit**

VIF: All values remain low (max 1.68 for RaceEthnicity, 1.59 for english.learner) — no multicollinearity concerns.

```{r}
model2_dv5 <- glm(
  highest.cred.level ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    edr +
    Dem_Desc +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.dv5
)

summary(model2_dv5)

nobs(model2_dv5)

AIC(model2_dv5)

exp(cbind(OR = coef(model2_dv5), confint(model2_dv5)))

anova(model1_dv5, model2_dv5, test = "Chisq")

1 - (model2_dv5$deviance / model2_dv5$null.deviance)

vif(model2_dv5)


or_plot <- tidy(model2_dv5, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## Model 3: Demographics + High School Characteristics + High School Accomplishments

**DV5 — Model 3: + High School Accomplishments** **Sub-Baccalaureate Credential Attainment, Completers Only**

Reference categories: RaceEthnicity = White; Gender = Male; economic.status = 0; SpecialEdStatus = 0; english.learner = 0; Dem_Desc = reference level (not shown); edr = EDR 9; pseo.participant = 0; took.ACT = No; sat.taken = 0; ap.exam = 0; cte.achievement = reference level (not shown)

Sample: n = 33,816

**Model comparison (vs. Model 2)**

Likelihood ratio test: χ²(8) = 6,408.9, p \< 0.001 — Model 3 is a very large, statistically significant improvement over Model 2. AIC improved substantially: 38,509 → 32,116. Pseudo-R² increased from 0.0593 → 0.2160 — the second-largest single-block jump seen across all five DV sequences (after DV3's Model 3).

**Notable shifts from Model 2**

- **GenderF**: strengthens from OR = 0.707 (Model 2) to OR = 0.811 (95% CI: 0.767–0.859, p \< 0.001) — direction unchanged, but the gap narrows somewhat.
- **RaceEthnicityAsian/PI**: attenuates from OR = 0.534 (Model 2) to OR = 0.753 (95% CI: 0.623–0.907, p = 0.003) — gap narrows substantially.
- **RaceEthnicityBlack**: attenuates only slightly (OR = 0.763 → 0.573, 95% CI: 0.477–0.687, p \< 0.001) — gap widens here rather than narrowing.
- **RaceEthnicityHispanic**: fully attenuates to null (OR = 1.068, 95% CI: 0.910–1.252, p = 0.423), from marginal in Model 2 (p = 0.069).
- **economic.status1**: attenuates from OR = 2.368 (Model 2) to OR = 1.738 (95% CI: 1.619–1.866, p \< 0.001) — still large and significant, but notably smaller.
- **SpecialEdStatus1**: attenuates substantially from OR = 4.191 (Model 2) to OR = 1.806 (95% CI: 1.581–2.065, p \< 0.001) — still a large positive effect, but roughly cut in half.
- **english.learner1**: fully attenuates to null (OR = 1.046, 95% CI: 0.894–1.224, p = 0.575), from significant positive in Models 1–2 (OR ≈ 1.88–1.91).
- **Dem_DescEntirely urban** (OR = 0.816, 95% CI: 0.746–0.891, p \< 0.001): remains significant and negative, similar to Model 2.
- **Dem_DescUrban/town/rural mix** (OR = 1.079, 95% CI: 1.001–1.163, p = 0.048): newly significant here, was non-significant in Model 2.
- **edrEDR 10 - Southeast** (OR = 0.897, 95% CI: 0.842–0.956, p \< 0.001): remains significant and negative, consistent with Model 2.
- **avg.wages.pct.state**: remains non-significant.
- **avg.unemp.rate** (OR = 0.894, 95% CI: 0.874–0.914, p \< 0.001): strengthens from Model 2 (OR = 0.977).

**New: High school accomplishments**

- **pseo.participant1** (OR = 0.696, 95% CI: 0.656–0.738, p \< 0.001): PSEO participants have lower odds of sub-baccalaureate attainment (i.e., higher odds of bachelor's or higher).
- **took.ACTYes** (OR = 6.587, 95% CI: 5.429–7.998, p \< 0.001): Very large positive effect — ACT test-takers have substantially higher odds of sub-baccalaureate attainment relative to non-takers.
- **ACTCompositeScore.filled** (OR = 0.857 per point, 95% CI: 0.850–0.864, p \< 0.001): Among the achievement gradient, higher composite scores associated with lower odds of sub-baccalaureate attainment (i.e., higher odds of bachelor's or higher) — the steepest per-point ACT gradient seen across any DV sequence.
- **ap.exam1** (OR = 0.519, 95% CI: 0.484–0.556, p \< 0.001): AP exam participants have substantially lower odds of sub-baccalaureate attainment.
- **sat.taken1** (OR = 0.617, 95% CI: 0.494–0.766, p \< 0.001): SAT takers have lower odds of sub-baccalaureate attainment.
- **cte.achievementCTE Concentrator or Completor** (OR = 1.144, 95% CI: 1.045–1.253, p = 0.004): significant and positive.
- **cte.achievementCTE Participant** (OR = 1.014, 95% CI: 0.941–1.092, p = 0.717): Not statistically significant.
- **total.cte.courses.taken** (OR = 1.063 per course, 95% CI: 1.051–1.076, p \< 0.001): significant and positive — the strongest total.cte.courses.taken effect seen across any DV sequence.

**Model fit**

VIF: `took.ACT` (6.69) and `ACTCompositeScore.filled` (7.16) remain elevated for the expected structural reason. All other variables remain below conventional concern thresholds.

```{r}
model3_dv5 <- glm(
  highest.cred.level ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam + 
    sat.taken +
    cte.achievement +
    total.cte.courses.taken,
    family = binomial,
  data = master.model.dv5
)

summary(model3_dv5)

nobs(model3_dv5)

AIC(model3_dv5)

exp(cbind(OR = coef(model3_dv5), confint(model3_dv5)))

anova(model1_dv5, model2_dv5, model3_dv5, test = "Chisq")

1 - (model3_dv5$deviance / model3_dv5$null.deviance)

vif(model3_dv5)


or_plot <- tidy(model3_dv5, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.M" = "MCA Math",
      "highest.cred.level" = "Highest credential"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )


ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )
```

<br>

## DV5 Summary: What Predicts Sub-Baccalaureate Credential Attainment (College Completers Only)

**Overview**

Dependent variable: `highest.cred.level` (binary — 1 = associate degree or less-than-associate credential, 0 = bachelor's degree or higher) Population: Postsecondary completers only (2011–2019 graduating cohorts), n = 33,816 Purpose: Confirmatory multivariate testing of associations identified in earlier bivariate analysis; not causal or predictive.

**Model Sequence**

Model 1 (Demographics): AIC 38,717 \| Pseudo-R² 0.0540 \| — (baseline) Model 2 (+ HS characteristics): AIC 38,509 \| Pseudo-R² 0.0593 \| χ²(5) = 218.86, p \< 0.001 Model 3 (+ HS accomplishments): AIC 32,116 \| Pseudo-R² 0.2160 \| χ²(8) = 6,408.9, p \< 0.001

This completes the DV5 model sequence per the methodology (Demographics → HS characteristics → HS accomplishments, with no further blocks specified for this outcome).

**Variables Retained (Final Model 3)**

Demographics: Gender, RaceEthnicity, economic.status, SpecialEdStatus, english.learner HS characteristics: Dem_Desc, edr, avg.wages.pct.state, avg.unemp.rate HS accomplishments: pseo.participant, took.ACT, ACTCompositeScore.filled, ap.exam, sat.taken, cte.achievement, total.cte.courses.taken

No variables were dropped across the DV5 sequence.

**Headline Findings**

1.  Academic-orientation variables show their steepest associations of any DV sequence here. ACT composite score (OR = 0.857 per point), AP exam participation (OR = 0.519), and SAT-taking (OR = 0.617) are all more strongly negative predictors of sub-baccalaureate attainment than the equivalent coefficients in DV1, DV2, or DV4 — suggesting academic orientation is even more tightly linked to credential level than to regional attachment or completion location.

2.  ACT participation remains a very large positive predictor (OR = 6.587) — consistent with the "two forces" pattern (participation positive, score negative) seen throughout DV1, DV2, and DV4.

3.  total.cte.courses.taken shows its strongest effect of any DV sequence (OR = 1.063 per course), while the CTE Participant category loses significance — suggesting course dosage may matter more than category threshold for this particular outcome, a pattern not seen elsewhere.

4.  SpecialEdStatus and economic.status remain the largest demographic effects, though both attenuate substantially once HS accomplishments enter. SpecialEdStatus drops from OR = 4.191 (Model 2) to OR = 1.806 (Model 3) — still large, but roughly half its earlier magnitude; economic.status drops from OR = 2.368 to OR = 1.738.

5.  english.learner and RaceEthnicityHispanic both fully attenuate to null by Model 3, while RaceEthnicityBlack's gap widens slightly rather than narrowing — a mixed attenuation pattern, unlike the more uniform narrowing seen in some other DV sequences.

6.  Urbanicity and region both show negative, stable associations with sub-baccalaureate attainment. Entirely-urban high schools (OR = 0.816) and EDR 10 (OR = 0.897) are both associated with lower odds of sub-baccalaureate attainment (i.e., higher odds of bachelor's or higher) — this is the opposite direction from DV4, where urbanicity predicted staying in-region for postsecondary study generally.

7.  avg.unemp.rate strengthens into a clear negative predictor by Model 3 (OR = 0.894) — higher local unemployment associated with lower odds of sub-baccalaureate attainment (i.e., higher odds of bachelor's or higher), a directionally counterintuitive pattern worth noting alongside similar unemployment-rate findings in other DVs.

**Diagnostics & Data Quality Notes**

VIF: `took.ACT` and `ACTCompositeScore.filled` show elevated VIFs (≈6.7–7.2) in Model 3 — expected and structural, consistent with the pattern across all five DV sequences. All other variables remain within acceptable VIF range throughout.

avg.wages.pct.state is non-significant in both Model 2 and Model 3 — the only DV sequence where this variable never reaches significance at any stage.

cte.achievement (Participant category) loses significance in Model 3 while total.cte.courses.taken strengthens — the reverse of the general CTE pattern seen in DV1/DV2, where both CTE terms held up together.

**Open Items**

No further specification (interactions, mediation checks) has been run yet.

<br>

**Executive Summary: Multivariate Confirmation Analysis of Workforce Participation Pathways** **Southeast Minnesota (EDR 9 and 10)**

**Purpose and Approach**

This analysis used sequential block-entry logistic regression to test whether the strongest relationships identified in earlier bivariate analysis — particularly the role of postsecondary credential attainment, credential level, and location of completion — remain directionally consistent once considered simultaneously with demographic, high school context, and accomplishment variables. Five outcomes were modeled across the 2011–2019 graduating cohorts from Southeast Minnesota (EDR 9 and 10): sustained regional workforce attachment (both for the full cohort and for college completers only), postsecondary completion, completion location, and credential level. Consistent with the confirmatory (not causal, not predictive) framing of this project, findings are described throughout as associations holding other variables constant, not as causal effects.

**The Central Story: Two Divergent Pathways**

A college-preparatory, mobility-oriented pathway — marked by higher ACT composite scores (conditional on taking the test), AP exam participation, and SAT participation — consistently predicts completing a postsecondary credential (DV3), completing a bachelor's degree rather than an associate's or less (DV5), and completing that credential outside the region (DV4). Among people who complete a credential, this same pathway predicts lower odds of sustained regional employment (DV1/DV2). These variables mark students headed toward more advanced credentials and greater geographic mobility, with regional workforce attachment as the trade-off.

A CTE-anchored, regionally-oriented pathway runs largely in the opposite direction. CTE engagement (achievement level and course count) showed a modest but real positive relationship with completing a postsecondary credential at all (DV3) — smaller in magnitude than most other accomplishment effects, but statistically significant, not absent. It consistently and more strongly predicted completing that credential locally rather than out-of-region (DV4), and predicted sustained regional employment among both the full cohort and completers (DV1/DV2). For credential level (DV5), CTE engagement (particularly total courses taken, its strongest effect of any outcome tested) predicted sub-baccalaureate rather than bachelor's attainment — placing CTE consistently on the regionally-anchored, sub-baccalaureate side of every outcome where it appears.

PSEO participation tells a distinct story from CTE rather than a parallel one. It was the single strongest predictor of postsecondary completion found anywhere in the project (nearly double the odds, DV3), and it predicted bachelor's completion over sub-baccalaureate attainment in DV5 — aligning it with the mobility-track variables on that outcome. But PSEO showed no reliable independent relationship with regional employment attachment once other accomplishment variables were controlled (DV1's Model 3 effect fully attenuated to null once postsecondary completion itself entered; DV2 and DV4 showed no significant PSEO effect at any stage). In this data, PSEO functions primarily as a general attainment accelerant leaning toward bachelor's completion, rather than as a regionally-anchoring variable in its own right.

A recurring ACT-specific nuance: `took.ACT` and ACT composite score point in opposite directions once combined in every model where both appear. Taking the test at all is associated with one direction, but each additional point pulls back the other way — meaning a low scorer who took the test can look more "regionally anchored" or more likely to complete than someone who never tested, while a high scorer looks considerably more academically oriented and mobile. This threshold effect should be described explicitly rather than reported as a single flat "ACT participation" finding.

**The Most Important Unresolved Finding: The Black-White Gap**

Across three of the five outcomes — sustained regional workforce attachment for the full cohort (DV1), for completers only (DV2), and sub-baccalaureate credential attainment (DV5) — the disparity between Black and White graduates did not narrow once accomplishment variables (PSEO, ACT, AP, SAT, CTE) were controlled. In DV1 and DV2 the gap remained essentially stable across all model blocks; in DV5 it widened once accomplishments were added (OR moving from 0.763 to 0.573 between Models 2 and 3). No accomplishment variable, credential level, or completion location tested anywhere in this project explains this gap.

This same disparity behaved differently on the other two outcomes: it narrowed substantially once accomplishments were controlled for whether someone completes a postsecondary credential at all (DV3, OR moving from roughly 0.71 to 0.88), and it was non-significant at every stage for completion location (DV4). Taken together, this suggests Black graduates' rates of completing a credential are reasonably well explained by measured differences in accomplishment and preparation access — but a real, unexplained disparity remains specifically in long-term regional employment attachment and in credential level, independent of everything else tested. This is the most important equity finding to lead with in any final report, precisely because it is not explained by the very access-oriented variables that explain most other disparities in this project.

American Indian graduates showed a related but distinct pattern: a persistent, largely stable negative gap across DV1, DV2, DV3, and DV4, but no significant relationship at all with credential level (DV5) — suggesting the AI disparity operates on whether and where people work and complete credentials, but not on which credential level they reach.

**Demographic Disadvantage: Two Different Underlying Mechanisms**

Economic status, SpecialEdStatus, and English learner status showed two distinct structural patterns depending on the outcome:

A suppression pattern, seen in DV1 (full-cohort employment attachment): economic.status was null-to-positive in the demographics-only baseline, turned significantly negative once HS accomplishments were added, and then attenuated fully to null once postsecondary completion itself entered the model. This indicates the early, more optimistic reading was an artifact of omitted variables rather than the true underlying relationship.

An access pattern, seen in DV3 and DV5 (completion and credential level): economic.status and SpecialEdStatus both showed clear disadvantage immediately in the demographics-only baseline, and that disadvantage shrank substantially — without fully disappearing — once accomplishment variables entered. This is consistent with a real share of each gap running through unequal access to PSEO, ACT, AP, and CTE opportunities. SpecialEdStatus showed the single largest demographic effect size found anywhere in the project (DV5 Model 1: more than four times the odds of sub-baccalaureate attainment), attenuating to roughly 1.8x by the final model.

English learner status followed a more outcome-dependent pattern than in a typical access story: it attenuated to full non-significance in DV5 (a clean access-pattern result) and reversed sign entirely in DV3 (negative in the demographics-only model, positive once accomplishments were controlled). It also showed positive associations in DV2 and DV4 at the demographics stage that attenuated toward null as accomplishments entered — the opposite starting point from the typical "early disadvantage" pattern, worth treating as its own distinct finding rather than folding into either mechanism above.

**Gender: Different Stories on Different Outcomes**

Female graduates were consistently more likely to complete a postsecondary credential (DV3) and, to an attenuated but still significant degree, more likely to complete a bachelor's rather than a sub-baccalaureate credential (DV5). For regional employment attachment, gender behaved differently depending on the population: among the full cohort (DV1), an early negative association for women fully attenuated to null once HS accomplishments were controlled. Among completers specifically (DV2), gender moved in the opposite direction — a null association in the demographics-only model became a significant positive association for women once HS accomplishments entered, and this held through the final model. These are not contradictory findings; they reflect different mechanisms across different populations and outcomes, and should be reported as distinct results rather than reconciled into one general "gender effect."

**The Single Strongest Lever Identified: Location of Postsecondary Completion**

Among every variable tested across all five outcomes, location of postsecondary completion (DV2, Model 5) produced the largest effect size and the largest single-block improvement in model fit found anywhere in the project. Completing a credential outside the region was associated with roughly a 69% reduction in the odds of sustained regional employment — exceeding the effect of credential level itself (roughly a 19–37% reduction for bachelor's vs. sub-baccalaureate, depending on model specification), though both were independently significant and not fully redundant with one another. If this project's findings point to one clear, actionable lever for regional workforce policy, it is this one: where people complete their postsecondary education matters more than almost anything else tested for whether they end up working in the region afterward.

**Regional Geography (EDR): Consistently Present, Modestly Shaped by Outcome**

EDR 10 (relative to EDR 9) showed a stable, significant positive association with regional employment attachment across DV1 and DV2, and a significant negative association with sub-baccalaureate attainment in DV5 (i.e., EDR 10 graduates were more likely to complete a bachelor's than an associate's or less). It was not a significant predictor of postsecondary completion (DV3) or completion location (DV4) at any stage — the only two outcomes where region showed no independent association once other variables were included.

**Outstanding Technical Item**

`avg.wages.pct.state` reached significance in most models across four of the five DVs (DV1 marginal only, DV2/DV3/DV4 significant, DV5 never significant), but its odds ratio is not directly interpretable as computed — it reflects a full 1-unit (100 percentage point) change on a variable that realistically ranges only about 0.7–1.0. This should be rescaled (e.g., per 0.10 increment or per standard deviation) before any odds ratio involving this variable appears in a final report table. The direction and significance of the findings involving this variable are sound; only the magnitude as currently expressed needs correction. It's also worth noting this variable showed unusual instability across blocks in DV2 specifically (non-significant → significant → growing substantially larger), which may warrant a closer look before being treated as a fully stable finding there.

**Bottom Line**

Two coherent, largely opposing pathways run through this analysis: a college-preparatory track associated with completing more advanced credentials and greater geographic mobility, and a CTE-anchored track associated with regional employment and regional completion, with PSEO functioning as a distinct attainment-accelerant that leans toward the mobility track rather than fitting neatly into either pathway. Layered on top of this structural finding are two distinct patterns of demographic disparity — one where disadvantage is suppressed until accomplishments are controlled, another where disadvantage is visible early and partly explained by unequal access to those same accomplishments — and, cutting across three of the five outcomes examined, a Black-White gap in long-term employment attachment and credential level that neither pattern explains. Of every lever tested, location of postsecondary completion stands out as the one with by far the largest, most policy-relevant effect on whether graduates remain in the regional workforce.
