---
title: "Local Employment Cross-tabs"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

master.factor.names <- master.original %>%
  select(PersonID, Gender, economic.status, SpecialEdStatus, RaceEthnicity, LimitedEnglishProficiencyIndicator, english.learner, non.english.home, Dem_Desc, edr, pseo.participant, took.ACT, ap.exam, cte.achievement, attended.ps.within.first.year.hsgrad, attended.ps, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location) 

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states)

master <- master.states %>%
  left_join(master.factor.names, by = "PersonID")

kable(head(master))

kable(names(master))
```

```{r independent and dependent variables}
updated.factor.var.names <-  master.factor.names %>%
  select(-PersonID) %>%
  names() %>%
  c("ps.grad.in.same.ruca", "ps.grad.in.same.edr", "ps.grad.in.same.pr", "ps.grad.in.MN")

numeric.var.names <- master.original %>%
  select(ACTCompositeScore, total.cte.courses.taken, `ACTE.SPED`:Youth.Apprenticeship,  MCA.M:MCA.S, n.institutions, avg.unemp.rate, wages.3year.avg, avg.wages.pct.state) %>%
  names()
  
updated.independent.var.names <- c(updated.factor.var.names, numeric.var.names)

cte.cc.var.names <- master.original %>%
  select(ACTE.SPED:Youth.Apprenticeship) %>%
  names()

dep.var.names <- master %>%
  select(PersonID, grad.year.1, grad.year.5, grad.year.10)

updated.master <- master.original %>%
  select(PersonID, all_of(updated.independent.var.names)) %>%
  left_join(dep.var.names, by = "PersonID") %>%
  mutate_at(vars(Dem_Desc:ps.grad.in.MN), as.factor) %>%
  mutate_at(vars(grad.year.1:grad.year.10), as.factor) %>%
  mutate(grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Did not grad", "Some college", as.character(highest.cred.level)),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Never attended ps", "No college", as.character(highest.cred.level)),
         highest.cred.level = as.factor(highest.cred.level),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix", "Urban/town/rural mix", "Entirely urban")) %>%
  droplevels()

kable(head(updated.master))

kable(names(updated.master))
```

<br>

There are a number of "states" in which an individual in this dataset are assigned;

1.  Has meaningful employment
    1.  In the same county as their high school.
    2.  In the same EDR as their high school, but not county.
    3.  In Minnesota, but not in Southern Minnesota.
2.  Is attending post-secondary.
3.  Has a MN employment record but it's not meaningful, and not attending post-secondary.
4.  Does not have a MN employment record and is not attending post-secondary.

There are three "times" in which we checked to see if the individual had meaningful employment within these geographies;

1.  one year after graduating high school
2.  five years after graduation, and
3.  ten years after graduation.

Meaningful employment is determined by whether an individual worked 1,000 hours for an employer during time X. In addition, if the individual is working for an employer but not meaningful at time X but worked 1,000 hours for that employer during another year it's still considered "meaningful".

Due to time x potentially being passed the date of the latest data (2023) for some individuals, the analysis below will filter out all individuals where time x is after 2023.

From the multiple correspondence analysis and the principle component analysis we determined that the following independent variables will be included in the analysis.

```{r list of ind var}
updated.master %>%
  select(-PersonID, -grad.year.1, -grad.year.5, -grad.year.10) %>%
  lapply(class) 

```

# Methodology

For this analysis we are keeping it simple. For categorical variables I’ll be using the Cramer’s v-square measure of association. The V\^2 value is from 0 to 1 with 0 meaning there’s no association and a value closer to 1 indicates a strong association.

| **Cramér’s V Value** | **Interpretation**               |
|----------------------|----------------------------------|
| 0.00–0.10            | Negligible association           |
| 0.10–0.20            | Weak                             |
| 0.20–0.40            | Moderate                         |
| 0.40–0.60            | Relatively strong                |
| 0.60–0.80            | Strong                           |
| 0.80–1.00            | Very strong (possibly redundant) |

For the variables with continuous values I’ll be using the Wilcoxon two paired groups test. P-value will be used to determine whether we can reject the NULL hypothesis at a .01 level. I’ll also include boxplots to see in which direction there might be influence from the variables.

We are going to split up the analysis into two different datasets. The first dataset will combine 3 of the previous states into one.

- Has meaningful employment anywhere in Southeast Minnesota, which combines the following states;
  - Meaningful emp County
  - Meaningful emp EDR
  - Meaningful emp region

The remaining states will stay the same.

- Has meaningful employment in Minnesota, but not in Southern Minnesota
- Is attending a post-secondary institution.
- Has a MN employment record but it's not meaningful, and they are not attending post-secondary.
- Does not have a MN employment record and is not attending post-secondary.

We have quite a few variables remaining, even after MCA analysis. So let's take a look at how we are going to proceed and group all of this together to see how these relationships work.

- Demographics
  - `Gender`
  - `economic.status`
  - `SpecialEdStatus`
  - `RaceEthnicity`
  - `LimitedEnglishProficiencyIndicator`
  - `english.learner`
  - `non.english.home`
- High school characteristics
  - `Dem_Desc`
  - `edr`
  - `avg.unemp.rate`
  - `avg.wages.pct.state`
- High school accomplishments
  - `pseo.participant`
  - `took.ACT`
  - `ACTCompositeScore`
  - `ap.exam`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`
- Post-secondary
  - `attended.ps.within.first.year.hsgrad`
  - `attended.ps`
  - `ps.grad.`
  - `ps.grad.InstitutionSector`
  - `highest.cred.level`
  - `ps.grad.location`

<br>

# Meaningful employment in Southern Minnesota

```{r prep meaningful employment in Southern MN}
meaningful.emp.southern <- updated.master %>%
  mutate(grad.year.1 = ifelse(grad.year.1 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.1)),
         grad.year.1 = as_factor(grad.year.1),
         grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.5)),
         grad.year.5 = as_factor(grad.year.5),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.10 = ifelse(grad.year.10 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.10)),
         grad.year.10 = as_factor(grad.year.10),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         highest.cred.level = fct_relevel(highest.cred.level, "No college", "Some college", "Less than Bachelors", "Bachelors or higher"))

write_csv(meaningful.emp.southern, "Data/SLEDS/Masters/After analysis/Master-meaningful-emp-Southern.csv")
```

**WORKFORCE PARTICIPATION OUTCOMES — SOUTHEAST MINNESOTA**

The chart below provides the percentage of students that fall into each workforce participation category at 1, 5, and 10-years after graduating high school.

At Year 1, the distribution is heavily shaped by postsecondary enrollment. 54.4% are attending postsecondary, 14.0% are in meaningful regional employment, 4.6% in meaningful MN employment, 16.8% in non-meaningful employment, and 10.2% have no MN employment record.

By Year 5, postsecondary attendance has dropped sharply to 18.5% as graduates complete or exit credentials. Meaningful regional employment rises to 27.2% and meaningful MN employment to 16.3%. Non-meaningful employment holds at 17.7% and those with no MN record rise modestly to 20.4%.

By Year 10, postsecondary attendance is essentially zero (n=1). Meaningful regional employment reaches 29.7% and meaningful MN employment 21.1%. However, the most notable feature is the sharp rise in no MN employment record to 33.4%, while non-meaningful employment declines to 15.8%.

The large and growing no-MN-employment-record group at Year 10 is the most important pattern to flag. This likely reflects a combination of out-of-state migration, self-employment, and genuine labor force disconnection — and will be a central interpretive challenge throughout the analysis.

```{r proportion of states meaningful emp SW}
prop.states.meaningful.emp.southern <- meaningful.emp.southern %>%
  select(grad.year.1, grad.year.5, grad.year.10) %>%
  pivot_longer(names_to = "grad.year", values_to = "states", 1:3) %>%
  group_by(grad.year, states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  filter(states != "After 2023") %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()),
         grad.year = fct_relevel(grad.year, "grad.year.1", "grad.year.5", "grad.year.10")) %>%
  write_csv(., "Data/Claude/analysis_local_employment_ct/workforce-state-freq.csv")

prop.states.meaningful.emp.southern.plot <- ggplot(prop.states.meaningful.emp.southern, aes(grad.year, pct, fill = states, group = states)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nState of individuals: ", states, "\nNumber of people in states: ", comma(n), "\nPercent of people in state: ", percent(pct, accuracy = .1), sep = ""))) +
    geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
   labs(x="", y = "", color="", title = "Percent of people in each state")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu"),
                    guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = prop.states.meaningful.emp.southern.plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

Let's examine the variables to see if there are any significant differences in the proportion of individuals in these states across the variable categories or values.

<br>

# Demographics

In the demographics section we will examine the relationship between that paths taken by individuals and the following variables;

- Demographics
  - `Gender`
  - `economic.status`
  - `SpecialEdStatus`
  - `RaceEthnicity`
  - `LimitedEnglishProficiencyIndicator`
  - `english.learner`
  - `non.english.home`

## Summary

The demographic variables as a group show moderate associations with workforce outcomes, with effect sizes generally strongest at Year 1 — where postsecondary attendance patterns dominate the distribution — and weakening over time as graduates sort into more stable labor market positions. The three language-related variables show a consistent and notable pattern across all time points: graduates from non-English language backgrounds show elevated meaningful MN employment at Year 1, likely reflecting concentration in specific industries or urban labor markets outside the Southeast region rather than broad workforce advantage.

**RANKING BY STRENGTH OF ASSOCIATION (Cramer's V at Year 5 as primary sort, with Year 1 and Year 10 as tiebreakers)**

1.  SpecialEdStatus (V=0.274 / 0.145 / 0.086) Special education status is the strongest demographic predictor at Year 1 and remains the most consistently differentiated variable across all three time points. The postsecondary attendance gap is the largest of any demographic variable, and the non-meaningful employment disadvantage persists through Year 10. Notably, regional employment reaches full parity by Year 10, suggesting that special education graduates who connect to the regional workforce do so successfully — but a larger share remains in non-meaningful or marginal employment throughout the observation window.

2.  economic.status (V=0.260 / 0.121 / 0.083) Economic disadvantage is the second strongest predictor and shows the most complete convergence over time of any variable in this group. The large Year 1 postsecondary gap closes substantially by Year 5, and meaningful regional employment reaches near parity by Year 5 and full parity by Year 10. The most durable disadvantage is in meaningful MN employment, pointing to differential access to college-driven career pathways rather than a broad regional workforce deficit.

3.  english.learner (V=0.164 / 0.078 / 0.059) English learner status is the strongest of the three language variables and the primary language indicator for multivariate modeling. The Year 1 postsecondary gap and elevated no-MN-record rate are the defining features, with gaps persisting — though narrowing — through Year 10. The elevated meaningful MN employment at Year 1 is an analytically interesting anomaly shared across all three language variables.

4.  Gender (V=0.112 / 0.058 / 0.048) Gender shows a moderate Year 1 association driven primarily by differential postsecondary attendance rather than direct workforce sorting. Males enter regional employment earlier; females attend postsecondary at higher rates and enter the labor market later. By Year 10, differences are small and consistent — males retain a modest advantage in both meaningful regional and MN employment, while females show higher non-meaningful employment rates.

5.  non.english.home (V=0.110 / 0.054 / 0.060) Non-english home largely mirrors english.learner with consistently weaker effect sizes. It is retained for descriptive bivariate purposes but excluded from multivariate models. The notable absence of a meaningful regional employment gap at Year 1 — unlike english.learner — suggests the two variables capture partially distinct subpopulations despite their conceptual overlap.

6.  RaceEthnicity (V=0.073 / 0.044 / 0.041) Race and ethnicity shows weak overall associations but contains substantively important within-group variation. The Year 10 no-MN-record rates for Asian/PI (45.1%), American Indian (41.1%), and Black (39.7%) graduates are the most striking findings and likely reflect differential out-of-state migration following postsecondary completion rather than labor force disconnection. Hispanic graduates show a distinctive pattern of relatively strong regional employment at Years 5 and 10 despite lower postsecondary attendance at Year 1. AI findings should be interpreted with caution throughout given small cell sizes.

7.  LimitedEnglishProficiencyIndicator (V=0.079 / 0.032 / 0.031) The weakest demographic predictor and the most rapidly declining in association over time. Its patterns are broadly consistent with english.learner but substantially attenuated, confirming the MCA finding that it occupies a secondary position relative to the other language variables. It adds limited analytical value beyond what english.learner already captures and should be treated as supplementary throughout the remaining stages.

**CROSS-CUTTING THEMES**

Two disconnection signatures are visible across the demographic group. The local underutilization signature — present in the labor market but below the sustained threshold — is most pronounced for special education and low-income graduates and persists through Year 10. The geographic absence signature — no MN employment record — is most elevated for language minority and racial and ethnic minority graduates, particularly at Year 10, and is more likely to reflect out-of-state migration than genuine disconnection for those groups. These two signatures track different populations and will require differentiated interpretation throughout the remaining analytical stages.

<br>

## `Gender`

At Year 1, gender shows its strongest association with workforce outcomes (Cramer's V=0.112), driven primarily by differences in postsecondary attendance and early employment. Females attend postsecondary at a notably higher rate (59.4%) than males (49.4%), while males are more likely to be in meaningful regional employment (16.0% vs. 12.0%) and substantially more likely to have no MN employment record (12.4% vs. 8.0%).

By Year 5, the association weakens considerably (V=0.058). The postsecondary gap narrows as females complete credentials and enter the labor market. Males remain slightly more likely to hold meaningful regional employment (28.3% vs. 26.0%), while females are more likely to still be attending postsecondary (20.5% vs. 16.6%). Females show a modestly higher rate of non-meaningful employment (18.2% vs. 17.1%).

By Year 10, the association is statistically significant but very weak (V=0.048). Males hold a slight advantage in meaningful regional employment (30.5% vs. 28.8%) and meaningful MN employment (22.2% vs. 20.1%), while females are more likely to be in non-meaningful employment (17.3% vs. 14.3%). The no-MN-employment-record group is similar across genders (33.9% female, 33.0% male).

Overall, gender is a modest predictor of workforce outcomes in Southeast Minnesota. The Year 1 association largely reflects differential postsecondary attendance patterns rather than direct workforce sorting. By Years 5 and 10 the gender gap narrows substantially, with males retaining a small but consistent advantage in sustained regional employment across all three time points.

::: panel-tabset
### One year after graduating

```{r demographics gender one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, Gender) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/gender-1.html")

```

### Five years after graduating

```{r demographics gender five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, Gender) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/gender-5.html")

```

### Ten years after graduating

```{r demographics gender ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, Gender) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/gender-10.html")

```
:::

<br>

## `economic.status`

Economic status is one of the stronger demographic predictors at Year 1 (V=0.260), driven by a stark postsecondary attendance gap. Non-low-income graduates attend postsecondary at nearly twice the rate of low-income graduates (62.8% vs. 35.0%). Low-income graduates are more likely to be in meaningful regional employment (18.3% vs. 12.2%), meaningful MN employment (6.8% vs. 3.7%), non-meaningful employment (25.4% vs. 13.0%), and without any MN employment record (14.4% vs. 8.3%).

By Year 5, the association weakens substantially (V=0.121). The postsecondary gap narrows as non-low-income graduates complete credentials and enter the labor market. Meaningful regional employment rates converge — low-income graduates actually show a slightly higher rate (28.1% vs. 26.8%). However, low-income graduates remain more likely to be in non-meaningful employment (22.6% vs. 15.6%) and less likely to be in meaningful MN employment (14.5% vs. 17.0%), suggesting lower rates of college-driven out-of-region employment.

By Year 10, the association weakens further (V=0.083) and meaningful regional employment reaches near parity (29.5% vs. 29.7%). The most persistent gap is in meaningful MN employment (17.4% vs. 22.5%) and non-meaningful employment (20.1% vs. 14.2%), with the no-MN-record group essentially equal across economic status categories (33.0% vs. 33.6%).

Overall, economic disadvantage is strongly associated with reduced postsecondary participation at Year 1 and elevated non-meaningful employment through Year 10. Notably, the regional employment gap essentially closes by Year 5, suggesting that low-income graduates who enter the workforce directly find regional footing at rates comparable to their non-low-income peers. The persistent gap in meaningful MN employment points to differential access to college-driven career pathways rather than a broad regional employment disadvantage.

::: panel-tabset
### One year after graduating

```{r demographics economic.status one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, economic.status) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/economic-status-1.html")
```

### Five years after graduating

```{r demographics economic.status five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, economic.status) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/economic-status-5.html")
```

### Ten years after graduating

```{r demographics economic.status ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, economic.status) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/economic-status-10.html")
```
:::

<br>

## `SpecialEdStatus`

Special education status is among the stronger demographic predictors at Year 1 (V=0.274). The most striking difference is in postsecondary attendance — non-special-ed graduates attend at nearly three times the rate of special education graduates (59.1% vs. 21.9%). Special education graduates are substantially more likely to be in non-meaningful employment (31.1% vs. 14.7%) and without any MN employment record (24.3% vs. 8.1%), while showing a modestly higher rate of meaningful regional employment (17.2% vs. 13.6%).

By Year 5, the association weakens but remains meaningful (V=0.145). Meaningful regional employment reaches near parity (27.1% vs. 27.2%), but special education graduates remain considerably more likely to be in non-meaningful employment (26.6% vs. 16.5%) and without a MN employment record (27.8% vs. 19.3%). The continued postsecondary attendance gap (7.5% vs. 20.0%) reflects lower credential attainment among special education graduates.

By Year 10, the association weakens further (V=0.086) and meaningful regional employment is identical across groups (29.7%). The persistent gaps are in meaningful MN employment (13.9% vs. 22.1%) and non-meaningful employment (22.5% vs. 14.9%), while the no-MN-record group is essentially equivalent (33.9% vs. 33.3%).

The pattern across time is notable. Special education graduates who enter the regional workforce do so at rates comparable to their peers by Year 5, and reach full parity in regional employment by Year 10. However, they show persistently lower rates of meaningful MN employment throughout the observation window, consistent with lower postsecondary credential attainment limiting access to college-driven out-of-region career pathways. The elevated non-meaningful employment share is the most durable disadvantage across all three time points.

::: panel-tabset
### One year after graduating

```{r demographics SpecialEdStatus one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, SpecialEdStatus) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/SpecialEdStatus-1.html")
```

### Five years after graduating

```{r demographics SpecialEdStatus five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, SpecialEdStatus) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/SpecialEdStatus-5.html")
```

### Ten years after graduating

```{r demographics SpecialEdStatus ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, SpecialEdStatus) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/SpecialEdStatus-10.html")
```
:::

<br>

## `RaceEthnicity`

Race and ethnicity shows modest overall associations across all three time points (V=0.073 at Year 1, V=0.044 at Year 5, V=0.041 at Year 10), with meaningful variation in how specific groups fare. Interpret AI findings with caution given small cell sizes at all time points.

At Year 1, the most striking patterns are the very low postsecondary attendance rate for American Indian graduates (25.5%) and Hispanic graduates (33.6%) compared to White (57.0%) and Asian/PI (57.1%) graduates. American Indian graduates show the highest rates of non-meaningful employment (29.7%) and no MN employment record (26.8%). Hispanic graduates show the highest rate of meaningful regional employment (18.1%), suggesting earlier workforce entry.

By Year 5, racial and ethnic gaps in meaningful regional employment are notable. White (27.9%) and Hispanic (27.3%) graduates reach near-average rates, while Asian/PI (19.2%), Black (16.5%), and American Indian (16.9%) graduates lag behind. The no-MN-record rate is strikingly high for American Indian graduates (36.0%) and elevated for Asian/PI (26.5%), Black (24.2%), and Hispanic (25.4%) graduates relative to White graduates (19.6%).

By Year 10, the no-MN-employment-record gap widens further. Asian/PI graduates show the highest rate at 45.1%, followed by American Indian (41.1%), Black (39.7%), and Hispanic (37.1%) graduates, all substantially above the White rate (32.7%). Meaningful regional employment is considerably lower for Asian/PI (20.4%), Black (17.3%), and American Indian (19.0%) graduates than for White (30.4%) and Hispanic (28.4%) graduates.

The Year 10 no-MN-record pattern is the most important finding here. The elevated rates for Asian/PI and Black graduates in particular likely reflect differential out-of-state migration following postsecondary completion rather than labor force disconnection, and warrants careful interpretation in the narrative synthesis.

::: panel-tabset
### One year after graduating

```{r demographics RaceEthnicity one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, RaceEthnicity) %>%
  filter(grad.year.1 != "After 2023",
         RaceEthnicity != "Unknown") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/RaceEthnicity-1.html")
```

### Five years after graduating

```{r demographics RaceEthnicity five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, RaceEthnicity) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/RaceEthnicity-5.html")
```

### Ten years after graduating

```{r demographics RaceEthnicity ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, RaceEthnicity) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/RaceEthnicity-10.html")
```
:::

<br>

## `LimitedEnglishProficiencyIndicator`

The association between limited English proficiency and workforce outcomes is statistically significant at all three time points but weak and declining (V=0.079 at Year 1, V=0.032 at Year 5, V=0.031 at Year 10).

At Year 1, the most notable differences are a substantially lower postsecondary attendance rate for LEP-identified graduates (32.5% vs. 54.8%), a higher rate of no MN employment record (21.0% vs. 10.0%), and a notably elevated meaningful MN employment rate (11.4% vs. 4.5%). The latter is an unexpected finding and may reflect concentration in specific industries or geographic clusters that warrant further examination.

By Year 5, differences narrow considerably. LEP-identified graduates show modestly lower meaningful regional employment (21.5% vs. 27.2%) and higher no-MN-record rates (29.5% vs. 20.2%), but differences across other outcome categories are small.

By Year 10, the association is very weak (V=0.031). The primary remaining gap is in meaningful regional employment (22.4% vs. 29.8%) and the no-MN-record rate (42.3% vs. 33.3%), the latter likely reflecting out-of-state migration or informal employment patterns.

Overall, limited English proficiency shows a modest and diminishing association with workforce outcomes over time. The Year 1 pattern of lower postsecondary attendance and higher labor force absence is consistent with other disadvantage indicators, but the variable adds limited explanatory power beyond what english.learner and non.english.home already capture. Its weak Cramer's V values across all time points confirm the MCA finding that it occupies a distinct but secondary position relative to the other language variables.

::: panel-tabset
### One year after graduating

```{r demographics LimitedEnglishProficiencyIndicator one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, LimitedEnglishProficiencyIndicator) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$LimitedEnglishProficiencyIndicator, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/LimitedEnglishProficiencyIndicator-1.html")
```

### Five years after graduating

```{r demographics LimitedEnglishProficiencyIndicator five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, LimitedEnglishProficiencyIndicator) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$LimitedEnglishProficiencyIndicator, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/LimitedEnglishProficiencyIndicator-5.html")
```

### Ten years after graduating

```{r demographics LimitedEnglishProficiencyIndicator ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, LimitedEnglishProficiencyIndicator) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$LimitedEnglishProficiencyIndicator, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/LimitedEnglishProficiencyIndicator-10.html")
```
:::

<br>

## `english.learner`

English learner status shows a moderate association at Year 1 (V=0.164) that weakens substantially by Years 5 and 10 (V=0.078 and V=0.059). The pattern closely mirrors the LimitedEnglishProficiencyIndicator findings but with stronger effect sizes throughout, consistent with the MCA finding that english.learner is the more structurally dominant of the two language variables.

At Year 1, english learner graduates attend postsecondary at roughly half the rate of non-english learner graduates (31.5% vs. 56.6%) and are more than twice as likely to have no MN employment record (21.8% vs. 9.1%). They show elevated non-meaningful employment (22.7% vs. 16.2%) and meaningfully higher meaningful MN employment (8.3% vs. 4.3%), a pattern also seen in the LEP indicator and worth noting as a potential clustering effect in specific industries or labor markets.

By Year 5, gaps narrow across all categories. English learner graduates show lower meaningful regional employment (23.0% vs. 27.5%) and higher no-MN-record rates (29.5% vs. 19.7%), with modest differences elsewhere.

By Year 10, the association weakens further but persists. Meaningful regional employment remains lower for english learner graduates (21.5% vs. 30.0%), non-meaningful employment remains elevated (21.6% vs. 15.5%), and the no-MN-record gap, while narrowed, remains present (41.6% vs. 33.1%).

Compared to the LimitedEnglishProficiencyIndicator, english.learner shows consistently stronger associations at all three time points and captures a broader and more analytically useful slice of the language minority population. This confirms the framework decision to use english.learner as the primary language variable in multivariate models while treating LimitedEnglishProficiencyIndicator as supplementary.

::: panel-tabset
### One year after graduating

```{r demographics english.learner one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, english.learner) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/english-learner-1.html")
```

### Five years after graduating

```{r demographics english.learner five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, english.learner) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/english-learner-5.html")
```

### Ten years after graduating

```{r demographics english.learner ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, english.learner) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/english-learner-10.html")
```
:::

<br>

## `non.english.home`

Non-english home shows a moderate association at Year 1 (V=0.110) that weakens at Years 5 and 10 (V=0.054 and V=0.060). The overall pattern closely parallels english.learner but with consistently weaker effect sizes, confirming the expected redundancy between the two variables.

At Year 1, graduates from non-english-speaking homes attend postsecondary at a lower rate (41.0% vs. 55.6%) and are twice as likely to have no MN employment record (18.5% vs. 9.4%). Meaningful regional employment is essentially equivalent across groups (14.4% vs. 14.0%), and meaningful MN employment is notably higher for non-english home graduates (8.1% vs. 4.3%), consistent with the pattern observed for both LEP and english learner graduates.

By Year 5, the most persistent gap is in meaningful regional employment (22.1% vs. 27.6%) and the no-MN-record rate (27.6% vs. 19.8%), with other categories largely converging.

By Year 10, gaps remain in meaningful regional employment (22.2% vs. 30.2%) and the no-MN-record rate (42.3% vs. 32.8%), while non-meaningful employment shows a modest difference (18.2% vs. 15.6%).

Compared to english.learner, non.english.home produces weaker Cramer's V values at all three time points and captures a slightly different population — notably, the Year 1 meaningful regional employment gap is essentially absent here while it is present for english.learner. This reinforces the framework decision to exclude non.english.home from multivariate models in favor of english.learner, while retaining it for descriptive bivariate purposes.

::: panel-tabset
### One year after graduating

```{r demographics non.english.home one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, non.english.home) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/non-english-home-1.html")
```

### Five years after graduating

```{r demographics non.english.home five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, non.english.home) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/non-english-home-5.html")
```

### Ten years after graduating

```{r demographics non.english.home ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, non.english.home) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Demographics/non-english-home-10.html")
```
:::

```{r echo = FALSE}
meaningful.emp.southern %>%
  select(grad.year.1, grad.year.5, grad.year.10, non.english.home) %>%
  write_csv(., "Data/ChatGPT/Employment ct/Demographics/non-english-home.csv")

```

<br>

# High school characteristics

We will take a look at the following variables;

- `Dem_Desc`
- `edr`
- `avg.unemp.rate`
- `avg.wage.pct.state`

## Summary

The high school characteristics group — comprising community demographic description (Dem_Desc), economic development region (edr), average local unemployment rate (avg.unemp.rate), and average wages as a percent of state average (avg.wages.pct.state) — is collectively the weakest variable group examined so far. Effect sizes are small throughout, and two of the four variables add negligible practical explanatory power despite reaching statistical significance given the large sample size.

RANKING BY STRENGTH OF ASSOCIATION

1.  avg.unemp.rate (F=107.7 / 21.7 / 59.0) The average local unemployment rate is the strongest variable in this group and the only one with substantively meaningful effect sizes. At Year 1, graduates in meaningful regional employment come from lower unemployment areas, while the no-MN-record group comes from the highest unemployment areas — consistent with local labor market conditions shaping early workforce entry. By Year 5 and 10, a clearer long-run mechanism emerges: stronger local labor markets anchor graduates regionally, while higher local unemployment is associated with out-of-region employment or labor force marginality. This variable is retained in multivariate models and represents the most analytically useful signal in this group.

2.  edr (V=0.017 / 0.049 / 0.055) EDR shows weak but growing associations over time — unusual relative to most variables. The two EDRs are essentially indistinguishable at Year 1 but diverge modestly by Years 5 and 10, with EDR 9 graduates showing higher meaningful MN employment and EDR 10 graduates showing higher no-MN-record rates. Regional employment reaches near parity across both EDRs at all time points. The emerging gap is concentrated entirely in out-of-region outcomes and may reflect the geographic pull of the Twin Cities metro for EDR 9 graduates. Despite statistical significance, practical magnitude remains small throughout.

3.  Dem_Desc (V=0.041 / 0.046 / 0.030) Community demographic description shows the weakest and flattest association of any variable in the group. The urban/rural character of a graduate's high school has virtually no bearing on long-run workforce outcomes in Southeast Minnesota. The only substantively interesting signal is a modestly higher postsecondary attendance rate for entirely urban graduates at Years 1 and 5, consistent with greater access to four-year institutions, but this advantage dissipates entirely by Year 10.

4.  avg.wages.pct.state (F=78.4 / 58.3 / 16.2) Despite reaching statistical significance at all three time points, avg.wages.pct.state produces effect sizes so small — raw mean differences on the order of 0.003 to 0.013 percentage points — as to be analytically meaningless. This confirms the framework decision to exclude it from multivariate models. It is noted here for completeness only.

CROSS-CUTTING THEMES

The high school characteristics group as a whole tells a consistent story: where a graduate went to high school matters very little for long-run workforce outcomes in Southeast Minnesota, with the partial exception of the local unemployment rate, which reflects genuine labor market conditions rather than institutional or geographic sorting. The absence of strong place-based effects in this region contrasts with what might be expected and suggests that individual-level factors — particularly postsecondary pathways and high school accomplishments — are likely to carry much more explanatory weight in the multivariate stages. The two disconnection signatures identified in the demographics group do not map cleanly onto geographic characteristics here, reinforcing the interpretation that disconnection in Southeast Minnesota is driven more by individual circumstances than by place.

<br>

## `Dem_Desc`

Dem_Desc shows weak associations with workforce outcomes across all three time points (V=0.041 at Year 1, V=0.046 at Year 5, V=0.030 at Year 10), and the association is essentially flat over time — unusual relative to most demographic variables which show stronger Year 1 effects driven by postsecondary sorting.

At Year 1, the most notable difference is in postsecondary attendance, where entirely urban graduates attend at a higher rate (57.8%) than urban/town/rural mix (51.9%) or town/rural mix (54.6%) graduates. Entirely urban graduates show the lowest no-MN-record rate (8.8%) and the lowest non-meaningful employment rate (15.5%), while urban/town/rural mix graduates show the highest no-MN-record rate (11.3%).

By Year 5, the most distinctive pattern is that entirely urban graduates show higher continued postsecondary attendance (21.6% vs. 17.1% and 16.7%), consistent with higher rates of four-year credential pursuit, and modestly lower meaningful MN employment (14.4% vs. 17.2% and 17.6%), suggesting that credential completion for urban graduates extends further into the five-year window.

By Year 10, differences across groups are negligible. Meaningful regional employment ranges narrowly from 28.4% to 31.5%, non-meaningful employment is essentially identical across all three groups (15.7%-15.8%), and the no-MN-record rate shows the largest remaining difference — town/rural mix graduates show the lowest rate (31.9%) compared to urban/town/rural mix (33.8%) and entirely urban (33.7%).

Overall, community demographic description shows the weakest association with workforce outcomes of any high school characteristics variable and the pattern is largely flat across time. The urban/rural character of the high school from which a graduate came matters very little for long-run workforce outcomes in Southeast Minnesota, with the modest Year 5 postsecondary attendance difference being the most substantively interesting signal in the data.

::: panel-tabset
### One year after graduating

```{r hs characteristics Dem_Desc one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, Dem_Desc) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school characteristics/Dem_Desc-1.html")
```

### Five years after graduating

```{r hs characteristics Dem_Desc five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, Dem_Desc) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school characteristics/Dem_Desc-5.html")
```

### Ten years after graduating

```{r hs characteristics Dem_Desc ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, Dem_Desc) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school characteristics/Dem_Desc-10.html")
```
:::

<br>

## `edr`

EDR shows very weak associations with workforce outcomes across all three time points (V=0.017 at Year 1, V=0.049 at Year 5, V=0.055 at Year 10), with the unusual pattern of strengthening slightly over time rather than weakening — the opposite of most variables examined so far.

At Year 1, the two EDRs are virtually indistinguishable across all outcome categories. Postsecondary attendance is identical (54.4%), meaningful regional employment is essentially equivalent (14.0% vs. 14.2%), and differences in non-meaningful employment and no-MN-record rates are negligible.

By Year 5, a meaningful MN employment gap emerges — EDR 9 graduates show a notably higher rate (18.7% vs. 15.3%) while EDR 10 graduates show a higher no-MN-record rate (21.0% vs. 18.7%). Meaningful regional employment remains nearly identical across the two EDRs (27.0% vs. 27.5%).

By Year 10, the meaningful MN employment gap widens further (24.3% for EDR 9 vs. 19.8% for EDR 10), and the no-MN-record gap also grows (34.6% for EDR 10 vs. 30.6% for EDR 9). Meaningful regional employment reaches near parity (29.7% vs. 29.6%).

The emerging EDR gap at Years 5 and 10 is concentrated entirely in meaningful MN employment and the no-MN-record category, with regional employment essentially equivalent across both EDRs at all time points. This suggests that EDR 9 graduates are more likely to find sustained employment outside the region but within Minnesota — potentially reflecting the geographic pull of the Twin Cities metro for EDR 9 graduates — while EDR 10 graduates are more likely to leave the state entirely or exit formal employment records. Despite statistical significance, the practical magnitude remains small throughout.

::: panel-tabset
### One year after graduating

```{r hs characteristics edr one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, edr) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school characteristics/edr-1.html")
```

### Five years after graduating

```{r hs characteristics edr five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, edr) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school characteristics/edr-5.html")
```

### Ten years after graduating

```{r hs characteristics edr ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, edr) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE, 
         file = "Data/Claude/analysis_local_employment_ct/High school characteristics/edr-10.html")
```
:::

<br>

## `avg.unemp.rate`

The average unemployment rate of the high school's local area shows a statistically significant association with workforce outcomes at all three time points, with the strongest effect at Year 1 (F=107.7) and weakening at Years 5 (F=21.7) and 10 (F=59.0).

At Year 1, Tukey HSD comparisons show that graduates in meaningful regional employment come from areas with notably lower average unemployment than all other outcome groups. The largest gaps are between meaningful regional employment and no-MN-record (difference=0.190) and meaningful regional employment and non-meaningful employment (difference=0.146). Meaningful MN employment graduates come from the lowest unemployment areas of all — sitting 0.168 below meaningful regional employment — suggesting early out-of-region employment is associated with the most economically distressed home communities, possibly reflecting limited local opportunity driving earlier geographic mobility. Attending ps and non-meaningful employment are not significantly different from each other (adj. p=0.125).

By Year 5, the pattern partially reverses. Meaningful regional employment graduates no longer come from the lowest unemployment areas — meaningful MN employment graduates now come from significantly higher unemployment areas (difference=0.112 above regional). Non-meaningful employment is modestly above meaningful regional employment (difference=0.046), while attending ps and no-MN-record are not significantly different from meaningful regional employment. This suggests that by Year 5 higher local unemployment is associated with out-of-region employment rather than regional attachment.

By Year 10, meaningful regional employment graduates come from the lowest unemployment areas among the three active workforce categories. Meaningful MN employment (difference=0.174) and non-meaningful employment (difference=0.124) both come from significantly higher unemployment areas, and the no-MN-record group sits between these extremes (difference=0.066 above regional). The pattern is consistent with a long-run mechanism where stronger local labor markets anchor graduates regionally, while higher local unemployment pushes graduates toward out-of-region employment or labor force marginality.

::: panel-tabset
### One year after graduation

```{r hs characteristics avg.unemp.rate one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, avg.unemp.rate) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

summary <- aov(avg.unemp.rate ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school characteristics/avg-unemp-rate-1.xlsx", overwrite = TRUE)
```

### Five years after graduating

```{r hs characteristics avg.unemp.rate five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, avg.unemp.rate) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()


summary <- aov(avg.unemp.rate ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school characteristics/avg-unemp-rate-5.xlsx", overwrite = TRUE)

```

### Ten years after graduating

```{r hs characteristics avg.unemp.rate ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, avg.unemp.rate) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

summary <- aov(avg.unemp.rate ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school characteristics/avg-unemp-rate-10.xlsx", overwrite = TRUE)
```
:::

<br>

## `avg.wages.pct.state`

While statistically significant at all three time points (F=78.4 at Year 1, F=58.3 at Year 5, F=16.2 at Year 10), the effect sizes are extremely small throughout — the raw differences between group means are on the order of 0.003 to 0.013 percentage points of the state wage average. This variable adds essentially no practical explanatory power despite reaching statistical significance, consistent with its exclusion from multivariate models in the established framework.

At Year 1, graduates attending postsecondary come from the highest-wage communities, sitting above meaningful regional employment (difference=0.010), which in turn sits above non-meaningful employment (difference=0.003) and the no-MN-record group (not significantly different from regional). The pattern likely reflects the socioeconomic composition of communities rather than a direct labor market mechanism.

By Year 5, the pattern shifts slightly. Attending ps graduates still come from the highest-wage communities. Meaningful regional employment now sits above meaningful MN employment (difference=-0.004), non-meaningful employment is not significantly different from regional, and the no-MN-record group comes from modestly higher-wage communities (difference=0.003).

By Year 10, only two contrasts are significant: meaningful MN employment graduates come from modestly lower-wage communities than meaningful regional employment graduates (difference=-0.005), and the no-MN-record group comes from modestly higher-wage communities than meaningful MN employment (difference=0.007).

The differences are negligible in practical terms across all three time points. This confirms the framework decision to exclude avg.wages.pct.state from multivariate models. It will be noted in the group summary but warrants no further analytical attention.

::: panel-tabset
### One year after graduating

```{r hs characteristics avg.wages.pct.state one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, avg.wages.pct.state) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()


summary <- aov(avg.wages.pct.state ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school characteristics/avg-wage-pct-state-1.xlsx", overwrite = TRUE)

```

### Five years after graduating

```{r hs characteristics avg.wages.pct.state five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, avg.wages.pct.state) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

summary <- aov(avg.wages.pct.state ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school characteristics/avg-wage-pct-state-5.xlsx", overwrite = TRUE)
```

### Ten years after graduating

```{r hs characteristics avg.wages.pct.state ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, avg.wages.pct.state) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

summary <- aov(avg.wages.pct.state ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school characteristics/avg-wage-pct-state-10.xlsx", overwrite = TRUE)
```
:::

<br>

# High School Accomplishments

We are going to analyze the following variables;

- High school accomplishments
  - `pseo.participant`
  - `took.ACT`
  - `ACTCompositeScore`
  - `ap.exam`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`

<br>

## Summary

The high school accomplishments group is the most analytically rich variable group examined so far, containing the strongest predictors of workforce outcomes in the entire analysis. The group divides clearly into two distinct clusters: academic achievement and orientation variables (took.ACT, ACTCompositeScore, ap.exam, MCA.R, MCA.M, MCA.S, pseo.participant) that are strongly associated with postsecondary pathways and out-of-region employment, and CTE engagement variables (cte.achievement, total.cte.courses.taken) that are strongly and uniquely associated with regional workforce attachment.

RANKING BY STRENGTH OF ASSOCIATION (Cramer's V or F-statistic at Year 5 as primary sort)

1.  took.ACT (V=0.419 / 0.203 / 0.123) The strongest single categorical predictor in the entire analysis to this point. ACT participation functions as a near-perfect proxy for postsecondary orientation at Year 1, sorting graduates sharply into postsecondary attendance versus workforce entry. By Year 10, the no-MN record rate converges completely across ACT and non-ACT takers, confirming that geographic absence is not driven by academic disengagement but rather by out-of-state migration following credential completion among the highest achievers.

2.  MCA Math (F=4955.4 / 1487.9 / 317.0) The strongest continuous predictor and the highest F-statistics of any academic achievement variable at Years 1 and 5. MCA Math scores differentiate postsecondary attendees from workforce entrants at Year 1, and by Years 5 and 10 produce the clearest separation between the regional employment cluster (lower scores) and the MN employment plus no-MN-record cluster (higher scores). The convergence of meaningful MN employment and no-MN-record at similar high scores by Year 10 is the sharpest expression of this pattern across all academic measures.

3.  ap.exam (V=0.281 / 0.165 / 0.130) AP participation produces the largest postsecondary attendance gap of any variable at Year 1 (80.0% vs. 46.7%) and the largest no-MN-record gap at Year 10 (40.0% vs. 31.7%) — both reflections of AP participation selecting for the most academically oriented graduates who are most likely to complete four-year credentials and leave Minnesota. Its exclusion from non-college completer models is well-justified by the small group size in that subpopulation.

4.  MCA Reading (F=3405.6 / 395.2 / 812.6) MCA Reading is notable for its increasing F-statistic from Year 5 to Year 10 — the only academic achievement variable to show this pattern. The Year 10 finding is analytically important: meaningful MN employment and the no-MN-record group converge at virtually identical scores, both substantially above regional employment and non-meaningful employment. This two-cluster structure at Year 10 is the clearest evidence in the entire analysis that geographic mobility — whether captured in Minnesota records or not — is driven by high academic achievement.

5.  MCA Science (F=2162.5 / 440.1 / 181.4) MCA Science closely mirrors MCA Reading in structure throughout but with a more steeply declining F-statistic, consistent with voluntary assessment self-selection introducing composition effects. The Year 10 two-cluster structure — regional and non-meaningful at lower scores, MN employment and no-MN-record at higher scores — is identical to MCA Reading, reinforcing confidence in the finding. MCA Science is excluded from multivariate models due to the self-selection concern but serves as useful corroborating evidence in bivariate stages.

6.  pseo.participant (V=0.257 / 0.097 / 0.071) PSEO participation shows a strong Year 1 association driven by the postsecondary orientation of PSEO participants, but weakens more sharply than other variables over time. Its cohort restriction (2011+ only) and modest long-run effect size relative to other variables confirm the framework decision to treat it as supplementary rather than a primary analytical variable.

7.  cte.achievement (V=0.096 / 0.099 / 0.089) CTE achievement shows the flattest association trajectory of any variable in the group — its Cramer's V values are virtually identical across all three time points. This stability is analytically distinctive: while academic orientation variables peak early and fade, CTE engagement maintains consistent associative strength across the full observation window. The dose-response gradient in meaningful regional employment (concentrators 34.9%, participants 30.5%, No CTE 22.2% at Year 10) is monotonic and durable, and the No CTE group's elevated no-MN-record rate (39.3% at Year 10) confirms that absence of CTE engagement is the defining characteristic of the disconnected pathway.

8.  total.cte.courses.taken (F=1292.2 / 613.1 / 335.9) Total CTE courses taken produces the most consistent dose-response relationship of any variable in the group. Regional employment graduates take the most CTE courses at every time point, and the no-MN-record group takes the fewest — a gradient that strengthens rather than weakens over time. The Year 10 finding that the no-MN-record group takes significantly fewer CTE courses than even non-meaningful employment graduates is the strongest single piece of evidence that geographic absence from Minnesota's labor market is associated with low CTE engagement rather than high academic achievement alone. This confirms the framework decision to use total.cte.courses.taken in non-college completer multivariate models.

CROSS-CUTTING THEMES

Two distinct and opposing pathways are visible across the accomplishment group. The college completer and geographic mobility pathway is defined by high academic achievement (ACT, MCA scores, AP participation, PSEO) and is associated with postsecondary attendance, credential completion, and sustained employment outside the region. The workforce-first pathway is defined by CTE engagement and is associated with early and sustained regional employment. These two pathways are not simply mirror images — they reflect genuinely different orientations toward education and work that manifest clearly by Year 1 and remain stable through Year 10.

The disconnected pathway, characterized by absence of both credential completion and sustained employment, is most strongly associated with low CTE engagement and low academic achievement simultaneously — the group that falls between both pathways rather than choosing one. The no-MN-record group at Year 10 shows a bimodal character: some portion reflects successful out-of-state migration among high academic achievers, while another portion reflects the geographic absence disconnection signature among low CTE engagers. Disentangling these two subpopulations within the no-MN-record category will be a central interpretive challenge in the narrative synthesis.

<br>

## `pseo.participant`

PSEO participation shows a strong association at Year 1 (V=0.257) that weakens substantially by Years 5 and 10 (V=0.097 and V=0.071). Note that this variable is restricted to 2011+ cohorts and the NA group has been excluded, so the analysis covers approximately 91,000 graduates at Year 1 rather than the full population.

At Year 1, the dominant pattern is a striking postsecondary attendance gap — PSEO participants attend postsecondary at 72.6% compared to 45.6% for non-participants. This is expected given that PSEO participation is itself a postsecondary experience during high school and selects for academically oriented students. Non-participants show substantially higher rates of meaningful regional employment (16.6% vs. 9.7%), non-meaningful employment (19.7% vs. 10.0%), and no-MN-record (12.4% vs. 4.3%), all consistent with earlier workforce entry among the non-PSEO population.

By Year 5, the postsecondary attendance gap narrows but remains visible (22.2% vs. 16.3%). PSEO participants now show higher meaningful MN employment (19.4% vs. 15.6%) — consistent with college completion driving out-of-region career pathways — while non-participants hold a modest advantage in meaningful regional employment (28.2% vs. 24.9%). Non-participants also show higher non-meaningful employment (18.5% vs. 14.5%).

By Year 10, differences are modest (V=0.071). PSEO participants show higher meaningful MN employment (25.9% vs. 19.7%) and lower meaningful regional employment (25.6% vs. 29.7%), while no-MN-record rates are essentially equivalent (33.8% vs. 34.5%). Non-meaningful employment is slightly higher for non-participants (16.2% vs. 14.7%).

The overall pattern is consistent with PSEO participation serving as a strong proxy for postsecondary orientation and four-year credential pursuit. PSEO participants are more likely to complete credentials and find sustained employment outside the region, while non-participants enter the regional workforce earlier but at similar long-run regional employment rates. This confirms the framework decision to treat PSEO as a supplementary variable given its cohort restriction, noting its modest effect size relative to other accomplishment variables.

::: panel-tabset
### One year after graduating

```{r hs accomplishments pseo.participant one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, pseo.participant) %>%
  filter(grad.year.1 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/pseo-participant-1.html")
```

### Five years after graduating

```{r hs accomplishments pseo.participant five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, pseo.participant) %>%
   filter(grad.year.5 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/pseo-participant-5.html")
```

### Ten years after graduating

```{r hs accomplishments pseo.participant ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, pseo.participant) %>%
  filter(grad.year.10 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/pseo-participant-10.html")
```
:::

<br>

## `took.ACT`

ACT participation shows one of the strongest associations of any variable examined so far, with a very high Cramer's V at Year 1 (V=0.419) that weakens substantially but remains meaningful at Years 5 and 10 (V=0.203 and V=0.123).

At Year 1, the pattern is dominated by a dramatic postsecondary attendance gap. ACT takers attend postsecondary at 67.8% compared to just 24.3% for non-takers — a gap of over 43 percentage points. Non-takers show dramatically higher rates of meaningful regional employment (20.3% vs. 11.2%), non-meaningful employment (27.7% vs. 11.9%), and no-MN-record (21.2% vs. 5.2%). ACT participation is functioning here as a near-perfect proxy for postsecondary orientation, reflecting self-selection rather than a direct causal mechanism.

By Year 5, the postsecondary gap narrows as ACT takers complete credentials and enter the labor market, but remains visible (23.2% vs. 8.8%). Non-takers now show higher meaningful regional employment (30.9% vs. 25.4%) and higher non-meaningful employment (21.7% vs. 15.7%) and no-MN-record rates (25.7% vs. 17.8%). ACT takers show meaningfully higher meaningful MN employment (17.9% vs. 13.0%), consistent with college-completion-driven out-of-region career pathways.

By Year 10, the no-MN-record rate converges completely (33.5% vs. 33.4%). Non-takers retain a modest advantage in meaningful regional employment (32.7% vs. 27.7%), while ACT takers show substantially higher meaningful MN employment (24.9% vs. 15.4%) and lower non-meaningful employment (14.1% vs. 18.4%).

The long-run picture is clear: ACT participation is strongly associated with college-completion pathways and sustained out-of-region employment, while non-ACT graduates who connect to the workforce are more likely to do so regionally. The complete convergence of no-MN-record rates by Year 10 is notable — the geographic absence signature is not systematically associated with ACT participation in the long run, suggesting it reflects out-of-state migration among college completers rather than disconnection among non-takers.

::: panel-tabset
### One year after graduating

```{r hs accomplishments took.ACT one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, took.ACT) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/took-ACT-1.html")
```

### Five years after graduating

```{r hs accomplishments took.ACT five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, took.ACT) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/took-ACT-5.html")
```

### Ten years after graduating

```{r hs accomplishments took.ACT ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, took.ACT) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/took-ACT-10.html")
```
:::

<br>

## `ACTCompositeScore`

ACT composite score shows strong and consistent associations with workforce outcomes at all three time points (F=2177.4 at Year 1, F=639.8 at Year 5, F=455.8 at Year 10), with effect sizes remaining meaningful across the full observation window — unlike many variables that weaken sharply over time.

At Year 1, postsecondary attendees score 3.83 points above meaningful regional employment graduates — by far the largest gap — confirming that ACT score is strongly associated with postsecondary orientation. Meaningful regional and MN employment graduates score nearly identically (difference=-0.064, not significant), while non-meaningful employment graduates score 0.85 points below regional and no-MN-record graduates score 1.49 points below regional. This suggests that at Year 1, ACT score differentiates postsecondary attendees from the rest but does not meaningfully distinguish between the workforce outcome categories.

By Year 5, a more differentiated picture emerges. Meaningful regional employment graduates now have the lowest mean ACT score of any active workforce group. Meaningful MN employment (difference=1.84) and attending ps (difference=2.43) graduates score substantially higher, while the no-MN-record group also scores notably higher (difference=2.28) — suggesting higher-scoring graduates who have not yet found sustained MN employment may still be completing credentials or pursuing opportunities outside the state. Non-meaningful employment graduates score only modestly above regional (difference=0.71).

By Year 10, the pattern consolidates. Meaningful regional employment graduates have the lowest mean ACT scores, with meaningful MN employment (difference=1.64), non-meaningful employment (difference=0.81), and no-MN-record (difference=2.37) all scoring above them. The no-MN-record group now has the highest mean ACT scores of any group — substantially above even meaningful MN employment (difference=0.73) — suggesting that the highest-scoring graduates are the most likely to have left Minnesota's formal employment records entirely, consistent with out-of-state migration following credential completion.

The long-run ACT pattern tells a consistent story: higher academic achievement is associated with college completion and out-of-region careers, while regional workforce attachment is most characteristic of graduates with lower ACT scores. This does not indicate a deficiency in regional employment — it reflects the sorting function of postsecondary education and the geography of college-driven career pathways.

::: panel-tabset
### One year after graduating

```{r hs accomplishments ACTCompositeScore one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, ACTCompositeScore) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

summary <- aov(ACTCompositeScore ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/ACTCompositeScore-1.xlsx", overwrite = TRUE)

```

### Five years after graduating

```{r hs accomplishments ACTCompositeScore five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, ACTCompositeScore) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

summary <- aov(ACTCompositeScore ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/ACTCompositeScore-5.xlsx", overwrite = TRUE)

```

### Ten years after graduating

```{r hs accomplishments ACTCompositeScore ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, ACTCompositeScore) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

summary <- aov(ACTCompositeScore ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/ACTCompositeScore-10.xlsx", overwrite = TRUE)

```
:::

<br>

## `ap.exam`

AP exam participation shows a strong association at Year 1 (V=0.281) that weakens but remains meaningful at Years 5 and 10 (V=0.165 and V=0.130). The pattern closely parallels took.ACT but with one important distinction at Year 10.

At Year 1, AP exam takers attend postsecondary at 80.0% compared to 46.7% for non-takers — the largest postsecondary attendance gap of any accomplishment variable examined so far. Non-takers show substantially higher rates of meaningful regional employment (16.5% vs. 5.8%), non-meaningful employment (19.4% vs. 8.1%), and no-MN-record (12.0% vs. 4.0%). AP participation is functioning as an even stronger proxy for postsecondary orientation than ACT-taking, reflecting the highly academically selective nature of AP course enrollment.

By Year 5, AP exam takers show markedly higher continued postsecondary attendance (27.9% vs. 15.7%) and meaningful MN employment (19.4% vs. 15.4%), while non-takers show substantially higher meaningful regional employment (29.8% vs. 18.3%). The no-MN-record rate is essentially identical across groups (20.3% vs. 20.5%), and non-meaningful employment is modestly higher for non-takers (18.8% vs. 14.0%).

By Year 10, the most striking finding is the elevated no-MN-record rate for AP exam takers (40.0% vs. 31.7%) — the largest such gap of any accomplishment variable at Year 10. AP takers show substantially lower meaningful regional employment (20.1% vs. 32.2%) and higher meaningful MN employment (26.8% vs. 19.6%) and lower non-meaningful employment (13.1% vs. 16.5%).

The Year 10 no-MN-record gap for AP exam takers is particularly noteworthy. Combined with their elevated meaningful MN employment rate, this pattern strongly suggests that AP participation selects for graduates who complete four-year credentials and pursue careers outside Minnesota rather than returning to the region — the college completer pathway operating at its most geographically mobile extreme. This reinforces the framework decision to include ap.exam in the full population multivariate model while excluding it from the non-college completer models where its small group size limits analytical utility.

::: panel-tabset
### One year after graduating

```{r hs accomplishments ap.exam one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, ap.exam) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/ap-exam-1.html")
```

### Five years after graduating

```{r hs accomplishments ap.exam five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, ap.exam) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/ap-exam-5.html")
```

### Ten years after graduating

```{r hs accomplishments ap.exam ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, ap.exam) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/ap-exam-10.html")
```
:::

<br>

## `MCA.R`

MCA Reading score shows strong and sustained associations with workforce outcomes at all three time points (F=3405.6 at Year 1, F=395.2 at Year 5, F=812.6 at Year 10), with effect sizes remaining meaningful throughout — and notably increasing from Year 5 to Year 10.

At Year 1, postsecondary attendees score 0.342 points above meaningful regional employment graduates on the standardized MCA Reading scale — by far the largest gap. Meaningful regional and MN employment graduates score nearly identically (difference=0.033, not significant), while non-meaningful employment graduates score 0.261 below regional and no-MN-record graduates score 0.362 below regional. The Year 1 pattern mirrors the ACT composite score findings: MCA Reading differentiates postsecondary attendees from workforce entrants but does not distinguish meaningfully between the two sustained employment categories.

By Year 5, a more differentiated structure emerges. Meaningful regional employment graduates now have the lowest mean MCA Reading scores among active workforce participants. Meaningful MN employment (difference=0.170) and attending ps (difference=0.261) graduates score notably higher. The no-MN-record group also scores above regional (difference=0.145), while non-meaningful employment graduates score only modestly above regional (difference=0.026).

By Year 10, the pattern sharpens considerably. Meaningful regional employment graduates remain at the bottom of the score distribution. Meaningful MN employment and the no-MN-record group both score approximately 0.40 points above regional — and are not significantly different from each other (difference=0.007, p=0.928). Non-meaningful employment graduates are not significantly different from meaningful regional employment (difference=0.012, p=0.747).

The Year 10 finding is analytically important. The convergence of meaningful MN employment and the no-MN-record group at similar MCA Reading scores — both substantially above regional employment and non-meaningful employment — suggests that higher academic achievement as measured in high school is associated with geographic mobility and out-of-region career trajectories, whether or not those trajectories appear in Minnesota's formal employment records. This reinforces the interpretation that a substantial portion of the no-MN-record group represents successful out-of-state migration rather than disconnection.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.R one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, MCA.R) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

summary <- aov(MCA.R ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-R-1.xlsx", overwrite = TRUE)

```

### Five years after graduating

```{r hs accomplishments MCA.R five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, MCA.R) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

summary <- aov(MCA.R ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-R-5.xlsx", overwrite = TRUE)

```

### Ten years after graduating

```{r hs accomplishments MCA.R ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, MCA.R) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

summary <- aov(MCA.R ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-R-10.xlsx", overwrite = TRUE)

```
:::

<br>

## `MCA.S`

MCA Science score shows strong associations at all three time points (F=2162.5 at Year 1, F=440.1 at Year 5, F=181.4 at Year 10), though the F-statistic declines more steeply over time than MCA Reading — consistent with MCA Science being a voluntary and self-selected assessment that introduces sample composition effects.

At Year 1, postsecondary attendees score 0.490 points above meaningful regional employment graduates — the largest gap of any academic achievement variable at this time point. Meaningful regional and MN employment graduates score nearly identically (difference=-0.016, not significant), while non-meaningful employment graduates score 0.072 below regional and no-MN-record graduates score 0.047 below regional (though the latter is only marginally significant). The Year 1 pattern mirrors MCA Reading closely.

By Year 5, meaningful regional employment graduates have the lowest mean MCA Science scores among active workforce participants. Meaningful MN employment (difference=0.221) and attending ps (difference=0.362) score substantially higher. The no-MN-record group scores above regional (difference=0.181) but below meaningful MN employment (difference=-0.039). Non-meaningful employment scores modestly below regional (difference=-0.036).

By Year 10, the pattern simplifies. Meaningful regional employment and non-meaningful employment graduates score essentially identically (difference=-0.013, p=0.833). Meaningful MN employment and the no-MN-record group both score approximately 0.23 points above regional, and are not significantly different from each other (difference=-0.013, p=0.767).

The Year 10 structure of MCA Science closely mirrors MCA Reading — regional employment and non-meaningful employment converge at lower scores, while meaningful MN employment and no-MN-record converge at higher scores. This parallel structure reinforces confidence in the finding and confirms that academic achievement in high school is consistently associated with geographic mobility rather than regional workforce attachment in the long run, regardless of which assessment is used as the measure.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.S five year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, MCA.S) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

summary <- aov(MCA.S ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-S-1.xlsx", overwrite = TRUE)

```

### Five years after graduating

```{r}
data <- meaningful.emp.southern %>%
  select(grad.year.5, MCA.S) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

summary <- aov(MCA.S ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-S-5.xlsx", overwrite = TRUE)
```

### Ten years after graduating

```{r hs accomplishments MCA.S ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, MCA.S) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

summary <- aov(MCA.S ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-S-10.xlsx", overwrite = TRUE)

```
:::

<br>

## `MCA.M`

MCA Math shows the strongest F-statistics of any academic achievement variable at Years 1 and 5 (F=4955.4 and F=1487.9), and remains strong at Year 10 (F=317.0). The overall pattern closely parallels MCA Reading and Science, with some notable differences in magnitude.

At Year 1, postsecondary attendees score 0.662 points above meaningful regional employment graduates — substantially larger than the comparable gaps for MCA Reading (0.342) and Science (0.490). Meaningful regional and MN employment graduates score essentially identically (difference=0.018, not significant). Non-meaningful employment graduates score 0.141 below regional and no-MN-record graduates score 0.216 below regional — both larger gaps than seen for MCA Reading, suggesting MCA Math is a somewhat stronger differentiator of early disconnection.

By Year 5, MCA Math produces the largest score gaps of any academic achievement variable. Meaningful MN employment (difference=0.534) and attending ps (difference=0.594) both score substantially above meaningful regional employment. The no-MN-record group scores modestly above regional (difference=0.135) but well below both MN employment and attending ps. Non-meaningful employment scores below regional (difference=-0.079).

By Year 10, the familiar two-cluster structure emerges. Meaningful regional employment and non-meaningful employment score essentially identically (difference=-0.010, p=0.917). Meaningful MN employment scores 0.333 above regional, and the no-MN-record group scores 0.300 above regional — these two are not significantly different from each other (difference=-0.033, p=0.096). The Year 10 gap magnitudes are larger for MCA Math than for MCA Reading (0.333/0.300 vs. 0.400/0.407) but the structural pattern is identical.

MCA Math is the strongest single academic achievement predictor in this analysis and shows the clearest separation between the college-completion and geographic mobility pathway on one hand, and the regional workforce and non-meaningful employment pathway on the other. The consistent convergence of meaningful regional employment and non-meaningful employment at the lower end of the score distribution across all three time points is a durable finding that will carry into the multivariate stages.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.M one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, MCA.M) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

summary <- aov(MCA.M ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-M-1.xlsx", overwrite = TRUE)

```

### Five years after graduating

```{r hs accomplishments MCA.M five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, MCA.M) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

summary <- aov(MCA.M ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-M-5.xlsx", overwrite = TRUE)
```

### Ten years after graduating

```{r hs accomplishments MCA.M ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, MCA.M) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

summary <- aov(MCA.M ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/MCA-M-10.xlsx", overwrite = TRUE)

```
:::

<br>

## `cte.achievement`

CTE achievement shows a remarkably stable association with workforce outcomes across all three time points (V=0.096 at Year 1, V=0.099 at Year 5, V=0.089 at Year 10) — the flattest trajectory of any accomplishment variable examined. Unlike the academic achievement variables which peak at Year 1 and decline, CTE engagement maintains essentially the same associative strength across the entire observation window.

At Year 1, CTE concentrators and completors show the highest rate of meaningful regional employment (18.3%) — more than double the rate for No CTE graduates (8.7%) and well above CTE participants (12.9%). The inverse is true for postsecondary attendance: No CTE graduates attend at 59.9%, CTE participants at 57.2%, and concentrators at just 48.8%. The no-MN-record rate is notably lower for CTE participants (8.5%) than for both concentrators (10.1%) and No CTE graduates (12.1%), a pattern worth watching over time.

By Year 5, the CTE gradient in meaningful regional employment sharpens further. Concentrators reach 33.1%, participants 27.6%, and No CTE graduates just 19.6% — a 13.5 percentage point spread across the three categories. No CTE graduates show the highest no-MN-record rate (24.5%) compared to concentrators (18.2%) and participants (18.7%). Meaningful MN employment is essentially flat across all three CTE categories (15.7%, 16.3%, 17.1%), confirming that CTE engagement is specifically associated with regional workforce attachment rather than out-of-region employment.

By Year 10, the regional employment gradient remains strong and consistent: concentrators at 34.9%, participants at 30.5%, and No CTE graduates at 22.2%. The no-MN-record gap widens further — No CTE graduates reach 39.3% compared to 29.6% for concentrators and 32.5% for participants. Meaningful MN employment shows a modest upward gradient from concentrators to No CTE graduates (19.8%, 20.9%, 23.1%), consistent with No CTE graduates being more likely to pursue college-completion pathways that lead out of the region.

CTE achievement is the defining variable for the workforce-first pathway in Southeast Minnesota. Its association with regional employment is monotonic, dose-responsive, and stable across all three time points — a pattern unmatched by any other variable in this group. The absence of a CTE engagement advantage in meaningful MN employment confirms that CTE's effect is specifically regional in character. The No CTE group's elevated no-MN-record rate at Year 10 (39.3%) adds further weight to the interpretation that absence of CTE engagement is the defining characteristic of the disconnected pathway.

::: panel-tabset
### One year after graduating

```{r hs accomplishments cte.achievement one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, cte.achievement) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/cte-achievement-1.html")
```

### Five years after graduating

```{r hs accomplishments cte.achievement five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, cte.achievement) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/cte-achievement-5.html")
```

### Ten years after graduating

```{r hs accomplishments cte.achievement ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, cte.achievement) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/High school accomplishments/cte-achievement-10.html")
```
:::

<br>

## `total.cte.courses.taken`

Total CTE courses taken shows strong and sustained associations with workforce outcomes at all three time points (F=1292.2 at Year 1, F=613.1 at Year 5, F=335.9 at Year 10), with meaningful regional employment graduates consistently taking the most CTE courses of any outcome group across the entire observation window.

At Year 1, meaningful regional employment graduates have the highest mean CTE course count of any group. All other groups score below them: meaningful MN employment (difference=-0.892), non-meaningful employment (difference=-0.927), no-MN-record (difference=-1.248), and postsecondary attendees (difference=-1.729). Notably, meaningful MN employment and non-meaningful employment are not significantly different from each other (difference=-0.035, p=0.935), while the no-MN-record group takes significantly fewer courses than either employment group.

By Year 5, the ordering is maintained and the gaps are substantively meaningful. Meaningful regional employment remains at the top. Meaningful MN employment (difference=-0.836) and attending ps (difference=-1.215) score below regional, while non-meaningful employment (difference=-0.405) falls between regional and MN employment. The no-MN-record group scores substantially below regional (difference=-1.081) and also below meaningful MN employment (difference=-0.245).

By Year 10, the pattern is fully consolidated. Meaningful regional employment graduates have taken approximately 0.81 more CTE courses than meaningful MN employment graduates, 0.55 more than non-meaningful employment graduates, and 0.96 more than the no-MN-record group. The no-MN-record group now takes significantly fewer courses than even non-meaningful employment graduates (difference=-0.404).

Total CTE courses taken produces the clearest and most consistent dose-response relationship of any variable in the accomplishment group. Higher CTE course-taking is monotonically associated with regional workforce attachment at all three time points, and the absence of CTE engagement is most strongly associated with geographic absence from Minnesota's formal labor market. This confirms the framework decision to use total.cte.courses.taken rather than cte.achievement in the non-college completer multivariate models, where its continuous dose-response structure provides superior sensitivity.

::: panel-tabset
### One year after graduating

```{r hs accomplishments total.cte.courses.taken one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, total.cte.courses.taken) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

summary <- aov(total.cte.courses.taken ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/total-cte-courses-taken-1.xlsx", overwrite = TRUE)

```

### Five years after graduating

```{r hs accomplishments total.cte.courses.taken five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, total.cte.courses.taken) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

summary <- aov(total.cte.courses.taken ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/total-cte-courses-taken-5.xlsx", overwrite = TRUE)

```

### Ten years after graduating

```{r hs accomplishments total.cte.courses.taken ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, total.cte.courses.taken) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

summary <- aov(total.cte.courses.taken ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_local_employment_ct/High school accomplishments/total-cte-courses-taken-10.xlsx", overwrite = TRUE)

```
:::

<br>

# Post-secondary

We will analyze the following variables;

- Post-secondary
  - `attended.ps.within.first.year.hsgrad`
  - `attended.ps`
  - `ps.grad.`
  - `ps.grad.InstitutionSector`
  - `highest.cred.level`
  - `ps.grad.location`

<br>

## Summary

The postsecondary variable group is the most structurally complex of the four groups, containing variables that are both outcomes and predictors, overlapping by construction, and operating through fundamentally different mechanisms depending on credential type and geographic location. The group's defining contribution to the analysis is clarifying how the college completer pathway operates — and specifically how graduation location, credential level, and institution type jointly determine whether postsecondary completion translates into regional workforce attachment or geographic departure.

RANKING BY STRENGTH OF ASSOCIATION (Cramer's V at Year 5 as primary sort)

1.  ps.grad.location (V=0.164 / 0.208 / 0.171) The strongest and most analytically important postsecondary variable, and the only one in the entire analysis whose association strengthens from Year 1 to Year 5. Graduation location is the critical hinge of the college completer pathway: regional graduation is strongly associated with meaningful regional employment at all three time points (63.0%, 52.7%, 43.2%), MN graduation is strongly associated with meaningful MN employment (26.3%, 43.4%, 37.9%), and outside MN graduation is associated with geographic absence from Minnesota's formal labor market — reaching a 54.2% no-MN-record rate by Year 10. This variable provides the clearest evidence in the entire analysis that the no-MN-record group is heterogeneous, with outside MN graduates representing successful out-of-state career establishment rather than disconnection.

2.  attended.ps.within.first.year.hsgrad (V=0.254 / 0.118 / 0.074) The timing of postsecondary enrollment matters most at Year 1, where immediate enrollees show substantially higher rates of concurrent workforce participation than delayed or non-enrollees. By Year 10 the distinction between immediate and delayed enrollment largely disappears, with postsecondary attendance of any timing providing a modest advantage over non-attendance primarily through access to out-of-region career pathways. This variable will be excluded from later analytical stages given its cohort restriction and overlap with attended.ps.

3.  attended.ps (V=0.285 / 0.164 / 0.095) Postsecondary attendance shows a moderate and declining association driven primarily by its role as an academic orientation indicator. Its most durable signal is in meaningful MN employment — postsecondary attenders show consistently higher out-of-region employment through Year 10 — confirming that attending postsecondary, even without completing a credential, provides some access to out-of-region career pathways. Its primary analytical role is as a predictor in the non-college completer multivariate models.

4.  ps.grad.InstitutionSector (V=0.184 / 0.150 / 0.117) Institution sector reveals that the college completer pathway operates very differently by institution type. Public 2-year graduates show the strongest regional employment at all three time points (55.1%, 41.4%, 44.0%), while private not-for-profit 4-year graduates show the highest no-MN-record rates (26.0%, 22.4%, 41.8%) and meaningful MN employment. The sector findings directly anticipate the ps.pathway combined variable structure for the multivariate stages and confirm the importance of disaggregating postsecondary pathways by both credential type and institution selectivity.

5.  ps.grad (V=0.176 / 0.119 / 0.120) Postsecondary graduation shows a stable moderate association across all three time points — unusual in its non-declining trajectory. The key finding is that credential completion is specifically associated with meaningful MN employment rather than regional employment, which converges between graduates and non-graduates by Year 10. Non-graduates show persistently elevated non-meaningful employment — the local underutilization disconnection signature — suggesting that postsecondary attendance without completion carries a durable labor market penalty.

6.  highest.cred.level (V=0.171 / 0.138 / 0.112) Credential level reveals a non-linear relationship with regional employment that is one of the most important findings in the postsecondary group. Sub-baccalaureate credential completion — the less than bachelor's category — is most strongly associated with regional employment at all three time points (52.5%, 47.6%, 40.6%), while bachelor's or higher completion is most strongly associated with meaningful MN employment and the highest no-MN-record rates. This non-linearity confirms that the college completer pathway bifurcates sharply by credential level and reinforces the sub-baccalaureate versus baccalaureate distinction built into the ps.pathway variable structure.

CROSS-CUTTING THEMES

The postsecondary group tells a coherent story that organizes around three mechanisms. First, sub-baccalaureate credential completion in the region is the most direct postsecondary route to sustained regional employment — more reliable than four-year completion and more durable than workforce entry without a credential. Second, four-year credential completion outside the region is the primary driver of geographic departure from Minnesota, with outside-MN graduation producing a 54.2% no-MN-record rate at Year 10 that should be interpreted as out-of-state career establishment rather than disconnection. Third, postsecondary attendance without completion produces the most persistently negative outcomes — elevated non-meaningful employment and moderate no-MN-record rates — representing a failed attempt at the college completer pathway that leaves graduates without either the credential advantages of completers or the early workforce advantages of those who never attempted postsecondary.

The confirmed redundancy across attended.ps, ps.grad, highest.cred.level, ps.grad.location, and ps.grad.InstitutionSector — visible in both the MCA and across the bivariate stage — reinforces the framework decision to consolidate these variables into the ps.pathway combined variable for multivariate modeling rather than entering them simultaneously. The bivariate findings here provide the substantive foundation for how ps.pathway categories should be defined and interpreted in Stages 7 and 8.

<br>

## `attended.ps.within.first.year.hsgrad`

This variable has three categories: Did not attend ps (never attended or attended only to withdraw), No (attended postsecondary but not within the first year), and Yes (attended within the first year). The attending ps outcome category is excluded from these tables given that virtually all postsecondary attendees are captured within the independent variable itself. Note that the analysis covers approximately 52,000 graduates rather than the full population due to exclusion of the NA group. The association weakens over time (V=0.254 at Year 1, V=0.118 at Year 5, V=0.074 at Year 10).

At Year 1, graduates who attended postsecondary within the first year show dramatically higher rates of meaningful regional employment (50.0%) and meaningful MN employment (15.9%) than those who did not attend ps (22.2% and 7.6%) or who attended later (20.4% and 7.1%). The no-MN-record rate is very low for immediate enrollees (7.4%) compared to delayed enrollees (27.1%) and non-attenders (29.5%). The delayed enrollment group (No) shows a pattern very similar to the did-not-attend group at Year 1, suggesting that delayed postsecondary entry provides little early workforce advantage.

By Year 5, the immediate enrollment advantage in meaningful regional employment narrows but persists (35.5% vs. 35.0% for delayed and 28.1% for non-attenders). Immediate enrollees show substantially higher meaningful MN employment (23.1% vs. 17.6% for delayed and 13.4% for non-attenders) and lower no-MN-record rates (21.3% vs. 22.9% for delayed and 33.7% for non-attenders). The delayed enrollment group begins to separate more clearly from non-attenders at this point.

By Year 10, differences across groups narrow considerably (V=0.074). Meaningful regional employment is essentially equivalent across immediate and delayed enrollees (30.3% vs. 27.7%) but both remain above non-attenders (28.3%). The most persistent gap is in meaningful MN employment — immediate enrollees (23.4%) and delayed enrollees (18.3%) both exceed non-attenders (15.0%), consistent with postsecondary attendance of any timing providing some access to out-of-region career pathways. Non-attenders show the highest no-MN-record rate (38.9%) followed by delayed enrollees (34.0%) and immediate enrollees (31.5%).

The overall pattern confirms that timing of postsecondary enrollment matters most at Year 1, where immediate enrollment is associated with early workforce entry alongside postsecondary attendance. By Year 10, postsecondary attendance of any timing provides a modest advantage in sustained employment outcomes relative to non-attendance, but the distinction between immediate and delayed enrollment largely disappears.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps.within.first.year one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(grad.year.1 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/attended-ps-within-first-year-hsgrad-1.html")
```

### Five years after graduating

```{r post-secondary attended.ps.within.first.year five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(grad.year.5 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/attended-ps-within-first-year-hsgrad-5.html")
```

### Ten years after graduating

```{r post-secondary attended.ps.within.first.year ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(grad.year.10 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/attended-ps-within-first-year-hsgrad-10.html")
```
:::

<br>

## `attended.ps`

Postsecondary attendance shows a moderate and declining association with workforce outcomes (V=0.285 at Year 1, V=0.164 at Year 5, V=0.095 at Year 10). Note that the attending ps outcome category is excluded from these tables, and the analysis covers approximately 52,000 graduates at Year 1 with the NA group removed.

At Year 1, the contrast between attenders and non-attenders is sharp. Among those not attending postsecondary at Year 1, postsecondary attenders show substantially higher rates of meaningful regional employment (42.8% vs. 22.2%) and meaningful MN employment (13.8% vs. 7.6%), and much lower no-MN-record rates (12.2% vs. 29.5%). This seemingly counterintuitive pattern — postsecondary attenders showing higher workforce attachment at Year 1 — reflects the fact that this analysis excludes those currently attending postsecondary, so the attender group here consists of those who attended and then entered the labor market, likely credential completers or early exiters.

By Year 5, postsecondary attenders continue to show higher meaningful regional employment (35.5% vs. 28.1%) and substantially higher meaningful MN employment (22.7% vs. 13.4%). Non-attenders show higher no-MN-record rates (33.7% vs. 21.4%) and higher non-meaningful employment (24.8% vs. 20.4%).

By Year 10, the association weakens considerably (V=0.095). Meaningful regional employment is nearly equivalent (30.1% vs. 28.3%). The most durable gap is in meaningful MN employment (23.0% vs. 15.0%) and the no-MN-record rate (31.7% vs. 38.9%), consistent with postsecondary attendance providing sustained access to out-of-region career pathways while non-attendees show higher rates of both regional workforce attachment and geographic absence from Minnesota's formal labor market.

The attended.ps variable will serve an important role in the non-college completer multivariate models as an academic orientation indicator — capturing whether graduates attempted postsecondary even if they did not complete a credential. Its long-run association with meaningful MN employment confirms it is picking up postsecondary orientation effects that persist well beyond the credential completion window.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, attended.ps) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(grad.year.1 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/atteded-ps-1.html")
```

### Five years after graduating

```{r post-secondary attended.ps five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, attended.ps) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(grad.year.5 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/atteded-ps-5.html")
```

### Ten years after graduating

```{r post-secondary attended.ps ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, attended.ps) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(grad.year.10 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/atteded-ps-10.html")
```
:::

<br>

## `ps.grad`

Postsecondary graduation status shows a moderate association at Year 1 (V=0.176) that remains relatively stable at Years 5 and 10 (V=0.119 and V=0.120) — an unusual pattern of stability more reminiscent of CTE achievement than the declining trajectory seen for most postsecondary variables. Note that the attending ps outcome category is excluded from these tables and the Year 10 table collapses the attending ps ps.grad category due to near-zero cell counts.

At Year 1, postsecondary graduates show the highest rates of meaningful regional employment (49.4%) and meaningful MN employment (15.3%) of any ps.grad category, and the lowest no-MN-record rate (10.5%). The No group — those who attended but did not graduate — shows the weakest outcomes, with the highest non-meaningful employment (40.0%) and no-MN-record rates (25.3%). The attending ps group shows a pattern similar to graduates, with high regional (45.2%) and MN employment (18.4%) — likely reflecting concurrent enrollment and part-time employment rather than credential completion effects.

By Year 5, graduates continue to show higher meaningful MN employment (26.1% vs. 14.9% for non-graduates) but regional employment differences narrow (35.1% vs. 31.7%). Non-graduates show substantially higher non-meaningful employment (25.3% vs. 17.3%) and no-MN-record rates (28.1% vs. 21.5%). The attending ps group remains small (n=765) and shows outcomes broadly similar to graduates.

By Year 10, the most striking finding is the near-convergence of meaningful regional employment across graduates and non-graduates (28.8% vs. 30.6%). The durable gap is in meaningful MN employment — graduates show 25.3% compared to 16.6% for non-graduates — and in non-meaningful employment (13.0% vs. 18.7%). The no-MN-record rate is essentially equivalent (32.8% vs. 34.0%).

The long-run picture confirms that postsecondary graduation is specifically associated with sustained out-of-region employment rather than regional employment, which converges regardless of credential completion by Year 10. Non-graduates show persistently elevated non-meaningful employment — the local underutilization disconnection signature — suggesting that postsecondary attendance without completion carries a durable labor market penalty relative to both credential completers and those who never attended.

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, ps.grad) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(grad.year.1 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-1.html")
```

### Five years after graduating

```{r post-secondary ps.grad five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, ps.grad) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(grad.year.5 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-5.html")
```

### Ten years after graduating

```{r post-secondary ps.grad ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, ps.grad) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(grad.year.10 != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-10.html")
```
:::

<br>

## `ps.grad.InstitutionSector`

Institution sector shows moderate associations across all three time points (V=0.184 at Year 1, V=0.150 at Year 5, V=0.117 at Year 10). The five retained categories — public 4-year (1), multiple sectors (10), other/collapsed (11), private not-for-profit 4-year (2), and public 2-year (4) — show markedly different workforce outcome profiles that clarify the postsecondary pathway picture considerably.

At Year 1, among those not currently attending postsecondary, public 2-year graduates (sector 4) show the highest meaningful regional employment rate (55.1%), followed closely by public 4-year graduates (52.4%) and multiple sector graduates (59.3%). Private not-for-profit 4-year graduates (sector 2) show by far the lowest regional employment (17.8%) and the highest non-meaningful employment (46.7%), suggesting early credential incompletion or delayed workforce entry. The did-not-grad and never-attended groups show the expected lower employment rates and elevated non-meaningful and no-MN-record rates.

By Year 5, the sector differentiation sharpens considerably. Public 2-year graduates continue to show the highest meaningful regional employment (41.4%) — substantially above public 4-year (19.8%) and private not-for-profit 4-year (15.1%) graduates. Public 4-year graduates show the highest continued postsecondary attendance (33.0%), consistent with longer credential timelines. Private not-for-profit 4-year graduates show the highest meaningful MN employment (22.0%) and the highest no-MN-record rate (22.4%), suggesting both greater out-of-region career success and greater geographic mobility. The multiple sectors group shows notably high postsecondary attendance (48.3%) at Year 5, likely reflecting sequential enrollment patterns.

By Year 10, public 2-year graduates maintain the highest meaningful regional employment (44.0%) — more than double the rate for private not-for-profit 4-year graduates (16.5%). Public 4-year graduates show substantial meaningful MN employment (27.8%) and the highest no-MN-record rate (35.5%), consistent with four-year credential completion driving out-of-region and potentially out-of-state career trajectories. Private not-for-profit 4-year graduates show the highest no-MN-record rate among all sectors (41.8%), suggesting the strongest geographic mobility of any credential type.

The institution sector findings provide the clearest evidence yet that the college completer pathway operates differently depending on credential type and institution selectivity. Public 2-year completion is strongly associated with regional workforce attachment, while 4-year completion — particularly at private not-for-profit institutions — is associated with out-of-region employment and geographic mobility. This distinction anticipates the ps.pathway variable structure for the multivariate stages and reinforces the importance of disaggregating postsecondary pathways by both credential level and graduation location.

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.InstitutionSector one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, ps.grad.InstitutionSector) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(grad.year.1 != "Attending ps") %>%
  filter(ps.grad.InstitutionSector != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-InstitutionSector-1.html")
```

### Five years after graduating

```{r post-secondary ps.grad.InstitutionSector five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, ps.grad.InstitutionSector) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-InstitutionSector-5.html")
```

### Ten years after graduating

```{r post-secondary ps.grad.InstitutionSector ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, ps.grad.InstitutionSector) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-InstitutionSector-10.html")
```
:::

<br>

## `highest.cred.level`

Highest credential level shows a moderate and declining association with workforce outcomes (V=0.171 at Year 1, V=0.138 at Year 5, V=0.112 at Year 10). The four categories — no college, some college, less than bachelor's, and bachelor's or higher — reveal a complex and non-linear pattern in which sub-baccalaureate credential completion is most strongly associated with regional employment, while bachelor's or higher completion is most strongly associated with out-of-region employment.

At Year 1, among those not currently attending postsecondary, less than bachelor's credential holders show the highest meaningful regional employment (52.5%), followed by bachelor's or higher graduates (45.4%), some college (36.9%), and no college (22.2%). The less than bachelor's group also shows the lowest no-MN-record rate (8.5%), while the no-college group shows the highest (29.5%). Meaningful MN employment increases monotonically with credential level — from 7.6% for no college to 16.3% for bachelor's or higher.

By Year 5, the credential gradient in meaningful regional employment sharpens. Less than bachelor's holders reach 47.6% — the highest of any credential category — while bachelor's or higher graduates show only 29.0%, actually below the some college group (35.9%). Bachelor's or higher graduates show by far the highest meaningful MN employment (29.1% vs. 20.2% for less than bachelor's and 16.6% for some college), and a relatively high no-MN-record rate (24.5%) suggesting substantial out-of-state migration.

By Year 10, the non-linear credential pattern is fully consolidated. Less than bachelor's holders show the highest meaningful regional employment (40.6%), followed by some college (32.8%), no college (28.3%), and bachelor's or higher (24.0%). Bachelor's or higher graduates show the highest meaningful MN employment (27.6%) and the highest no-MN-record rate (36.3%), consistent with four-year credential completion driving both out-of-region career success and out-of-state migration. Non-meaningful employment declines monotonically with credential level (17.8%, 19.5%, 15.5%, 12.1%), with bachelor's or higher showing the lowest rate.

The non-linear relationship between credential level and regional employment is the central finding for this variable. Sub-baccalaureate credential completion — the less than bachelor's category, which includes associate degrees and technical certificates — is most strongly associated with sustained regional employment at every time point. This directly supports the institution sector findings and confirms that the college completer pathway operates through very different mechanisms depending on whether the credential is sub-baccalaureate or baccalaureate in nature.

::: panel-tabset
### One year after graduating

```{r post-secondary highest.cred.level one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, highest.cred.level) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(grad.year.1 != "Attending ps") %>%
  filter(highest.cred.level != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/highest-cred-level-1.html")
```

### Five years after graduating

```{r post-secondary highest.cred.level five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, highest.cred.level) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(grad.year.5 != "Attending ps") %>%
  filter(highest.cred.level != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/highest-cred-level-5.html")
```

### Ten years after graduating

```{r post-secondary highest.cred.level ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, highest.cred.level) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(grad.year.10 != "Attending ps") %>%
  filter(highest.cred.level != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/highest-cred-level-10.html")
```
:::

<br>

## `ps.grad.location`

Graduation location is the strongest postsecondary variable examined and shows an unusual pattern of strengthening association over time (V=0.164 at Year 1, V=0.208 at Year 5, V=0.171 at Year 10). It is also the variable most directly relevant to the college completer pathway framework, where graduation location serves as the critical hinge determining whether credential completion translates into regional or out-of-region workforce attachment.

At Year 1, regional graduates show by far the highest meaningful regional employment (63.0%) — more than double the rate for MN graduates (32.6%) and outside MN graduates (30.9%). Regional graduates also show the lowest no-MN-record rate (5.3%) and lowest non-meaningful employment (18.6%). MN graduates show notably high meaningful MN employment (26.3%), consistent with out-of-region but in-state credential completion translating into in-state employment. Outside MN graduates show an elevated no-MN-record rate (27.3%) at Year 1, likely reflecting concurrent enrollment or recent graduation rather than sustained out-of-state employment.

By Year 5, the graduation location signal is at its strongest. Regional graduates maintain a dominant meaningful regional employment advantage (52.7% vs. 23.9% for MN graduates and 21.4% for outside MN graduates). MN graduates show a striking meaningful MN employment rate (43.4%) — by far the highest of any location category — confirming that in-state out-of-region graduation is strongly associated with sustained in-state out-of-region employment. Outside MN graduates show the highest no-MN-record rate of any credential group at Year 5 (45.7%), consistent with sustained out-of-state employment following out-of-state credential completion.

By Year 10, the patterns consolidate further. Regional graduates maintain the highest meaningful regional employment (43.2%) — nearly twice the rate for MN graduates (20.5%) and outside MN graduates (17.8%). MN graduates continue to show the highest meaningful MN employment (37.9%), and outside MN graduates show the highest no-MN-record rate of any category (54.2%) — more than half of outside-MN credential completers have no Minnesota employment record a full decade after high school graduation. Did-not-graduate graduates show the highest non-meaningful employment (18.7%) and a substantial no-MN-record rate (34.0%).

Graduation location is the clearest expression of the college completer pathway's geographic sorting mechanism in the entire analysis. Regional graduation produces regional employment; MN graduation produces MN employment; outside MN graduation produces geographic absence from Minnesota's formal labor market. The outside MN no-MN-record rate of 54.2% at Year 10 is not a disconnection indicator but rather the most direct evidence that a substantial portion of Southeast Minnesota's high school graduates are successfully establishing careers outside the state following out-of-state credential completion. This distinction will be essential to the narrative synthesis.

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.location one year}
data <- meaningful.emp.southern %>%
  select(grad.year.1, ps.grad.location) %>%
  filter(grad.year.1 != "After 2023",
         grad.year.1 != "Attending ps",
         ps.grad.location != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-location-1.html")
```

### Five years after graduating

```{r post-secondary ps.grad.location five year}
data <- meaningful.emp.southern %>%
  select(grad.year.5, ps.grad.location) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         ps.grad.location != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-location-5.html")
```

### Ten years after graduating

```{r post-secondary ps.grad.location ten year}
data <- meaningful.emp.southern %>%
  select(grad.year.10, ps.grad.location) %>%
  filter(grad.year.10 != "After 2023",
         grad.year.10 != "Attending ps",
         ps.grad.location != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_local_employment_ct/Post-secondary/ps-grad-location-10.html")
```
:::

<br>

# Summary

**OVERVIEW**

The bivariate analysis of Southeast Minnesota high school graduates across four variable groups — demographics, high school characteristics, high school accomplishments, and postsecondary — reveals a consistent and coherent structure that maps onto the three-pathway framework with considerable clarity. The findings are organized around two fundamental sorting mechanisms: postsecondary orientation, which drives geographic departure from the region, and CTE engagement, which drives regional workforce attachment. A third group — those who neither complete credentials nor achieve sustained employment — is defined primarily by the absence of both mechanisms.

**WORKFORCE OUTCOME DISTRIBUTION**

The baseline distribution establishes the interpretive context for all that follows. At Year 1, the population is dominated by postsecondary attendance (54.4%), reflecting the high educational orientation of Southeast Minnesota graduates. By Year 5, postsecondary attendance contracts sharply to 18.5% as graduates complete or exit credentials, and meaningful regional employment rises to 27.2%. By Year 10, postsecondary attendance is essentially zero, meaningful regional employment reaches 29.7%, and the most notable feature is the large and growing no-MN-record group (33.4%). This group is the central interpretive challenge of the analysis — it is heterogeneous, comprising both successful out-of-state migrants and genuinely disconnected individuals, and its composition varies systematically across independent variables in ways that distinguish these two populations.

**VARIABLE GROUP SUMMARIES**

Demographics produced moderate associations, strongest at Year 1 where postsecondary sorting dominates the distribution. Special education status and economic disadvantage are the strongest demographic predictors, both associated with reduced postsecondary participation and elevated non-meaningful employment. Language minority variables — english.learner, non.english.home, and LimitedEnglishProficiencyIndicator — show a consistent and analytically interesting anomaly: elevated meaningful MN employment at Year 1, likely reflecting concentration in specific industries or urban labor markets. The most important demographic finding at Year 10 is the elevated no-MN-record rate for racial and ethnic minority graduates, particularly Asian/PI (45.1%), American Indian (41.1%), and Black (39.7%) graduates — a pattern more consistent with out-of-state migration following postsecondary completion than with genuine disconnection.

High school characteristics is the weakest variable group in the analysis, with the partial exception of average local unemployment rate. Community demographic description and EDR show negligible practical associations with workforce outcomes, and average wages as a percent of state average adds no meaningful explanatory power. The unemployment rate finding is substantively meaningful: stronger local labor markets anchor graduates regionally, while higher local unemployment is associated with out-of-region employment or labor force marginality — a long-run mechanism that becomes clearer at Years 5 and 10. The near-absence of place-based effects in this group suggests that individual-level characteristics and postsecondary pathways carry the explanatory weight in Southeast Minnesota.

High school accomplishments is the richest variable group and contains the strongest predictors in the entire analysis. The group divides into two distinct clusters with opposing relationships to workforce outcomes. The academic orientation cluster — took.ACT, ACT composite score, ap.exam, MCA Reading, MCA Math, MCA Science, and PSEO participation — is strongly and consistently associated with postsecondary attendance, credential completion, and sustained out-of-region employment. Higher academic achievement at the extreme is associated with geographic departure from Minnesota entirely — the no-MN-record group at Year 10 has among the highest ACT and MCA scores of any outcome category. The CTE engagement cluster — cte.achievement and total.cte.courses.taken — shows the opposite pattern: a stable, monotonic, dose-responsive association with meaningful regional employment that persists across all three time points with essentially no decay. CTE concentrators show 34.9% meaningful regional employment at Year 10; No CTE graduates show 22.2%. This 12.7 percentage point gap, stable from Year 1 through Year 10, is the single most consistent finding in the entire bivariate analysis.

The postsecondary group clarifies the mechanisms through which postsecondary pathways translate into workforce outcomes. The central finding is the non-linear relationship between credential level and regional employment — sub-baccalaureate completion is most strongly associated with regional workforce attachment, while baccalaureate and higher completion is most strongly associated with out-of-region employment and geographic departure. Graduation location is the strongest postsecondary variable and the only one that strengthens over time: regional graduation produces regional employment, MN graduation produces MN employment, and outside MN graduation produces a 54.2% no-MN-record rate at Year 10 — the clearest single data point establishing that geographic absence from Minnesota's labor market reflects out-of-state career establishment among college completers rather than disconnection.

## THE THREE PATHWAYS

The bivariate evidence supports a clear three-pathway structure.

The college completer pathway is defined by academic orientation in high school — ACT participation, high MCA scores, AP and PSEO participation — followed by postsecondary enrollment and credential completion. The critical hinge within this pathway is graduation location. Regional sub-baccalaureate completion produces sustained regional employment and is the postsecondary route most closely aligned with the workforce-first pathway. Out-of-region completion — particularly four-year completion outside Minnesota — produces sustained out-of-state employment and accounts for a substantial portion of the no-MN-record population at Year 10.

The workforce-first pathway is defined by CTE engagement in high school and direct entry into the regional labor market. It is the most stable pathway in the analysis — its defining variable (total CTE courses taken) shows the most consistent dose-response relationship across all three time points of any predictor examined. CTE concentrators and completors achieve meaningful regional employment at rates that exceed all other high school accomplishment profiles by Year 5 and Year 10. This pathway is not defined by academic disengagement — it coexists with moderate academic achievement — but by a deliberate orientation toward vocational preparation and regional labor market entry.

The disconnected pathway is defined by the simultaneous absence of both mechanisms — neither substantial CTE engagement nor postsecondary credential completion. It manifests as two distinct signatures. The local underutilization signature — present in the labor market but below the sustained employment threshold — is most strongly associated with special education status, economic disadvantage, and the No CTE group, and persists through Year 10 as non-meaningful employment. The geographic absence signature — no MN employment record — is the more complex of the two: it is driven partly by successful out-of-state migration among high academic achievers and partly by genuine disconnection among low-CTE, low-achievement graduates. The No CTE group shows a 39.3% no-MN-record rate at Year 10 — the highest of any CTE category and substantially above what can be explained by out-of-state migration alone — confirming that absence of CTE engagement is the defining characteristic of the genuine disconnection signature.

**IMPLICATIONS FOR MULTIVARIATE STAGES**

The bivariate analysis establishes clear expectations for the multivariate stages. CTE course-taking should emerge as the dominant predictor of sustained regional employment in the non-college completer models, replicating the Southwest Minnesota benchmark (OR≈1.055 per course at Years 1, 5, and 10). Graduation location should emerge as the dominant structural variable in the full population model, with regional sub-baccalaureate graduation showing the strongest regional employment odds. Academic achievement variables — particularly MCA Reading and took.ACT — should show negative associations with regional employment net of postsecondary pathway, reflecting the sorting function of academic orientation toward out-of-region careers. The unemployment rate should show a modest but consistent negative association with regional employment across both model sets. Demographic variables are expected to show attenuated effects in multivariate context given their substantial overlap with high school accomplishment and postsecondary pathway variables, with economic disadvantage and special education status the most likely to retain independent associations.

<br>

# VARIABLE REVIEW PRIOR TO MULTIVARIATE ANALYSIS

**VARIABLES TO EXCLUDE — ALREADY ESTABLISHED BY FRAMEWORK**

The following exclusions are already built into the analytical framework and confirmed by the bivariate findings:

`MCA.S`: Voluntary and self-selected assessment excluding roughly 30% of the population. Its bivariate pattern was essentially identical to MCA.R and MCA.M throughout, adding no independent information.

`non.english.home`: Redundant with english.learner in both the MCA and bivariate stages. Weaker effect sizes at every time point. Excluded from multivariate models but retained for descriptive purposes.

`avg.wages.pct.state`: Negligible practical effect sizes throughout despite statistical significance. Confirmed as analytically meaningless.

`pseo.participant`: Cohort restriction (2011+) limits coverage. Confirmed as a modest predictor with effect sizes well below other accomplishment variables.

`ap.exam`: Excluded from non-college completer multivariate models due to small group size in that subpopulation. Retained in full population models.

**VARIABLES TO EXCLUDE — SUPPORTED BY BIVARIATE FINDINGS**

`LimitedEnglishProficiencyIndicator`: The weakest demographic predictor throughout (V=0.079 / 0.032 / 0.031), adding no explanatory power beyond english.learner. Its MCA loading on a separate dimension suggested it captures a distinct but narrow subpopulation. Not worth the degrees of freedom in a multivariate model.

`attended.ps.within.first.year.hsgrad`: Overlaps structurally with attended.ps and becomes essentially indistinguishable from it by Year 5. Its Year 1 signal is largely an artifact of concurrent enrollment timing rather than a substantive independent predictor. Exclude from all multivariate stages.

`Dem_Desc`: The weakest variable in the high school characteristics group (V=0.041 / 0.046 / 0.030) and essentially flat across time. Its bivariate associations are negligible in practical terms and it is likely to be absorbed entirely by edr in a multivariate context given their geographic overlap. Not worth including.

VARIABLES TO EXAMINE IN MULTIVARIATE CONFIRMATION ANALYSIS

- Demographics
  - `Gender`
  - `economic.status`
  - `SpecialEdStatus`
  - `RaceEthnicity`
- High school characteristics
  - `edr`
  - `avg.unemp.rate`
- High school accomplishments
  - `took.ACT`
  - `ACTCompositeScore`
  - `MCA.R`
  - `cte.achievement`
  - `total.cte.courses.taken`
- Post-secondary
  - `attended.ps`
  - `ps.grad`
  - `ps.grad.location`
  - `ps.grad.InstitutionSector`
  - `highest.cred.level`

<br>

```{r create new master after analysis}
master.after.ct <- meaningful.emp.southern %>%
  select(PersonID, Gender, economic.status, SpecialEdStatus, RaceEthnicity, edr, avg.unemp.rate, took.ACT, ACTCompositeScore, MCA.R, cte.achievement, total.cte.courses.taken, attended.ps, ps.grad, ps.grad.location, ps.grad.InstitutionSector, highest.cred.level, grad.year.1, grad.year.5, grad.year.10)
  
write_csv(master.after.ct, "Data/SLEDS/Masters/After analysis/Master-after-ct.csv")
```
