---
title: "Multiple Correspondence Analysis"
editor: visual
---

```{r library, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(leaps)
library(bestglm)
library(fastDummies)
library(car)
library(FactoMineR)
library(factoextra)
library(psych)
library(corrplot)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         mif = ifelse(is.na(mif), "TC", mif),
         mif = as.factor(mif),
         mif = fct_relevel(mif, "NW", "NE", "WC", "EC", "SW", "SE", "TC"),
Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))



color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master original}
master.original <- read.csv("Data/SLEDS/Masters/Master.csv") %>%
  drop_na(Dem_Desc)
```

<br>

The purpose of multiple correspondence analysis is to examine any patterns among the independent variables with the goal of eliminating variables that are redundant - essentially two variables that are explaining the same thing. This type of analysis is not meant to confirm any predictions, but rather explore the data and make decisions on what data should be included in the prediction model.

<br>

# Independent variables

The master dataset is composed of `r comma(nrow(master.original))` rows. Each row represents an individual that graduated (StatusEnd = 8 or 9) from a high school located in the Central region (defined by project) between 2008 and 2023.

The master dataset is also composed of `r comma(ncol(master.original))` columns. Columns mainly represent various socio-economic, demographic, and educational career characteristics of the individual as well as a few characteristics of the high school geographic from which they graduated (i.e. county unemployment). The variables for each individual is composed of a mix of categorical and numerical values. In this analysis, we will only be exploring categorical data.

Here is a list of all the categorical variables with a description of each.

<br>

```{r primary independent variables}
qual.var.names <-  c("PersonID", "Gender", "LimitedEnglishProficiencyIndicator", "HomelessIndicator", "economic.status", "pseo.participant", "SpecialEdStatus", "non.english.home", "RaceEthnicity", "Dem_Desc", "edr", "took.ACT", "ap.exam", "cte.achievement", "english.learner", "sat.taken", "attended.ps.within.first.year.hsgrad", "attended.ps", "ps.grad", "ps.grad.InstitutionSector", "highest.cred.level", "ps.grad.location")

quan.var.names <- c("total.cte.courses.taken", "cte.0", "cte.1", "cte.2", "cte.3", "cte.4", "cte.5", "cte.6", "cte.7", "cte.8", "cte.9", "cte.10", "cte.11", "cte.12", "cte.14", "cte.21", "cte.22", "cte.23", "cte.24", "cte.25", "cte.26", "avg.cte.intensity", "MCA.M", "MCA.R", "MCA.S", "avg.unemp.rate", "wages.3year.avg", "avg.wages.pct.state")

independent.var.names <- c("hs.grad.year", "Gender", "LimitedEnglishProficiencyIndicator", "non.english.home", "HomelessIndicator", "economic.status", "pseo.participant", "SpecialEdStatus", "non.english.home", "RaceEthnicity", "Dem_Desc", "edr", "took.ACT", "ap.exam", "total.cte.courses.taken", "cte.achievement", "avg.cte.intensity", "MCA.M", "MCA.R", "MCA.S", "sat.taken", "attended.ps", "ps.grad", "highest.cred.level", "ps.grad.location", "avg.unemp.rate", "wages.3year.avg")

cte.cc.var.names <- c("cte.1", "cte.2", "cte.3", "cte.4", "cte.5", "cte.6", "cte.7", "cte.8", "cte.9", "cte.10", "cte.11", "cte.12", "cte.14", "cte.21", "cte.22", "cte.23", "cte.24", "cte.25")

ind.qual.var <- master.original %>%
  select(all_of(`qual.var.names`)) %>%
  mutate_at(2:ncol(.), as.factor)

```

<br>

- **Demographics**
  - *`Gender`:*
    - M = male
    - F = female
  - *`RaceEthnicity`:*
    - AI = American Indian
    - Asian/PI = Asian or Pacific Islander
    - Black = African American
    - Hispanic = any race any Hispanic or Latino
    - Unknown = unknown race or ethnicity
    - White = Caucasian and not Hispanic
  - *`LimitedEnglishProficiencyIndicator`:*
    - Y = identified as having limited English proficiency at some point between 10th and 12th grade, otherwise N.
  - *`HomelessIndicator`:*
    - Y = the individual was identified as homeless at any point between 10th and 12th grade, otherwise N.
  - *`economic.status`:*
    - 1 = the individual was eligible for free or reduced lunch (codes 1,2,4,5) at any point between 10th and 12th grade, otherwise 0.
  - *`SpecialEdStatus`:*
    - 1 = the individual required special education services at some point between 10th and 12th grade, otherwise 0.
  - *`non.english.home`:*
    - 1 = the individual was identified as having English as not the primary language spoken at home between 10th and 12th grad, otherwise 0.
  - *`english.learner`:*
    - 1 = the individual was identified as an “English learner” at least one time between 10th and 12th grade, otherwise 0.
- **High school characteristics**
  - *`Dem_Desc`:* the RUCA category of the high school from which the individual graduated.
  - `edr`: there are two EDR's in this region - EDR 9 and EDR 10.
- **High school accomplishments**
  - *`pseo.participant`:*
    - 1 = the individual participated in a PSEO course sometime between 10th and 12th grade, otherwise 0. There are also NA values in this due to missing data.
  - *`took.ACT:`* whether an individual took the ACT exam
    - Yes
    - No
  - *`ap.exam`:*
    - 1 = the individual took an AP exam at some point,
    - 0 = the individual did not take an AP exam.
  - *`cte.achievement`:* three indicators with
    - “CTE concentrator or completor” = self-explanatory,
    - “CTE participant” = they took a CTE course but was not a concentrator or completor, and,
    - “No CTE” = they didn’t take a CTE course
  - *`sat.taken`:*
    - 1 = individual took the SAT at some point between 10th and 12th grade, otherwise 0.
- **Post-secondary**
  - *`attended.ps.within.first.year.hsgrad`:* the individual attended a post-secondary institution within the first year of graduating high school.
    - Yes = Attended PS within first year
    - No = Attended PS but not within first year
    - NA = never attended PS or attended but only to withdraw.
  - *`attended.ps`:*
    - Yes = the individual attended a post-secondary education institution.

    - No = never attended post-secondary, or attended but only “W” or “D”.
  - *`ps.grad`:*
    - Yes = Graduated from a post-secondary institution by 2023

    - Attending ps = attending a post-secondary institution as of 2023

    - No = Not attending a post-secondary institution nor graduated from one as of 2023.
  - *`ps.grad.InstitutionSector`:* the institution sector of the post-secondary school from which an individual has graduated.
    - 1 - Public, 4-year or above

    - 2 - Private not-for-profit, 4-year or above

    - 3 - Private for-profit, 4-year or above

    - 4 - Public, 2-year

    - 5 - Private not-for-profit, 2-year

    - 6 - Private for-profit, 2-year

    - 7 - Public, less-than 2-year

    - 8 - Private not-for-profit, less-than 2-year

    - 9 - Private for-profit, less-than 2-year

    - 10 - multiple sectors (if attended multiple institutions)

    - 99 - Sector unknown (not active)

    - “Never attended ps” = the individual never attended a post-secondary institution

    - “Did not grad” = the individual attended a post-secondary institution but did not graduate.
  - *`highest.cred.level`:* an identifier of the highest credential earned by the PersonID. The categories are;
    - Did not graduate ps

    - Attending ps

    - Less than associate degree

    - Associate degree

    - Bachelors degree

    - Higher than bachelors degree.
  - `ps.grad.location`: this is the following;
    - ps.grad.region: graduated from a college in the Southeaset region.

    - ps.grad.MN: graduated from a college in MN but not located in Southeast.

    - ps.grad.outside: graduated from a college outside MN

    - Did not grad: attended but did not graduate college or attended but did not graduate

    - Attending ps: Currently attending post-secondary at grad.year.x

<br>

# Frequencies

First, lets check to see there aren't any variable categories with very low frequencies which can distort the analysis.

**VARIABLES TO ELIMINATE OUTRIGHT**

1.  `HomelessIndicator`:
    - Only 0.9% of graduates (n=1,136) were ever identified as homeless.
    - Too sparse for meaningful MCA contribution or stable bivariate estimates.
    - Eliminate
2.  `sat.taken`
    - Only 1.8% of graduates (n=2,191) took the SAT.
    - Far too rare to distinguish from noise. Not part of the core framework.
    - Eliminate.

**Variables to consolidate**

- `ps.grad.InstitutionSector`: everal categories need collapsing before MCA
  - Sectors 5, 6, 7, 8, 9 are effectively empty (n=7, 52, 3, 1, 60 respectively).
  - Sector 3 (private for-profit 4-year) is marginal at n=939 (0.8%).
  - NA has only n=3.
  - The new categories will be the following;
    - 1 - Public, 4-year or above
    - 2 - Private not-for-profit, 4-year or above
    - 4 - Public, 2-year
    - 10 - multiple sectors (if attended multiple institutions)
    - 11 - Other type of college
      - 3 - Private for-profit, 4-year or above
      - 5 - Private not-for-profit, 2-year
      - 6 - Private for-profit, 2-year
      - 7 - Public, less-than 2-year
      - 8 - Private not-for-profit, less-than 2-year
      - 9 - Private for-profit, less-than 2-year
      - 99 - Sector unknown (not active)
    - “Never attended ps” = the individual never attended a post-secondary institution
    - “Did not grad” = the individual attended a post-secondary institution but did not graduate.
- `ps.grad.location` — "PS grad EDR" category is very small and I will consolidate those so they just read "PS grad region"
- `highest.cred.level`: I will collapse some of these categories and have the following;
  - Did not graduate ps
  - Less than a bachelor's
  - Bachelor's or higher

**VARIABLES TO WATCH BUT KEEP**

- `RaceEthnicity/AI`: n=642 (0.5%). Small but present in the framework; retain for bivariate stages with appropriate caution in interpretation. May be unstable in multivariate models.
- `LimitedEnglishProficiencyIndicator`: n=2,625 (2.1%) identified as Y. Small but not negligible. Keep for now; its relationship to english.learner (8.9%) and non.english.home (8.8%) will be assessed in the MCA.

<br>

```{r ind qual frequencies}
ind.qual.freq <- ind.qual.var %>%
  gather(key = "key", value = "value", 2:ncol(.)) %>%
  group_by(key, value) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / nrow(master.original),
         pct = percent(pct, accuracy = .1)) %>%
  mutate(data_id = seq(n())) %>%
  write_csv(., "Data/Claude/analysis_mca/variable-frequencies.csv")

ind.qual.freq.plot <- ggplot(ind.qual.freq, aes(value, n)) +
  facet_wrap(~key, ncol = 4, scales = "free") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(key, "\n", value, "\nN = ", comma(n, accuracy = 1), "\nPercent: ", pct, sep = ""))) +
  labs(x="", y = "", color="", title = "N for each independent, categorical variable")+
  scale_y_continuous(labels=scales::comma)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 5, "RdYlBu"),
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 14),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = ind.qual.freq.plot, width_svg = 10, height_svg = 20) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

```{r updated ind qual var}
master.ind.qual.var <- ind.qual.var %>%
  select(-PersonID, -HomelessIndicator, -sat.taken) %>%
  mutate(ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector %in% c("3", "5", "6", "7", "8", "9"), "11", as.character(ps.grad.InstitutionSector)),
         ps.grad.location = ifelse(ps.grad.location %in% c("PS grad region", "PS grad EDR"), "PS grad region", as.character(ps.grad.location)),
         ps.grad.location = as.factor(ps.grad.location),
         highest.cred.level = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Less than Bachelors",
                                     ifelse(highest.cred.level %in% c("Bachelor degree", "Master degree or higher"), "Bachelors or higher", as.character(highest.cred.level))),
         highest.cred.level = as.factor(highest.cred.level))
```

<br>

# MCA

For the MCA analysis, I'm going to group these variables together to make sure that they aren't explaining the same thing. The goal is to eliminate any variables that are essentially the same. This will help me make my model smaller when analyzing the pathways taken after graduating high school.

Here are the groupings I'm going to use;

- Demographics
  - `Gender`
  - `LimitedEnglishProficiency`
  - `economic.status`
  - `SpecialEdStatus`
  - `non.english.home`
  - `RaceEthnicity`
  - `english.learner`
  - `non.english.home`
- High school characteristics
  - `Dem_Desc`
  - `EDR`
- High school accomplishments
  - `pseo.participant`
  - `took.ACT`
  - `ap.exam`
  - `cte.achievement`
- Post-secondary
  - `attended.ps.within.first.year.hsgrad`
  - `attended.ps`
  - `ps.grad`
  - `ps.grad.InstitutionSector`
  - `highest.cred.level`
  - `ps.grad.location`

<br>

## Demographics

Below are the variables that are categorized as "demographics".

- Demographics
  - `Gender`
  - `LimitedEnglishProficiency`
  - `economic.status`
  - `SpecialEdStatus`
  - `non.english.home`
  - `RaceEthnicity`
  - `english.learner`
  - `non.english.home`

```{r demographics prep, echo=TRUE}
demographics <- master.ind.qual.var %>%
  select(Gender, LimitedEnglishProficiencyIndicator, economic.status, SpecialEdStatus, non.english.home, RaceEthnicity, english.learner, non.english.home) %>%
  filter(RaceEthnicity != "Unknown")

kable(head(demographics))

kable(names(demographics))
```

<br>

As expected, we have `r comma(nrow(demographics))` individuals in the dataset.

<br>

### MCA analysis

#### Eigenvalues

The eigenvalues measure the amount of variation retained by each principle component. This will help us determine how many principle components to be considered.

Dimension 1 explains 27.4% of variance. It takes 5 dimensions to reach 70% and 7 to reach 85%. No single dominant structure — demographic variables are capturing several independent axes of variation. No redundancy concerns.

<br>

```{r demographics eigenvalues}
demographics.mca <- FactoMineR::MCA(demographics, graph = FALSE)

eig.val <- get_eigenvalue(demographics.mca)

eig.val


```

<br>

#### COS^2^

All variables show total cos² above 0.43, most above 0.65. Solid representation across five dimensions. Nothing to eliminate.

\<br\>

```{r demogrpahics mca correlation between variables}
table <- demographics.mca$var$cos2 %>%
  as.data.frame() %>%
  select(`Dim 1`, `Dim 2`, `Dim 3`, `Dim 4`, `Dim 5`) %>%
  mutate(total = `Dim 1` + `Dim 2` + `Dim 3` + `Dim 4` + `Dim 5`) %>%
  write_csv(., "Data/Claude/analysis_mca/Demographics/cos2.csv")


  
datatable(table) %>%
  formatCurrency(1:5, "", digits = 3)


fviz_cos2(demographics.mca, choice = "var", axes = 1:5)

```

<br>

#### Contribution to dimensions

Dimension 1: Driven by non.english.home/english.learner and economic.status — a language/poverty axis.

Dimension 2: Dominated by SpecialEdStatus — stands largely alone.

Dimension 3: RaceEthnicity (Hispanic, Black) and LimitedEnglishProficiencyIndicator.

Dimension 4: RaceEthnicity (White, AI) and Gender.

Dimension 5: RaceEthnicity (AI) and Hispanic — capturing remaining minority group variation.

No variable contributes nothing across all five dimensions. No eliminations warranted.

<br>

```{r demographics mca contribution}
table <- demographics.mca$var$contrib %>%
  as.data.frame() %>%
  mutate(id = seq(n())) %>%
  select(id, `Dim 1`, `Dim 2`, `Dim 3`, `Dim 4`, `Dim 5`) %>%
  mutate(`Dim 1` = `Dim 1` / 100,
         `Dim 2` = `Dim 2` / 100,
         `Dim 3` = `Dim 3` / 100,
         `Dim 4` = `Dim 4` / 100,
         `Dim 5` = `Dim 5` / 100) %>%
  write_csv(., "Data/Claude/analysis_mca/Demographics/contribution-to-dimensions.csv")

datatable(table) %>%
  formatPercentage(2:6, digits = 1)
  

```

<br>

For demographics we will keep all variables.

- Eliminate - none

<br>

## High school characteristics

Below are the variables that are categorized as "high school characteristics".

- High school characteristics
  - `Dem_Desc`
  - `EDR`

```{r hs characteristics prep, echo=TRUE}
hs.char <- master.ind.qual.var %>%
  select(Dem_Desc, edr) 

kable(head(hs.char))

kable(names(hs.char))
```

<br>

We have `r comma(nrow(demographics))` individuals in the dataset.

<br>

### MCA analysis

#### Eigenvalues

The eigenvalues measure the amount of variation retained by each principle component. This will help us determine how many principle components to be considered.

Three dimensions exhaust the variance, which is expected given there are only two variables (Dem_Desc and edr). Dimension 1 explains 39% and the first two dimensions together explain 73%. Simple structure — no redundancy concerns.

<br>

```{r hs characteristics eigenvalues}
hs.char.mca <- FactoMineR::MCA(hs.char, graph = FALSE)

eig.val <- get_eigenvalue(hs.char.mca)

eig.val
```

<br>

#### COS^2^

Let's take a look to see which variables are going to be important. The chart and table below shows that all the variables are significant in the first two dimensions.

**Dimension 1:** More rural

**Dimension 2:** More urban

**Dimension 3:** ?????

```{r hs characteristics mca correlation between variables}
table <- hs.char.mca$var$cos2 %>%
  as.data.frame() %>%
  select(`Dim 1`, `Dim 2`, `Dim 3`) %>%
  mutate(total = `Dim 1` + `Dim 2` + `Dim 3`) %>%
  write_csv("Data/Claude/analysis_mca/High school characteristics/cos2.csv")
  
datatable(table) %>%
  formatCurrency(1:3, digits = 3, "")

fviz_cos2(hs.char.mca, choice = "var", axes = 1:2)


```

<br>

#### Contribution to dimensions

Dim_Desc drives Dimension 2 (entirely urban category dominant). EDR drives Dimensions 1 and 3. The two variables are capturing distinct geographic axes — urban/rural character versus EDR membership. No redundancy, no eliminations warranted.

<br>

```{r hs characteristics mca contribution}
table <- hs.char.mca$var$contrib %>%
  as.data.frame() %>%
  select(`Dim 1`, `Dim 2`, `Dim 3`) %>%
  mutate(`Dim 1` = `Dim 1` / 100,
         `Dim 2` = `Dim 2` / 100,
         `Dim 3` = `Dim 3` / 100) %>%
  write_csv(., "Data/Claude/analysis_mca/High school characteristics/contribution-to-dimensions.csv")

datatable(table) %>%
  formatPercentage(1:3, digits = 1)
```

<br>

So regarding any variables to eliminate in the dataset, I don't see one. It looks important to keep them all.

- Eliminate: none

<br>

## High school accomplishments

Below are the variables that are categorized as "high school accomplishments".

- High school accomplishments
  - `pseo.participant`
  - `took.ACT`
  - `ap.exam`
  - `cte.achievement`

```{r high school accomplishments prep, echo=TRUE}
hs.accomplishments <- master.ind.qual.var %>%
  select(pseo.participant, took.ACT, ap.exam, cte.achievement)

kable(head(hs.accomplishments))

kable(names(hs.accomplishments))
```

<br>

There are `r comma(nrow(demographics))` individuals in the dataset.

<br>

### MCA analysis

#### Eigenvalues

Dimension 1 explains 25.5% of variance. It takes 4 dimensions to reach 75% and all 6 to exhaust the variance. Moderate complexity — no single dominant structure. Consistent with a group of variables capturing distinct aspects of academic engagement. No redundancy concerns at this stage.

```{r high school accomplishments eigenvalues}
hs.accomplishments.mca <- FactoMineR::MCA(hs.accomplishments, graph = FALSE)

eig.val <- get_eigenvalue(hs.accomplishments.mca)

eig.val
```

<br>

#### COS^2^

All variables show total cos² above 0.52, most above 0.63. Solid representation across four dimensions. The one relatively lower value (took.ACT No at 0.526) is still acceptable. No eliminations warranted.

```{r high school accomplishments mca correlation between variables}
table <- hs.accomplishments.mca$var$cos2 %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  mutate(total = `Dim 1` + `Dim 2` + `Dim 3` + `Dim 4` + `Dim 5`) %>%
  write_csv(., "Data/Claude/analysis_mca/High school accomplishments/cos2.csv")
  
datatable(table) %>%
  formatCurrency(1:5, "", digits = 3)

fviz_cos2(hs.accomplishments.mca, choice = "var", axes = 1:5)
```

<br>

#### Contribution to dimensions

Dimension 1: Driven by took.ACT (No and Yes) — academic testing axis.

Dimension 2: Driven by pseo.participant NA and CTE Participant — separating non-PSEO-eligible cohorts from moderate CTE engagement.

Dimension 3: Dominated by No CTE — absence of CTE engagement stands apart.

Dimension 4: CTE Participant and pseo.participant NA — further separating CTE engagement levels.

Dimension 5: pseo.participant_1 and ap.exam_1 — advanced academic participation axis.

No eliminations warranted among retained variables.

<br>

```{r hs accomplishments mca contribution}
table <- hs.accomplishments.mca$var$contrib %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  mutate(`Dim 1` = `Dim 1` / 100,
         `Dim 2` = `Dim 2` / 100,
         `Dim 3` = `Dim 3` / 100,
         `Dim 4` = `Dim 4` / 100,
         `Dim 5` = `Dim 5` / 100) %>%
  write_csv(., "Data/Claude/analysis_mca/High school accomplishments/contribution-to-dimensions.csv")

datatable(table) %>%
  formatPercentage(2:6, digits = 1)
```

<br>

So regarding any variables to eliminate in the dataset, I don't see one. It looks important to keep them all.

- Eliminate: none

<br>

## Post secondary

Below are the variables that are categorized as "high school accomplishments".

- Post-secondary
  - `attended.ps.within.first.year.hsgrad`
  - `attended.ps`
  - `ps.grad`
  - `ps.grad.InstitutionSector`
  - `highest.cred.level`
  - `ps.grad.location`

```{r ps prep, echo=TRUE}
ps <- master.ind.qual.var %>%
  select(attended.ps.within.first.year.hsgrad, attended.ps, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location)

write_csv(ps, "Data/ps-data.csv")

kable(head(ps))

kable(names(ps))
```

<br>

There are `r comma(nrow(demographics))` individuals in the dataset.

<br>

### MCA analysis

#### Eigenvalues

The eigenvalues measure the amount of variation retained by each principle component. This will help us determine how many principle components to be considered.

Dimension 1 explains 24.8% of variance. It takes 5 dimensions to reach 72% and 8 to reach 88%. Dimensions 15-21 are effectively zero — these are artifacts of category redundancy among the postsecondary variables, which is expected given the structural overlap between attended.ps, ps.grad, highest.cred.level, and ps.grad.location. Complex but not surprising for this variable group. No concerns at this stage.

<br>

```{r ps eigenvalues}
ps.mca <- FactoMineR::MCA(ps, graph = FALSE)

eig.val <- get_eigenvalue(ps.mca)

eig.val
```

<br>

#### COS^2^

Let's take a look to see which variables are going to be important. From the total cos^2^ we can see that there are a few that just aren't make the cut (below .2);

Several categories show notably low total cos² across five dimensions and warrant attention:

ps.grad.InstitutionSector_10 (total = 0.071) and ps.grad.InstitutionSector_11 (total = 0.024) are poorly represented — the collapsed "other" category and "multiple sectors" are not well captured by the MCA dimensions. This is partly a consequence of their heterogeneity.

ps.grad.InstitutionSector_2 (total = 0.173) is also weak.

ps.grad.location categories — PS grad MN (0.210), PS grad outside MN (0.337), and PS grad region (0.467) — are all moderate to low, suggesting graduation location is not strongly structured relative to the other postsecondary variables.

attended.ps.within.first.year.hsgrad_No (total = 0.160) is very poorly represented.

Everything else is well represented (total \> 0.77). The weak representation of several ps.grad.InstitutionSector categories and attended.ps.within.first.year.hsgrad_No suggests these variables may not be contributing meaningfully to the MCA structure. Worth flagging before reviewing contributions.

<br>

```{r ps mca correlation between variables}
table <- ps.mca$var$cos2 %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  mutate(total = `Dim 1` + `Dim 2` + `Dim 3` + `Dim 4` + `Dim 5`)  %>%
  write_csv(., "Data/Claude/analysis_mca/Post secondary/cos2.csv")

  
datatable(table) %>%
  formatCurrency(1:6, "", digits = 3)

fviz_cos2(ps.mca, choice = "var", axes = 1:5)
```

<br>

#### Contribution to dimensions

Next, lets see how much each variable is contributing to each dimension. There are 29 categories so any percent above 3.4% would be quite influential. There are a few categories that don't reach that threshold for any of the four dimensions.

- 

<br>

```{r ps mca contribution}
table <- ps.mca$var$contrib %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  mutate(`Dim 1` = `Dim 1` / 100,
         `Dim 2` = `Dim 2` / 100,
         `Dim 3` = `Dim 3` / 100,
         `Dim 4` = `Dim 4` / 100,
         `Dim 5` = `Dim 5` / 100) %>%
  write_csv(., "Data/Claude/analysis_mca/Post secondary/contributions-to-dimensions.csv")

datatable(table) %>%
  formatPercentage(2:6, digits = 1)
```

<br>

Dimension 1: Driven by attended.ps/attended.ps_No and ps.grad_No — a non-participation axis.

Dimension 2: Dominated by ps.grad_Attending ps, highest.cred.level_Attending ps, ps.grad.location_Attending ps, and ps.grad.InstitutionSector_Attending ps — a current enrollment axis. These four categories are essentially the same construct appearing across multiple variables, confirming the expected redundancy.

Dimension 3: Driven by ps.grad.InstitutionSector_4 (public 2-year), highest.cred.level_Less than Bachelors, and highest.cred.level_Bachelors or higher — a credential level and institution type axis.

Dimension 4: Dominated entirely by ps.grad.InstitutionSector.NA and ps.grad.location.NA — an artifact of missing data, not a meaningful substantive dimension.

Dimension 5: Dominated by ps.grad.InstitutionSector_Did not grad — non-completion axis.

The redundancy across attended.ps, ps.grad, highest.cred.level, ps.grad.location, and ps.grad.InstitutionSector is clearly visible. These variables are structurally overlapping as expected. No eliminations needed for the MCA itself, but this confirms they should not all enter multivariate models simultaneously.

<br>

# SUMMARY AND RECOMMENDATIONS

**VARIABLES ELIMINATED PRIOR TO MCA**

Two variables were eliminated based on frequency review before the MCA was conducted. `HomelessIndicator` was dropped due to extremely low prevalence (0.9%), and `sat.taken` was dropped due to both low prevalence (1.8%) and its absence from the core analytical framework.

**DEMOGRAPHIC VARIABLES**

The demographic MCA produced ten dimensions, with Dimension 1 explaining 27.4% of variance. No single dominant structure emerged, indicating that demographic variables capture several relatively independent axes of variation. Dimension 1 reflected a language/poverty axis driven by `non.english.home`, `english.learner`, and `economic.status`. Dimension 2 was dominated by `SpecialEdStatus`. Dimensions 3 and 4 captured racial/ethnic variation. Notably, the three language-related variables — `english.learner`, `non.english.home`, and `LimitedEnglishProficiencyIndicator` — loaded on different dimensions despite their conceptual overlap, suggesting they capture meaningfully distinct subpopulations. All demographic variables showed adequate cos² representation. No eliminations are warranted.

Recommendation: Retain all demographic variables. Consistent with the established framework, `non.english.home` will be excluded from multivariate models due to redundancy with `english.learner`, but retained for bivariate stages.

**HIGH SCHOOL CHARACTERISTICS**

With only two variables (`Dem_Desc` and `edr`), three dimensions exhausted all variance. `Dem_Desc` drove Dimension 2, reflecting urban/rural character, while `edr` drove Dimensions 1 and 3, reflecting EDR membership. The two variables capture distinct geographic axes with no redundancy. Recommendation: Retain both variables.

**HIGH SCHOOL ACCOMPLISHMENTS**

Dimension 1 explained 25.5% of variance across six dimensions. `took.ACT` anchored Dimension 1 as a general academic testing axis. `ap.exam` and `pseo.participant` defined an advanced academic coursework axis across Dimensions 2 and 5. `cte.achievement` dominated Dimensions 3 and 4, with the absence of CTE engagement (No CTE) standing particularly apart. All variables showed adequate cos² representation.

Recommendation: Retain all variables.

**POSTSECONDARY VARIABLES**

The postsecondary MCA revealed the expected structural redundancy across `attended.ps`, `ps.grad`, `highest.cred.level`, `ps.grad.location`, and `ps.grad.InstitutionSector` — these variables are overlapping by construction. Dimension 1 captured non-participation, Dimension 2 captured current enrollment (appearing identically across four variables), Dimension 3 captured credential level and institution type, and Dimension 4 was an artifact of missing data rather than a substantive structure. Several `ps.grad.InstitutionSector` categories showed weak cos² representation, particularly the collapsed other category (11, total = 0.024) and multiple sectors (10, total = 0.071). `ps.grad.location` graduation destination categories were also moderately weak.

Recommendation: Retain all postsecondary variables for bivariate stages. The confirmed redundancy reinforces the framework decision to consolidate these variables into the ps.pathway combined variable for multivariate modeling rather than entering them simultaneously. The weak representation of `ps.grad.InstitutionSector` collapsed categories should be noted when interpreting bivariate results for that variable.

VARIABLES TO CARRY FORWARD WITH CAUTION

`RaceEthnicity/AI` (n=642, 0.5%) should be retained but interpreted cautiously in bivariate stages and may be unstable in multivariate models. `LimitedEnglishProficiencyIndicator` (n=2,625, 2.1%) will be retained and its relationship to `english.learner` and `non.english.home` monitored throughout the bivariate analysis.

<br>

```{r final master}
master.final.after.mca <- master.original %>%
  select(-HomelessIndicator, -sat.taken) %>%
  mutate(ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector %in% c("3", "5", "6", "7", "8", "9"), "11", as.character(ps.grad.InstitutionSector)),
         ps.grad.location = ifelse(ps.grad.location %in% c("PS grad region", "PS grad EDR"), "PS grad region", as.character(ps.grad.location)),
         ps.grad.location = as.factor(ps.grad.location),
         highest.cred.level = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Less than Bachelors",
                                     ifelse(highest.cred.level %in% c("Bachelor degree", "Master degree or higher"), "Bachelors or higher", as.character(highest.cred.level))),
         highest.cred.level = as.factor(highest.cred.level)) %>%
  select(1:ps.grad.in.MN, ps.grad.location, mn.emp.record.hs.grad.year:ncol(.))

names(master.final.after.mca)

write_csv(master.final.after.mca, "Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

```
