---
title: "Non-post-secondary graduates"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(rpart)
library(rpart.plot)
library(tidymodels)
library(caret)
library(themis)
library(ranger)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.emp.states <- read_csv("Data/SLEDS/Masters/After analysis/Master-meaningful-emp-Southern.csv") %>%
  select(PersonID, grad.year.1, grad.year.5, grad.year.10)

master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv") %>%
  left_join(master.emp.states, by = "PersonID") %>%
  mutate(ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "1", "Public, 4-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "4", "Public, 2-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "11", "Other", as.character(ps.grad.InstitutionSector)),
         ps.grad.location = ifelse(ps.grad.location == "PS grad EDR", "PS grad region", ps.grad.location),
         highest.cred.level = as.factor(highest.cred.level))

kable(names(master.original))

kable(head(master.original))
```

# Define the subset

One of the aspects noticed during analysis is the percentage of individuals that never graduated college that also had meaningful employment in the region. The table below shows that 54% of the high school graduates in Southeast Minnesota never graduated college.

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps") %>%
  group_by(ps.grad) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)

```

We are going to examine, more closely, this 54% of the population and see if there are patterns in which they end up in different employment outcomes 1-, 5- and 10-years after graduating high school.

<br>

# Employment outcomes distribution

The following tables provide the percentage of non-college graduates that were categorized in the various employment outcomes at 1-, 5-, and 10-years after graduating high school.

**WORKFORCE OUTCOME DISTRIBUTION — NON-COLLEGE COMPLETERS**

At Year 1, the non-completer population is still partially sorted — 21.8% are attending postsecondary, indicating delayed or part-time enrollment that did not ultimately result in credential completion. Among those not attending, meaningful regional employment accounts for 20.4% and meaningful MN employment for 6.7%. Non-meaningful employment is the largest single category at 31.2%, and no-MN-record sits at 19.8%.

By Year 5, postsecondary attendance drops to 5.1% as the last enrollees exit without completing. Meaningful regional employment rises substantially to 30.1% and becomes the largest active workforce category. Non-meaningful employment remains high at 24.0% and no-MN-record at 26.6%. Meaningful MN employment reaches 14.1%.

By Year 10, postsecondary attendance is zero. Meaningful regional employment holds at 30.6% — essentially stable from Year 5, suggesting this group has found its floor. The no-MN-record group rises to 34.0%, becoming the largest single category. Non-meaningful employment declines modestly to 18.7% and meaningful MN employment falls to 16.6%.

The most important pattern is the large and growing no-MN-record group — from 19.8% at Year 1 to 34.0% at Year 10. Among non-completers this group is more likely to reflect genuine disconnection or informal employment than the out-of-state migration interpretation that applies to college completers. The persistence of non-meaningful employment through Year 10 (18.7%) is the clearest expression of the local underutilization disconnection signature in this subgroup.

::: panel-tabset
## 1 year after graduation

```{r}
data <- master.original %>%
  filter(ps.grad == "No") %>%
  filter(grad.year.1 != "After 2023") %>%
  group_by(grad.year.1) %>%
  summarize(n = n()) %>%
  ungroup %>%
  mutate(pct = n / sum(n)) %>%
  write_csv(., "Data/Claude/analysis_non_post_secondary_graduates/grad-year-1-wf-outcomes.csv")

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)
```

## 5 years after graduation

```{r}
data <- master.original %>%
  filter(ps.grad == "No") %>%
  filter(grad.year.5 != "After 2023") %>%
  group_by(grad.year.5) %>%
  summarize(n = n()) %>%
  ungroup %>%
  mutate(pct = n / sum(n)) %>%
  write_csv(., "Data/Claude/analysis_non_post_secondary_graduates/grad-year-5-wf-outcomes.csv")

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)
```

## 10 years after graduation

```{r}
data <- master.original %>%
  filter(ps.grad == "No") %>%
  filter(grad.year.10 != "After 2023") %>%
  group_by(grad.year.10) %>%
  summarize(n = n()) %>%
  ungroup %>%
  mutate(pct = n / sum(n)) %>%
  write_csv(., "Data/Claude/analysis_non_post_secondary_graduates/grad-year-10-wf-outcomes.csv")

datatable(data, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)

```
:::

This analysis is going to focus on that 54% of the graduates, and explore relationships and patterns of the portion of this population that found meaningful employment in the region vs. other employment outcomes.

<br>

# Methodology

The above tables show the importance of understanding how non-college grads are navigating their post-high school pathways with a significant portion of them being disconnected from the regional workforce. To better understand these pathways, we will explore the relationships between independent variables (Demographic, High school characteristics, and High school accomplishments) along with workforce outcomes at 1-, 5- and 10-years after graduating high school.

Since we provided the descriptive statistics above, we now need to analyze which factors relate most strongly to outcomes, and then see if we can use those factors to build a predictive model.

For exploring the relationships we will do the following;

- For **categorical predictors** (e.g., Race/Ethnicity, Gender, Economic status, CTE achievement, PSEO participation), use cross-tabs with Cramer’s V and chi-square.
- For **continuous predictors** (e.g., ACT, MCA scores, total CTE courses, county-level variables), use ANOVA to compare means across the employment states.
- Run these analyses separately for grad.year.1, grad.year.5, grad.year.10 to see how relationships shift over time.

<br>

# Independent variables

We are going to continue using the variables that we have used throughout this project.

- **Demographic:**
  - `Gender`
  - `RaceEthnicity`
  - `economic.status`
  - `SpecialEdStatus`
  - `english.learner`
  - `non.english.home`
  - `LimitedEnglishProficiencyIndicator`
- **High school characteristics:**
  - `Dem_Desc`
  - `edr`
  - `avg.unemp.rate`
  - `avg.wages.pct.state`
- **High school accomplishments:**
  - `took.ACT`
  - `ACTCompositeScore`
  - `ap.exam`
  - `pseo.participant`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`
- **Post-Secondary**
  - `attended.ps`
  - `attended.ps.within.first.year.hsgrad`

<br>

# Relationships

We will first explore the relationships between the independent variables and the 1-, 5- and 10-year outcomes.

## Demographics

Here are the independent variables we are exploring;

- **Demographic:**
  - `Gender`
  - `RaceEthnicity`
  - `economic.status`
  - `SpecialEdStatus`
  - `english.learner`
  - `non.english.home`
  - `LimitedEnglishProficiencyIndicator`

### Summary

**OVERVIEW**

The demographic variable group among non-college completers tells a distinctly different story from the full population analysis. With postsecondary sorting removed from the picture, the dominant patterns shift from academic orientation and credential completion toward two competing disconnection signatures — local underutilization, expressed as elevated non-meaningful employment, and geographic absence, expressed as elevated no-MN-record rates. The relative prominence of these two signatures varies systematically across demographic groups in ways that are analytically important for understanding which populations face which types of labor market disadvantage.

**RANKING BY STRENGTH OF ASSOCIATION (V at Year 5 as primary sort)**

1.  SpecialEdStatus (V=0.176 / 0.113 / 0.084) The strongest demographic predictor in the non-completer subgroup and the only variable that shows a dual disconnection signature — elevated non-meaningful employment and elevated no-MN-record rates simultaneously. At Year 1, special education non-completers show a 28.9% no-MN-record rate and a 36.5% non-meaningful employment rate, both substantially above non-special education non-completers. The regional employment gap persists throughout the observation window (26.8% vs. 31.0% at Year 5, 28.3% vs. 31.2% at Year 10). By Year 10, the non-meaningful employment gap (24.0% vs. 17.4%) is the most durable disadvantage signal of any demographic variable in this subgroup. Special education non-completers are simultaneously more likely to be locally underemployed and more likely to have no Minnesota employment record — the most severe disadvantage profile in the non-completer demographic group.

2.  economic.status (V=0.115 / 0.079 / 0.076) The second strongest predictor, but with a very different profile from the full population. Among non-completers, economic disadvantage does not produce a meaningful regional employment gap — rates are essentially identical at Year 1 (20.3% non-low-income vs. 20.5% low-income) and the gap remains modest at Years 5 and 10. Instead, economic disadvantage concentrates graduates in non-meaningful employment — low-income non-completers show 34.9% non-meaningful employment at Year 1 compared to 28.4% for non-low-income non-completers, a gap that persists through Year 10 (22.1% vs. 16.7%). The no-MN-record rate is essentially identical across economic status groups throughout the observation window, confirming that geographic absence is not primarily an economic status phenomenon among non-completers. Economic disadvantage in this subgroup produces local underutilization, not geographic departure.

3.  english.learner (V=0.122 / 0.093 / 0.078) The third strongest predictor, with the most distinctive profile in the subgroup. English learner non-completers show the largest and most persistent no-MN-record rates of any demographic variable — 30.7% at Year 1, 36.8% at Year 5, and 44.6% at Year 10 — combined with substantially lower meaningful regional employment throughout. The no-MN-record gap grows over the observation window rather than declining, reaching 11.4 percentage points by Year 5 and remaining stable through Year 10. Non-meaningful employment is not substantially elevated, confirming that geographic absence rather than local underutilization is the primary disconnection signature. The english learner pattern in this subgroup is the clearest expression of the geographic absence signature and is analytically distinctive because the no-MN-record rate is more likely to reflect informal employment, return migration, or genuine geographic absence than out-of-state career establishment given the absence of postsecondary credentials.

4.  Gender (V=0.080 / 0.057 / 0.075) Gender shows a consistent but modest association. Males show higher meaningful regional employment throughout (22.1% vs. 18.2% at Year 1, 31.7% vs. 27.9% at Year 5, 32.4% vs. 28.2% at Year 10). Females show persistently higher non-meaningful employment at all three time points, which is the dominant gender signal in this subgroup. The no-MN-record pattern reverses over the observation window — females show lower no-MN-record rates at Year 1 but higher rates by Year 10 (35.3% vs. 33.0%), a pattern that may reflect differential labor force exit among females by Year 10. The gender association is weaker here than in the full population, confirming that much of the full population gender gap was driven by differential postsecondary sorting.

5.  non.english.home (V=0.089 / 0.081 / 0.079) Mirrors the english.learner pattern with smaller effect sizes, as expected given the conceptual overlap. The geographic absence signature dominates — no-MN-record rates reach 44.9% at Year 10 compared to 32.9% for english home graduates. Non-meaningful employment shows essentially no gap by Year 10, confirming a clean geographic absence rather than local underutilization pattern. Non-english home adds limited independent information beyond english.learner but confirms the geographic absence signature for the broader language minority non-completer population.

6.  RaceEthnicity (V=0.053 / 0.052 / 0.053) The weakest demographic predictor, with a flat association across all three time points — a distinctive pattern absent from the full population where racial and ethnic gaps shifted substantially as postsecondary graduates entered the labor market. Among non-completers, White graduates show the highest regional employment throughout (20.5%, 31.6%, 31.8%), while Black (10.9%, 15.6%, 16.6%) and American Indian (12.3%, 18.4%, 17.7%) graduates show the largest regional employment deficits. Asian/PI and American Indian non-completers show the highest no-MN-record rates at Year 10 (45.3% and 40.3%), while Black non-completers show the highest non-meaningful employment at Year 10 (27.3%) — the clearest racial expression of the local underutilization signature.

7.  LimitedEnglishProficiencyIndicator (V=0.055 / 0.045 / 0.038) The weakest and most rapidly declining association. Confirms the geographic absence signature of the language minority variables but adds no independent analytical value beyond english.learner and non.english.home. Small and declining group sizes limit interpretive confidence at Year 10 (n=525). Retained for descriptive completeness but not a primary variable for the narrative synthesis.

**CROSS-CUTTING THEMES**

Three findings stand out across the demographic group as analytically important for the narrative synthesis.

First, the two disconnection signatures divide along demographic lines in ways that are not visible in the full population analysis. Economic disadvantage and special education status primarily produce local underutilization — elevated non-meaningful employment among graduates who remain in Minnesota but cannot achieve sustained employment. Language minority status primarily produces geographic absence — elevated no-MN-record rates among graduates who leave Minnesota's formal employment records without postsecondary credentials to facilitate out-of-state career establishment. Special education status is the only group showing both signatures simultaneously.

Second, racial and ethnic gaps are weaker and flatter among non-completers than in the full population. The postsecondary sorting that amplifies racial and ethnic gaps in the full population is absent here, leaving smaller but more structurally revealing differences. The Black non-completer profile — concentrated in non-meaningful employment rather than geographic absence — is analytically distinct from the Asian/PI and American Indian profiles where geographic absence dominates.

Third, gender differences narrow substantially when postsecondary sorting is removed. The male regional employment advantage persists but is smaller, and the female non-meaningful employment disadvantage emerges more clearly as the primary gender signal in this subgroup — a pattern that is obscured in the full population by differential postsecondary attendance rates.

<br>

### `Gender`

Gender shows a modest and declining association with workforce outcomes among non-college completers (V=0.080 at Year 1, V=0.057 at Year 5, V=0.075 at Year 10).

At Year 1, the primary gender difference is in meaningful regional employment — males enter at 22.1% compared to 18.2% for females — and in non-meaningful employment, where females show a substantially higher rate (34.7% vs. 28.7%). Postsecondary attendance is slightly higher for females (22.6% vs. 21.2%) and the no-MN-record rate is notably lower for females (17.8% vs. 21.3%), suggesting females who do not complete credentials are more likely to remain in Minnesota's labor market even if below the sustained threshold.

By Year 5, the regional employment gap narrows (27.9% female vs. 31.7% male) and meaningful MN employment is essentially identical (14.1% vs. 14.2%). Females show higher non-meaningful employment (26.6% vs. 22.1%) and lower no-MN-record rates (26.1% vs. 27.0%), continuing the pattern of females remaining more attached to Minnesota's formal labor market even when not achieving sustained employment.

By Year 10, the male regional employment advantage persists (32.4% vs. 28.2%) and males show higher meaningful MN employment (17.8% vs. 15.1%). Females show notably higher non-meaningful employment (21.4% vs. 16.8%) and higher no-MN-record rates (35.3% vs. 33.0%) — a reversal of the Year 1 no-MN-record pattern that may reflect differential labor force exit among females by Year 10.

Compared to the full population bivariate findings, the gender association among non-college completers is similar in direction but slightly weaker — the male regional employment advantage is present throughout but the postsecondary attendance difference that amplified the full population Year 1 gap is much smaller here. The persistent female non-meaningful employment disadvantage is the most distinctive finding in this subgroup and warrants attention in the narrative synthesis.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/gender-1.html")

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/gender-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$Gender,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/gender-10.html")
```
:::

<br>

### `RaceEthnicity`

Race and ethnicity shows weak but consistent associations with workforce outcomes among non-college completers across all three time points (V=0.053 at Year 1, V=0.052 at Year 5, V=0.053 at Year 10). The flatness of the association over time is notable — unlike the full population where racial and ethnic gaps shifted substantially as postsecondary graduates entered the labor market, the non-completer subgroup shows a stable pattern throughout the observation window. Interpret AI findings with caution at all time points given small and declining cell sizes (n=408 at Year 1, n=124 at Year 10).

At Year 1, the most striking pattern is the elevated no-MN-record rate for American Indian (33.6%) and Asian/PI (26.2%) non-completers, and the high non-meaningful employment rates for American Indian (36.8%) and Black (34.3%) non-completers. Hispanic non-completers show the highest meaningful regional employment rate among minority groups (19.7%), essentially equivalent to White non-completers (20.5%). Black non-completers show notably lower regional employment (10.9%) and higher postsecondary attendance (28.8%) — suggesting a larger share of Black non-completers are attempting postsecondary without completing.

By Year 5, Black non-completers show the most concerning profile — the highest non-meaningful employment rate (32.8%) and a no-MN-record rate of 31.5%, both substantially above White non-completers (23.7% and 25.5%). Asian/PI non-completers show a very high no-MN-record rate (35.4%) consistent with out-of-region migration. American Indian non-completers show the highest no-MN-record rate of any group (41.4%), which among non-completers is more likely to reflect genuine disconnection than out-of-state career establishment. Hispanic non-completers show relatively strong regional employment (27.0%) but elevated no-MN-record rates (31.2%). White non-completers show the highest regional employment (31.6%) of any group.

By Year 10, racial and ethnic disparities in regional employment are substantial. White non-completers reach 31.8% meaningful regional employment, Hispanic non-completers 28.6%, while Black (16.6%), Asian/PI (22.1%), and American Indian (17.7%) non-completers lag considerably behind. The no-MN-record rate is highest for Asian/PI (45.3%), American Indian (40.3%), and Black (40.8%) non-completers — all substantially above White non-completers (33.0%). Black non-completers show the highest non-meaningful employment rate at Year 10 (27.3%), the clearest expression of the local underutilization disconnection signature among minority non-completers.

Compared to the full population findings, the racial and ethnic gaps among non-completers are more stable over time and more concentrated in the disconnection signatures — particularly for Black and American Indian non-completers. The postsecondary sorting that shifts the full population racial and ethnic patterns over time is absent here, leaving the structural disadvantages more directly visible.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No") %>%
  filter(RaceEthnicity != "Unknown")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/RaceEthnicity-1.html")

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No") %>%
  filter(RaceEthnicity != "Unknown")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/RaceEthnicity-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No") %>%
  filter(RaceEthnicity != "Unknown")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/RaceEthnicity-10.html")
```
:::

<br>

### `economic.status`

Economic status shows a moderate and declining association with workforce outcomes among non-college completers (V=0.115 at Year 1, V=0.079 at Year 5, V=0.076 at Year 10). The pattern here is meaningfully different from both the full population findings and what might be expected — economic disadvantage among non-completers does not produce a regional employment gap, but instead concentrates graduates in non-meaningful employment while leaving the no-MN-record rate essentially unchanged.

At Year 1 the most striking finding is that meaningful regional employment is essentially identical across economic status groups (20.3% non-low-income vs. 20.5% low-income) — a complete absence of the regional employment gap seen in the full population. The primary difference is in postsecondary attendance — non-low-income non-completers are more likely to still be attending at Year 1 (25.7% vs. 16.7%), reflecting delayed credential attempts. Low-income non-completers show substantially higher non-meaningful employment (34.9% vs. 28.4%), which is the dominant economic status signal at this time point. Meaningful MN employment is modestly higher for low-income graduates (7.5% vs. 6.1%), and no-MN-record rates are similar (20.3% vs. 19.4%).

By Year 5, a regional employment gap emerges but remains modest — non-low-income non-completers show 31.5% meaningful regional employment compared to 28.2% for low-income non-completers, a gap of 3.3 percentage points. The non-meaningful employment gap widens — low-income non-completers show 27.5% compared to 21.6% for non-low-income non-completers. The no-MN-record rate is essentially identical across groups (26.4% vs. 26.9%), confirming that geographic absence is not primarily an economic status phenomenon among non-completers.

By Year 10, the regional employment gap narrows slightly (31.5% vs. 29.1%) while the non-meaningful employment gap persists (16.7% non-low-income vs. 22.1% low-income). The no-MN-record rate remains essentially equivalent across groups (33.9% vs. 34.3%) — one of the flattest no-MN-record patterns of any demographic variable examined.

The most analytically important finding here is what does not happen — economic disadvantage does not produce a meaningful regional employment gap among non-completers the way it does in the full population. The primary economic disadvantage signal in this subgroup is concentrated in non-meaningful employment, not in geographic departure or regional employment deficits. This suggests that among non-completers, economic disadvantage produces labor market underutilization — the local disconnection signature — rather than the geographic departure or postsecondary sorting that drives economic status gaps in the full population.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/economic-status-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/economic-status-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/economic-status-10.html")
```
:::

<br>

### `SpecialEdStatus`

Special education status shows the strongest association with workforce outcomes of any demographic variable among non-college completers, and the association declines but remains meaningful over time (V=0.176 at Year 1, V=0.113 at Year 5, V=0.084 at Year 10).

At Year 1, the most striking pattern is the no-MN-record gap — special education non-completers show a 28.9% no-MN-record rate compared to 17.3% for non-special education non-completers, an 11.6 percentage point gap that is the largest single-category difference at this time point. Non-meaningful employment is also substantially higher for special education non-completers (36.5% vs. 29.8%). Meaningful regional employment is modestly lower for special education non-completers (18.2% vs. 21.0%), and postsecondary attendance is substantially lower (10.6% vs. 24.9%), confirming that special education non-completers have largely already exited the postsecondary pathway by Year 1.

By Year 5, the no-MN-record gap remains large — special education non-completers show 31.3% compared to 25.4% for non-special education non-completers. Non-meaningful employment remains elevated (29.3% vs. 22.7%) and meaningful regional employment shows a 4.2 percentage point gap (26.8% vs. 31.0%). The gap in meaningful MN employment is notable — special education non-completers show 10.3% compared to 15.1% for non-special education non-completers, a 4.8 percentage point gap suggesting special education non-completers are not only less likely to achieve regional employment but also less likely to achieve any form of sustained employment in Minnesota.

By Year 10, the regional employment gap narrows modestly (28.3% vs. 31.2%) and the no-MN-record rates converge somewhat (35.4% vs. 33.7%). The most persistent gap at Year 10 is in non-meaningful employment — special education non-completers show 24.0% compared to 17.4% for non-special education non-completers, a 6.6 percentage point gap that represents the most durable expression of the local underutilization disconnection signature in this subgroup.

The special education pattern among non-completers is distinct from the economic status pattern in an important way — special education non-completers show both elevated no-MN-record rates and elevated non-meaningful employment, while economic disadvantage primarily elevates non-meaningful employment without a corresponding no-MN-record gap. This dual disconnection signature for special education non-completers — present in both geographic absence and local underutilization — is the most severe disadvantage profile of any demographic variable in the non-completer subgroup and warrants specific attention in the narrative synthesis.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/SpecialEdStatus-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/SpecialEdStatus-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/SpecialEdStatus-10.html")
```
:::

<br>

### `english.learner`

English learner status shows a moderate and declining association with workforce outcomes among non-college completers (V=0.122 at Year 1, V=0.093 at Year 5, V=0.078 at Year 10). The dominant and most analytically distinctive pattern is the very large no-MN-record rate among english learner non-completers that persists across all three time points — a finding that is stronger and more persistent here than in the full population analysis.

At Year 1, english learner non-completers show a no-MN-record rate of 30.7% compared to 18.2% for non-english learner non-completers — a 12.5 percentage point gap that is the largest single-category difference at this time point and the largest no-MN-record gap of any demographic variable at Year 1. Meaningful regional employment is substantially lower for english learner non-completers (16.7% vs. 21.0%), and postsecondary attendance is lower (13.7% vs. 23.0%). Meaningful MN employment is modestly higher (8.3% vs. 6.5%), and non-meaningful employment is essentially identical across groups (30.6% vs. 31.3%).

By Year 5, the no-MN-record gap remains the dominant signal — english learner non-completers show 36.8% compared to 25.4% for non-english learner non-completers, an 11.4 percentage point gap that has narrowed only marginally from Year 1. Meaningful regional employment shows a large gap — 21.8% for english learner non-completers versus 31.1% for non-english learner non-completers, a 9.3 percentage point deficit that is the largest regional employment gap of any demographic variable in the non-completer subgroup at Year 5. Non-meaningful employment is slightly elevated for english learner non-completers (25.5% vs. 23.9%).

By Year 10, the no-MN-record gap widens further — english learner non-completers show 44.6% compared to 33.2% for non-english learner non-completers, a 11.4 percentage point gap that is essentially unchanged from Year 5. Meaningful regional employment remains substantially lower (20.1% vs. 31.4%), a 11.3 percentage point gap that has grown from Year 5. Non-meaningful employment is modestly elevated (22.0% vs. 18.5%).

The english learner pattern among non-completers is analytically distinctive in two ways. First, the no-MN-record rate is the highest of any demographic variable across all three time points and grows over the observation window — reaching 44.6% at Year 10. Among non-completers, this is more likely to reflect genuine geographic absence or informal employment than out-of-state career establishment, given the absence of postsecondary credentials that facilitate formal out-of-state employment. Second, the regional employment gap is large and persistent — the combination of very high no-MN-record rates and low regional employment rates represents the clearest dual disconnection signature in the non-completer demographic group. English learner non-completers are neither establishing regional employment nor showing the formal out-of-state career establishment pattern seen among credential completers — they are the most geographically absent demographic group in the non-completer subgroup.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$english.learner,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/english-learner-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$english.learner,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/english-learner-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$english.learner,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/english-learner-10.html")
```
:::

<br>

### `non.english.home`

Non-english home shows a moderate and declining association with workforce outcomes among non-college completers (V=0.089 at Year 1, V=0.081 at Year 5, V=0.079 at Year 10). The pattern closely mirrors english learner status, which is expected given the conceptual overlap between the two variables, but with somewhat smaller effect sizes throughout — consistent with the full population findings where non.english.home was consistently weaker than english.learner.

At Year 1, the dominant signal is again the no-MN-record rate — non-english home graduates show 28.7% compared to 18.7% for english home graduates, a 10.0 percentage point gap. Meaningful regional employment is lower for non-english home graduates (16.5% vs. 20.9%), and meaningful MN employment is modestly higher (8.9% vs. 6.5%). Non-meaningful employment is notably lower for non-english home graduates (26.9% vs. 31.8%), which combined with the elevated no-MN-record rate suggests geographic absence rather than local underutilization is the primary disconnection signature for this group.

By Year 5, the no-MN-record gap remains large — non-english home graduates show 36.2% compared to 25.5% for english home graduates, a 10.7 percentage point gap that is essentially unchanged from Year 1. Meaningful regional employment remains substantially lower (22.1% vs. 31.0%), a 8.9 percentage point gap. Non-meaningful employment is modestly lower for non-english home graduates (22.4% vs. 24.2%), confirming that geographic absence rather than local underutilization is the primary outcome differentiator.

By Year 10, the no-MN-record gap is the largest at any time point — non-english home graduates show 44.9% compared to 32.9% for english home graduates, a 12.0 percentage point gap. Meaningful regional employment remains substantially lower (21.8% vs. 31.5%). Non-meaningful employment is essentially identical across groups (18.6% vs. 18.8%), confirming that the entire non-english home disadvantage is concentrated in geographic absence rather than local labor market underutilization.

The non-english home pattern among non-college completers is analytically consistent with english.learner across all three time points and adds relatively little independent information. The key distinction is that non-english home shows a cleaner geographic absence signature — the no-MN-record gap is large and growing while non-meaningful employment shows essentially no gap by Year 10. This suggests that non-english home non-completers are leaving Minnesota's formal employment records entirely rather than remaining locally underemployed, a pattern that may reflect return migration, informal employment, or geographic mobility patterns specific to immigrant and language minority communities in Southeast Minnesota.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$non.english.home,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/non-english-home-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$non.english.home,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/non-english-home-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$non.english.home,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/non-english-home-10.html")
```
:::

<br>

### `LimitedEnglishProficiencyIndicator`

LimitedEnglishProficiencyIndicator shows weak and declining associations with workforce outcomes among non-college completers (V=0.055 at Year 1, V=0.045 at Year 5, V=0.038 at Year 10). The group is small — n=1,587 at Year 1, declining to n=525 at Year 10 — and estimates should be interpreted with appropriate caution. The pattern is directionally consistent with english.learner and non.english.home but weaker throughout, as expected given that LimitedEnglishProficiencyIndicator captures a more severely language-constrained subset of the broader language minority population.

At Year 1, the dominant signals are the elevated no-MN-record rate (29.0% vs. 19.5% for non-LEP graduates) and the elevated meaningful MN employment rate (11.2% vs. 6.6%). The latter is consistent with the anomalous meaningful MN employment pattern seen for language minority variables throughout the full population analysis — LEP graduates show a concentration in out-of-region but in-state employment at Year 1 that likely reflects industry and geographic clustering in specific Minnesota labor markets. Meaningful regional employment is lower for LEP graduates (16.5% vs. 20.5%) and non-meaningful employment is lower (25.6% vs. 31.4%), with the no-MN-record gap absorbing the difference.

By Year 5, the no-MN-record gap dominates — LEP graduates show 37.7% compared to 26.3% for non-LEP graduates, an 11.4 percentage point gap consistent with the english.learner pattern. Meaningful regional employment is substantially lower (20.3% vs. 30.3%), and non-meaningful employment is essentially equivalent (22.8% vs. 24.1%). The Year 1 meaningful MN employment advantage has largely disappeared (13.3% vs. 14.2%).

By Year 10, the no-MN-record gap remains large (44.2% vs. 33.8%) and meaningful regional employment remains substantially lower (21.9% vs. 30.8%). Non-meaningful employment is modestly elevated (20.2% vs. 18.7%). The pattern by Year 10 is essentially identical in structure to english.learner at Year 10, confirming that LEP status and english learner status are capturing the same underlying geographic absence signature among non-completers.

This variable adds limited independent information beyond english.learner and non.english.home in the non-completer subgroup. It confirms the geographic absence signature for language minority non-completers but at smaller effect sizes and with declining group sizes that limit interpretive confidence. Its primary value in this subgroup is confirmatory — the three language minority variables all tell a consistent story of geographic absence rather than local underutilization as the dominant disconnection signature for language minority non-completers.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$LimitedEnglishProficiencyIndicator,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/LimitedEnglishProficiencyIndicator-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$LimitedEnglishProficiencyIndicator,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/LimitedEnglishProficiencyIndicator-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$LimitedEnglishProficiencyIndicator,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Demographics/LimitedEnglishProficiencyIndicator-10.html")
```
:::

<br>

## High school characteristics

- **High school characteristics:**
  - `Dem_Desc`
  - `edr`
  - `avg.unemp.rate`
  - `avg.wages.pct.state`

<br>

### Summary

**OVERVIEW**

The high school characteristics group among non-college completers confirms the full population finding that place-based factors have limited independent associations with individual workforce outcomes. However, the one exception — average local unemployment rate — shows a more nuanced and analytically interesting pattern in the non-completer subgroup than in the full population, revealing a labor market substitution mechanism that operates differently when postsecondary sorting is removed from the picture.

**RANKING BY STRENGTH OF ASSOCIATION**

1.  avg.unemp.rate (F=110.8 / 31.2 / 23.8) The strongest variable in the group and the only one with meaningful analytical content. The association declines over time but remains significant through Year 10. The key finding is that stronger local labor markets are consistently associated with meaningful regional employment among non-completers — graduates from lower-unemployment communities achieve regional workforce attachment at higher rates, while graduates from higher-unemployment communities are more likely to appear in the no-MN-record or non-meaningful employment categories. By Year 5, meaningful MN employment emerges as the outcome most strongly associated with higher local unemployment, suggesting that non-completers from weaker local markets who do achieve sustained employment are more likely to find it outside the region within Minnesota. By Year 10 the pattern consolidates — lower unemployment predicts regional employment, higher unemployment predicts out-of-region or absent outcomes. This mechanism is consistent across the observation window and represents the most substantive place-based finding in the non-completer subgroup.

2.  edr (V=0.022 / 0.030 / 0.038) Negligible and essentially flat across all three time points. EDR differences in workforce outcomes among non-completers are trivially small — the largest single-category gap across any time point is the 2.0 percentage point meaningful MN employment difference at Year 5 (15.5% EDR 9 vs. 13.5% EDR 10). The EDR effects visible in the full population — driven by differential postsecondary pathway sorting — are entirely absent here. Where a non-completer attended high school within Southeast Minnesota has no meaningful bearing on their workforce trajectory.

3.  Dem_Desc (V=0.035 / 0.038 / 0.032) Negligible throughout and slightly weaker than EDR at Year 10. Urban, town/rural mix, and urban/town/rural mix communities produce essentially identical workforce outcome distributions across all three time points. The modest Year 5 finding that entirely urban communities show slightly lower meaningful MN employment (12.4%) is the only pattern worth noting, but it is too small to carry analytical weight. Community demographic description adds no value in this subgroup.

4.  avg.wages.pct.state (F=12.3 / 9.8 / 1.1) Statistically significant at Years 1 and 5 but with absolute mean differences measured in fractions of a percentage point of wage parity. Non-significant by Year 10. This variable has no practical analytical value in the non-completer subgroup and its full population exclusion from multivariate modeling is fully confirmed.

**CROSS-CUTTING THEMES**

The high school characteristics group tells a simple and consistent story among non-college completers: place does not determine workforce outcomes, with one partial exception. Local unemployment rate captures a genuine labor market substitution mechanism — stronger local economies facilitate regional workforce attachment for graduates who enter the labor market without credentials. This mechanism is plausible and theoretically coherent, and it becomes cleaner over the observation window as the postsecondary sorting noise present at Year 1 dissipates.

The absence of EDR and Dem_Desc effects in this subgroup is analytically informative. In the full population, geographic sorting through postsecondary pathways created apparent EDR differences in workforce outcomes. Among non-completers, those mechanisms are absent and the underlying place-based effects are revealed to be negligible. This confirms that the EDR and community type associations seen in the full population were artifacts of postsecondary sorting rather than independent place-based effects on workforce outcomes.

For multivariate modeling, only avg.unemp.rate warrants inclusion from this group. EDR retains marginal theoretical justification given its significance in the full population workforce model but its near-zero associations here suggest it will add little to a non-completer multivariate model.

<br>

### `edr`

EDR shows negligible associations with workforce outcomes among non-college completers across all three time points (V=0.022 at Year 1, V=0.030 at Year 5, V=0.038 at Year 10). The association is the weakest of any variable examined in the non-completer subgroup and barely grows over the observation window. The practical differences between EDR 9 and EDR 10 non-completers are minimal at every time point.

At Year 1, the outcome distributions are nearly identical across EDRs. Meaningful regional employment (20.5% EDR 10 vs. 20.2% EDR 9), non-meaningful employment (31.4% vs. 31.0%), and no-MN-record rates (20.2% vs. 19.0%) all show differences of less than 2 percentage points. Postsecondary attendance is modestly lower for EDR 10 non-completers (21.3% vs. 22.9%), consistent with a slightly higher proportion of EDR 10 graduates having already fully exited the postsecondary pathway by Year 1.

By Year 5, the most notable difference is in meaningful MN employment — EDR 9 non-completers show 15.5% compared to 13.5% for EDR 10 non-completers, a 2.0 percentage point gap. Meaningful regional employment is essentially identical (30.0% vs. 30.3%), as are no-MN-record rates (27.1% vs. 25.4%) and non-meaningful employment (24.1% vs. 24.0%).

By Year 10, EDR 9 non-completers show modestly higher meaningful MN employment (18.7% vs. 15.7%) — the most consistent EDR difference across the observation window — while regional employment remains essentially equivalent (30.1% vs. 30.8%). No-MN-record rates are similar (33.3% vs. 34.4%) and non-meaningful employment is modestly lower for EDR 9 (17.9% vs. 19.1%).

The EDR finding among non-completers is analytically important for what it confirms rather than what it reveals. In the full population analysis, EDR showed a growing association driven by differential postsecondary pathway sorting — EDR 10 graduates were more likely to complete credentials out of state and subsequently show higher no-MN-record rates. Among non-completers, that mechanism is absent and EDR effects are negligible. The modest EDR 9 advantage in meaningful MN employment that persists through Year 10 likely reflects the different labor market geography of the two EDRs — EDR 9 graduates without credentials have slightly better access to in-state out-of-region employment opportunities — but the practical magnitude is too small to carry substantial analytical weight.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/edr-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/edr-5.html")
```

#### 10-years after gradudation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$edr,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/edr-10.html")
```
:::

<br>

### `Dem_Desc`

Community demographic description shows negligible associations with workforce outcomes among non-college completers across all three time points (V=0.035 at Year 1, V=0.038 at Year 5, V=0.032 at Year 10). This is the weakest variable in the non-completer subgroup and confirms the full population finding that urban/rural community type has essentially no bearing on individual workforce outcomes once other characteristics are accounted for.

At Year 1, the outcome distributions are nearly identical across all three community types. Meaningful regional employment ranges from 20.3% to 20.6%, non-meaningful employment from 31.2% to 31.3%, and meaningful MN employment from 6.5% to 6.8% — all differences of less than 1 percentage point. The only modest signal is in no-MN-record rates, where the urban/town/rural mix category shows the highest rate (21.3%) compared to entirely urban (17.9%) and town/rural mix (19.1%). This pattern is consistent with more geographically diverse communities producing somewhat higher geographic absence rates, though the magnitude is small.

By Year 5, the community type differences remain minimal. Meaningful regional employment ranges from 28.8% to 31.5%, with entirely urban and town/rural mix communities showing modestly higher rates than the urban/town/rural mix category. Meaningful MN employment is modestly lower in entirely urban communities (12.4% vs. 14.4% and 15.1%), suggesting that urban non-completers without credentials are slightly less likely to achieve sustained in-state out-of-region employment — perhaps the most analytically interesting finding in this variable, though the magnitude remains small.

By Year 10, the community type differences are similarly minimal. Regional employment ranges from 29.0% to 33.1%, with entirely urban communities showing modestly higher regional employment — a reversal of the Year 5 pattern that suggests urban non-completers eventually achieve comparable or slightly stronger regional employment than those from mixed or rural communities. No-MN-record rates range from 32.2% to 35.1% with no meaningful gradient.

Dem_Desc adds no meaningful analytical value in the non-completer subgroup and confirms its exclusion from multivariate modeling was appropriate. Where graduates grew up in Southeast Minnesota — whether in entirely urban, town/rural mix, or urban/town/rural mix communities — has no practical bearing on their workforce trajectories as non-completers.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/Dem_Desc-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/Dem_Desc-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/Dem_Desc-10.html")
```
:::

<br>

### `avg.unemp.rate`

Average local unemployment rate shows a significant and meaningful association with workforce outcomes among non-college completers at all three time points (F=110.8 at Year 1, F=31.2 at Year 5, F=23.8 at Year 10), though the association weakens over time as expected.

At Year 1, postsecondary attendees come from communities with the highest average unemployment — consistent with the full population finding that higher unemployment increases postsecondary orientation among non-completers still attempting credentials. Among those not attending, meaningful regional employment comes from communities with the second lowest unemployment (0.319 below attending ps), while meaningful MN employment comes from the lowest unemployment communities (0.394 below attending ps). The no-MN-record group comes from communities with substantially higher unemployment than either meaningful employment group (0.164 above meaningful regional, 0.238 above meaningful MN), and non-meaningful employment sits between (0.107 above meaningful regional). The ordering at Year 1 is: attending ps (highest unemployment) \> no-MN-record \> non-meaningful \> meaningful regional \> meaningful MN (lowest unemployment).

This pattern is analytically important. Among non-completers at Year 1, stronger local labor markets are associated with both meaningful regional employment and meaningful MN employment, while weaker labor markets are associated with geographic absence and labor market underutilization. This is the labor market substitution mechanism operating in reverse — stronger local economies provide immediate employment opportunities that reduce both postsecondary enrollment and geographic departure.

By Year 5, the structure changes substantially. Attending ps still comes from the highest unemployment communities. But now the three active non-attending outcomes — meaningful regional employment, no-MN-record, and non-meaningful employment — are not significantly different from each other (differences all p\>0.76). Meaningful MN employment comes from communities with significantly higher unemployment than all other non-attending outcomes (0.112 above meaningful regional, 0.133 above no-MN-record, 0.119 above non-meaningful). This is a notable finding — by Year 5, meaningful MN employment is associated with higher local unemployment among non-completers, suggesting that graduates from weaker local markets who do achieve sustained employment are more likely to find it outside the region within Minnesota rather than locally.

By Year 10, the structure simplifies further. Meaningful MN employment comes from the highest unemployment communities, well above the other three outcomes (0.178 above meaningful regional, 0.088 above no-MN-record, 0.070 above non-meaningful). Meaningful regional employment comes from the lowest unemployment communities, significantly below no-MN-record (0.091) and non-meaningful employment (0.108), which are not significantly different from each other. By Year 10 the pattern is clear — stronger local labor markets produce regional employment attachment among non-completers, while weaker local markets are associated with either in-state out-of-region employment or geographic and labor market absence.

Across all three time points, average local unemployment rate shows a consistent and theoretically coherent pattern among non-college completers. Stronger local economies facilitate regional workforce attachment — the mechanism most directly relevant to the workforce-first pathway. Weaker local economies produce geographic absence and out-of-region employment among those who do achieve sustained employment, consistent with graduates from disadvantaged labor markets needing to travel further to find work without the postsecondary credentials that would otherwise facilitate that mobility.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(avg.unemp.rate ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/avg-unemp-rate-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(avg.unemp.rate ~ grad.year.5, data = data)


summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/avg-unemp-rate-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(avg.unemp.rate ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/avg-unemp-rate-10.xlsx", overwrite = TRUE)

```
:::

<br>

### `avg.wages.pct.state`

Average wages as a percent of state average shows a statistically significant but negligible practical association with workforce outcomes at Year 1 (F=12.3) and Year 5 (F=9.8), and becomes completely non-significant by Year 10 (F=1.1, p=0.329). The mean differences across outcome categories are extremely small throughout — measured in fractions of a percentage point of wage parity with the state average — and carry no meaningful analytical weight.

At Year 1, meaningful regional employment graduates come from the highest-wage communities (0.007 above attending ps), while meaningful MN employment graduates come from communities essentially equivalent to attending ps graduates. No-MN-record and non-meaningful employment graduates come from modestly lower-wage communities than meaningful regional graduates, but these differences are not significant from each other. The statistically significant F-statistic is driven entirely by the attending ps versus meaningful regional contrast — a difference so small in absolute terms (0.007 percentage points of wage parity) as to be analytically meaningless.

By Year 5, the structure shifts — attending ps now comes from the highest-wage communities, with meaningful MN employment coming from significantly lower-wage communities than all other outcomes. No-MN-record and non-meaningful employment are not significantly different from meaningful regional employment. Again the absolute differences are trivially small.

By Year 10, the variable is completely non-significant (p=0.329) and no pairwise comparisons approach significance. All outcome categories come from communities with essentially identical average wage levels relative to the state.

This variable adds no analytical value in the non-completer subgroup and confirms its exclusion from multivariate modeling. Community wage levels relative to the state average have no meaningful bearing on the workforce trajectories of non-college completers in Southeast Minnesota at any time point after the trivially small Year 1 and Year 5 associations are accounted for.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(avg.wages.pct.state ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/avg-wages-pct-state-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(avg.wages.pct.state ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/avg-wages-pct-state-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(avg.wages.pct.state ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school characteristics/avg-wages-pct-state-10.xlsx", overwrite = TRUE)

```
:::

<br>

## High school accomplishments

- **High school accomplishments:**
  - `took.ACT`
  - `ACTCompositeScore`
  - `ap.exam`
  - `pseo.participant`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`

<br>

### Summary

**OVERVIEW**

The high school accomplishments group among non-college completers reveals a fundamental inversion of what might be expected from the full population findings. Academic achievement and orientation variables — which predict credential completion and out-of-region careers in the full population — predict geographic mobility and in-state out-of-region employment rather than regional workforce attachment among non-completers. CTE engagement variables show the opposite pattern — a stable, dose-responsive, and durable association with regional employment that is the defining finding of the entire non-completer subgroup analysis. The two mechanisms that organize the full population analysis operate with equal clarity but opposite regional employment implications in this subgroup.

**RANKING BY STRENGTH AND DURABILITY OF ASSOCIATION**

1.  total.cte.courses.taken (F=261.1 / 183.0 / 93.6) The strongest and most durable predictor of regional employment in the non-completer subgroup and the variable with the clearest dose-response relationship. Meaningful regional employment graduates take approximately 1.2 more CTE courses than no-MN-record graduates across all three time points — a gap that is essentially unchanged from Year 1 through Year 10. The ordering by CTE courses is consistent: meaningful regional employment highest, meaningful MN employment second, non-meaningful employment third, and no-MN-record lowest at Years 5 and 10. The convergence of attending ps and no-MN-record at similar low CTE course-taking levels by Year 5 is analytically important — both groups represent delayed workforce entry, but through very different mechanisms. No other continuous variable matches the stability or interpretive clarity of total CTE courses taken across the observation window.

2.  took.ACT (V=0.273 / 0.116 / 0.092) The strongest categorical predictor at Year 1 but one of the most rapidly declining. The large Year 1 association is primarily driven by postsecondary attendance sorting — ACT takers attend postsecondary at 32.2% compared to 11.6% for non-takers — rather than by direct workforce outcome differences. Once that sorting resolves, meaningful regional employment converges almost completely (30.2% vs. 29.9% at Year 5). The durable ACT signal is in meaningful MN employment — ACT takers show 16.4% at Year 5 and 21.1% at Year 10 compared to 12.1% and 14.0% for non-takers — confirming that academic orientation without credential completion produces in-state geographic mobility rather than regional employment.

3.  ACTCompositeScore (F=264.3 / 111.2 / 56.4) The strongest F-statistic in the accomplishment group at Year 1 but restricted to ACT takers only. The score structure reveals a clear and consistent pattern: higher ACT scores predict geographic mobility (no-MN-record and meaningful MN employment) while lower scores predict regional employment and non-meaningful employment. By Year 10 the ordering is: meaningful MN ≈ no-MN-record (highest) \>\> meaningful regional ≈ non-meaningful (lowest). This two-cluster structure — geographic mobility at high scores, regional attachment and local underutilization at low scores — is the most analytically important structural finding in the academic achievement variables.

4.  MCA Reading (F=252.8 / 70.1 / 53.7) Closely mirrors ACT Composite Score and produces an almost identical Year 10 structure: meaningful MN ≈ no-MN-record (highest) \> meaningful regional ≈ non-meaningful (lowest). The Year 10 convergence of meaningful regional employment and non-meaningful employment at the lower end of the reading distribution is the clearest expression of the academic achievement paradox among non-completers — reading skill does not differentiate between regional employment success and local underutilization. Among non-completers, regional employment is driven by CTE engagement and labor market orientation, not by academic achievement.

5.  cte.achievement (V=0.096 / 0.095 / 0.088) The flattest association trajectory of any accomplishment variable — essentially unchanged across all three time points — and the clearest categorical expression of the CTE engagement mechanism. The dose-response gradient in regional employment is stable and monotonic: concentrators (35.3%) \> participants (30.2%/30.6%) \> No CTE (22.1%/ 22.7%) at Years 5 and 10. The growing no-MN-record gap for No CTE graduates — reaching 41.2% at Year 10 — identifies the absence of CTE engagement as the defining characteristic of geographic absence among non-completers. Non-meaningful employment is essentially flat across CTE categories at all time points, confirming that CTE engagement specifically predicts regional employment rather than general labor market success.

6.  MCA Math (F=132.0 / 58.5 / 60.9) Broadly consistent with MCA Reading but with a notable Year 10 distinction — meaningful regional employment and non-meaningful employment converge even more completely (difference=-0.042, p=0.197), confirming that math achievement does not differentiate between regional employment success and local underutilization among non-completers. The Year 5 finding that meaningful MN employment scores above meaningful regional is the earliest emergence of the geographic mobility pattern in the MCA assessments.

7.  ap.exam (V=0.133 / 0.095 / 0.082) Directionally consistent with took.ACT and ACT Composite Score — AP participation predicts geographic mobility and away from regional employment — but with a small group size (n=5,256 at Year 1 declining to n=2,174 at Year 10) that limits interpretive confidence. The regional employment deficit for AP exam takers is larger and more persistent than for ACT takers, consistent with AP participation representing a stronger academic orientation signal. Best understood as confirmatory of the academic orientation pattern rather than an independent primary finding.

8.  MCA Science (F=170.8 / 60.9 / 38.8) Broadly consistent with MCA Math and MCA Reading but with two distinguishing features. First, the complete absence of differentiation among active workforce outcomes at Year 1 — all four non-attending categories cluster at identical science scores — is unique in the non-completer accomplishment group. Second, by Year 10 meaningful regional employment scores above non-meaningful employment (difference=-0.088, p=0.000), unlike MCA Math and MCA Reading where these converge. Self-selection concerns apply given the voluntary nature of science assessment. Best treated as supplementary to MCA Math and MCA Reading.

9.  pseo.participant (V=0.122 / 0.052 / 0.059) Weak and rapidly declining. The 2011+ cohort restriction limits coverage and the associations are consistently weaker than all other academic orientation variables. The direction is consistent — PSEO participation predicts in-state geographic mobility and away from regional employment — but effect sizes are too small and declining too rapidly to carry substantial analytical weight. Confirmatory only.

**CROSS-CUTTING THEMES**

Three findings define the high school accomplishments group among non-college completers.

First, the CTE engagement mechanism operates independently and durably. Both cte.achievement and total.cte.courses.taken show stable, dose-responsive associations with regional employment that do not decay over the observation window. This is the most consequential finding in the non-completer subgroup and directly confirms the workforce-first pathway mechanism — CTE engagement in high school produces regional employment attachment without requiring credential completion.

Second, academic achievement and orientation produce geographic mobility, not regional employment. Higher ACT scores, AP exam participation, and higher MCA scores all consistently predict in-state out-of-region employment and to a lesser extent geographic absence — not regional employment attachment. This inversion relative to the full population is the defining structural feature of the non-completer subgroup and establishes that academic orientation without credential completion leads graduates out of the region rather than into it.

Third, academic achievement does not differentiate between regional employment success and local underutilization. The convergence of meaningful regional employment and non-meaningful employment at the lower end of MCA Math and MCA Reading distributions by Year 10 is among the most analytically important findings in the entire analysis. Among non-completers, the distinction between achieving regional employment and being locally underemployed is not predicted by academic achievement — it is predicted by CTE engagement. This finding has direct implications for policy: improving academic outcomes alone will not reduce local underutilization among non- completers if CTE engagement does not accompany it.

<br>

### `took.ACT`

took.ACT shows a strong but rapidly declining association with workforce outcomes among non-college completers (V=0.273 at Year 1, V=0.116 at Year 5, V=0.092 at Year 10). The large Year 1 association is primarily driven by postsecondary attendance sorting rather than direct workforce outcome differences, and the association weakens substantially as that sorting resolves over the observation window.

At Year 1, the dominant signal is postsecondary attendance — ACT takers attend postsecondary at 32.2% compared to just 11.6% for non-takers, a 20.6 percentage point gap that accounts for most of the overall association. Among those not attending, the most notable difference is in no-MN-record rates — non-takers show 26.6% compared to 12.9% for ACT takers, a 13.7 percentage point gap. Non-takers also show higher non-meaningful employment (33.5% vs. 29.0%). Meaningful regional employment is modestly higher for non-takers (21.5% vs. 19.3%), suggesting that among those not attending postsecondary, graduates without ACT preparation show slightly stronger immediate regional employment attachment.

By Year 5, the postsecondary attendance sorting has largely resolved and meaningful regional employment converges almost completely (30.2% non-takers vs. 29.9% ACT takers). The most meaningful remaining differences are in no-MN-record rates — non-takers show 29.6% compared to 23.3% for ACT takers, a 6.3 percentage point gap — and in meaningful MN employment, which is higher for ACT takers (16.4% vs. 12.1%). Non-meaningful employment is essentially equivalent (24.7% vs. 23.4%).

By Year 10, meaningful regional employment remains essentially equivalent across groups (31.4% non-takers vs. 29.2% ACT takers). The most notable difference is in meaningful MN employment — ACT takers show 21.1% compared to 14.0% for non-takers, a 7.1 percentage point gap that has actually grown from Year 5. Non-takers show higher no-MN-record rates (34.9% vs. 32.5%) and higher non-meaningful employment (19.6% vs. 17.3%).

The ACT trajectory among non-completers is analytically important for what it reveals about the nature of academic orientation in this subgroup. ACT participation does not predict regional employment among non-completers — the gap closes completely by Year 5 and stays closed at Year 10. What ACT participation does predict is in-state geographic mobility — ACT takers who never complete credentials show persistently higher meaningful MN employment throughout the observation window, suggesting they have sufficient academic orientation to achieve sustained employment outside the region within Minnesota but not enough postsecondary follow-through to complete credentials that would facilitate out-of-state careers. Non-takers without credentials are more likely to either achieve regional employment or have no MN employment record at all.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/took-ACT-1.html")

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/took-ACT-5.html")

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$took.ACT,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/took-ACT-10.html")

```
:::

<br>

### `ACTCompositeScore`

ACT Composite Score shows a strong and persistent association with workforce outcomes among non-college completers (F=264.3 at Year 1, F=111.2 at Year 5, F=56.4 at Year 10). Unlike the full population where high ACT scores predicted postsecondary completion and out-of- region careers, among non-completers the score structure reveals a consistent and analytically interesting ordering that persists through the decade of observation. Note that this analysis is restricted to ACT takers only — approximately half the non-completer population — so these findings reflect academic achievers among non-completers rather than the full subgroup.

At Year 1, postsecondary attendees score highest, approximately 2.44- 2.56 composite points above both meaningful regional and MN employment groups — which are not significantly different from each other. The no-MN-record group scores 1.62 points above meaningful regional employment, and non-meaningful employment scores 1.18 points above meaningful regional. This produces a clear ordering at Year 1: attending ps (highest) \> no-MN-record \> non-meaningful \> meaningful regional ≈ meaningful MN (lowest). Higher ACT scores among non- completers at Year 1 are associated with geographic absence rather than regional employment — an important inversion of the expected academic achievement advantage.

By Year 5, the structure changes substantially and becomes more analytically coherent. Attending ps still scores highest. Meaningful regional employment now scores the lowest of all active outcomes — 1.15 points below meaningful MN employment and 1.42 points below no-MN-record. Meaningful MN employment and no-MN-record are not significantly different from each other (difference=0.278, p=0.084). Non-meaningful employment scores 0.45 points above meaningful regional but 0.97 points below no-MN-record. The Year 5 ordering is: attending ps \> no-MN-record ≈ meaningful MN \> non-meaningful \> meaningful regional (lowest).

By Year 10, the structure is cleanest. Meaningful MN employment and no-MN-record are not significantly different from each other (difference=0.014, p=1.000) and both score substantially above meaningful regional employment (1.37 and 1.38 points respectively) and non-meaningful employment (1.04 points above meaningful regional, not significantly different, p=0.127). The Year 10 ordering is: meaningful MN ≈ no-MN-record (highest) \>\> meaningful regional ≈ non-meaningful (lowest).

The ACT score trajectory among non-completers is the clearest expression of a consistent pattern across the accomplishment variables — higher academic achievement among non-completers is not associated with better regional employment outcomes. Instead, higher ACT scores predict geographic mobility (no-MN-record and meaningful MN employment) while lower ACT scores predict regional employment and non-meaningful employment. This is the inverse of the credential completer pathway and confirms that among non-completers, academic orientation without credential completion produces geographic departure rather than regional workforce attachment.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(ACTCompositeScore ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/ACTCompositeScore-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(ACTCompositeScore ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/ACTCompositeScore-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(ACTCompositeScore ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/ACTCompositeScore-10.xlsx", overwrite = TRUE)
```
:::

<br>

### `ap.exam`

ap.exam shows a moderate and declining association with workforce outcomes among non-college completers (V=0.133 at Year 1, V=0.095 at Year 5, V=0.082 at Year 10). Note that AP exam takers represent a small minority of non-completers — n=5,256 at Year 1, declining to n=2,174 at Year 10 — so these findings should be interpreted with appropriate caution given the small group size and strong selection effects.

At Year 1, AP exam takers show dramatically higher postsecondary attendance (37.3% vs. 20.1% for non-AP takers) — the largest attendance gap of any binary accomplishment variable. Among those not attending, AP exam takers show substantially lower meaningful regional employment (12.3% vs. 21.3%) and lower no-MN-record rates (15.4% vs. 20.3%), but essentially equivalent non-meaningful employment (30.5% vs. 31.3%). The regional employment deficit for AP exam takers at Year 1 is consistent with the ACT composite score finding — academic orientation among non-completers is associated with delayed workforce entry, not enhanced regional employment.

By Year 5, the postsecondary attendance gap narrows but remains notable (9.9% AP takers vs. 4.6% non-takers). The most distinctive finding is in meaningful MN employment — AP exam takers show 17.9% compared to 13.7% for non-takers, a 4.2 percentage point gap consistent with the ACT score finding that academic orientation predicts in-state geographic mobility among non-completers. Meaningful regional employment remains substantially lower for AP exam takers (22.3% vs. 31.0%), a 8.7 percentage point gap that is the largest single-category difference at Year 5. No-MN-record rates are modestly higher for AP exam takers (27.9% vs. 26.5%).

By Year 10, the pattern consolidates. AP exam takers show higher meaningful MN employment (24.3% vs. 15.8%) and higher no-MN-record rates (37.1% vs. 33.7%), while showing substantially lower meaningful regional employment (22.4% vs. 31.5%). Non-meaningful employment is modestly lower for AP exam takers (16.2% vs. 19.0%).

The ap.exam pattern among non-completers is directionally consistent with took.ACT and ACT composite score — academic orientation predicts geographic mobility and in-state out-of-region employment rather than regional workforce attachment. The regional employment deficit for AP exam takers is larger and more persistent than for ACT takers, consistent with AP participation representing a stronger academic orientation signal that produces more pronounced geographic departure even without credential completion. Given the small AP exam taker group among non-completers, these findings are better understood as confirmatory of the academic orientation pattern than as primary analytical findings in their own right.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/ap-exam-1.html")

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/ap-exam-5.html")

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$ap.exam,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/ap-exam-10.html")

```
:::

<br>

### `pseo.participant`

pseo.participant shows a moderate and rapidly declining association with workforce outcomes among non-college completers (V=0.122 at Year 1, V=0.052 at Year 5, V=0.059 at Year 10). All findings are restricted to the 2011+ cohort and reflect a selected population — approximately 78% of Year 1, 70% of Year 5, and 47% of Year 10 non-completers have pseo.participant data. Estimates should be interpreted with these restrictions in mind.

At Year 1, PSEO participants show substantially higher postsecondary attendance (29.8% vs. 18.1% for non-participants) — consistent with PSEO participation being a strong signal of continued postsecondary intent even among eventual non-completers. Among those not attending, PSEO participants show lower no-MN-record rates (13.7% vs. 20.7%) and modestly lower meaningful regional employment (19.0% vs. 21.7%). Non-meaningful employment is essentially equivalent (30.9% vs. 31.9%).

By Year 5, the association weakens substantially (V=0.052). The most notable remaining difference is in meaningful MN employment — PSEO participants show 17.7% compared to 14.1% for non-participants, a 3.6 percentage point gap consistent with the ACT and AP exam patterns where academic orientation among non-completers predicts in-state geographic mobility rather than regional employment. Meaningful regional employment is modestly lower for PSEO participants (28.4% vs. 30.0%) and no-MN-record rates are essentially equivalent (26.1% vs. 27.5%).

By Year 10, the association remains weak (V=0.059). PSEO participants show higher meaningful MN employment (22.3% vs. 16.1%) and lower meaningful regional employment (26.9% vs. 29.5%). No-MN-record (33.9% vs. 35.9%) and non-meaningful employment (16.9% vs. 18.4%) are both modestly lower for PSEO participants.

The pseo.participant pattern among non-completers is directionally consistent with all other academic orientation variables — PSEO participation predicts in-state geographic mobility and away from regional employment, even among graduates who ultimately do not complete credentials. The effect is weaker than ACT and AP exam participation and declines more rapidly over time. Given the cohort restriction and weak associations, this variable will not be carried into multivariate modeling but confirms the broader pattern that any form of academic orientation in high school is associated with reduced regional employment attachment among non-completers.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/pseo-participant-1.html")

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/pseo-participant-5.html")

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/pseo-participant-10.html")

```
:::

<br>

### `cte.achievement`

cte.achievement shows a moderate and remarkably stable association with workforce outcomes among non-college completers (V=0.096 at Year 1, V=0.095 at Year 5, V=0.088 at Year 10). The near-flat trajectory of the association across the observation window is the most analytically distinctive feature of this variable in the non-completer subgroup — and it mirrors the full population finding precisely. CTE engagement produces a durable regional employment advantage that does not erode over time.

At Year 1, the regional employment gradient across CTE categories is clear and monotonic. CTE concentrators and completors show 24.9% meaningful regional employment, participants show 19.0%, and No CTE graduates show 13.7% — an 11.2 percentage point gap between concentrators and No CTE. The no-MN-record gradient runs in the opposite direction — No CTE graduates show 25.8% compared to 18.0% for concentrators and 17.5% for participants. Non-meaningful employment is modestly higher for participants (33.1%) than for either concentrators (30.4%) or No CTE graduates (30.7%), which is the only departure from a clean monotonic pattern. Postsecondary attendance is lower for concentrators (19.9%) than for participants (23.8%) or No CTE (23.0%), confirming that CTE concentration is associated with earlier workforce entry.

By Year 5, the regional employment gradient strengthens and clarifies. CTE concentrators show 35.3% meaningful regional employment, participants 30.2%, and No CTE graduates 22.1% — a 13.2 percentage point gap between concentrators and No CTE that has grown from Year 1. The no-MN-record gradient also sharpens dramatically — No CTE graduates show 33.7% compared to 23.3% for concentrators and 24.7% for participants, a 10.4 percentage point gap between concentrators and No CTE. Non-meaningful employment is essentially equivalent across CTE categories (23.0%, 25.2%, 24.3%). The meaningful MN employment rates are also similar across groups (13.6%, 14.8%, 14.2%).

By Year 10, the regional employment gradient remains the dominant signal and is the most durable finding in the accomplishment group. CTE concentrators show 35.3% meaningful regional employment, participants 30.6%, and No CTE graduates 22.7% — a 12.6 percentage point gap between concentrators and No CTE that is essentially unchanged from Year 5. The no-MN-record gradient is now the starkest single finding — No CTE graduates show 41.2% no-MN-record compared to 30.0% for concentrators and 33.7% for participants. Non-meaningful employment is essentially equivalent across CTE categories (18.4%, 19.4%, 18.7%). Meaningful MN employment is similar across groups (16.3%, 16.4%, 17.4%).

The cte.achievement findings among non-college completers are among the most analytically important in the entire subgroup analysis. The dose-response gradient in regional employment — concentrators highest, participants intermediate, No CTE lowest — is stable across all three time points and shows essentially no decay from Year 5 to Year 10. This 12-13 percentage point gap between concentrators and No CTE graduates represents the most durable and consistent finding in the non-completer subgroup and directly replicates the full population pattern. The absence of a meaningful MN employment or non-meaningful employment gradient confirms that CTE engagement specifically predicts regional employment attachment rather than general labor market success. The growing no-MN-record gap for No CTE graduates — reaching 41.2% by Year 10 — is the clearest expression of genuine disconnection in the non-completer subgroup and identifies the absence of CTE engagement as the defining characteristic of geographic absence among non-completers who lack credentials.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/cte-achievement-1.html")

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/cte-achievement-5.html")

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/cte-achievement-10.html")
```
:::

<br>

### `total.cte.courses.taken`

Total CTE courses taken shows a strong and highly stable association with workforce outcomes among non-college completers (F=261.1 at Year 1, F=183.0 at Year 5, F=93.6 at Year 10). The ordering of outcome categories by CTE course-taking is consistent and interpretable across all three time points, and the structure becomes cleaner over the observation window as postsecondary sorting resolves.

At Year 1, meaningful regional employment comes from graduates with the most CTE courses — 1.23 courses above attending ps graduates and 0.71 courses above meaningful MN employment graduates. The no-MN-record group takes the fewest courses — 1.41 below meaningful regional employment and 0.70 below meaningful MN employment. Non-meaningful employment takes 0.90 fewer courses than meaningful regional employment. The Year 1 ordering by CTE course-taking is: meaningful regional (highest) \> meaningful MN \> attending ps \> non-meaningful \> no-MN-record (lowest). The most striking contrast is between meaningful regional employment and no-MN-record — a difference of 1.41 courses that is the largest gap in the analysis and confirms that geographic absence at Year 1 is concentrated among graduates with the least CTE investment.

By Year 5, the structure is remarkably similar. Meaningful regional employment still takes the most CTE courses — 1.22 above attending ps and 0.64 above meaningful MN employment. The no-MN-record group remains the lowest — 1.17 below meaningful regional and 0.53 below meaningful MN. Non-meaningful employment is 0.72 below meaningful regional. Attending ps and no-MN-record are not significantly different from each other (difference=0.051, p=0.973). Meaningful MN and non-meaningful employment are also not significantly different (difference=-0.082, p=0.609). The Year 5 ordering is: meaningful regional (highest) \> meaningful MN ≈ non-meaningful \> attending ps ≈ no-MN-record (lowest).

By Year 10, the structure simplifies to its clearest form. Meaningful regional employment takes the most CTE courses — 0.56 above meaningful MN employment, 0.92 above no-MN-record, and 0.61 above non-meaningful employment. Meaningful MN and non-meaningful employment are not significantly different from each other (difference=-0.048, p=0.920). No-MN-record takes significantly fewer courses than all other outcomes and is significantly lower than non-meaningful employment (difference= 0.309, p\<0.001). The Year 10 ordering is: meaningful regional (highest) \> meaningful MN ≈ non-meaningful \> no-MN-record (lowest).

Total CTE courses taken produces the clearest and most consistent outcome ordering of any continuous accomplishment variable in the non-completer subgroup. The regional employment advantage associated with CTE course-taking is visible at Year 1, maintained at Year 5, and still fully present at Year 10 — with meaningful regional employment graduates consistently taking approximately 1.2 more CTE courses than no-MN-record graduates across all three time points. This dose-response relationship between CTE investment and regional employment attachment is the defining finding of the accomplishment group and directly confirms the workforce-first pathway mechanism identified throughout the analysis. The consistently low CTE course- taking of the no-MN-record group — significantly below even non-meaningful employment graduates at Year 10 — identifies the absence of CTE investment as the single strongest predictor of geographic absence among non-college completers.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(total.cte.courses.taken ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/total-cte-courses-taken-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(total.cte.courses.taken ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/total-cte-courses-taken-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(total.cte.courses.taken ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/total-cte-courses-taken-10.xlsx", overwrite = TRUE)
```
:::

<br>

### `MCA.M`

MCA Math shows a significant and evolving association with workforce outcomes among non-college completers (F=132.0 at Year 1, F=58.5 at Year 5, F=60.9 at Year 10). The association is restricted to MCA takers only — approximately 92% of the non-completer population at Year 1 declining to about 91% at Year 10 — so coverage is substantially better than ACT composite score. The structure of associations shifts meaningfully across the observation window in ways that are analytically important.

At Year 1, postsecondary attendees score highest on MCA Math. Meaningful regional employment and meaningful MN employment are not significantly different from each other (difference=-0.038, p=0.359) and both score substantially below attending ps. No-MN-record scores the lowest of all outcomes — 0.154 below meaningful regional and 0.192 below meaningful MN employment. Non-meaningful employment scores 0.073 below meaningful regional. The Year 1 ordering is: attending ps (highest) \> meaningful regional ≈ meaningful MN \> non-meaningful \> no-MN-record (lowest). Higher MCA Math scores are associated with postsecondary attendance, while lower scores are associated with geographic absence — consistent with the ACT and total CTE courses patterns.

By Year 5, the structure shifts substantially. Attending ps and meaningful MN employment are now not significantly different (difference= 0.004, p=1.000) and both score highest. Meaningful regional employment scores 0.176 below meaningful MN employment — a meaningful reversal from Year 1 where the two were equivalent. No-MN-record scores 0.122 below meaningful MN but 0.053 above meaningful regional (p=0.005). Non-meaningful employment scores the lowest — 0.253 below meaningful MN and 0.077 below meaningful regional (p=0.000). The Year 5 ordering is: attending ps ≈ meaningful MN (highest) \> no-MN-record \> meaningful regional \> non-meaningful (lowest).

By Year 10, the structure clarifies to its most interpretable form. Meaningful MN employment scores highest — 0.225 above meaningful regional, 0.077 above no-MN-record, and 0.267 above non-meaningful employment. Meaningful regional and non-meaningful employment are not significantly different from each other (difference=-0.042, p=0.197). No-MN-record scores 0.148 above meaningful regional and 0.191 above non-meaningful employment. The Year 10 ordering is: meaningful MN (highest) \> no-MN-record \> meaningful regional ≈ non-meaningful (lowest).

The MCA Math trajectory among non-college completers is analytically consistent with ACT Composite Score — higher academic achievement predicts geographic mobility (meaningful MN employment and to a lesser extent no-MN-record) rather than regional employment attachment. By Year 10 the pattern is clear: non-completers with the strongest math skills achieve in-state out-of-region employment, those with moderate skills show geographic absence, and those with lower scores are split between regional employment and non-meaningful employment. The convergence of meaningful regional employment and non-meaningful employment at the lower end of the MCA Math distribution by Year 10 is one of the most analytically distinctive findings in this subgroup — math achievement does not differentiate between regional employment success and local underutilization among non-completers.

::: panel-tabset
#### 1-year after graduating

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.M ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-M-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.M ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-M-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.M ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-M-10.xlsx", overwrite = TRUE)
```
:::

<br>

### `MCA.R`

MCA Reading shows a significant and stable association with workforce outcomes among non-college completers (F=252.8 at Year 1, F=70.1 at Year 5, F=53.7 at Year 10). The structure evolves meaningfully across the observation window and by Year 10 produces one of the clearest and most analytically distinctive outcome orderings in the non-completer subgroup.

At Year 1, postsecondary attendees score highest on MCA Reading, substantially above all other outcomes. Among non-attending outcomes, meaningful regional employment and meaningful MN employment are not significantly different from each other (difference=-0.044, p=0.171) and both score substantially below attending ps. Non-meaningful employment and meaningful regional employment are also not significantly different (difference=0.006, p=0.986). No-MN-record scores the lowest — 0.101 below meaningful regional and 0.145 below meaningful MN employment. The Year 1 ordering is: attending ps (highest) \> meaningful MN ≈ meaningful regional ≈ non-meaningful \> no-MN-record (lowest). The most notable Year 1 finding is that MCA Reading does not differentiate between regional employment, MN employment, and non-meaningful employment — all three active workforce outcomes cluster together at similar reading score levels.

By Year 5, the structure differentiates substantially. Attending ps still scores highest. Meaningful MN employment scores 0.159 below attending ps but 0.159 above meaningful regional employment — establishing MN employment as the intermediate academic achievement outcome. No-MN-record scores 0.061 above meaningful regional (p=0.000) but 0.099 below meaningful MN employment. Non-meaningful employment scores the lowest of any non-attending outcome — 0.187 below meaningful MN, 0.028 below meaningful regional (not significant, p=0.281), and 0.089 below no-MN-record. The Year 5 ordering is: attending ps (highest) \> meaningful MN \> no-MN-record \> meaningful regional ≈ non-meaningful (lowest).

By Year 10, the structure reaches its most interpretable and analytically important form. Meaningful MN employment scores highest — 0.201 above meaningful regional, 0.212 above non-meaningful employment, and 0.042 above no-MN-record (not significant, p=0.204). No-MN-record scores 0.159 above meaningful regional and 0.171 above non-meaningful employment. Meaningful regional employment and non-meaningful employment are not significantly different from each other (difference=-0.012, p=0.940). The Year 10 ordering is: meaningful MN ≈ no-MN-record (highest) \> meaningful regional ≈ non-meaningful (lowest).

The MCA Reading trajectory among non-college completers closely mirrors MCA Math and ACT Composite Score — higher academic achievement predicts geographic mobility rather than regional employment attachment. The Year 10 convergence of meaningful MN employment and no-MN-record at the high end, and meaningful regional employment and non-meaningful employment at the low end, is the clearest expression of the academic achievement paradox in this subgroup. Among non-completers, reading skill does not differentiate between regional employment success and local underutilization — graduates at the lower end of the reading distribution are equally likely to achieve regional employment or remain non-meaningfully employed. This convergence confirms that among non-completers, regional employment is primarily driven by CTE engagement and labor market orientation rather than academic achievement.

::: panel-tabset
#### 1-year after graduating

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.R ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-R-1.xlsx", overwrite = TRUE)

```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.R ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-R-5.xlsx", overwrite = TRUE)

```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.R ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-R-10.xlsx", overwrite = TRUE)
```
:::

<br>

### `MCA.S`

MCA Science shows a significant association with workforce outcomes among non-college completers (F=170.8 at Year 1, F=60.9 at Year 5, F=38.8 at Year 10). However, the structure of associations is analytically distinct from MCA Math and MCA Reading in an important way — at Year 1 the four non-attending outcome categories are statistically indistinguishable from each other, revealing a flat structure among active workforce outcomes that does not appear in the other MCA assessments. The self-selection caveat applies here as well — MCA Science is a voluntary assessment, so coverage is lower than MCA Math and Reading and the analytic sample is more academically selected.

At Year 1, postsecondary attendees score substantially above all other outcomes. Among non-attending outcomes, all four categories show essentially identical MCA Science scores — none of the pairwise comparisons between meaningful regional, meaningful MN, no-MN-record, and non-meaningful employment approaches significance (all p\>0.28). This complete absence of differentiation among active workforce outcomes at Year 1 is unique to MCA Science in the non-completer subgroup and suggests that science achievement has no independent bearing on early workforce trajectory among non-completers beyond its association with postsecondary attendance.

By Year 5, differentiation emerges. Attending ps still scores highest. Meaningful MN employment and no-MN-record are not significantly different from each other (difference=-0.031, p=0.400) and both score above the other outcomes. Meaningful regional employment scores 0.118 below meaningful MN and 0.087 above no-MN-record (p=0.000). Non-meaningful employment scores the lowest — 0.176 below meaningful MN, 0.058 below meaningful regional (p=0.001). The Year 5 ordering is: attending ps (highest) \> meaningful MN ≈ no-MN-record \> meaningful regional \> non-meaningful (lowest).

By Year 10, the structure clarifies. Meaningful MN employment scores highest — 0.162 above meaningful regional, 0.083 above no-MN-record, and 0.250 above non-meaningful employment. No-MN-record scores 0.079 above meaningful regional and 0.167 above non-meaningful employment. Meaningful regional and non-meaningful employment are significantly different (difference=-0.088, p=0.000) unlike in MCA Math and MCA Reading where they converge. The Year 10 ordering is: meaningful MN (highest) \> no-MN-record \> meaningful regional \> non-meaningful (lowest).

The MCA Science trajectory among non-completers is broadly consistent with MCA Math and MCA Reading — higher science achievement predicts geographic mobility rather than regional employment attachment — but with two distinguishing features. First, the complete absence of differentiation among active workforce outcomes at Year 1 suggests science achievement is more weakly related to early labor market sorting than the other assessments. Second, by Year 10 meaningful regional employment scores above non-meaningful employment — unlike MCA Math and MCA Reading where these two outcomes converge — suggesting science achievement modestly differentiates between regional employment success and local underutilization in a way that reading and math do not. Given the self-selection issues with MCA Science, these findings should be treated as supplementary to rather than independent of the MCA Math and MCA Reading patterns.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.S ~ grad.year.1, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-S-1.xlsx", overwrite = TRUE)
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.S ~ grad.year.5, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-S-5.xlsx", overwrite = TRUE)
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

summary <- aov(MCA.S ~ grad.year.10, data = data)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/analysis_non_post_secondary_graduates/High school accomplishments/MCA-S-10.xlsx", overwrite = TRUE)
```
:::

<br>

## Post-Secondary

- **Post-Secondary**
  - `attended.ps`
  - `attended.ps.within.first.year.hsgrad`

<br>

### Summary

**OVERVIEW**

The postsecondary orientation variables — attended.ps and attended.ps.within.first.year.hsgrad — capture a dimension of the non-completer subgroup that no other variable group addresses directly: whether these graduates engaged with postsecondary at all, and if so, when. Both variables show strong associations at Year 1 that are substantially mechanical and decline rapidly over the observation window. The most analytically important findings emerge at Years 5 and 10, where the associations reflect genuine labor market differences rather than artifacts of current enrollment status.

**RANKING BY STRENGTH AND DURABILITY**

1.  attended.ps (V=0.633 / 0.272 / 0.102) The strongest Year 1 association of any variable in the entire non-completer subgroup analysis, though this is primarily mechanical — postsecondary attendees by definition include all current enrollees at Year 1. The substantive finding is in the no-MN-record gap: those who never attempted postsecondary show 29.5% no-MN-record at Year 1, compared to 6.7% for those who did attempt. This 22.8 percentage point gap narrows to 14.3 points at Year 5 (33.7% vs. 19.4%) and 9.6 points at Year 10 (39.0% vs. 29.4%), but never closes. The persistent no-MN-record advantage for postsecondary attempters — even those who ultimately do not complete credentials — is the central substantive finding of this variable. A postsecondary attempt, even unsuccessful, provides some geographic orientation and institutional connection that reduces the probability of complete absence from Minnesota's formal labor market. Meaningful regional employment differences are more modest — 4.2 percentage points at Year 5 (32.2% vs. 28.0%) and 4.6 points at Year 10 (32.8% vs. 28.2%) — confirming that the primary attended.ps signal is in geographic absence reduction rather than regional employment enhancement.

2.  attended.ps.within.first.year.hsgrad (V=0.504 / 0.209 / 0.079) Adds meaningful analytical value beyond attended.ps by disaggregating postsecondary timing into three categories — never attended, attended but delayed, and attended immediately. The key finding is that immediate enrollment produces meaningfully better outcomes than delayed enrollment, and by Year 10 delayed enrollment produces outcomes essentially indistinguishable from never having attended at all. At Year 5, immediate enrollees show 33.3% meaningful regional employment compared to 25.9% for delayed enrollees and 28.0% for non-attenders — and 19.2% no-MN-record compared to 20.1% for delayed enrollees and 33.7% for non-attenders. The convergence of delayed enrollment and non-attendance by Year 10 (27.1% vs. 28.2% regional employment, 34.4% vs. 39.0% no-MN-record) suggests that the timing of postsecondary engagement matters as much as whether engagement occurs at all. The delayed enrollment group (n=2,970 at Year 1) is small and its profiles at Year 10 should be interpreted with appropriate caution given the declining sample sizes.

**CROSS-CUTTING THEMES**

Two findings define the postsecondary orientation group among non-college completers.

First, any postsecondary attempt reduces geographic absence. The most consistent and durable finding across both variables is that non-completers who never attempted postsecondary show substantially higher no-MN-record rates throughout the observation window. This cannot be explained by credential completion — by definition, all individuals in this subgroup are non-completers. The reduction in geographic absence for postsecondary attempters likely reflects institutional engagement that provides geographic orientation, social networks, and labor market information that direct-to-workforce graduates lack.

Second, the timing of postsecondary engagement matters. Immediate enrollment is associated with better regional employment and lower no-MN-record rates at Years 5 and 10, while delayed enrollment converges with non-attendance by Year 10. This finding suggests that postsecondary engagement loses its labor market orientation benefit when delayed — graduates who enter the workforce first and attempt postsecondary later do not gain the same geographic anchoring that immediate enrollment provides.

For multivariate modeling, attended.ps is the more analytically tractable of the two variables and the one recommended for inclusion. attended.ps.within.first.year.hsgrad adds nuance but introduces a three-category structure that complicates interpretation and produces very small cell sizes in the delayed enrollment category by Year 10. The attended.ps binary cleanly captures the most important distinction — attempted postsecondary versus did not — and is preferred for the confirmatory multivariate stage.

<br>

### `attended.ps`

attended.ps shows a very strong but rapidly declining association with workforce outcomes among non-college completers (V=0.633 at Year 1, V=0.272 at Year 5, V=0.102 at Year 10). The large Year 1 association is almost entirely mechanical — by definition, all current postsecondary attendees have attended.ps=Yes, so the variable perfectly predicts the attending ps outcome category at Year 1. As postsecondary sorting resolves over time, the association reflects a more substantive distinction between those who attempted postsecondary and those who entered the labor market directly from high school.

At Year 1, the mechanical relationship dominates. Among attended.ps=No graduates, the outcome distribution is entirely composed of non-attending outcomes — no-MN-record is the largest category (29.5%), followed by non-meaningful employment (40.7%) and meaningful regional employment (22.2%). Among attended.ps=Yes graduates, 51.2% are currently attending, with the remaining distributed across non-meaningful employment (18.5%), meaningful regional employment (18.0%), no-MN-record (6.7%), and meaningful MN employment (5.5%). The most analytically important Year 1 finding is the very large no-MN-record gap — those who never attended postsecondary show 29.5% no-MN-record compared to just 6.7% for those who attempted postsecondary, a 22.8 percentage point gap that represents the clearest early disconnection signal in the non-completer subgroup.

By Year 5, the association is more substantively interpretable as postsecondary attendees have largely exited enrollment. Those who attended postsecondary show higher meaningful regional employment (32.2% vs. 28.0%) and meaningful MN employment (14.9% vs. 13.4%), and substantially lower no-MN-record rates (19.4% vs. 33.7%) — a 14.3 percentage point gap. Non-meaningful employment is modestly lower for postsecondary attempters (23.1% vs. 24.9%). The no-MN-record gap is the dominant signal at Year 5 — those who never attempted postsecondary are substantially more likely to have no Minnesota employment record, consistent with direct labor market entry without either postsecondary credentials or the geographic orientation that postsecondary attendance provides.

By Year 10, the association weakens substantially (V=0.102). Postsecondary attempters show modestly higher meaningful regional employment (32.8% vs. 28.2%) and meaningful MN employment (18.3% vs. 14.9%). The no-MN-record gap narrows but persists — those who never attended postsecondary show 39.0% no-MN-record compared to 29.4% for postsecondary attempters, a 9.6 percentage point gap. Non-meaningful employment is essentially equivalent (19.5% vs. 17.9%).

The attended.ps findings among non-college completers are analytically important for what they reveal about the heterogeneity of the non-completer subgroup. Non-completers who at least attempted postsecondary — even without completing — show meaningfully better workforce outcomes than those who never attempted, primarily through lower no-MN-record rates and modestly higher meaningful employment. The postsecondary attempt, even without completion, provides some labor market orientation and institutional connection that reduces geographic absence. Graduates who never attempted postsecondary are the most likely to appear in the no-MN-record category — a finding that directly connects to the disconnection signatures identified throughout the demographic and accomplishment variable analyses.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$attended.ps,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Post-secondary/attended-ps-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No",
         attended.ps != "After 2023")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$attended.ps,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Post-secondary/attended-ps-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$attended.ps,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Post-secondary/attended-ps-10.html")
```
:::

<br>

### `attended.ps.within.first.year.hsgrad`

attended.ps.within.first.year.hsgrad has three categories — did not attend ps (n=30,496), attended ps but not within first year (No, n=2,970 at Year 1), and attended ps within first year (Yes, n=19,612 at Year 1). The variable shows a strong and declining association (V=0.504 at Year 1, V=0.209 at Year 5, V=0.079 at Year 10). As with attended.ps, the large Year 1 association is substantially mechanical, and the most analytically meaningful findings emerge at Years 5 and 10 once postsecondary sorting has resolved.

At Year 1, the three categories produce strikingly different outcome profiles. Did-not-attend graduates — by definition not currently enrolled — show 40.7% non-meaningful employment, 29.5% no-MN-record, and 22.2% meaningful regional employment as their primary outcomes. Delayed enrollees (No category, n=2,970) show the worst Year 1 profile of any postsecondary orientation group — 46.6% non-meaningful employment, 26.8% no-MN-record, and only 19.8% meaningful regional employment. Immediate enrollees (Yes) are mostly still attending postsecondary (59.0%) and among those not attending, show substantially lower non-meaningful employment (14.3%) and no-MN-record (3.7%) than either other group.

By Year 5, the structure is more substantively interpretable. Immediate enrollees show the highest meaningful regional employment (33.3%) and meaningful MN employment (15.3%), and the lowest no-MN-record of any group (19.2%). Did-not-attend graduates show 28.0% regional employment and 33.7% no-MN-record. Delayed enrollees — the most analytically interesting category — show 25.9% regional employment, 20.1% no-MN-record, and 22.4% non-meaningful employment, positioning them between the other two groups on most outcomes but closer to did-not- attend graduates than to immediate enrollees.

By Year 10, the differences narrow substantially (V=0.079). Immediate enrollees retain the highest meaningful regional employment (33.9%) and meaningful MN employment (18.5%), and the lowest no-MN-record (28.4%). Did-not-attend graduates show 28.2% regional employment and 39.0% no-MN-record — the largest geographic absence of any enrollment timing group. Delayed enrollees show 27.1% regional employment, 34.4% no-MN-record, and 21.3% non-meaningful employment — essentially indistinguishable from did-not-attend graduates on most outcomes by Year 10.

The timing of postsecondary enrollment adds modest analytical value beyond attended.ps alone. Immediate enrollment in the first year after high school graduation is associated with the best long-run workforce outcomes among non-completers — most clearly in regional employment at Year 5 (33.3% vs. 28.0% for did-not-attend) and in substantially lower no-MN-record rates throughout. Delayed enrollment produces outcomes by Year 10 that are essentially indistinguishable from never attending at all, suggesting that the timing benefit of postsecondary engagement is concentrated among those who begin immediately. The primary value of this variable in the non-completer context is confirming that the geographic absence signature is most concentrated among those who never attempted postsecondary, and that even unsuccessful immediate postsecondary attempts provide some long-run labor market orientation that delayed or absent attempts do not.

::: panel-tabset
#### 1-year after graduation

```{r}
data <- master.original %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.1,
         var.col = data$attended.ps.within.first.year.hsgrad,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Post-secondary/attended-ps-within-first-year-hsgrad-1.html")
```

#### 5-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.5,
         var.col = data$attended.ps.within.first.year.hsgrad,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Post-secondary/attended-ps-within-first-year-hsgrad-5.html")
```

#### 10-years after graduation

```{r}
data <- master.original %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad == "No")

tab_xtab(var.row = data$grad.year.10,
         var.col = data$attended.ps.within.first.year.hsgrad,
         show.exp = TRUE,
         show.col.prc = TRUE,
         file = "Data/Claude/analysis_non_post_secondary_graduates/Post-secondary/attended-ps-within-first-year-hsgrad-10.html")
```
:::

<br>

# Summary of analysis

**OVERVIEW**

The non-college completer analysis examined workforce participation outcomes at Years 1, 5, and 10 for the 56,448 Southeast Minnesota high school graduates who never completed a postsecondary credential — 54% of the total graduate population. This subgroup is not a residual category of academic failure. It is a structurally distinct population with its own internal heterogeneity, its own pathway mechanisms, and its own disconnection signatures. The analysis reveals that the variables which predict success in this subgroup are fundamentally different from those that predict success in the full population — and that the differences are analytically important for understanding both who achieves regional workforce attachment and who does not.

**THE OUTCOME LANDSCAPE**

By Year 5, non-college completers distribute across four active workforce categories: meaningful regional employment (30.1%), no-MN-record (26.6%), non-meaningful employment (24.0%), and meaningful MN employment (14.1%). By Year 10, meaningful regional employment holds at 30.6%, no-MN-record rises to 34.0%, non-meaningful employment declines to 18.7%, and meaningful MN employment falls to 16.6%. The most analytically significant feature of this distribution is the large and growing no-MN-record group — reaching 34.0% by Year 10 — which among non-completers is more likely to reflect genuine disconnection or informal employment than the out-of-state career establishment that drives no-MN-record among credential completers.

**THE WORKFORCE-FIRST PATHWAY**

The clearest and most important story the data tells about non-college completers is about CTE engagement. Total CTE courses taken is the strongest and most durable predictor of regional employment in the entire non-completer analysis — stronger than any demographic variable, stronger than any academic achievement measure, and essentially unchanged in its effect across the decade of observation. Graduates who take more CTE courses show meaningfully regional employment at Year 1, maintain that advantage at Year 5, and still show approximately 1.2 more CTE courses than no-MN-record graduates at Year 10. The dose-response gradient in cte.achievement is equally clear and equally stable — CTE concentrators show 35.3% meaningful regional employment at Year 10 compared to 22.7% for No CTE graduates, a 12.6 percentage point gap that has been essentially constant since Year 1.

This finding has two analytically important dimensions. First, CTE engagement predicts regional employment directly — not through credential completion, not through academic achievement, and not through postsecondary orientation. Among graduates who never complete a credential, those who concentrated in CTE achieve sustained regional employment at rates that exceed graduates with higher academic achievement and lower CTE engagement. Second, the absence of CTE engagement is the defining characteristic of geographic absence. No CTE graduates show 41.2% no-MN-record at Year 10 — the highest of any CTE category and the most direct bivariate expression of genuine disconnection in the non-completer subgroup. The workforce- first pathway works when CTE engagement is present, and it fails — producing geographic absence rather than regional employment — when CTE engagement is absent.

**THE ACADEMIC ORIENTATION PARADOX**

The second major story the data tells is about what academic achievement and orientation do not produce among non-completers. Every academic achievement and orientation variable examined — ACT participation, ACT composite score, AP exam participation, MCA Math, MCA Reading, MCA Science, PSEO participation — shows the same pattern: higher academic orientation predicts geographic mobility and in-state out-of-region employment rather than regional workforce attachment. By Year 10, the ACT composite score structure is unambiguous — meaningful MN employment and no-MN-record graduates score approximately 1.4 points above meaningful regional employment graduates, which in turn score at essentially the same level as non-meaningful employment graduates. MCA Reading and MCA Math show the same convergence — academic achievement does not differentiate between regional employment success and local underutilization among non-completers.

This is the academic orientation paradox of the non-completer subgroup. Among credential completers, academic achievement produces out-of- region careers. Among non-completers, academic achievement produces geographic mobility without career establishment — graduates who were oriented toward postsecondary education but did not complete credentials are more likely to either achieve in-state out-of-region employment or disappear from Minnesota's formal employment records than to achieve sustained regional employment. The academic orientation mechanism that drives geographic departure in the full population operates equally among non-completers, but without the credential completion that facilitates successful out-of-state career establishment it produces geographic absence rather than out-of-state success.

**THE TWO DISCONNECTION SIGNATURES**

The non-completer analysis clarifies the two disconnection signatures that appear throughout the broader analysis, revealing their demographic and preparation profiles with greater precision than the full population analysis allows.

The local underutilization signature — present in Minnesota's labor market but below the sustained employment threshold — is concentrated among graduates with economic disadvantage and special education status, and among those with low CTE engagement. Low-income non- completers show 27.5% non-meaningful employment at Year 5 compared to 21.6% for non-low-income non-completers. Special education non- completers show 29.3% at Year 5. No CTE graduates show non-meaningful employment rates consistently above CTE participants and concentrators at all three time points. The defining characteristic of this signature is the absence of both CTE engagement and postsecondary orientation — graduates who neither developed vocational skills for direct regional employment nor attempted postsecondary credentials that might facilitate geographic mobility.

The geographic absence signature — no Minnesota employment record — is concentrated among language minority graduates, graduates who never attempted postsecondary, and No CTE graduates. English learner non-completers show 44.6% no-MN-record at Year 10 — the highest of any demographic group and growing throughout the observation window. Non-english home graduates show 44.9%. These are not out-of-state career establishment rates — they are more likely to reflect return migration, informal employment, or genuine disconnection from formal labor markets. The no-MN-record rate for No CTE graduates (41.2% at Year 10) reinforces that geographic absence is most concentrated among those who entered the labor market without either vocational preparation or postsecondary orientation.

Special education non-completers show both signatures simultaneously — elevated no-MN-record (31.3% at Year 5, 35.4% at Year 10) and elevated non-meaningful employment (29.3% at Year 5, 24.0% at Year 10). This dual disconnection profile is the most severe disadvantage pattern in the non-completer subgroup and identifies special education non-completers as a population facing structural barriers in both dimensions.

**THE ROLE OF POSTSECONDARY ORIENTATION**

A finding that cuts across the analysis is the modest but persistent benefit of postsecondary attempt — even unsuccessful — for geographic attachment. Non-completers who attempted postsecondary show substantially lower no-MN-record rates throughout the observation window compared to those who never attempted. The gap is 22.8 percentage points at Year 1 (6.7% vs. 29.5%), narrows to 14.3 at Year 5 (19.4% vs. 33.7%), and remains 9.6 points at Year 10 (29.4% vs. 39.0%). This benefit is concentrated among immediate enrollees — graduates who attempted postsecondary in the year after high school graduation. Delayed enrollment produces outcomes by Year 10 that are essentially indistinguishable from never attending at all.

The postsecondary attempt, even without completion, provides institutional engagement and geographic orientation that reduces the probability of complete absence from Minnesota's formal labor market. This is not a credential effect — all individuals in this subgroup are non-completers. It is an engagement effect, and it dissipates when enrollment is delayed beyond the first year after high school.

**THE PLACE OF PLACE**

The high school characteristics group confirms that where non-completers attended high school within Southeast Minnesota has essentially no bearing on their workforce trajectories. EDR and community demographic description show negligible associations at all three time points. The EDR differences visible in the full population — driven by differential postsecondary pathway sorting — are entirely absent in this subgroup. The one partial exception is average local unemployment rate, which shows a consistent and theoretically coherent association: stronger local labor markets produce regional employment attachment, while weaker local economies produce geographic departure or in-state out-of-region employment. This mechanism is present but modest, and it confirms that local economic conditions shape the direction of labor market entry for non-completers without credential advantages to facilitate mobility.

**WHAT THE DATA CANNOT TELL US**

The non-completer analysis faces an important interpretive limitation in the no-MN-record category. For credential completers, no-MN-record at Year 10 is reasonably interpreted as successful out-of-state career establishment — particularly for those who completed credentials outside Minnesota. For non-completers, the same category is heterogeneous and more difficult to interpret. It may include graduates employed informally or in cash economies not captured by Minnesota's wage records, graduates who returned to countries of origin, graduates who are genuinely disconnected from any formal labor market, and a smaller number who achieved out-of-state employment without credentials. The analysis cannot distinguish among these possibilities, and the growing no-MN-record rate among non-completers over time — reaching 34.0% at Year 10 — should be interpreted with this heterogeneity explicitly in mind.

**THE CENTRAL CONCLUSION**

The non-college completer subgroup is not defined by what its members lack — credentials, academic achievement, postsecondary orientation. It is defined by the presence or absence of CTE engagement and the resulting trajectory into or away from regional workforce attachment. Non-completers who concentrated in CTE achieve regional employment at rates that rival credential completers through the sub-baccalaureate regional pathway. Non-completers without CTE engagement — particularly those who also never attempted postsecondary — are the most likely to be absent from Minnesota's formal labor market a decade after high school graduation.

The policy implication is direct: for the 54% of Southeast Minnesota graduates who do not complete postsecondary credentials, CTE engagement is not a secondary consideration or an alternative pathway. It is the primary mechanism through which regional workforce attachment is achieved. Strengthening CTE engagement — particularly at the concentrator level — among students who are unlikely to complete postsecondary credentials is the most evidence-supported intervention for reducing both the local underutilization and geographic absence signatures that define disconnection in this population.
